### 2013-05-08 Manuscript/manuscript.tex
Manuscript

%
\documentclass{svjour3}

\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{tabularx}
\usepackage{lineno}

\linenumbers

\begin{document}

\title{Environmental pathogen reservoirs and habitat heterogeneity in a metapopulation}
\subtitle{Manuscript v.0}

\author{Clint Leach \and Paul Cross \and Colleen Webb}

%\institute{}
%\journalname{}

\date{October 21, 2012}

\titlerunning{Pathogen Reservoirs}

\maketitle

\section{Introduction}
\label{intro}

Many wildlife diseases are caused by pathogens that can persist for long periods of time in the environment.  Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\cite{Dragon1995}), plague (\cite{Eisen2008}), and recently white nose syndrome (\cite{Lindner2011}), among others.  This environmental longevity can have important consequences for disease dynamics, especially in spatially structured populations.

In general the presence of a pathogen can have significant effects on host dynamics in a metapopulation.  Increased movement among patches in a metapopulation is thought to improve metapopulation stability and reduce the likelihood of extinction, but, as shown by \cite{Hess1996}, this can also lead to widespread infection, thereby increasing the probability of extinction.  Studies by \cite{Gog2002} and \cite{McCallum2002} add infection from an alternate host and patches resistant to the pathogen, respectively, and find that increased landscape connectivity (e.g. through increased host movement) always improves the persistence of the host population, regardless of pathogen spread.  These studies highlight the potential for important interactions between spatial structure, host movement, and pathogen transmission, and the importance of understanding these interactions for conservation and management.

The above studies demonstrate the importance of host movement in governing metapopulation and transmission dynamics, but none incorporate patch level heterogeneity.  In particular, the quality of a habitat patch can affect its extinction and colonization probabilities, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).  As an example, in a study of the Glanville fritillary butterfly metapopulation, \cite{Moilanen1998} found that increased patch quality, as measured by the density of flowering plants, decreased emigration and increased immigration.  Thus, patterns of host, and therefore pathogen, movement are determined at least in part by patch quality.  It then follows that the role played by an environmental pathogen reservoir is also a function of patch quality and its distribution.

Though both disease and habitat heterogeneity have been identified independently as potentially important factors in metapopulation dynamics, how these two factors interact is poorly understood.  In particular, at the metapopulation scale, we ask how pathogen spread is influenced by environmental longevity and the variance (or heterogeneity) in the patch quality distribution.  In addition, at the patch level, we suspect that high quality habitat patches might become contaminated with the pathogen, yet still attract colonists, thereby creating an infectious trap for the host.  In this situation, low quality patches might then serve as refuges that allow susceptible hosts to escape infection.  Therefore, we are also interested in how individual patch quality affects host and pathogen dynamics.

\section{Methods}
\label{methods}

\subsection{Model Structure}
To address the above questions, we developed a theoretical stochastic patch occupancy model (SPOM) in which each patch can be in one of four possible states: occupied by susceptibles (S), occupied by infecteds (I), pathogen reservoir (R), and unoccupied (by both pathogen and host) (E).  State transitions are governed by host colonization, extinction, and infection rates.  Using the framework developed by \cite{Hanski1994}, the rate at which patch \emph{i}, with quality $A_i$, is colonized by individuals in state $X \in (S,I)$ is given by its connectivity:
\begin{equation}
S_{Xi}=A_i^{\xi_{im}} \sum_{i\neq j }X_jA_j^be^{-D d_{ij}}.
\end{equation}
where $X_j$ is an indicator function that is $1$ if patch $j$ is in state $X$ and is $0$ otherwise, and the other parameters are defined in Table 1.  Essentially, $S_{Xi}$ sums the colonization effort from all patches in state $X$ to patch $i$.  Infected colonists then infect susceptible patches via direct contact with probability $\delta$.

The extinction rate of patch $i$ in state $X$ is given by:
\begin{equation}
E_{Xi}=\frac{e_X}{A_i^x},
\end{equation}
where $e_X$ is the extinction rate of a patch of unit quality in state $X$ (here we assume that $e_I>e_S$).  When an infected patch goes extinct, we assume that the infected hosts leave behind an infectious pathogen reservoir.  When susceptibles  colonize a reservoir patch, they become infected at rate:
\begin{equation}
\gamma=\gamma_0exp(-rt_I),
\end{equation}
where $t_I$ is the time since last infected occupancy, and $r$ is the pathogen's decay rate in the environment, as determined by the pathogen's longevity.

\begin{table}[h!]
\label{parameters}      
\caption{Parameters of SPOM model and their meanings.}
\begin{tabular}{l p{11cm} r}
Parameter & Interpretation &  Default Value\\
\hline
$\xi_{im}$ & Scaling parameter for effect of target patch quality on immigration & 0.5\\
$b$ & Scaling parameter for how patch quality affects the number of colonists produced & 0.5\\
$D$& Inverse of mean dispersal distance & 5\\
$d_{ij}$ & Distance between patch $i$ and $j$ & 1 for all $i,j$\\
\hline
$e_S$ & Extinction rate of unit quality susceptible patch & 0.1\\
$e_I$ & Extinction rate of unity quality infected patch & 0.5\\
$x$ & Strength of environmental stochasticity & 1\\
\hline
$\delta$ & Probability of direct infection & 0.5\\
$\gamma_0$ & Initial rate of infection from reservoir patch & 0.5
\end{tabular}
\end{table}

\subsection{Model Implementation and Analysis}

The model outlined above was parameterized (Table 2) according to the following assumptions. (1) High quality patches maintain higher population sizes and therefore produce more colonists than low quality patches and have lower extinction rates than low quality patches ($x=1$,$b=0.5$, \cite{Hanski2003}).  (2) High quality patches attract more colonists than low quality patches ($\xi_{im}=0.5$, \cite{Hanski2003}).  (3) All patches are equidistant from one another (i.e. connectivity is determined entirely by quality, $d_{ij}=1$).  In addition to these assumptions, the model was tuned so that without infection, approximately 0.75 of the metapopulation is occupied ($e_S = 0.1$, $D=5$).  Infection parameters were tuned so that the pathogen persists endemically at intermediate longevity ($e_I = 0.5$, $\delta = 0.5$, $\gamma_0 = 0.5$).

Continuous time stochastic simulations were implemented using the Gillespie algorithm (\cite{Gillespie1977}).  For each simulation, qualities for 100 patches were drawn from a uniform distribution with mean 1 and variance ranging from 0.02 to 0.2.  Pathogen longevity was also varied across simulations, ranging from 1 to 200 time steps (with values representing the half-life of the pathogen's infectivity).  For each model run (50 replicates for each of the 30 longevity-variance combinations), an entirely susceptible population was simulated until it reached a steady state, at which point a randomly chosen occupied patch was infected.  The state of the metapopulation was then tracked for 5000 time steps.  For each simulation, we recorded results at both the metapopulation level (proportion of patches in each state) and patch level (proportion of time each patch spends in each state).

\section{Results}
\label{results}

The pathogen persisted in 404 of the 1500 simulations, spreading through the entire susceptible host population (i.e. a pandemic) in 147 simulations.  The probability of pathogen persistence increases significantly with pathogen longevity, though patch quality variance has relatively little effect (Figure 1a).  Within cases where the pathogen persists however, the probability of a pandemic increases significantly with both pathogen longevity and patch quality variance (\ref{poutcome}, Figure 1b).

\begin{figure}
\label{poutcome}
\centering
\includegraphics[height=5.5cm]{poutcome.pdf}
\caption{Logistic regression surface showing the probability of (a) pathogen persistence, and (b) pandemic spread given persistence, as a function of pathogen environmental longevity (half-life of infectivity decay) and patch quality variance.}
\end{figure}

To separate the effects of high and low quality patches in high variance simulations, we selectively replace the ten highest or ten lowest quality patches with patches of unit quality. This reveals that the presence of high quality patches significantly increases infected occupancy, while the presence of low quality patches significantly reduces infected occupancy (Figure 2a).  Similarly, high quality patches decrease susceptible occupancy, while low quality patches increase susceptible occupancy (Figure 2b).  Thus, the increase in the likelihood of a pandemic with increased patch quality variance seems to be attributable to the presence of high quality patches.

\begin{figure}
\label{sens}
\centering
\includegraphics[height=5cm]{sens.pdf}
\caption{The effect of the presence of high and low quality patches on (a) mean infected occupancy and (b) mean susceptible occupancy. The black bar corresponds to a low variance metapopulation.  The \textit{+high} treatment represents the replacement of 10 unit quality patches with 10 high quality patches, while the \textit{+low} treatment represents the replacement of 10 unit quality patches with 10 low quality patches.  Error bars show $\pm$ one standard deviation.}
\end{figure}

To understand the mechanism behind this effect, we look to within patch dynamics, namely the proportion of time spent by individual patches in each of the four possible states.  We limit our analysis to high variance endemic simulations (where the pathogen persists but does not infect the entire susceptible population).  This suggests that high quality patches are more frequently occupied by both infected and susceptible populations (Figure 3).  However, while the proportion of time occupied by infecteds increases steadily with patch quality (Figure 3b), the proportion of time occupied by susceptibles plateaus (Figure 3a).  Moreover, high quality patches tend to experience a greater number of infection events than low quality patches (Figure 4).

\begin{figure}
\label{qualityraw}
\centering
\includegraphics[height=7cm]{qualityraw.pdf}
\caption{Scatter cloud showing the proportion of time spent occupied by infecteds and susceptibles for individual patches in endemic simulations. Darker shades of blue represent higher point density. The red line represents the loess curve fit to the data.}
\end{figure}

\begin{figure}
\label{infections}
\centering
\includegraphics[height=5.5cm]{infectionevents.pdf}
\caption{Scatter cloud showing the number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations. Darker shades of blue represent higher point density. The red line represents the loess curve fit to the data.}
\end{figure}

\section{Discussion}
\label{discussion}

This model framework demonstrates that environmental longevity increases pathogen persistence and spread.  This agrees both with our intuition -- environmental longevity should provide greater opportunity for transmission -- and with models of chronic wasting disease (\cite{Almberg2011},\cite{Sharp2011}).

When longevity is sufficient for the pathogen to persist, the variance in the patch quality distribution (i.e. habitat heterogeneity) controls how and where it spreads.  In particular, increased variance facilitates pandemic dynamics, increasing the likelihood that the pathogen will spread through the entire host population.  The mechanisms underlying this behavior can best be understood by drawing parallels between our metapopulation model and models of pathogen spread through a contact network.  In this framework, our metapopulation becomes a network with patches as nodes connected through immigration (with the strength of the connection measured by connectivity).  Many studies have shown that networks with high degree variance (where a node's degree is roughly equivalent to its connectivity in our metapopulation model) are easily invaded by pathogens, which are then able to spread much more rapidly than in less heterogeneous networks (\cite{Pastor-Satorras2001}).

This increase in pathogen spread in high variance metapopulations is largely attributable to the presence of high quality patches (Figure 2).  Since metapopulation connectivity is determined entirely by patch quality (recall $d_{ij}=1$ for all $i, j$), high quality patches function as highly connected hubs.  Borrowing vocabulary from network theory, high quality patches are effectively high degree nodes (i.e. many connections).  Specifically, the assumptions of the model mean that high quality patches have both high in-degree (attract colonists from other patches) and a high out-degree (produce colonists that spread to other patches).  These two properties have important implications for pathogen spread in the metapopulation. The high in-degree helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptibles repeatedly colonize high quality patches and subsequently become infected (Figure 4).  This infection pressure is the result of both the environmental reservoir and direct infection from infected immigrants attracted to the high quality patch.  This reflects the results from the contact network literature that show that highly connected (i.e. high degree) nodes have higher infection risk than less connected nodes (\cite{Christley2005}, \cite{Keeling2005}).  This trap effect, coupled with the relatively low extinction rate on high quality patches helps to create a stable platform from which the pathogen can spread through the rest of the metapopulation.  Indeed, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}).  Once the pathogen infects high quality patches, it is able to spread to the rest of the metapopulation relatively easily.  This process helps to maintain infected patches throughout the metapopulation, which in turn feed back on high quality patches to ensure a steady stream of infected colonists that maintain the trap effect.  Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches in high variance metapopulations serves to significantly increase pathogen spread.

Despite the dominating effects of high quality patches, low quality patches still play an important role in some cases by providing refuges on which susceptibles can escape infection. Even though individual low quality patches are unable to support consistent occupancy (susceptible or otherwise), at the metapopulation level, the presence of low quality patches increases overall susceptible occupancy, suggesting that in aggregate, low quality patches can provide susceptible refuges.  This phenomenon represents the other side of the hub role played by high quality patches in that low quality patches are relatively weakly connected to the metapopulation and are therefore infrequently colonized by infected individuals.  This is consistent with the result that low degree nodes tend to be the last infected in models of pathogen spread on contact networks (\cite{Barthelemy2004}).  Moreover, the instability of low quality patches (i.e. the high extinction rate), together with the increased mortality of infected populations, means that infected individuals do not persist long on low quality patches.   However, it is important to caveat this refuge effect by noting that it appears to be swamped in the presence of too many high quality patches, as seen by the fact that the likelihood of pandemic increases with patch quality variance despite the fact that high variance metapopulations have equal proportions of high and low quality patches.  This suggests that once enough high quality patches are added, the force of infection in the system becomes large enough to overwhelm the weak connectivity of low quality patches.

In metapopulations without disease, high quality patches generally serve as sources, while low quality patches are sinks.  The presence of a pathogen then considerably alters, and even inverts, these patterns.  Not only do high quality patches become sinks in the presence of a pathogen with an environmental reservoir, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).  From a management perspective, it is critical to understand this shift, as previous theoretical work suggests that management should focus on patches where conditions are most favorable (\cite{Strasser2010}).  Generally, this means that maintaining high quality habitat should form the focus of managers, but Strasser et al. (2010) show that stochastic disturbance (e.g. disease) can lead to cases where focusing on low quality patches is more effective in increasing population growth rate.

Habitat fragmentation and disease represent two major threats to wildlife populations.  As we have seen, these two factors can interact in complex and interesting ways, especially when the pathogen is able to persist in the environment.  Failing to consider these interactions and their effect on the roles played by high and low quality habitat overlooks the importance of low-quality habitat for the persistence of wildlife populations and the importance of high-quality habitat for the persistence of wildlife disease.

\bibliographystyle{spbasic}     
\bibliography{metapop}   % name your BibTeX data base

\end{document}

### 2013-05-31 Manuscript/manuscript.tex
Introduction rewrites

@@
 \title{Environmental pathogen reservoirs and habitat heterogeneity in a metapopulation}
-\subtitle{Manuscript v.0}
 
-\author{Clint Leach \and Paul Cross \and Colleen Webb}
+\author{Clint B. Leach \and Paul C. Cross \and Colleen T. Webb}
 
@@
 
-\date{October 21, 2012}
+\date{May 30, 2013}
 
@@
 
-Many wildlife diseases are caused by pathogens that can persist for long periods of time in the environment.
-Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\cite{Dragon1995}), plague (\cite{Eisen2008}), and recently white nose syndrome (\cite{Lindner2011}), among others.
-This environmental longevity can have important consequences for disease dynamics, especially in spatially structured populations.
+Many wildlife diseases are caused by pathogens that can persist, and remain infectious, for long periods of time in the environment.
+Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\cite{Dragon1995}), plague (\cite{Eisen2008}), and white nose syndrome (\cite{Lindner2011}), among others.
+This environmental persistence creates environmental pathogen reservoirs from which hosts can become infected without direct contact with an infected individual.
+This additional transmission pathway can have important consequences for disease dynamics, especially in spatially structured populations (\cite{Park2012}).
 
-In general the presence of a pathogen can have significant effects on host dynamics in a metapopulation.
-Increased movement among patches in a metapopulation is thought to improve metapopulation stability and reduce the likelihood of extinction, but, as shown by \cite{Hess1996}, this can also lead to widespread infection, thereby increasing the probability of extinction.
-Studies by \cite{Gog2002} and \cite{McCallum2002} add infection from an alternate host and patches resistant to the pathogen, respectively, and find that increased landscape connectivity (e.g.
-through increased host movement) always improves the persistence of the host population, regardless of pathogen spread.
-These studies highlight the potential for important interactions between spatial structure, host movement, and pathogen transmission, and the importance of understanding these interactions for conservation and management.
+Patterns and rates of host movement also play an important role in pathogen spread through a metapopulation.
+As shown by \cite{Hess1996}, for pathogens spread by direct contact alone, increased host movment enhances pathogen spread through a metapopulation.
+Further studies by \cite{Gog2002} and \cite{McCallum2002} explore the role of additional transmission pathways by including background infection from an alternate host, while \cite{Park2012} includes transmission from an environmental reservoir.
+These studies show that the the effect of host movement on occupancy (and pathogen spread) depends on the relative strength of the different transmission pathways and highlight the potential for patterns of host movement and pathways of pathogen transmission to interact and influence disease dynamics.
 
-The above studies demonstrate the importance of host movement in governing metapopulation and transmission dynamics, but none incorporate patch level heterogeneity.
+However, the above studies assume that all patches in the metapopulation are identical, and consider movement only at the metapopulation scale.
+Since environmental transmission is spatially explicit (i.e.
+environmental reservoirs can only infect residents of the local patch), its role in disease dynamics may depend on the attributes of the specific habitat patches in which reservoirs become established.
 In particular, the quality of a habitat patch can affect its extinction and colonization probabilities, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).
 As an example, in a study of the Glanville fritillary butterfly metapopulation, \cite{Moilanen1998} found that increased patch quality, as measured by the density of flowering plants, decreased emigration and increased immigration.
-Thus, patterns of host, and therefore pathogen, movement are determined at least in part by patch quality.
-It then follows that the role played by an environmental pathogen reservoir is also a function of patch quality and its distribution.
+Thus, patterns of host movement, and consequently the role of environmental pathogen transmission, are determined at least in part by patch quality.
 
-Though both disease and habitat heterogeneity have been identified independently as potentially important factors in metapopulation dynamics, how these two factors interact is poorly understood.
-In particular, at the metapopulation scale, we ask how pathogen spread is influenced by environmental longevity and the variance (or heterogeneity) in the patch quality distribution.
-In addition, at the patch level, we suspect that high quality habitat patches might become contaminated with the pathogen, yet still attract colonists, thereby creating an infectious trap for the host.
-In this situation, low quality patches might then serve as refuges that allow susceptible hosts to escape infection.
-Therefore, we are also interested in how individual patch quality affects host and pathogen dynamics.
+Many of the pathogens listed above affect spatially structured host populations (e.g.
+plague in prairie dog colonies), and thus understanding the consquences of environmental pathogen persistence and habitat quality for pathogen spread is critical to managing disease in these systems.
+In this study, we ask how pathogen longevity (how long the pathogen can persist and remain infections in the environment), and habitat heterogeneity (the variance in the distribution of patch quality across the metapopulation) interact to influence pathogen spread through the metapopulation.
+Within a metapopulation, we ask what specific role low and high quality patches play in controlling pathogen spread.
+
+We expect pathogen spread to increase with both environmental longevity and with habitat heterogeneity.
+Increased environmental longevity should provide greater opportunities for transmission, and thus should favor pathogen spread.
+The role of habitat heterogeneity is likely to be more complex, emerging from the behavior of high and low quality patchs in heterogeneous metapopulations.
+Specifically, we expect that high quality habitat patches, which support greater host traffic than lower quality habitat, might be more likely to form pathogen reservoirs.
+As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (citations, examples?).
+In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (citation).
+Similarly, we predict that low quality patches, which see relatively less host traffic, will be less likely to develop pathogen reservoirs and and thus may serve as refuges for susceptible hosts to escape infection.
+
+To address the above questions and evaluate our hypotheses, we build on the metapopulation framework developed by \cite{Hanski1994} to develop and explore a theoretical stochastic patch occupancy model that incorporates pathogen transmission from environmental reservoirs and habitat heterogeneity.
 
@@
 \centering
-\includegraphics[height=5.5cm]{poutcome.pdf}
+\includegraphics[height=5.5cm]{metapop(full).pdf}
 \caption{Logistic regression surface showing the probability of (a) pathogen persistence, and (b) pandemic spread given persistence, as a function of pathogen environmental longevity (half-life of infectivity decay) and patch quality variance.}
@@
 \centering
-\includegraphics[height=5cm]{sens.pdf}
+\includegraphics[height=5cm]{highlow(full).pdf}
 \caption{The effect of the presence of high and low quality patches on (a) mean infected occupancy and (b) mean susceptible occupancy.
@@
 \centering
-\includegraphics[height=7cm]{qualityraw.pdf}
+\includegraphics[height=7cm]{qualityraw(full).pdf}
 \caption{Scatter cloud showing the proportion of time spent occupied by infecteds and susceptibles for individual patches in endemic simulations.
@@
 \centering
-\includegraphics[height=5.5cm]{infectionevents.pdf}
+\includegraphics[height=5.5cm]{infevents(full).pdf}
 \caption{Scatter cloud showing the number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.

### 2013-06-03 Manuscript/manuscript.tex
Methods edits

@@
 \usepackage{lineno}
+\usepackage{hyperref}
 
@@
 \subsection{Model Structure}
-To address the above questions, we developed a theoretical stochastic patch occupancy model (SPOM) in which each patch can be in one of four possible states: occupied by susceptibles (S), occupied by infecteds (I), pathogen reservoir (R), and unoccupied (by both pathogen and host) (E).
-State transitions are governed by host colonization, extinction, and infection rates.
-Using the framework developed by \cite{Hanski1994}, the rate at which patch \emph{i}, with quality $A_i$, is colonized by individuals in state $X \in (S,I)$ is given by its connectivity:
+To address the above questions, we developed a theoretical stochastic patch occupancy model (SPOM) in which each patch can be in one of four possible states: occupied by susceptibles (S), occupied by infecteds (I), pathogen reservoir without host population (Z), and unoccupied (by both pathogen and host) ($\oslash$).
+State transitions are governed by host colonization, extinction, and infection rates, where a susceptible population can become infected either through direct contact with infected immigrants, or through a local pathogen reservoir.
+
+Building on the framework developed by \cite{Hanski1994}, the rate at which patch \emph{i}, with quality $A_i$, is colonized by individuals in state $X \in (S,I)$ is given by its connectivity:
 \begin{equation}
-S_{Xi}=A_i^{\xi_{im}} \sum_{i\neq j }X_jA_j^be^{-D d_{ij}}.
+C_{Xi}=A_i^{\xi_{im}} \sum_{i\neq j }\phi_jA_j^{\xi_{em}}e^{-D d_{ij}},
 \end{equation}
-where $X_j$ is an indicator function that is $1$ if patch $j$ is in state $X$ and is $0$ otherwise, and the other parameters are defined in Table 1.
-Essentially, $S_{Xi}$ sums the colonization effort from all patches in state $X$ to patch $i$.
-Infected colonists then infect susceptible patches via direct contact with probability $\delta$.
+where $\phi_j$ is an indicator function that is $1$ if patch $j$ is in state $X$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality; $D$ is the inverse of mean dispersal distance; and $d_{ij}$ is the distance between patches $i$ and $j$.
+Essentially, $C_{Xi}$ sums the colonization effort from all patches in state $X$ to patch $i$.
 
@@
 \begin{equation}
-E_{Xi}=\frac{e_X}{A_i^x},
+E_{Xi}=\frac{e_X}{A_i^\alpha},
 \end{equation}
@@
 When an infected patch goes extinct, we assume that the infected hosts leave behind an infectious pathogen reservoir.
-When susceptibles  colonize a reservoir patch, they become infected at rate:
+
+Susceptible populations are infected via direct contact with infected colonists at rate $\delta C_{Ii}$, where $\delta$ is the transmission probability.
+Environmental transmission occurs when a susceptible population occupies a reservoir patch, and occurs at rate:
 \begin{equation}
@@
 \end{equation}
-where $t_I$ is the time since last infected occupancy, and $r$ is the pathogen's decay rate in the environment, as determined by the pathogen's longevity.
+where $\gamma_0$ is the initial infection rate of the pathogen reservoir, $t_I$ is the time since last infected occupancy, and $r$ is the pathogen's decay rate in the environment.
 
@@
 \caption{Parameters of SPOM model and their meanings.}
-\begin{tabular}{l p{11cm} r}
+\begin{tabular}{l p{8.5cm} r}
 Parameter & Interpretation &  Default Value\\
@@
 $\xi_{im}$ & Scaling parameter for effect of target patch quality on immigration & 0.5\\
-$b$ & Scaling parameter for how patch quality affects the number of colonists produced & 0.5\\
+$\xi_{em}$ & Scaling parameter for effect of target patch quality in emigration & 0.5\\
 $D$& Inverse of mean dispersal distance & 5\\
@@
 $e_I$ & Extinction rate of unity quality infected patch & 0.5\\
-$x$ & Strength of environmental stochasticity & 1\\
+$\alpha$ & Strength of environmental stochasticity & 1\\
 \hline
@@
 
-The model outlined above was parameterized (Table 2) according to the following assumptions.
-(1) High quality patches maintain higher population sizes and therefore produce more colonists than low quality patches and have lower extinction rates than low quality patches ($x=1$,$b=0.5$, \cite{Hanski2003}).
-(2) High quality patches attract more colonists than low quality patches ($\xi_{im}=0.5$, \cite{Hanski2003}).
-(3) All patches are equidistant from one another (i.e.
-connectivity is determined entirely by quality, $d_{ij}=1$).
-In addition to these assumptions, the model was tuned so that without infection, approximately 0.75 of the metapopulation is occupied ($e_S = 0.1$, $D=5$).
-Infection parameters were tuned so that the pathogen persists endemically at intermediate longevity ($e_I = 0.5$, $\delta = 0.5$, $\gamma_0 = 0.5$).
+The model outlined above was parameterized (Table 1) according to the following assumptions.
+(1) High quality patches maintain higher population sizes and therefore produce more colonists than low quality patches and have lower extinction rates than low quality patches ($\xi_{im}=0.5$, $\alpha=1$, \cite{Hanski2003}).
+(2) High quality patches attract more colonists than low quality patches ($\xi_{em}=0.5$, \cite{Hanski2003}).
+(3)  Connectivity is determined entirely by quality ($d_{ij}=1$).
+In addition to these assumptions, parameters were chosen so that without infection, approximately 0.75 of the metapopulation was occupied ($e_S = 0.1$, $D=5$), and a range of epidemiological behaviors were feasible ($e_I = 0.5$, $\delta = 0.5$, $\gamma_0 = 0.5$).
 
-Continuous time stochastic simulations were implemented using the Gillespie algorithm (\cite{Gillespie1977}).
-For each simulation, qualities for 100 patches were drawn from a uniform distribution with mean 1 and variance ranging from 0.02 to 0.2.
-Pathogen longevity was also varied across simulations, ranging from 1 to 200 time steps (with values representing the half-life of the pathogen's infectivity).
-For each model run (50 replicates for each of the 30 longevity-variance combinations), an entirely susceptible population was simulated until it reached a steady state, at which point a randomly chosen occupied patch was infected.
+With the above parameters fixed, we explored the effects of pathogen environmental persistence and habitat heterogeneity by varying pathogen longevity (10 values from 20 to 200 time steps, with values representing the half-life of the pathogen's infectivity), and the variance of the patch quality distribution (10 values ranging from 0.02 to 0.2).
+For each simulation (100 replicates for each of the 100 longevity-variance combinations), qualities for 100 patches were drawn from a uniform distribution with mean 1 and given variance.
+In each case, an entirely susceptible population was simulated until it reached a steady state, at which point a randomly chosen occupied patch was infected.
 The state of the metapopulation was then tracked for 5000 time steps.
-For each simulation, we recorded results at both the metapopulation level (proportion of patches in each state) and patch level (proportion of time each patch spends in each state).
+To address the metapopulation-level questions above, we recorded the proportion of patches in each state at the end of each simulation and classified the simulation as host extinction, disease-free (no infected populations after 5000 time steps), endemic (both susceptible and infected populations present), or pandemic (only infected populations).
+In addition, to address patch-level questions, for every simulation, we recorded the quality of each patch and the proportion of time spent in each of the four states.
+
+In the above simulations, the mean of the patch quality distribution was held constant, and thus the range was expanded symmetrically about one as the variance was increased.
+To further parse out the relative roles of high and low quality patches, we performed a series of simulations where we expanded the range of the quality distribution non-symmetrically to include high and low quality patches separately.
+For these simulations, we fixed longevity at an intermediate level (100) and ran 100 replicates for each of four patch quality distributions: low varaince, (range 0.75 to 1.25); intermediate variance, high quality (range 0.75 to 1.75); intermediate variance, low quality (range 0.25 to 1.25); and high variance (range 0.25 to 1.75).
+
+Continuous time stochastic simulations were implemented in the R language (citation) using the Gillespie algorithm (\cite{Gillespie1977}).
+Code is available at \url{https://github.com/clint-leach/Metapop-Disease}.
 

### 2013-06-05 Manuscript/manuscript.tex
Additional methods and results edits
Also changes to R code for highlow analysis -- producing bar chart based
on epidemiological outcome instead of looking at occupancy.

@@
 
-Continuous time stochastic simulations were implemented in the R language (citation) using the Gillespie algorithm (\cite{Gillespie1977}).
+Continuous time stochastic simulations of the above model were implemented in the R language (citation) using the Gillespie algorithm (\cite{Gillespie1977}).
 Code is available at \url{https://github.com/clint-leach/Metapop-Disease}.
@@
 
-The pathogen persisted in 404 of the 1500 simulations, spreading through the entire susceptible host population (i.e.
-a pandemic) in 147 simulations.
-The probability of pathogen persistence increases significantly with pathogen longevity, though patch quality variance has relatively little effect (Figure 1a).
-Within cases where the pathogen persists however, the probability of a pandemic increases significantly with both pathogen longevity and patch quality variance (\ref{poutcome}, Figure 1b).
+The pathogen persisted in 6954 of the 10000 simulations, spreading through the entire susceptible host population (i.e.
+a pandemic) in 3734 simulations.
+The probability of pathogen persistence increased substantially with pathogen longevity, though patch quality variance had relatively little effect (Figure 1a).
+Within cases where the pathogen persisted however, the probability of a pandemic increased with both pathogen longevity and patch quality variance (Figure 1b).
 
@@
 
-To separate the effects of high and low quality patches in high variance simulations, we selectively replace the ten highest or ten lowest quality patches with patches of unit quality.
-This reveals that the presence of high quality patches significantly increases infected occupancy, while the presence of low quality patches significantly reduces infected occupancy (Figure 2a).
-Similarly, high quality patches decrease susceptible occupancy, while low quality patches increase susceptible occupancy (Figure 2b).
-Thus, the increase in the likelihood of a pandemic with increased patch quality variance seems to be attributable to the presence of high quality patches.
+To separate the effects of high and low quality patches, we explore how patch occupancy changes when high and low quality patches are included in the patch quality distribution separately.
+In low variance metapopulations (where patch quality ranges from 0.75 to 1.25, corresponding to a variance of 0.02), the proportions of patches occupied by susceptible populations and infected populations are approximately equal, with occupancies of 0.26 and 0.24 respectively.
+However, when the patch quality distribution is expanded to include high quality patches (range from 0.75 to 1.75), susceptible occupancy drops to nearly zero, while infected occupancy increases to approximately half of the metapopulation (Figure 2).
+Conversely, when the patch quality distribution is expanded to include low quality patches (range from 0.25 to 1.25), susceptible occupancy increases to nearly half, while infected occupancy declines (Figure 2).
+When both high and low quality patches are included (range from 0.25 to 1.75, corresponding to a variance of 0.2), their effects approximately cancel, reducing both susceptible and infected occupancy slightly (Figure 2).
+Though overall occupancy changes little from low to high variance metapopulations, the proportion of simulations that are pandemics still increases (Figure 1), an effect that can be attributed largely to high quality patches.
+However, in high variance metapopulations, the number of disease-free simulations also increases, resulting in relatively little change in mean occupancy.
+\emph{Replace these results with outcome-level results for consistency.}
 
@@
 \includegraphics[height=5cm]{highlow(full).pdf}
-\caption{The effect of the presence of high and low quality patches on (a) mean infected occupancy and (b) mean susceptible occupancy.
-The black bar corresponds to a low variance metapopulation.
-The \textit{+high} treatment represents the replacement of 10 unit quality patches with 10 high quality patches, while the \textit{+low} treatment represents the replacement of 10 unit quality patches with 10 low quality patches.
-Error bars show $\pm$ one standard deviation.}
+\caption{The effect of the presence of high and low quality patches on (a) mean infected occupancy and (b) mean susceptible occupancy. Fix to represent the correct experimental design and results.}
 \end{figure}
@@
 \centering
-\includegraphics[height=5.5cm]{infevents(full).pdf}
+\includegraphics[height=7cm]{infevents(full).pdf}
 \caption{Scatter cloud showing the number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
@@
 This agrees both with our intuition -- environmental longevity should provide greater opportunity for transmission -- and with models of chronic wasting disease (\cite{Almberg2011},\cite{Sharp2011}).
+\emph{Develop further -- both the papers cited, and other studies including a reservoir.
+Remember to highlight not just consistency, but also novelty.}
 

### 2013-06-05 Code/highlow.analysis.r

@@ -16,8 +16,8 @@ highlow<-highlow[,-1]
 
 low<-highlow[which(highlow$treatment=="+low"),]
 high<-highlow[which(highlow$treatment=="+high"),]
-both<-highlow[which(highlow$treatment=="full"),]
-mid<-highlow[which(highlow$treatment=="low.var"),]
+full<-highlow[which(highlow$treatment=="full"),]
+low.var<-highlow[which(highlow$treatment=="low.var"),]
 
 
 outcome<-function(data){
@@ -28,20 +28,16 @@ outcome<-function(data){
   return(c(extinction,nodisease,endemic,pandemic))
 }
 
-outcomes<-rbind(
-  both.outcome<-outcome(both),
-  low.outcome<-outcome(low),
-  high.outcome<-outcome(high),
-  mid.outcome<-outcome(mid)
-  )
+outcomes<-data.frame(cbind(outcome(low.var), outcome(low), outcome(high), outcome(full)))
 
-outcomes<-as.data.frame(outcomes,row.names=c("both","low","high","mid"))
-names(outcomes)<-c("extinction","nodeisease","endemic","pandemic")
+colnames(outcomes)<-c("low.var","low","high","high.var")
 
-outcomes<-as.matrix(outcomes)
-dotchart(outcomes)
+outcomes<-cbind(factor(c("Extinction","Disease-Free","Endemic","Pandemic"), levels=c("Extinction","Disease-Free", "Endemic", "Pandemic")),stack(outcomes))
+names(outcomes)<-c("outcome", "number", "trt")
 
-dotchart(outcomes[,3:4], pch=19,lcolor="black",xlab="Number of outcomes")
+outcomes$trt<-factor(outcomes$trt, levels = c("low.var", "low", "high", "high.var"))
+
+ggplot(outcomes, aes(x=trt, y=number)) + geom_bar(stat="identity") + facet_grid(.~outcome)
 
 
 par(mfrow=c(2,4))

### 2013-06-05 Code/pubplots.R

@@ -10,8 +10,8 @@
 #   full = fully connected - all patches equally accessible from all other patches
 #   lattice = lattice structure - each patch has four accessible neighbors
 
-#conn<- "(full)"
-conn <- "(lattice)"
+conn <- "(full)"
+# conn <- "(lattice)"
 
 
 # Metapopulation-level results

### 2014-03-17 Manuscript/manuscript.tex
Modelrun output and discussion edits
Clean up modelrun output to avoid loops and streamline.  Initial edits
of discussion and some plotting changes.

@@
 
-\date{May 30, 2013}
+\date{November 26, 2013}
 
@@
 
-This model framework demonstrates that environmental longevity increases pathogen persistence and spread.
-This agrees both with our intuition -- environmental longevity should provide greater opportunity for transmission -- and with models of chronic wasting disease (\cite{Almberg2011},\cite{Sharp2011}).
-\emph{Develop further -- both the papers cited, and other studies including a reservoir.
-Remember to highlight not just consistency, but also novelty.}
+The positive relationship between pathogen persistence/spread and environmental longevity demonstrated in the theoretical model above is well documented in models of empirical systems.
+\cite{Almberg2011} find that, in plausible simulations of chronic wasting disease in deer, increasing prion survival in the environment increases the resulting force of infection and peak prevalence, with the  importance of reservoir transmission increasing through time.
+Similarly, Breban et al 2009 show that environmental transmission of avian influenza virus allows it to persist in situations where direct transmission alone is insufficient.
+\emph{Other examples.}
 
-When longevity is sufficient for the pathogen to persist, the variance in the patch quality distribution (i.e.
-habitat heterogeneity) controls how and where it spreads.
-In particular, increased variance facilitates pandemic dynamics, increasing the likelihood that the pathogen will spread through the entire host population.
+When longevity is sufficient for the pathogen to persist, we find that habitat heterogeneity controls how and where it spreads.
+In particular, increased habitat quality variance facilitates pandemic dynamics, increasing the likelihood that the pathogen will spread through the entire host population.
 The mechanisms underlying this behavior can best be understood by drawing parallels between our metapopulation model and models of pathogen spread through a contact network.

### 2014-03-17 Code/highlow.analysis.r

@@ -37,7 +37,7 @@ names(outcomes)<-c("outcome", "number", "trt")
 
 outcomes$trt<-factor(outcomes$trt, levels = c("low.var", "low", "high", "high.var"))
 
-ggplot(outcomes, aes(x=trt, y=number)) + geom_bar(stat="identity") + facet_grid(.~outcome)
+ggplot(outcomes, aes(x=outcome, y=number)) + geom_bar(stat="identity") + facet_grid(.~trt)
 
 
 par(mfrow=c(2,4))

### 2014-03-17 Code/modelrun3.r

@@ -16,154 +16,66 @@
 modelrun<-function(longevity, variance, k, parms, distance, initial, timesteps){
   
 	#Generates quality vector with desired variance
-	#quality<-rgamma(100,shape=1/variance,scale=variance
-  max<-0.5*sqrt(12*variance)+1
-  min<-2-max
-  quality<-runif(100,min=min,max=max)
+  max <- 0.5 * sqrt(12 * variance) + 1
+  min <- 2 - max
+  quality <- runif(100, min=min, max=max)
 	
 	#Converts longevity (half-life) to decay rate
-	r<-log(2)/longevity
+	r <- log(2) / longevity
 	
   #Runs simulation without infection until it reaches steady state
-	Sonly<-diseaseSPOM(distance, quality, initial, parms, r, timesteps,0.15)
-	Sonly<-na.omit(Sonly)
-	initial.inf<-Sonly[length(Sonly[,1]),2:101]
+	Sonly <- diseaseSPOM(distance, quality, initial, parms, r, timesteps, 0.15)
+	Sonly <- na.omit(Sonly)
+	initial.inf <- S only[length(Sonly[, 1]), 2:101]
   
   #Randomly infects an occupied patch
-	initial.inf[sample(which(initial.inf=="S"),1)]<-"I"
+	initial.inf[sample(which(initial.inf=="S"), 1)] <- "I"
 	
 	#Stores quality of initial infected patch
-	initial.quality<-quality[which(initial.inf=="I")]
+	initial.quality <- quality[which(initial.inf == "I")]
 	
 	#Calls SPOM to run with infection
-	output<-diseaseSPOM(distance, quality, initial.inf, parms, r, timesteps,0.05)
-	output<-na.omit(output)
-	
-	#Generates occupancy vectors for each state
-	S<-vector(mode="numeric",length=length(output[,1]))
-	I<-vector(mode="numeric",length=length(output[,1]))
-	R<-vector(mode="numeric",length=length(output[,1]))
-	
-	for(i in 1:length(S)){
-		S[i]<-length(which(output[i,]=="S"))/100}
-
-	for(i in 1:length(I)){
-		I[i]<-length(which(output[i,]=="I"))/100}
-
-	for(i in 1:length(R)){
-		R[i]<-length(which(output[i,]=="R"))/100}
-
-	
-	#Metaopulation-level output
-	#Occupancy at the final time step
[…]

### 2014-03-17 Code/pubplots.R

@@ -18,15 +18,16 @@ conn <- "(full)"
 #===============================================================
 
 
-metapop <- read.csv(paste(dirname(getwd()), "/Data", "/metapop", conn, ".csv", sep=""), header=FALSE)
-names(metapop) <- c("X", "var", "longevity", "ID", "S", "I", "R", "avgS", "avgI", "avgR", "maxI", "tmaxI", "quality")
+metapop <- read.csv(paste(getwd(), "/Data", "/metapop", conn, ".csv", sep=""), header=FALSE)
+metapop <- metapop[, -1]
+names(metapop) <- c("var", "longevity", "ID", "S", "I", "R", "avgS", "avgI", "avgR", "maxI", "tmaxI", "quality")
 
 
 # Plotting logistic regression surface for prob. of pathogen persistence
 
 library(lattice)
 
-pdf(paste(dirname(getwd()), "/Manuscript", "/metapop", conn, ".pdf", sep=""), height=5, width=10)
+pdf(paste(getwd(), "/Manuscript", "/metapop", conn, ".pdf", sep=""), height=5, width=10)
 
 # Generating pathogen persistence and pandemic indicator vectors
 persists <- (metapop$I>0)
@@ -117,7 +118,7 @@ dev.off()
 # Patch level results
 #========================================================
 
-patch<-read.csv(paste(dirname(getwd()), "/Data", "/patch", conn, ".csv", sep=""), header=TRUE)
+patch<-read.csv(paste(getwd(), "/Data", "/patch", conn, ".csv", sep=""), header=TRUE)
 
 names(patch)<-c("patchID","quality","tinf","S","I","R","E","S1","S2","S3","S4","S5","S6","S7","S8","S9","S10",
                      "I1","I2","I3","I4","I5","I6","I7","I8","I9","I10",
@@ -151,3 +152,12 @@ smoothScatter(patch$quality,patch$inf,ylab="Number of infection events",xlab="Pa
 lines(loess.smooth(patch$quality,patch$inf),lwd=3,col="red")
 
 dev.off()
+
+
+#===============================================================================
+bins <- cut(patch$quality, breaks = seq(0.2, 1.8, by = 0.1), labels = seq(0.25, 1.75, by = 0.1))
+
+plot(seq(0.25, 1.75, by = 0.1), c(1:16), ylim = c(0, 10), type = "n", xlab = "Quality", ylab = "Infections")
[…]

### 2014-04-15 Manuscript/manuscript.tex
Figure and results edits
Updated figures and corresponding text in results.

@@
 
-\author{Clint B. Leach \and Paul C. Cross \and Colleen T. Webb}
+\author{Clinton B. Leach \and Paul C. Cross \and Colleen T. Webb}
 
@@
 %\journalname{}
-
-\date{November 26, 2013}
 
@@
 
-The pathogen persisted in 6954 of the 10000 simulations, spreading through the entire susceptible host population (i.e.
-a pandemic) in 3734 simulations.
-The probability of pathogen persistence increased substantially with pathogen longevity, though patch quality variance had relatively little effect (Figure 1a).
-Within cases where the pathogen persisted however, the probability of a pandemic increased with both pathogen longevity and patch quality variance (Figure 1b).
+The pathogen persisted in 7087 of the 10000 simulations, spreading through the entire susceptible host population (i.e.
+a pandemic) in 3721 simulations.
+Pathogen persistence increased substantially with longevity, with intermediate longevities favoring endemic disease, and high longevities facilitating pandemic dynamics (Fig~ \ref{poutcome}).
+The effect of longevity, however, depended on the variance of the patch quality distribution, with low variance metapopulations favoring endemic dynamics for a wider range of longevities (Fig~ \ref{poutcome}a), and high variance metapopulations lowering the longevity required for pandemic dynamics (Fig~ \ref{poutcome}b).
 
 \begin{figure}
-\label{poutcome}
 \centering
 \includegraphics[height=5.5cm]{metapop(full).pdf}
-\caption{Logistic regression surface showing the probability of (a) pathogen persistence, and (b) pandemic spread given persistence, as a function of pathogen environmental longevity (half-life of infectivity decay) and patch quality variance.}
+\caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infecteds persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
+\label{poutcome}
 \end{figure}
 
-To separate the effects of high and low quality patches, we explore how patch occupancy changes when high and low quality patches are included in the patch quality distribution separately.
-In low variance metapopulations (where patch quality ranges from 0.75 to 1.25, corresponding to a variance of 0.02), the proportions of patches occupied by susceptible populations and infected populations are approximately equal, with occupancies of 0.26 and 0.24 respectively.
-However, when the patch quality distribution is expanded to include high quality patches (range from 0.75 to 1.75), susceptible occupancy drops to nearly zero, while infected occupancy increases to approximately half of the metapopulation (Figure 2).
-Conversely, when the patch quality distribution is expanded to include low quality patches (range from 0.25 to 1.25), susceptible occupancy increases to nearly half, while infected occupancy declines (Figure 2).
-When both high and low quality patches are included (range from 0.25 to 1.75, corresponding to a variance of 0.2), their effects approximately cancel, reducing both susceptible and infected occupancy slightly (Figure 2).
-Though overall occupancy changes little from low to high variance metapopulations, the proportion of simulations that are pandemics still increases (Figure 1), an effect that can be attributed largely to high quality patches.
-However, in high variance metapopulations, the number of disease-free simulations also increases, resulting in relatively little change in mean occupancy.
-\emph{Replace these results with outcome-level results for consistency.}
+To separate the influence of high and low quality patches, we explored how susceptible and infected patch occupancy changes when high and low quality patches are included separately in the patch quality distribution.
+Relative to a low-variance metapopulation (patch qualities roughly between 0.75 and 1.25), shifting the distribution toward low quality patches (distribution ranging from 0.25 to 1.25) increased mean susceptible occupancy, while shifting towards high quality patches (range from 0.75 to 1.75) decreased mean susceptible occupancy, with the largest effects observed at intermediate longevities (Fig~\ref{sens}(a)).
+Conversely, a shift towards low quality patches decreased infected occupancy, while a shift towards high quality patches increased infected occupancy (Fig~\ref{sens}(b)).
+
+However, at high pathogen longevities, high quality patch distributions had a net positive effect on total occupancy relative to the other patch quality distributions (Fig~\ref{sens}(c)).
+Due to their lower extinction rates, high quality patches are able to maintain stable occupancy despite widespread infection.
+On the other hand, at high pathogen longevities, the positive effect of low qualtiy distributions on susceptible occupancy is overwhelmed, and low quality patches are unable to support infected occupancy, resulting in an increase in host extinction events.
+Thus, the role of the patch quality distribution and the presence of high or low quality patches is generally greatest for low to intermediate pathogen longevity -- once longevity is high enough, the pathogen persists and spreads easily enough that the effects of individual patches are washed out.
+Then the role of patch quality lies primarily in its effect on extinction rates, and whether infected hosts can persist.
 
 \begin{figure}
+\centering
+\includegraphics[height=7cm]{highlow(full).pdf}
+\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infected occupancy, and (c) mean total occupancy, measured relative to  a low variance metapopulation.
+Red solid lines show hiqh quality metapopulations, blue dashed show low quality, and black shows high variance.}
 \label{sens}
-\centering
-\includegraphics[height=5cm]{highlow(full).pdf}
-\caption{The effect of the presence of high and low quality patches on (a) mean infected occupancy and (b) mean susceptible occupancy. Fix to represent the correct experimental design and results.}
 \end{figure}
 
-To understand the mechanism behind this effect, we look to within patch dynamics, namely the proportion of time spent by individual patches in each of the four possible states.
-We limit our analysis to high variance endemic simulations (where the pathogen persists but does not infect the entire susceptible population).
-This suggests that high quality patches are more frequently occupied by both infected and susceptible populations (Figure 3).
-However, while the proportion of time occupied by infecteds increases steadily with patch quality (Figure 3b), the proportion of time occupied by susceptibles plateaus (Figure 3a).
-Moreover, high quality patches tend to experience a greater number of infection events than low quality patches (Figure 4).
+To understand the mechanisms behind these effects, we further examined within patch dynamics on high variance metapopulations that produced endemic dynamics.
+Because of their lower extinction rate and higher recolonization rate, high quality patches support more consistent occupancy than low quality patches (Fig~\ref{simvis}).
+However, as an epidemic progresses, high quality patches increasingly support infected occupants, while low quality patches support infecteds infrequently (Fig~\ref{simvis}).
+In addition, while high quality patches do support susceptible occupancy, they experience a greater number of infection events (susceptible to infected transitions) than low quality patches (Fig~\ref{infections}).
+This relationship becomes weaker as longevity increases, due to the fact that as longevity increases, there are fewer susceptibles remaining in the metapopulation to become infected.
 
 \begin{figure}
-\label{qualityraw}
 \centering
-\includegraphics[height=7cm]{qualityraw(full).pdf}
-\caption{Scatter cloud showing the proportion of time spent occupied by infecteds and susceptibles for individual patches in endemic simulations.
-Darker shades of blue represent higher point density.
-The red line represents the loess curve fit to the data.}
+\includegraphics[height = 9cm]{simvis.png}
+\caption{Results from a single simulation with variance 0.2 and longevity 80.
+The top panel shows total occupancy of susceptible (blue) and infected (red) patches over time.
+The bottom panel shows the state -- susceptible (blue), infected (red), or unoccupied (white) -- of individual patches through time.
+Patches are stacked vertically with the lowest quality ($\sim$ 0.25) at the bottom and the highest quality ($\sim$ 1.75) at the top.}
+\label{simvis}
 \end{figure}
@@
 \begin{figure}
-\label{infections}
 \centering
 \includegraphics[height=7cm]{infevents(full).pdf}
-\caption{Scatter cloud showing the number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
-Darker shades of blue represent higher point density.
-The red line represents the loess curve fit to the data.}
+\caption{The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
+Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed fit over all simulations.
+The thin black lines show the smoothed fits for longevities of 40 (top) and 140 (bottom).}
+\label{infections}
 \end{figure}
@@
 
+\clearpage
+
+\section{Appendix}
+
+\subsection{Sensitivity Analyses}
+
+\subsubsection{Lattice structure}
+
+\subsubsection{Extinction scaling}
+
+\subsubsection{Weak direct transmission}
+
 \bibliographystyle{spbasic}     

### 2014-04-15 Code/highlow.analysis.r

@@ -8,32 +8,35 @@
 #   both = high quality variance (low.var+low+high)
 #=================================================================================
 
-highlow<-read.csv("highlowlong.csv",header=TRUE)
+highlow<-read.csv(paste(getwd(), "/Data/highlow", conn, ".csv", sep = ""), header=TRUE)
 
 highlow<-highlow[,-1]
 
 par(mfrow = c(1, 3), bty = "l")
 
 S.occ <- tapply(highlow$S, list(highlow$longevity, highlow$treatment), mean)
+S.occ <- S.occ - S.occ[, 4]
 
-plot(seq(20, 200, by = 20), S.occ[, 1], type = "b", lty = 1, ylim = c(-0.5, 0.5), pch = 20, col = "red", ylab = "S occupancy", xlab = "Longevity")
-lines(seq(20, 200, by = 20), S.occ[, 2], type = "b", lty = 2, pch = 20, col = "blue")
-lines(seq(20, 200, by = 20), S.occ[, 3], type = "b", lty = 3, pch = 20, col = "black")
-lines(seq(20, 200, by = 20), S.occ[, 4], type = "b", lty = 4, pch = 20, col = "grey")
+plot(seq(20, 200, by = 20), S.occ[, 1], type = "b", lty = 1, ylim = c(-0.5, 0.5), pch = 19, col = "red", ylab = "S occupancy", xlab = "Longevity")
+lines(seq(20, 200, by = 20), S.occ[, 2], type = "b", lty = 2, pch = 19, col = "blue")
+lines(seq(20, 200, by = 20), S.occ[, 3], type = "b", lty = 3, pch = 19, col = "black")
+lines(seq(20, 200, by = 20), S.occ[, 4], type = "l", lty = 4, pch = 19, col = "grey")
 
 I.occ <- tapply(highlow$I, list(highlow$longevity, highlow$treatment), mean)
+I.occ <- I.occ - I.occ[, 4]
 
-plot(seq(20, 200, by = 20), I.occ[, 1], type = "b", lty = 1, ylim = c(-0.5, 0.5), pch = 20, col = "red", ylab = "I occupancy", xlab = "Longevity")
-lines(seq(20, 200, by = 20), I.occ[, 2], type = "b", lty = 2, pch = 20, col = "blue")
-lines(seq(20, 200, by = 20), I.occ[, 3], type = "b", lty = 3, pch = 20, col = "black")
[…]

### 2014-04-15 Code/pubplots.R

@@ -17,141 +17,136 @@ conn <- "(full)"
 # Metapopulation-level results
 #===============================================================
 
-metapop <- read.csv(paste(getwd(), "/Data", "/metapop", conn, ".csv", sep=""), header=FALSE)
-metapop <- metapop[, -1]
-names(metapop) <- c("var", "longevity", "ID", "S", "I", "R", "avgS", "avgI", "avgR", "maxI", "tmaxI", "quality")
+load(paste(getwd(), "/Data/out", conn, ".Rdata", sep = ""))
 
-
-# Plotting logistic regression surface for prob. of pathogen persistence
-
+library(gridExtra)
 library(lattice)
+
+metapop.S <- tapply(out$Sfin, list(out$longevity, out$variance), mean)
+metapop.I <- tapply(out$Ifin, list(out$longevity, out$variance), mean)
+
+metapop.pan <- tapply(out$Sfin == 0 & out$Ifin > 0, list(out$longevity, out$variance), sum) / 100
+metapop.end <- tapply(out$Sfin > 0 & out$Ifin > 0, list(out$longevity, out$variance), sum) / 100
+metapop.ext <- tapply(out$Sfin == 0 & out$Ifin == 0, list(out$longevity, out$variance), sum) / 100
+metapop.nd <- tapply(out$Sfin > 0 & out$Ifin == 0, list(out$longevity, out$variance), sum) / 100
+
+p.pan <- levelplot(metapop.pan, col.regions = brewer.pal(9, "Reds"), xlab = "Longevity", ylab = "Variance", 
+                   at = seq(0, 100, by = 12.5), colorkey = F)
+p.end <- levelplot(metapop.end, col.regions = brewer.pal(9, "Reds"), cuts = 8, xlab = "Longevity", ylab = "Variance", 
+                   at = seq(0, 100, by = 12.5), colorkey = F)
+p.ext <- levelplot(metapop.ext, col.regions = brewer.pal(9, "Reds"), cuts = 8, xlab = "Longevity", ylab = "Variance")
+p.nd <- levelplot(metapop.nd, col.regions = brewer.pal(9, "Reds"), cuts = 8, xlab = "Longevity", ylab = "Variance")
+
+key <- draw.colorkey(list(col = brewer.pal(9, "Reds"), 
+                          at = seq(0, 100, by = 12.5),
+                          labels = list(at = seq(0, 100, by = 25)),
+                          height = 0.7), 
+                     draw = F)
 
[…]

### 2014-04-21 Manuscript/manuscript.tex
Discussion edits

@@
 
-To understand the mechanisms behind these effects, we further examined within patch dynamics on high variance metapopulations that produced endemic dynamics.
+To understand the mechanisms behind these effects, we further examined within patch dynamics in high variance metapopulations that produced endemic dynamics.
 Because of their lower extinction rate and higher recolonization rate, high quality patches support more consistent occupancy than low quality patches (Fig~\ref{simvis}).
@@
 
-The positive relationship between pathogen persistence/spread and environmental longevity demonstrated in the theoretical model above is well documented in models of empirical systems.
+The positive relationship between pathogen spread and environmental longevity demonstrated in the theoretical model above is well documented in models of empirical systems.
 \cite{Almberg2011} find that, in plausible simulations of chronic wasting disease in deer, increasing prion survival in the environment increases the resulting force of infection and peak prevalence, with the  importance of reservoir transmission increasing through time.
@@
 
-When longevity is sufficient for the pathogen to persist, we find that habitat heterogeneity controls how and where it spreads.
-In particular, increased habitat quality variance facilitates pandemic dynamics, increasing the likelihood that the pathogen will spread through the entire host population.
-The mechanisms underlying this behavior can best be understood by drawing parallels between our metapopulation model and models of pathogen spread through a contact network.
-In this framework, our metapopulation becomes a network with patches as nodes connected through immigration (with the strength of the connection measured by connectivity).
-Many studies have shown that networks with high degree variance (where a node's degree is roughly equivalent to its connectivity in our metapopulation model) are easily invaded by pathogens, which are then able to spread much more rapidly than in less heterogeneous networks (\cite{Pastor-Satorras2001}).
+Moreover, for a given longevity, these simulations suggest that high variance metapopulations are more prone to widespread pathogen spread than low variance metapopulations.
+The mechanisms underlying this behavior can be understood by drawing parallels with models of pathogen spread through a contact network.
+In this context, our metapopulation can be viewed a network with patches as nodes connected through immigration (with the strength of the connection measured by connectivity).
+Many studies have shown that networks with high degree variance (where a node's degree is roughly equivalent to its total connectivity in our metapopulation model) are easily invaded by pathogens, which are then able to spread much more rapidly than in less heterogeneous networks (\cite{Pastor-Satorras2001}).
 
-This increase in pathogen spread in high variance metapopulations is largely attributable to the presence of high quality patches (Figure 2).
-Since metapopulation connectivity is determined entirely by patch quality (recall $d_{ij}=1$ for all $i, j$), high quality patches function as highly connected hubs.
-Borrowing vocabulary from network theory, high quality patches are effectively high degree nodes (i.e.
-many connections).
-Specifically, the assumptions of the model mean that high quality patches have both high in-degree (attract colonists from other patches) and a high out-degree (produce colonists that spread to other patches).
+In addition to allowing us to explore the effects of longevity and variance independently, this experiment also allowed us to vary the two simultaneously and explore their interaction, suggesting that the specific effect of longevity on pathogen spread is mediated by the variance of the patch quality distribution.
+In particular, metapopulations with high patch quality variance require lower environmental longevities to experience pandemic dynamics.
+In this way, the effects of environmental longevity are exacerbated by a high variance patch quality distribution, especially for intermediate longevities.
+
+This interaction between environmental longevity and the variance of the patch quality distribution is largely attributable to the presence of high quality patches.
+Since a patch's connectivity is determined, at least in part, by its quality, high quality patches function as highly connected hubs in the metapopulation.
+Specifically, high quality patches have both high in-degree (attract colonists from other patches) and a high out-degree (produce colonists that spread to other patches).
 These two properties have important implications for pathogen spread in the metapopulation.
-The high in-degree helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptibles repeatedly colonize high quality patches and subsequently become infected (Figure 4).
-This infection pressure is the result of both the environmental reservoir and direct infection from infected immigrants attracted to the high quality patch.
-This reflects the results from the contact network literature that show that highly connected (i.e.
-high degree) nodes have higher infection risk than less connected nodes (\cite{Christley2005}, \cite{Keeling2005}).
+The high in-degree helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).
+This infection pressure results in part from direct transmission from infected immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}), but is further augemented by environmental transmission.
+
 This trap effect, coupled with the relatively low extinction rate on high quality patches helps to create a stable platform from which the pathogen can spread through the rest of the metapopulation.
@@
 This process helps to maintain infected patches throughout the metapopulation, which in turn feed back on high quality patches to ensure a steady stream of infected colonists that maintain the trap effect.
-Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches in high variance metapopulations serves to significantly increase pathogen spread.
+Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread.
 
 Despite the dominating effects of high quality patches, low quality patches still play an important role in some cases by providing refuges on which susceptibles can escape infection.
-Even though individual low quality patches are unable to support consistent occupancy (susceptible or otherwise), at the metapopulation level, the presence of low quality patches increases overall susceptible occupancy, suggesting that in aggregate, low quality patches can provide susceptible refuges.
+Even though individual low quality patches are unable to support consistent occupancy, at the metapopulation level, the presence of low quality patches increases overall susceptible occupancy, an effect that is strongest at intermediate longevities.
 This phenomenon represents the other side of the hub role played by high quality patches in that low quality patches are relatively weakly connected to the metapopulation and are therefore infrequently colonized by infected individuals.
-This is consistent with the result that low degree nodes tend to be the last infected in models of pathogen spread on contact networks (\cite{Barthelemy2004}).
 Moreover, the instability of low quality patches (i.e.
-the high extinction rate), together with the increased mortality of infected populations, means that infected individuals do not persist long on low quality patches.
-However, it is important to caveat this refuge effect by noting that it appears to be swamped in the presence of too many high quality patches, as seen by the fact that the likelihood of pandemic increases with patch quality variance despite the fact that high variance metapopulations have equal proportions of high and low quality patches.
-This suggests that once enough high quality patches are added, the force of infection in the system becomes large enough to overwhelm the weak connectivity of low quality patches.
+the high extinction rate), together with the increased mortality of infected populations, means that infected populations do not persist long on low quality patches.
+Because low quality patches are so infrequently colonized, the environmental reservoir left behind by these infected populations generally decays before it has the opportunity to infect new susceptible colonists.
+
+Due to their lower extinction rates, high quality patches are able to maintain stable occupancy despite widespread infection.
+On the other hand, at high pathogen longevities, the positive effect of low qualtiy distributions on susceptible occupancy is overwhelmed, and low quality patches are unable to support infected occupancy, resulting in an increase in host extinction events.
+Thus, the role of the patch quality distribution and the presence of high or low quality patches is generally greatest for low to intermediate pathogen longevity -- once longevity is high enough, the pathogen persists and spreads easily enough that the effects of individual patches are washed out.
+Then the role of patch quality lies primarily in its effect on extinction rates, and whether infected hosts can persist.
 

### 2014-04-23 Manuscript/manuscript.tex
More discussion edits
Added discussion of relationship to Hess, Gog, and Park papers and the
longevity dependence of the role of high and low quality patches.

@@
 
-Moreover, for a given longevity, these simulations suggest that high variance metapopulations are more prone to widespread pathogen spread than low variance metapopulations.
+Moreover, for a given longevity, these simulations suggest that high variance metapopulations are more prone to widespread pathogen spread than low variance metapopulations (Fig~\ref{poutcome}.
 The mechanisms underlying this behavior can be understood by drawing parallels with models of pathogen spread through a contact network.
@@
 
-In addition to allowing us to explore the effects of longevity and variance independently, this experiment also allowed us to vary the two simultaneously and explore their interaction, suggesting that the specific effect of longevity on pathogen spread is mediated by the variance of the patch quality distribution.
-In particular, metapopulations with high patch quality variance require lower environmental longevities to experience pandemic dynamics.
-In this way, the effects of environmental longevity are exacerbated by a high variance patch quality distribution, especially for intermediate longevities.
+These simulations also offer insight into the interaction between environmental longevity and habitat heterogeneity, suggesting that the specific effect of longevity on pathogen spread is mediated by the variance of the patch quality distribution.
+In particular, metapopulations with high patch quality variance require lower environmental longevities to experience pandemic dynamics (Fig~\ref{poutcome}).
+In this way, the effects of environmental longevity are exacerbated, or amplified, by a high variance patch quality distribution, especially at intermediate longevities.
 
-This interaction between environmental longevity and the variance of the patch quality distribution is largely attributable to the presence of high quality patches.
+This interaction between environmental longevity and the variance of the patch quality distribution is largely driven by the presence of high quality patches.
 Since a patch's connectivity is determined, at least in part, by its quality, high quality patches function as highly connected hubs in the metapopulation.
-Specifically, high quality patches have both high in-degree (attract colonists from other patches) and a high out-degree (produce colonists that spread to other patches).
+Specifically, high quality patches both attract more colonists from other patches and produce more colonists that spread to other patches.
 These two properties have important implications for pathogen spread in the metapopulation.
-The high in-degree helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).
+Attracting colonists helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).
 This infection pressure results in part from direct transmission from infected immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}), but is further augemented by environmental transmission.
@@
 Indeed, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}).
-Once the pathogen infects high quality patches, it is able to spread to the rest of the metapopulation relatively easily.
-This process helps to maintain infected patches throughout the metapopulation, which in turn feed back on high quality patches to ensure a steady stream of infected colonists that maintain the trap effect.
-Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread.
+Once the pathogen infects high quality patches, the larger number of colonists produced allow it to spread to the rest of the metapopulation relatively easily.
+This process helps to maintain infected occupancy throughout the metapopulation, which in turn feed back on high quality patches to ensure a steady stream of infected colonists that maintain the trap effect.
+Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread and infected occupancy in the metapopulation (Fig~\ref{sens}).
 
-Despite the dominating effects of high quality patches, low quality patches still play an important role in some cases by providing refuges on which susceptibles can escape infection.
-Even though individual low quality patches are unable to support consistent occupancy, at the metapopulation level, the presence of low quality patches increases overall susceptible occupancy, an effect that is strongest at intermediate longevities.
+Despite the dominating effects of high quality patches, low quality patches still play an important role in some cases by facilitating increased susceptible occupancy (Fig~\ref{sens}).
+Even though individual low quality patches are unable to support consistent occupancy (Fig~\ref{simvis}), at the metapopulation level, the presence of low quality patches increases overall susceptible occupancy relative to a lower variance metapopulation, an effect that is strongest at intermediate longevities (Fig~\ref{sens}).
 This phenomenon represents the other side of the hub role played by high quality patches in that low quality patches are relatively weakly connected to the metapopulation and are therefore infrequently colonized by infected individuals.
@@
 the high extinction rate), together with the increased mortality of infected populations, means that infected populations do not persist long on low quality patches.
-Because low quality patches are so infrequently colonized, the environmental reservoir left behind by these infected populations generally decays before it has the opportunity to infect new susceptible colonists.
+Because low quality patches are so infrequently colonized, at low to intermediate longevities, the environmental reservoir left behind by these infected populations generally decays before it has the opportunity to infect new susceptible colonists.
+As a result, low quality patches represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infected occupancy (Fig~\ref{sens}).
 
-Due to their lower extinction rates, high quality patches are able to maintain stable occupancy despite widespread infection.
-On the other hand, at high pathogen longevities, the positive effect of low qualtiy distributions on susceptible occupancy is overwhelmed, and low quality patches are unable to support infected occupancy, resulting in an increase in host extinction events.
-Thus, the role of the patch quality distribution and the presence of high or low quality patches is generally greatest for low to intermediate pathogen longevity -- once longevity is high enough, the pathogen persists and spreads easily enough that the effects of individual patches are washed out.
-Then the role of patch quality lies primarily in its effect on extinction rates, and whether infected hosts can persist.
+However, the importance of the patch quality distribution in pathogen spread depends on the pathogen's longevity, with the strongest effects of patch quality observed for intermediate longevities.
+For these longevities, the timescale of pathogen decay roughly matches the timescale of the colonization and extinction processes.
+In contrast, for low longevities, when the pathogen reservoir decays much faster than colonizations and extinctions occur, environmental transmission has little effect on disease dynamics and the effect of patch quality is relatively weak.
+Conversely, when the pathogen decays much slower than colonizations and extinctions occur, i.e.
+high longevity, the pathogen reservoir persists on nearly all patches and the effects of patch quality are washed out.
+In these cases, the role of patch quality lies primarily in its effect on extinction rates.
+Due to their lower extinction rates, high quality patches are able to maintain stable occupancy despite widespread (pandemic) infection, while low quality patches are unable to support infected occupancy, resulting in an increase in host extinction events at the highest longevities.
 
-In metapopulations without disease, high quality patches generally serve as sources, while low quality patches are sinks.
-The presence of a pathogen then considerably alters, and even inverts, these patterns.
-Not only do high quality patches become sinks in the presence of a pathogen with an environmental reservoir, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).
+These findings suggest the presence of a trade-off between facilitating metapopulation stability and inhibiting disease spread, with a pathogen's environmental longevity controlling the roles of different quality patches with respect to this trade-off.
+At low to intermediate longevities, the addition of high quality patches functions similarly to increasing movement in the model developed by \cite{Hess1996}, increasing pathogen spread and reducing overall occupancy (Fig~\ref{sens}c).
+In these cases, rather than serving as sources in the metapopulation, high quality patches  become sinks, and more importantly, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).
 From a management perspective, it is critical to understand this shift, as previous theoretical work suggests that management should focus on patches where conditions are most favorable (\cite{Strasser2010}).
@@
 
-Habitat fragmentation and disease represent two major threats to wildlife populations.
-As we have seen, these two factors can interact in complex and interesting ways, especially when the pathogen is able to persist in the environment.
-Failing to consider these interactions and their effect on the roles played by high and low quality habitat overlooks the importance of low-quality habitat for the persistence of wildlife populations and the importance of high-quality habitat for the persistence of wildlife disease.
+However, as pathogen longevity increases, the trade-off shifts and the addition of high quality patches is a net benefit to overall occupancy, reflecting the findings of \cite{Gog2002}, who find that under background infection from an alternative host, increasing movement rarely reduces occupancy, despite facilitating pathogen spread.
+\cite{Park2012} further develops this work by noting that the magnitude of the detrimental effects of increased moevement on occupancy depends on the relative strength of different transmission pathways.
+Specifically, he finds that as environmental transmission becomes stronger (e.g.
+higher longevity) relative to direct transmission, then total occupancy increases monotonically with movement, i.e.
+movement is a net benefit to the metapopulation.
+These studies, coupled with the results presented here, suggest that when there is a persistent source of infection throughout the metapopulation (e.g.
+when the pathogen is long-lived in the environment), factors that facilitate metapopulation stability (i.e.
+increased movement, high quality patches) are a greater benefit than factors that inhibit pathogen spread (i.e.
+decreased movement, low quality patches).
+
+Spatial structure and disease spread represent two major factors influencing the dynamics of wildlife populations.
+As we have seen, in cases where the pathogen is able to persist and remain infectious in the environment, these two factors can interact in complex and interesting ways.
+Failing to consider these interactions and their effect on the roles played by high and low quality habitat risks overlooking the importance of low-quality habitat for the persistence of wildlife populations and the importance of high-quality habitat for the persistence of wildlife disease.
 

### 2014-04-23 Manuscript/manuscript.tex
Introduction edits
Attempts to streamline introduction.

@@
 
-Patterns and rates of host movement also play an important role in pathogen spread through a metapopulation.
-As shown by \cite{Hess1996}, for pathogens spread by direct contact alone, increased host movment enhances pathogen spread through a metapopulation.
-Further studies by \cite{Gog2002} and \cite{McCallum2002} explore the role of additional transmission pathways by including background infection from an alternate host, while \cite{Park2012} includes transmission from an environmental reservoir.
-These studies show that the the effect of host movement on occupancy (and pathogen spread) depends on the relative strength of the different transmission pathways and highlight the potential for patterns of host movement and pathways of pathogen transmission to interact and influence disease dynamics.
-
-However, the above studies assume that all patches in the metapopulation are identical, and consider movement only at the metapopulation scale.
 Since environmental transmission is spatially explicit (i.e.
@@
 In particular, the quality of a habitat patch can affect its extinction and colonization probabilities, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).
-As an example, in a study of the Glanville fritillary butterfly metapopulation, \cite{Moilanen1998} found that increased patch quality, as measured by the density of flowering plants, decreased emigration and increased immigration.
-Thus, patterns of host movement, and consequently the role of environmental pathogen transmission, are determined at least in part by patch quality.
+Thus, we expect that patterns of host movement, and consequently the effect of environmental pathogen transmission, are determined at least in part by patch quality.
 
 Many of the pathogens listed above affect spatially structured host populations (e.g.
-plague in prairie dog colonies), and thus understanding the consquences of environmental pathogen persistence and habitat quality for pathogen spread is critical to managing disease in these systems.
-In this study, we ask how pathogen longevity (how long the pathogen can persist and remain infections in the environment), and habitat heterogeneity (the variance in the distribution of patch quality across the metapopulation) interact to influence pathogen spread through the metapopulation.
+plague in prairie dog colonies), and thus understanding how environmental pathogen persistence interacts with habitat quality is critical to managing disease in these systems.
+In this study, we ask how pathogen longevity (how long the pathogen can persist and remain infections in the environment), and habitat heterogeneity (the variance in the distribution of patch quality across the metapopulation) interact to influence pathogen spread and metapopulation occupancy.
 Within a metapopulation, we ask what specific role low and high quality patches play in controlling pathogen spread.
@@
 Specifically, we expect that high quality habitat patches, which support greater host traffic than lower quality habitat, might be more likely to form pathogen reservoirs.
-As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (citations, examples?).
+As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Almberg2011}, citations, examples?).
 In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (citation).

### 2014-04-24 Manuscript/manuscript.tex
More intro edits and appendix plots
Adding notes on lit into intro and adding sensitivity analysis plots to
appendix.

@@
 This environmental persistence creates environmental pathogen reservoirs from which hosts can become infected without direct contact with an infected individual.
-This additional transmission pathway can have important consequences for disease dynamics, especially in spatially structured populations (\cite{Park2012}).
+
+\cite{Almberg2011} find in models of CWD that increasing prior survival increases pathogen spread (using a number of measures) relative to a model that only accounts for direct transmission; highlights importance of accounting for environmental transmission.
+
+\cite{Breban2009} find that including environmental transmission better explains AIV in North American waterfowl than direct transmission only, and that it allows for persistence in cases where direct transmission alone would not.
+
+\cite{Clancy2005} shows that environmental transmission facilitates pathogen persistence compared to models with direct transmission only.
+
+\cite{Fuller2012} quantify persistence time baculovirus in gypsy moths and suggest that quantifying persistence times in empirical systems and including those estimates in models offers greater understanding of disease spread.
+(should read this more carefully)
+
+\cite{Sharp2011} model CWD and find that as longevity increases, critical carrying capacities (that trigger disease persistence first, and then stable limit cycles and recurrent epidemics) are reduced, and thus populations are more susceptible to disease outbreaks.
+Also suggest that management efforts that increase carrying capacity (e.g.
+high quality habitat) can facilitate disease persistence and possibly destabilize the population.
 
@@
 
-\subsubsection{Extinction scaling}
+\begin{figure}[h!]
+\centering
+\includegraphics[height=5.5cm]{metapop(x0).pdf}
+\caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infecteds persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
+\label{poutcome_x0}
+\end{figure}
 
-\subsubsection{Weak direct transmission}
+\begin{figure}
+\centering
+\includegraphics[height=7cm]{highlow(x0).pdf}
+\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infected occupancy, and (c) mean total occupancy, measured relative to  a low variance metapopulation.
+Red solid lines show hiqh quality metapopulations, blue dashed show low quality, and black shows high variance.}
+\label{sens_x0}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[height=7cm]{infevents(x0).pdf}
+\caption{The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
+Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed fit over all simulations.
+The thin black lines show the smoothed fits for longevities of 40 (top) and 140 (bottom).}
+\label{infections_x0}
+\end{figure}
+
+\clearpage
+
+\subsubsection{Extinction scaling, $\alpha = 0$}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[height=5.5cm]{metapop(x0).pdf}
+\caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infecteds persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
+\label{poutcome_x0}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[height=7cm]{highlow(x0).pdf}
+\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infected occupancy, and (c) mean total occupancy, measured relative to  a low variance metapopulation.
+Red solid lines show hiqh quality metapopulations, blue dashed show low quality, and black shows high variance.}
+\label{sens_x0}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[height=7cm]{infevents(x0).pdf}
+\caption{The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
+Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed fit over all simulations.
+The thin black lines show the smoothed fits for longevities of 40 (top) and 140 (bottom).}
+\label{infections_x0}
+\end{figure}
+
+\clearpage
+
+\subsubsection{Weak direct transmission, $\delta = 0.1$}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[height=5.5cm]{metapop(x0).pdf}
+\caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infecteds persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
+\label{poutcome_x0}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[height=7cm]{highlow(x0).pdf}
+\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infected occupancy, and (c) mean total occupancy, measured relative to  a low variance metapopulation.
+Red solid lines show hiqh quality metapopulations, blue dashed show low quality, and black shows high variance.}
+\label{sens_x0}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[height=7cm]{infevents(x0).pdf}
+\caption{The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
+Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed fit over all simulations.
+The thin black lines show the smoothed fits for longevities of 40 (top) and 140 (bottom).}
+\label{infections_x0}
+\end{figure}
+
+\clearpage
 

### 2014-04-24 Code/pubplots.R

@@ -10,8 +10,9 @@
 #   full = fully connected - all patches equally accessible from all other patches
 #   lattice = lattice structure - each patch has four accessible neighbors
 
-conn <- "(full)"
+# conn <- "(full)"
 # conn <- "(lattice)"
+conn <- "(x0)"
 
 
 # Metapopulation-level results
@@ -21,6 +22,7 @@ load(paste(getwd(), "/Data/out", conn, ".Rdata", sep = ""))
 
 library(gridExtra)
 library(lattice)
+library(RColorBrewer)
 
 metapop.S <- tapply(out$Sfin, list(out$longevity, out$variance), mean)
 metapop.I <- tapply(out$Ifin, list(out$longevity, out$variance), mean)
@@ -164,6 +166,8 @@ for(i in c(40, 60, 80, 100)){
 # Number of infection events
 library(scales)
 
+pdf(paste(getwd(), "/Manuscript", "/infevents", conn, ".pdf", sep=""), height=6, width=8)
+
 reps <- unique(patch$repID)
 
 par(mfrow = c(1, 1))
@@ -179,6 +183,8 @@ lines(loess.smooth(patch$quality[patch$longevity == 80], patch$inf.events.tot[pa
 lines(loess.smooth(patch$quality[patch$longevity == 100], patch$inf.events.tot[patch$longevity == 100]), lwd = 1, col = "black")
 lines(loess.smooth(patch$quality[patch$longevity == 120], patch$inf.events.tot[patch$longevity == 120]), lwd = 1, col = "black")
 lines(loess.smooth(patch$quality[patch$longevity == 140], patch$inf.events.tot[patch$longevity == 140]), lwd = 1, col = "black")
+
+dev.off()
 
 par(mfrow = c(2, 2))
 

### 2014-04-24 Manuscript/manuscript.tex
Still more introduction edits
Restructuring, streamlining, and adding a few citations to first
paragraph to justify importance of env. transmission.

--- 2014-04-23 Manuscript/manuscript.tex
@@
 This environmental persistence creates environmental pathogen reservoirs from which hosts can become infected without direct contact with an infected individual.
-This additional transmission pathway can have important consequences for disease dynamics, especially in spatially structured populations (\cite{Park2012}).
+This additional transmission pathway can have important consequences for disease dynamics, with models showing that increased environmental persistence generally facilitates increased pathogen persistence and spread relative to direct transmission alone (\cite{Almberg2011}, \cite{Sharp2011}, \cite{Breban2009}.
 
 Since environmental transmission is spatially explicit (i.e.
-environmental reservoirs can only infect residents of the local patch), its role in disease dynamics may depend on the attributes of the specific habitat patches in which reservoirs become established.
-In particular, the quality of a habitat patch can affect its extinction and colonization probabilities, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).
-Thus, we expect that patterns of host movement, and consequently the effect of environmental pathogen transmission, are determined at least in part by patch quality.
+environmental reservoirs can only infect local residents), its role in disease dynamics may further depend on the spatial structure of the host population.
+In particular, we expect that the host population structure and movement will be influenced by heterogeneity in quality among the habitat patches in a metapopulation.
+Indeed, the quality of a habitat patch can affect its extinction and colonization probabilities, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).
+These processes in turn may then influence where environmental pathogen reservoirs get established and how they affect disease dynamics and host occupancy throughout the metapopulation.
 
-Many of the pathogens listed above affect spatially structured host populations (e.g.
-plague in prairie dog colonies), and thus understanding how environmental pathogen persistence interacts with habitat quality is critical to managing disease in these systems.
-In this study, we ask how pathogen longevity (how long the pathogen can persist and remain infections in the environment), and habitat heterogeneity (the variance in the distribution of patch quality across the metapopulation) interact to influence pathogen spread and metapopulation occupancy.
-Within a metapopulation, we ask what specific role low and high quality patches play in controlling pathogen spread.
+Specifically, we expect that high quality habitat patches, which support greater host density and traffic than lower quality habitat, might be more likely to form pathogen reservoirs.
+As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Almberg2011}).
+In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (citation).
+Similarly, we expect that low quality patches, which see relatively less host traffic, will be less likely to develop pathogen reservoirs and thus may serve as refuges on which susceptible hosts can escape infection.
 
-We expect pathogen spread to increase with both environmental longevity and with habitat heterogeneity.
-Increased environmental longevity should provide greater opportunities for transmission, and thus should favor pathogen spread.
-The role of habitat heterogeneity is likely to be more complex, emerging from the behavior of high and low quality patchs in heterogeneous metapopulations.
-Specifically, we expect that high quality habitat patches, which support greater host traffic than lower quality habitat, might be more likely to form pathogen reservoirs.
-As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Almberg2011}, citations, examples?).
-In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (citation).
-Similarly, we predict that low quality patches, which see relatively less host traffic, will be less likely to develop pathogen reservoirs and and thus may serve as refuges for susceptible hosts to escape infection.
-
-To address the above questions and evaluate our hypotheses, we build on the metapopulation framework developed by \cite{Hanski1994} to develop and explore a theoretical stochastic patch occupancy model that incorporates pathogen transmission from environmental reservoirs and habitat heterogeneity.
+Many, if not all, of the pathogens listed above affect spatially structured host populations (e.g.
+plague in prairie dog colonies), and thus understanding how environmental transmission interacts with patterns of habitat quality is critical to managing disease in these systems.
+In this study then, we seek to explore how a pathogen's environmental longevity (how long it can persist and remain infections in the environment), and habitat heterogeneity (the variance in the distribution of patch quality in a metapopulation) interact to influence pathogen spread and host occupancy, with a specific interest in the roles played by high and low quality habitat.
 
@@
 
+Also discuss \cite{Sharp2011}, who show that management actions that increase carrying capacity can increase disease spread and destabilize population.
+And \cite{Hess1996} who finds that implementing a quarantine on centralized patches reduces pathogen spread.
+
 However, as pathogen longevity increases, the trade-off shifts and the addition of high quality patches is a net benefit to overall occupancy, reflecting the findings of \cite{Gog2002}, who find that under background infection from an alternative host, increasing movement rarely reduces occupancy, despite facilitating pathogen spread.

### 2014-06-25 Manuscript/manuscript.tex
Manuscript edits and sensitivity analysis
Edits throughout manuscript; incorporating sensitivity analysis into
results and discussion; adding sensitivity analysis results.

@@
 environmental reservoirs can only infect local residents), its role in disease dynamics may further depend on the spatial structure of the host population.
-In particular, we expect that the host population structure and movement will be influenced by heterogeneity in quality among the habitat patches in a metapopulation.
-Indeed, the quality of a habitat patch can affect its extinction and colonization probabilities, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).
+In particular, we expect that host population structure and movement will be influenced by heterogeneity in quality among the habitat patches in a metapopulation.
+Indeed, the quality of a habitat patch can affect its extinction and colonization rates, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).
 These processes in turn may then influence where environmental pathogen reservoirs get established and how they affect disease dynamics and host occupancy throughout the metapopulation.
@@
 \subsection{Model Structure}
-To address the above questions, we developed a theoretical stochastic patch occupancy model (SPOM) in which each patch can be in one of four possible states: occupied by susceptibles (S), occupied by infecteds (I), pathogen reservoir without host population (Z), and unoccupied (by both pathogen and host) ($\oslash$).
+To address the above questions, we developed a theoretical stochastic patch occupancy model (SPOM) in which each patch can be in one of three possible states: occupied by susceptible hosts (S), occupied by infected hosts (I), and unoccupied by the host.
 State transitions are governed by host colonization, extinction, and infection rates, where a susceptible population can become infected either through direct contact with infected immigrants, or through a local pathogen reservoir.
@@
 \end{equation}
-where $e_X$ is the extinction rate of a patch of unit quality in state $X$ (here we assume that $e_I>e_S$).
-When an infected patch goes extinct, we assume that the infected hosts leave behind an infectious pathogen reservoir.
+where $e_X$ is the extinction rate of a patch of unit quality in state $X$ (here we assume that $e_I>e_S$), and $\alpha$ controls how extinction rate scales with patch quality.
+When an infected population goes extinct, we assume that the infected hosts leave behind an infectious pathogen reservoir on the patch.
 
 Susceptible populations are infected via direct contact with infected colonists at rate $\delta C_{Ii}$, where $\delta$ is the transmission probability.
-Environmental transmission occurs when a susceptible population occupies a reservoir patch, and occurs at rate:
+In addition, environmental transmission can take place when a susceptible population occupies a patch with an environmental reservoir, and occurs at rate:
 \begin{equation}
-\gamma=\gamma_0exp(-rt_I),
+\gamma_i=\gamma_0exp(-rt_{Ii}),
 \end{equation}
-where $\gamma_0$ is the initial infection rate of the pathogen reservoir, $t_I$ is the time since last infected occupancy, and $r$ is the pathogen's decay rate in the environment.
+where $\gamma_0$ is the initial infection rate of the pathogen reservoir, $t_{Ii}$ is the time since last infected occupancy, and $r$ is the pathogen's decay rate in the environment.
 
@@
 \label{parameters}      
-\caption{Parameters of SPOM model and their meanings.}
+\caption{Parameters of SPOM model and their interpretations.}
 \begin{tabular}{l p{8.5cm} r}
@@
 (2) High quality patches attract more colonists than low quality patches ($\xi_{em}=0.5$, \cite{Hanski2003}).
-(3)  Connectivity is determined entirely by quality ($d_{ij}=1$).
+(3)  Connectivity among patches is determined entirely by quality ($d_{ij}=1$ for all $i \neq j$).
 In addition to these assumptions, parameters were chosen so that without infection, approximately 0.75 of the metapopulation was occupied ($e_S = 0.1$, $D=5$), and a range of epidemiological behaviors were feasible ($e_I = 0.5$, $\delta = 0.5$, $\gamma_0 = 0.5$).
 
-With the above parameters fixed, we explored the effects of pathogen environmental persistence and habitat heterogeneity by varying pathogen longevity (10 values from 20 to 200 time steps, with values representing the half-life of the pathogen's infectivity), and the variance of the patch quality distribution (10 values ranging from 0.02 to 0.2).
+With the above parameters fixed, we explored the effects of pathogen environmental persistence and habitat heterogeneity by varying pathogen longevity (10 values ranging from 20 to 200 time steps, with values representing the half-life of the pathogen's infectivity), and the variance of the patch quality distribution (10 values ranging from 0.02 to 0.2).
 For each simulation (100 replicates for each of the 100 longevity-variance combinations), qualities for 100 patches were drawn from a uniform distribution with mean 1 and given variance.
-In each case, an entirely susceptible population was simulated until it reached a steady state, at which point a randomly chosen occupied patch was infected.
+In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infected.
 The state of the metapopulation was then tracked for 5000 time steps.
 To address the metapopulation-level questions above, we recorded the proportion of patches in each state at the end of each simulation and classified the simulation as host extinction, disease-free (no infected populations after 5000 time steps), endemic (both susceptible and infected populations present), or pandemic (only infected populations).
-In addition, to address patch-level questions, for every simulation, we recorded the quality of each patch and the proportion of time spent in each of the four states.
+In addition, to address patch-level questions, for every simulation, we recorded the quality of each patch, the proportion of time spent in each state, and the number of transitions between each state.
 
@@
 To further parse out the relative roles of high and low quality patches, we performed a series of simulations where we expanded the range of the quality distribution non-symmetrically to include high and low quality patches separately.
-For these simulations, we fixed longevity at an intermediate level (100) and ran 100 replicates for each of four patch quality distributions: low varaince, (range 0.75 to 1.25); intermediate variance, high quality (range 0.75 to 1.75); intermediate variance, low quality (range 0.25 to 1.25); and high variance (range 0.25 to 1.75).
+For these simulations, at each value for longevity, we simulated 100 metapopulations for each of four patch quality distributions: low varaince, (range 0.75 to 1.25); intermediate variance, high quality (range 0.75 to 1.75); intermediate variance, low quality (range 0.25 to 1.25); and high variance (range 0.25 to 1.75).
+
+In addition, to explore the effects of the model assumptions and parameterization, we repeated the above analyses for three additional model parameterizations: with patches arranged in a lattice structure; with $\alpha = 0$ to remove the effect of patch quality on extinction rate; and with $\delta = 0.1$ to reduce the relative importance of direct transmission.
 
@@
 The effect of longevity, however, depended on the variance of the patch quality distribution, with low variance metapopulations favoring endemic dynamics for a wider range of longevities (Fig~ \ref{poutcome}a), and high variance metapopulations lowering the longevity required for pandemic dynamics (Fig~ \ref{poutcome}b).
+The three alternative parameterizations produced qualitatively similar results, though pathogen spread was generally more difficult under these assumptions (i.e.
+fewer pandemics), and the habitat quality variance exerted a relatively weaker influence on the epidemiological outcome (Figs~\ref{poutcome_lattice}, \ref{poutcome_x0}, \ref{poutcome_delta}).
 
@@
 Conversely, a shift towards low quality patches decreased infected occupancy, while a shift towards high quality patches increased infected occupancy (Fig~\ref{sens}(b)).
+These general trends were consistent across the three other parameterizations (Figs~\ref{sens_lattice}, \ref{sens_x0}, \ref{sens_delta}).
 
 However, at high pathogen longevities, high quality patch distributions had a net positive effect on total occupancy relative to the other patch quality distributions (Fig~\ref{sens}(c)).
-Due to their lower extinction rates, high quality patches are able to maintain stable occupancy despite widespread infection.
-On the other hand, at high pathogen longevities, the positive effect of low qualtiy distributions on susceptible occupancy is overwhelmed, and low quality patches are unable to support infected occupancy, resulting in an increase in host extinction events.
-Thus, the role of the patch quality distribution and the presence of high or low quality patches is generally greatest for low to intermediate pathogen longevity -- once longevity is high enough, the pathogen persists and spreads easily enough that the effects of individual patches are washed out.
-Then the role of patch quality lies primarily in its effect on extinction rates, and whether infected hosts can persist.
+Due to their lower extinction rates and high connectivity, high quality patches were able to maintain stable occupancy despite the widespread infection facilitated by high longevity.
+On the other hand, when high pathogen longevities lead to pandemic dynamics, low quality distributions were unable to support infected occupancy, resulting in an increase in host extinction events and a decrease in mean occupancy.
+The model in which the extinction rate does not scale with quality (i.e.
+$\alpha = 0$) produced very similar results (Fig~\ref{sens_x0}(c)).
+However, in simulations with weak direct infection ($\delta = 0.1$), the high quality patch distributions always produced higher occupancy than low quality patch distributions (Fig~\ref{sens_delta}(c)), while in the model with the lattice structure, low quality patch distributions actually produced higher occupancies than high quality distributions at high longevity (Fig~\ref{sens_lattice}(c)).
 
@@
 
-To understand the mechanisms behind these effects, we further examined within patch dynamics in high variance metapopulations that produced endemic dynamics.
-Because of their lower extinction rate and higher recolonization rate, high quality patches support more consistent occupancy than low quality patches (Fig~\ref{simvis}).
-However, as an epidemic progresses, high quality patches increasingly support infected occupants, while low quality patches support infecteds infrequently (Fig~\ref{simvis}).
-In addition, while high quality patches do support susceptible occupancy, they experience a greater number of infection events (susceptible to infected transitions) than low quality patches (Fig~\ref{infections}).
-This relationship becomes weaker as longevity increases, due to the fact that as longevity increases, there are fewer susceptibles remaining in the metapopulation to become infected.
+To understand the mechanisms behind these effects, we further examined within patch dynamics in high variance metapopulations (with qualities ranging from 0.25 to 1.75) that produced endemic dynamics.
+Because of their lower extinction rate and higher recolonization rate, high quality patches supported more consistent occupancy than low quality patches (Fig~\ref{simvis}).
+However, as the epidemic progressed, high quality patches increasingly supported infected occupants, while low quality patches supported infecteds only briefly and infrequently (Fig~\ref{simvis}).
+In addition, while high quality patches did support susceptible occupancy, they experienced a greater number of infection events (susceptible to infected transitions) than low quality patches (Fig~\ref{infections}).
+This relationship became weaker as longevity increased, due to the fact that as longevity increases and the pathogen spreads more easily, there are fewer susceptibles remaining in the metapopulation to become infected.
+A similar trend was observed across the other parameterizations, though the observed relationship between patch quality and number of infection events was shallower for the lattice and weak direct infection models (Figs~\ref{infections_lattice}, \ref{infections_x0}, \ref{infections_delta}).
 
@@
 \cite{Almberg2011} find that, in plausible simulations of chronic wasting disease in deer, increasing prion survival in the environment increases the resulting force of infection and peak prevalence, with the  importance of reservoir transmission increasing through time.
-Similarly, Breban et al 2009 show that environmental transmission of avian influenza virus allows it to persist in situations where direct transmission alone is insufficient.
+Similarly, \cite{Breban2009} show that environmental transmission of avian influenza virus allows it to persist in situations where direct transmission alone is insufficient.
 \emph{Other examples.}
 
-Moreover, for a given longevity, these simulations suggest that high variance metapopulations are more prone to widespread pathogen spread than low variance metapopulations (Fig~\ref{poutcome}.
-The mechanisms underlying this behavior can be understood by drawing parallels with models of pathogen spread through a contact network.
+Moreover, these simulations suggest that high variance metapopulations are more prone to widespread pathogen spread than low variance metapopulations (Fig~\ref{poutcome}).
+In particular, metapopulations with high patch quality variance require lower environmental longevities to experience pandemic dynamics (Fig~\ref{poutcome}).The mechanisms underlying this behavior can be understood by drawing parallels with models of pathogen spread through a contact network.
 In this context, our metapopulation can be viewed a network with patches as nodes connected through immigration (with the strength of the connection measured by connectivity).
 Many studies have shown that networks with high degree variance (where a node's degree is roughly equivalent to its total connectivity in our metapopulation model) are easily invaded by pathogens, which are then able to spread much more rapidly than in less heterogeneous networks (\cite{Pastor-Satorras2001}).
+However, as the alternative parameterizations show, the overall effect of patch quality variance depends on the structure of the metapopulation, the resulting colonization and extinction rates, and the relative balance of the effects of low and high quality patches.
+In the focal model, the dynamics of high variance metapopulations were driven largely by the presence of high quality patches, whereas in the models with the lattice structure and with equal extinction rates across all patches, high and low quality patches were more balanced in their effects, thus canceling each other out as variance increased.
 
-These simulations also offer insight into the interaction between environmental longevity and habitat heterogeneity, suggesting that the specific effect of longevity on pathogen spread is mediated by the variance of the patch quality distribution.
-In particular, metapopulations with high patch quality variance require lower environmental longevities to experience pandemic dynamics (Fig~\ref{poutcome}).
-In this way, the effects of environmental longevity are exacerbated, or amplified, by a high variance patch quality distribution, especially at intermediate longevities.
-
-This interaction between environmental longevity and the variance of the patch quality distribution is largely driven by the presence of high quality patches.
+The effects of patch quality variance, and its interaction with longevity, result from the combined influence of both low and high quality patches in pathogen spread and metapopulation stability.
 Since a patch's connectivity is determined, at least in part, by its quality, high quality patches function as highly connected hubs in the metapopulation.
@@
 Attracting colonists helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).
-This infection pressure results in part from direct transmission from infected immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}), but is further augemented by environmental transmission.
+This infection pressure results from both transmission from the environmental reservoir and from direct transmission from infected immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).
 
 This trap effect, coupled with the relatively low extinction rate on high quality patches helps to create a stable platform from which the pathogen can spread through the rest of the metapopulation.
-Indeed, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}).
-Once the pathogen infects high quality patches, the larger number of colonists produced allow it to spread to the rest of the metapopulation relatively easily.
-This process helps to maintain infected occupancy throughout the metapopulation, which in turn feed back on high quality patches to ensure a steady stream of infected colonists that maintain the trap effect.
+Indeed, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).
+Once the pathogen infects high quality patches, the larger number of host colonists produced by high quality patches allow it to spread to the rest of the metapopulation relatively easily.
+This process helps to maintain infected occupancy throughout the metapopulation, which in turn feeds back on high quality patches to ensure a steady stream of infected colonists that help maintain the trap effect.
 Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread and infected occupancy in the metapopulation (Fig~\ref{sens}).
 
-Despite the dominating effects of high quality patches, low quality patches still play an important role in some cases by facilitating increased susceptible occupancy (Fig~\ref{sens}).
-Even though individual low quality patches are unable to support consistent occupancy (Fig~\ref{simvis}), at the metapopulation level, the presence of low quality patches increases overall susceptible occupancy relative to a lower variance metapopulation, an effect that is strongest at intermediate longevities (Fig~\ref{sens}).
+In contrast, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{sens}).
+Even though individual low quality patches are unable to support consistent occupancy (Fig~\ref{simvis}), at the metapopulation level, the presence of low quality patches increases overall susceptible occupancy relative to a lower variance metapopulation (Fig~\ref{sens}).
 This phenomenon represents the other side of the hub role played by high quality patches in that low quality patches are relatively weakly connected to the metapopulation and are therefore infrequently colonized by infected individuals.
@@
 
-However, the importance of the patch quality distribution in pathogen spread depends on the pathogen's longevity, with the strongest effects of patch quality observed for intermediate longevities.
-For these longevities, the timescale of pathogen decay roughly matches the timescale of the colonization and extinction processes.
+However, the ability of low quality patches to influence disease dynamics depends on the relative strength of the different transmission pathways -- direct and environmental -- and on the structure of the metapopulation.
+At intermediate longevities, the timescale of pathogen decay roughly matches the timescale of the colonization and extinction processes.
 In contrast, for low longevities, when the pathogen reservoir decays much faster than colonizations and extinctions occur, environmental transmission has little effect on disease dynamics and the effect of patch quality is relatively weak.
@@
 \centering
-\includegraphics[height=5.5cm]{metapop(x0).pdf}
+\includegraphics[height=5.5cm]{metapop(lattice).pdf}
 \caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infecteds persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
-\label{poutcome_x0}
+\label{poutcome_lattice}
 \end{figure}
@@
 \centering
-\includegraphics[height=7cm]{highlow(x0).pdf}
+\includegraphics[height=7cm]{highlow(lattice).pdf}
 \caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infected occupancy, and (c) mean total occupancy, measured relative to  a low variance metapopulation.
 Red solid lines show hiqh quality metapopulations, blue dashed show low quality, and black shows high variance.}
-\label{sens_x0}
+\label{sens_lattice}
 \end{figure}
@@
 \centering
-\includegraphics[height=7cm]{infevents(x0).pdf}
+\includegraphics[height=7cm]{infevents(lattice).pdf}
 \caption{The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
 Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed fit over all simulations.
-The thin black lines show the smoothed fits for longevities of 40 (top) and 140 (bottom).}
-\label{infections_x0}
+The thin black lines show the smoothed fits for longevities of 40 (bottom) and 140 (top).}
+\label{infections_lattice}
 \end{figure}
@@
 \centering
-\includegraphics[height=5.5cm]{metapop(x0).pdf}
+\includegraphics[height=5.5cm]{metapop(delta).pdf}
 \caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infecteds persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
-\label{poutcome_x0}
+\label{poutcome_delta}
 \end{figure}
@@
 \centering
-\includegraphics[height=7cm]{highlow(x0).pdf}
+\includegraphics[height=7cm]{highlow(delta).pdf}
 \caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infected occupancy, and (c) mean total occupancy, measured relative to  a low variance metapopulation.
 Red solid lines show hiqh quality metapopulations, blue dashed show low quality, and black shows high variance.}
-\label{sens_x0}
+\label{sens_delta}
 \end{figure}
@@
 \centering
-\includegraphics[height=7cm]{infevents(x0).pdf}
+\includegraphics[height=7cm]{infevents(delta).pdf}
 \caption{The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
 Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed fit over all simulations.
-The thin black lines show the smoothed fits for longevities of 40 (top) and 140 (bottom).}
-\label{infections_x0}
+The thin black lines show the smoothed fits for longevities of 60 (bottom) and 120 (top).}
+\label{infections_delta}
 \end{figure}

### 2014-06-25 Code/highlow.analysis.r

@@ -44,28 +44,28 @@ lines(seq(20, 200, by = 20), occ[, 4], type = "l", lty = 4, pch = 19, col = "gre
 par(mfrow = c(1, 4), bty = "l")
 pandemic <- tapply(highlow$S == 0 & highlow$I > 0, list(highlow$longevity, highlow$treatment), sum)
 
-plot(seq(20, 200, by = 20), pandemic[, 1], type = "b", lty = 1, ylim = c(-100, 100), pch = 20, col = "red", xlab = "Longevity", ylab = "Pandemics")
+plot(seq(20, 200, by = 20), pandemic[, 1], type = "b", lty = 1, ylim = c(0, 100), pch = 20, col = "red", xlab = "Longevity", ylab = "Pandemics")
 lines(seq(20, 200, by = 20), pandemic[, 2], type = "b", lty = 2, pch = 20, col = "blue")
 lines(seq(20, 200, by = 20), pandemic[, 3], type = "b", lty = 3, pch = 20, col = "black")
 lines(seq(20, 200, by = 20), pandemic[, 4], type = "b", lty = 4, pch = 20, col = "grey")
 
 endemic <- tapply(highlow$S > 0 & highlow$I > 0, list(highlow$longevity, highlow$treatment), sum)
 
-plot(seq(20, 200, by = 20), endemic[, 1], type = "b", lty = 1, ylim = c(-100, 100), pch = 20, col = "red", xlab = "Longevity", ylab = "Endemics")
+plot(seq(20, 200, by = 20), endemic[, 1], type = "b", lty = 1, ylim = c(0, 100), pch = 20, col = "red", xlab = "Longevity", ylab = "Endemics")
 lines(seq(20, 200, by = 20), endemic[, 2], type = "b", lty = 2, pch = 20, col = "blue")
 lines(seq(20, 200, by = 20), endemic[, 3], type = "b", lty = 3, pch = 20, col = "black")
 lines(seq(20, 200, by = 20), endemic[, 4], type = "b", lty = 4, pch = 20, col = "grey")
 
 nodisease <- tapply(highlow$S > 0 & highlow$I == 0, list(highlow$longevity, highlow$treatment), sum)
 
-plot(seq(20, 200, by = 20), nodisease[, 1], type = "b", lty = 1, ylim = c(-100, 100), pch = 20, col = "red", xlab = "Longevity", ylab = "No disease")
[…]

### 2014-06-25 Code/pubplots.R

@@ -13,6 +13,7 @@
 # conn <- "(full)"
 # conn <- "(lattice)"
 conn <- "(x0)"
+# conn <- "(delta)"
 
 
 # Metapopulation-level results
@@ -174,7 +175,7 @@ par(mfrow = c(1, 1))
 plot(patch$quality, patch$inf.events.early, type = "n", ylab = "Number of infection events", xlab = "Patch quality", bty = "l")
 for(i in 1:length(reps)){
   sim <- patch[patch$repID == reps[i], ]
-  lines(loess.smooth(sim$quality, sim$inf.events.tot), type = "l", col = alpha("gray", 0.3))
+  try(lines(loess.smooth(sim$quality, sim$inf.events.tot), type = "l", col = alpha("gray", 0.3)))
 }
 lines(loess.smooth(patch$quality, patch$inf.events.tot), lwd = 2, col = "black")
 lines(loess.smooth(patch$quality[patch$longevity == 40], patch$inf.events.tot[patch$longevity == 40]), lwd = 1, col = "black")

### 2014-06-30 Manuscript/manuscript.tex
Manuscript edits
Addressing comments from Colleen and working to flesh out discussion.

@@
 Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\cite{Dragon1995}), plague (\cite{Eisen2008}), and white nose syndrome (\cite{Lindner2011}), among others.
-This environmental persistence creates environmental pathogen reservoirs from which hosts can become infected without direct contact with an infected individual.
+This environmental persistence creates environmental pathogen reservoirs from which hosts can become infected without direct contact with an infectious individual.
 This additional transmission pathway can have important consequences for disease dynamics, with models showing that increased environmental persistence generally facilitates increased pathogen persistence and spread relative to direct transmission alone (\cite{Almberg2011}, \cite{Sharp2011}, \cite{Breban2009}.
@@
 As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Almberg2011}).
-In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (citation).
+In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (\cite{Paull2012}).
 Similarly, we expect that low quality patches, which see relatively less host traffic, will be less likely to develop pathogen reservoirs and thus may serve as refuges on which susceptible hosts can escape infection.
@@
 Many, if not all, of the pathogens listed above affect spatially structured host populations (e.g.
-plague in prairie dog colonies), and thus understanding how environmental transmission interacts with patterns of habitat quality is critical to managing disease in these systems.
+plague in prairie dog colonies, cite George et al 2013), and thus understanding how environmental transmission interacts with patterns of habitat quality is critical to managing disease in these systems.
 In this study then, we seek to explore how a pathogen's environmental longevity (how long it can persist and remain infections in the environment), and habitat heterogeneity (the variance in the distribution of patch quality in a metapopulation) interact to influence pathogen spread and host occupancy, with a specific interest in the roles played by high and low quality habitat.
@@
 \subsection{Model Structure}
-To address the above questions, we developed a theoretical stochastic patch occupancy model (SPOM) in which each patch can be in one of three possible states: occupied by susceptible hosts (S), occupied by infected hosts (I), and unoccupied by the host.
-State transitions are governed by host colonization, extinction, and infection rates, where a susceptible population can become infected either through direct contact with infected immigrants, or through a local pathogen reservoir.
+To address the above questions, we developed a theoretical stochastic patch occupancy model (SPOM) in which each patch can be in one of three possible states: occupied by susceptible hosts (S), occupied by infectious hosts (I), unoccupied by the host, but with an infectious pathogen reservoir, and unoccupied by both host and pathogen.
+State transitions are governed by host colonization, extinction, and infection rates, where a susceptible population can become infectious either through direct contact with infectious immigrants, or through a local pathogen reservoir.
 
-Building on the framework developed by \cite{Hanski1994}, the rate at which patch \emph{i}, with quality $A_i$, is colonized by individuals in state $X \in (S,I)$ is given by its connectivity:
+Building on the framework developed by \cite{Hanski2003}, the rate at which patch \emph{i}, with quality $A_i$, is colonized by individuals in state $X \in (S,I)$ is given by its connectivity:
 \begin{equation}
 C_{Xi}=A_i^{\xi_{im}} \sum_{i\neq j }\phi_jA_j^{\xi_{em}}e^{-D d_{ij}},
+\label{connectivity}
 \end{equation}
@@
 \end{equation}
-where $e_X$ is the extinction rate of a patch of unit quality in state $X$ (here we assume that $e_I>e_S$), and $\alpha$ controls how extinction rate scales with patch quality.
-When an infected population goes extinct, we assume that the infected hosts leave behind an infectious pathogen reservoir on the patch.
+where $e_X$ is the extinction rate of a patch of unit quality in state $X$ (here we assume that $e_I>e_S$ to account for disease-induced mortality), and $\alpha$ controls how extinction rate scales with patch quality.
+When an infectious population goes extinct, we assume that the infectious hosts leave behind an infectious pathogen reservoir on the patch.
 
-Susceptible populations are infected via direct contact with infected colonists at rate $\delta C_{Ii}$, where $\delta$ is the transmission probability.
+Susceptible populations are infectious via direct contact with infectious colonists at rate $\delta C_{Ii}$, where $\delta$ is the transmission probability.
 In addition, environmental transmission can take place when a susceptible population occupies a patch with an environmental reservoir, and occurs at rate:
@@
 \end{equation}
-where $\gamma_0$ is the initial infection rate of the pathogen reservoir, $t_{Ii}$ is the time since last infected occupancy, and $r$ is the pathogen's decay rate in the environment.
+where $\gamma_0$ is the initial infection rate of the pathogen reservoir, $t_{Ii}$ is the time since last infectious occupancy, and $r$ is the pathogen's decay rate in the environment.
 
@@
 $e_S$ & Extinction rate of unit quality susceptible patch & 0.1\\
-$e_I$ & Extinction rate of unity quality infected patch & 0.5\\
+$e_I$ & Extinction rate of unity quality infectious patch & 0.5\\
 $\alpha$ & Strength of environmental stochasticity & 1\\
@@
 (3)  Connectivity among patches is determined entirely by quality ($d_{ij}=1$ for all $i \neq j$).
-In addition to these assumptions, parameters were chosen so that without infection, approximately 0.75 of the metapopulation was occupied ($e_S = 0.1$, $D=5$), and a range of epidemiological behaviors were feasible ($e_I = 0.5$, $\delta = 0.5$, $\gamma_0 = 0.5$).
+In addition to these assumptions, parameters were chosen so that without infection, approximately 0.75 of the metapopulation was occupied ($e_S = 0.1$, $D=5$), and a range of epidemiological behaviors -- disease-free, endemic disease, and pandemic disease (i.e.
+all host populations infectious) -- was feasible, depending on the pathogen's environmental longevity ($e_I = 0.5$, $\delta = 0.5$, $\gamma_0 = 0.5$).
 
@@
 For each simulation (100 replicates for each of the 100 longevity-variance combinations), qualities for 100 patches were drawn from a uniform distribution with mean 1 and given variance.
-In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infected.
+In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infectious.
 The state of the metapopulation was then tracked for 5000 time steps.
-To address the metapopulation-level questions above, we recorded the proportion of patches in each state at the end of each simulation and classified the simulation as host extinction, disease-free (no infected populations after 5000 time steps), endemic (both susceptible and infected populations present), or pandemic (only infected populations).
+To address the metapopulation-level questions above, we recorded the proportion of patches in each state at the end of each simulation and classified the simulation as host extinction, disease-free (no infectious populations after 5000 time steps), endemic (both susceptible and infectious populations present), or pandemic (only infectious populations).
 In addition, to address patch-level questions, for every simulation, we recorded the quality of each patch, the proportion of time spent in each state, and the number of transitions between each state.
@@
 
-In addition, to explore the effects of the model assumptions and parameterization, we repeated the above analyses for three additional model parameterizations: with patches arranged in a lattice structure; with $\alpha = 0$ to remove the effect of patch quality on extinction rate; and with $\delta = 0.1$ to reduce the relative importance of direct transmission.
+In addition, to explore the effects of the model assumptions and parameterization, we repeated the above analyses for three additional model parameterizations.
+The focal model assumes that all patches are equally accessible from all other patches, so to examine the effects of a more rigid spatial structure, we implemented a model with patches arranged in a lattice structure.
+In addition, we explored a model with $\alpha = 0$ so that the extinction rate was constant with patch quality, simulating the effecs of high environmental stochasticity (i.e.
+environmental stochasticity is strong enough that the population size of a patch does not affects its extinction rate).
+Lastly, we parameterized a model with $\delta = 0.1$ to reduce the relative importance of direct transmission and explore the disease dynamics when transmission is driven largely by the environmental reservoir.
 
@@
 
-The pathogen persisted in 7087 of the 10000 simulations, spreading through the entire susceptible host population (i.e.
-a pandemic) in 3721 simulations.
+The pathogen persisted in 70$\%$ of the 10000 simulations, spreading through the entire susceptible host population (i.e.
+a pandemic) in 37$\%$ of the simulations.
 Pathogen persistence increased substantially with longevity, with intermediate longevities favoring endemic disease, and high longevities facilitating pandemic dynamics (Fig~ \ref{poutcome}).
 The effect of longevity, however, depended on the variance of the patch quality distribution, with low variance metapopulations favoring endemic dynamics for a wider range of longevities (Fig~ \ref{poutcome}a), and high variance metapopulations lowering the longevity required for pandemic dynamics (Fig~ \ref{poutcome}b).
-The three alternative parameterizations produced qualitatively similar results, though pathogen spread was generally more difficult under these assumptions (i.e.
+The three alternative parameterizations (lattice structure, $\alpha = 0$, $\delta = 0.1$) produced qualitatively similar results, though pathogen spread in these models was generally more difficult (i.e.
 fewer pandemics), and the habitat quality variance exerted a relatively weaker influence on the epidemiological outcome (Figs~\ref{poutcome_lattice}, \ref{poutcome_x0}, \ref{poutcome_delta}).
@@
 \includegraphics[height=5.5cm]{metapop(full).pdf}
-\caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infecteds persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
+\caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infectiouss persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
 \label{poutcome}
@@
 
-To separate the influence of high and low quality patches, we explored how susceptible and infected patch occupancy changes when high and low quality patches are included separately in the patch quality distribution.
+To separate the influence of high and low quality patches, we explored how susceptible and infectious patch occupancy changes when high and low quality patches are included separately in the patch quality distribution.
 Relative to a low-variance metapopulation (patch qualities roughly between 0.75 and 1.25), shifting the distribution toward low quality patches (distribution ranging from 0.25 to 1.25) increased mean susceptible occupancy, while shifting towards high quality patches (range from 0.75 to 1.75) decreased mean susceptible occupancy, with the largest effects observed at intermediate longevities (Fig~\ref{sens}(a)).
-Conversely, a shift towards low quality patches decreased infected occupancy, while a shift towards high quality patches increased infected occupancy (Fig~\ref{sens}(b)).
+Conversely, a shift towards low quality patches decreased infectious occupancy, while a shift towards high quality patches increased infectious occupancy (Fig~\ref{sens}(b)).
 These general trends were consistent across the three other parameterizations (Figs~\ref{sens_lattice}, \ref{sens_x0}, \ref{sens_delta}).
@@
 Due to their lower extinction rates and high connectivity, high quality patches were able to maintain stable occupancy despite the widespread infection facilitated by high longevity.
-On the other hand, when high pathogen longevities lead to pandemic dynamics, low quality distributions were unable to support infected occupancy, resulting in an increase in host extinction events and a decrease in mean occupancy.
+On the other hand, when high pathogen longevities led to pandemic dynamics, low quality distributions were unable to support infectious occupancy, resulting in an increase in host extinction events and a decrease in mean occupancy.
 The model in which the extinction rate does not scale with quality (i.e.
@@
 \includegraphics[height=7cm]{highlow(full).pdf}
-\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infected occupancy, and (c) mean total occupancy, measured relative to  a low variance metapopulation.
+\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy, measured relative to  a low variance metapopulation.
 Red solid lines show hiqh quality metapopulations, blue dashed show low quality, and black shows high variance.}
@@
 Because of their lower extinction rate and higher recolonization rate, high quality patches supported more consistent occupancy than low quality patches (Fig~\ref{simvis}).
-However, as the epidemic progressed, high quality patches increasingly supported infected occupants, while low quality patches supported infecteds only briefly and infrequently (Fig~\ref{simvis}).
-In addition, while high quality patches did support susceptible occupancy, they experienced a greater number of infection events (susceptible to infected transitions) than low quality patches (Fig~\ref{infections}).
-This relationship became weaker as longevity increased, due to the fact that as longevity increases and the pathogen spreads more easily, there are fewer susceptibles remaining in the metapopulation to become infected.
+However, as the epidemic progressed, high quality patches increasingly supported infectious occupants, while low quality patches supported infectiouss only briefly and infrequently (Fig~\ref{simvis}).
+In addition, while high quality patches did support susceptible occupancy, they experienced a greater number of infection events (susceptible to infectious transitions) than low quality patches (Fig~\ref{infections}).
+This relationship became weaker as longevity increased, due to the fact that as longevity increases and the pathogen spreads more easily, there are fewer susceptibles remaining in the metapopulation to become infectious.
 A similar trend was observed across the other parameterizations, though the observed relationship between patch quality and number of infection events was shallower for the lattice and weak direct infection models (Figs~\ref{infections_lattice}, \ref{infections_x0}, \ref{infections_delta}).
@@
 \caption{Results from a single simulation with variance 0.2 and longevity 80.
-The top panel shows total occupancy of susceptible (blue) and infected (red) patches over time.
-The bottom panel shows the state -- susceptible (blue), infected (red), or unoccupied (white) -- of individual patches through time.
+The top panel shows total occupancy of susceptible (blue) and infectious (red) patches over time.
+The bottom panel shows the state -- susceptible (blue), infectious (red), or unoccupied (white) -- of individual patches through time.
 Patches are stacked vertically with the lowest quality ($\sim$ 0.25) at the bottom and the highest quality ($\sim$ 1.75) at the top.}
@@
 The positive relationship between pathogen spread and environmental longevity demonstrated in the theoretical model above is well documented in models of empirical systems.
-\cite{Almberg2011} find that, in plausible simulations of chronic wasting disease in deer, increasing prion survival in the environment increases the resulting force of infection and peak prevalence, with the  importance of reservoir transmission increasing through time.
+Using a model parameterized for chronic wasting disease in mule deer, \cite{Almberg2011} find that increasing prion survival in the environment increases the resulting force of infection and peak prevalence, with the importance of reservoir transmission increasing through time.
 Similarly, \cite{Breban2009} show that environmental transmission of avian influenza virus allows it to persist in situations where direct transmission alone is insufficient.
-\emph{Other examples.}
+\emph{Other examples.}  Look into George et al (2013).
 
-Moreover, these simulations suggest that high variance metapopulations are more prone to widespread pathogen spread than low variance metapopulations (Fig~\ref{poutcome}).
-In particular, metapopulations with high patch quality variance require lower environmental longevities to experience pandemic dynamics (Fig~\ref{poutcome}).The mechanisms underlying this behavior can be understood by drawing parallels with models of pathogen spread through a contact network.
-In this context, our metapopulation can be viewed a network with patches as nodes connected through immigration (with the strength of the connection measured by connectivity).
+By modeling the spatial spread of such an environmentally-transmitted pathogen, these simulations suggest that high variance metapopulations are more prone to widespread pathogen spread than low variance metapopulations (Fig~\ref{poutcome}).
+In particular, metapopulations with high patch quality variance require lower environmental longevities to experience pandemic dynamics (Fig~\ref{poutcome}).
+The mechanisms underlying this behavior can be understood by drawing parallels with models of pathogen spread through a contact network.
+In this context, our metapopulation can be viewed as a network with patches as nodes connected through immigration (with the strength of the connection measured by connectivity).
 Many studies have shown that networks with high degree variance (where a node's degree is roughly equivalent to its total connectivity in our metapopulation model) are easily invaded by pathogens, which are then able to spread much more rapidly than in less heterogeneous networks (\cite{Pastor-Satorras2001}).
@@
 
-The effects of patch quality variance, and its interaction with longevity, result from the combined influence of both low and high quality patches in pathogen spread and metapopulation stability.
-Since a patch's connectivity is determined, at least in part, by its quality, high quality patches function as highly connected hubs in the metapopulation.
+\emph{Emphasize own results relative to the contact network literature.
+Should try to avoid the "main model results suggest X, but sensitivity analysis models suggest Y" format and instead try to present a more complete picture/story.
+In the primary model, the increase in pathogen spread from high variance metapopulations likely results from the presence of high quality patches and their ability to overwhelm low quality patches at high longevity through a combination of their relative stability, the high connectivity due to the uniform spatial structure, and the relative strength of direct infection.
+The assumptions tested in the other models all weaken these effects -- by constraining movement through the lattice structure; by limiting the influence of direct infection and thus the influence of colonists from infected high quality patches; and by reducing the stability of high quality patches relative to low -- thus preventing low quality patches from being overwhelmed by the high quality patches.}
+
+The differences observed between the different parameterizations of the model highlight the fact that the effects of patch quality variance, and its interaction with longevity, result from the combined influence of both low and high quality patches in pathogen spread and metapopulation stability.
+Since a patch's connectivity is determined by its quality (Eqn~\ref{connectivity}), high quality patches function as highly connected hubs in the metapopulation.
 Specifically, high quality patches both attract more colonists from other patches and produce more colonists that spread to other patches.
@@
 Attracting colonists helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).
-This infection pressure results from both transmission from the environmental reservoir and from direct transmission from infected immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).
+This infection pressure on high quality patches results from both transmission from the environmental reservoir and from direct transmission from infectious immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).
+
+\emph{High qualily patches also experience more infection events because they are able to support susceptibles long enough for them to get infected, whereas susceptibles on low quality patches are probably going to go extinct before they can become infected (though the effect persists when $\alpha = 0$ and is substantially diminished in the lattice model and when $\delta = 0.1$, so it seems that it is driven largely by connectivity and direct infection}
 
@@
 Indeed, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).
-Once the pathogen infects high quality patches, the larger number of host colonists produced by high quality patches allow it to spread to the rest of the metapopulation relatively easily.
-This process helps to maintain infected occupancy throughout the metapopulation, which in turn feeds back on high quality patches to ensure a steady stream of infected colonists that help maintain the trap effect.
-Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread and infected occupancy in the metapopulation (Fig~\ref{sens}).
+Once the pathogen infects high quality patches, the larger number of host colonists produced by these patches allow it to spread to the rest of the metapopulation relatively easily.
+This process helps to maintain infectious occupancy throughout the metapopulation, which in turn feeds back on high quality patches to ensure a steady stream of infectious colonists that help maintain the trap effect.
+Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread and infectious occupancy in the metapopulation (Fig~\ref{sens}).
+
+And \cite{Hess1996} who finds that implementing a quarantine on centralized patches reduces pathogen spread.
 
 In contrast, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{sens}).
-Even though individual low quality patches are unable to support consistent occupancy (Fig~\ref{simvis}), at the metapopulation level, the presence of low quality patches increases overall susceptible occupancy relative to a lower variance metapopulation (Fig~\ref{sens}).
-This phenomenon represents the other side of the hub role played by high quality patches in that low quality patches are relatively weakly connected to the metapopulation and are therefore infrequently colonized by infected individuals.
+Even though individual low quality patches, due to their high extinction rates and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}), the presence of low quality patches increases overall susceptible occupancy relative to a lower variance metapopulation (Fig~\ref{sens}).
+This phenomenon represents the other side of the hub role played by high quality patches in that low quality patches are relatively weakly connected to the metapopulation and are therefore infrequently colonized by infectious individuals.
 Moreover, the instability of low quality patches (i.e.
-the high extinction rate), together with the increased mortality of infected populations, means that infected populations do not persist long on low quality patches.
-Because low quality patches are so infrequently colonized, at low to intermediate longevities, the environmental reservoir left behind by these infected populations generally decays before it has the opportunity to infect new susceptible colonists.
-As a result, low quality patches represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infected occupancy (Fig~\ref{sens}).
+the high extinction rate), together with the increased mortality of infectious populations, means that infectious populations do not persist long on low quality patches.
+Because low quality patches are so infrequently colonized, at low to intermediate longevities, the environmental reservoir left behind by these infectious populations generally decays before it has the opportunity to infect new susceptible colonists.
+As a result, low quality patches represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infectious occupancy (Fig~\ref{sens}).
 
-However, the ability of low quality patches to influence disease dynamics depends on the relative strength of the different transmission pathways -- direct and environmental -- and on the structure of the metapopulation.
-At intermediate longevities, the timescale of pathogen decay roughly matches the timescale of the colonization and extinction processes.
+%However, the net effect of low and high quality patches on metapopulation occupancy depends on the relative strength of the different transmission pathways -- direct and environmental -- and on the structure of the metapopulation.
+When there is little pathogen spread (e.g.
+low longevity, or weak direct infection), high quality patch distributions support greater total occupancy than low quality patch distributions
+%
+%At intermediate longevities, the timescale of pathogen decay roughly matches the timescale of the colonization and extinction processes.
 In contrast, for low longevities, when the pathogen reservoir decays much faster than colonizations and extinctions occur, environmental transmission has little effect on disease dynamics and the effect of patch quality is relatively weak.
@@
 In these cases, the role of patch quality lies primarily in its effect on extinction rates.
-Due to their lower extinction rates, high quality patches are able to maintain stable occupancy despite widespread (pandemic) infection, while low quality patches are unable to support infected occupancy, resulting in an increase in host extinction events at the highest longevities.
+Due to their lower extinction rates, high quality patches are able to maintain stable occupancy despite widespread (pandemic) infection, while low quality patches are unable to support infectious occupancy, resulting in an increase in host extinction events at the highest longevities.
 
-These findings suggest the presence of a trade-off between facilitating metapopulation stability and inhibiting disease spread, with a pathogen's environmental longevity controlling the roles of different quality patches with respect to this trade-off.
-At low to intermediate longevities, the addition of high quality patches functions similarly to increasing movement in the model developed by \cite{Hess1996}, increasing pathogen spread and reducing overall occupancy (Fig~\ref{sens}c).
-In these cases, rather than serving as sources in the metapopulation, high quality patches  become sinks, and more importantly, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).
-From a management perspective, it is critical to understand this shift, as previous theoretical work suggests that management should focus on patches where conditions are most favorable (\cite{Strasser2010}).
-Generally, this means that maintaining high quality habitat should form the focus of managers, but Strasser et al.
+Though this model provides evidence that high quality patches can serve as traps and superspreaders, and low quality patches can serve as susceptible refuges, the effects of these processes on total metapopulation occupancy depend on the strength and structure of the two transmission pathways.
+\emph{Need more here?}  In particular, these findings suggest the presence of a trade-off between facilitating metapopulation stability and inhibiting disease spread, with a pathogen's environmental longevity and ease of spread controlling the roles of different quality patches with respect to this trade-off.
+When pathogen spread is limited or slow (e.g.
+due to low longevity, or weak direct transmission), high quality patches are a net benefit to the metapopulation, as their connectivity and stability facilitate greater occupancy.
+
+\emph{Maybe lay out the big picture in this paragraph: that there are essentially three different scenarios, very low pathogen spread, intermediate/endemic pathogen spread, and easy/pandemic pathogen spread and then go through each one and its consequences on the roles played by high and low quality habitat.}
+
+However, in cases where the pathogen spreads more easily (i.e.
+at an endemic level), the addition of high quality patches functions similarly to increasing movement in the model developed by \cite{Hess1996}, increasing pathogen spread and reducing overall occupancy (Fig~\ref{sens}c).
+In these cases, where pathogen longevity is intermediate, or pathogen spread is limited by metapopulation structure, rather than serving as sources in the metapopulation, high quality patches become sinks.
+More importantly, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).
+From a management perspective, it is critical to understand this shift, as previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable (\cite{Strasser2010}).
+Generally, this means that managers should focus on maintaining high quality habitat, but Strasser et al.
 (2010) show that stochastic disturbance (e.g.
-disease) can lead to cases where focusing on low quality patches is more effective in increasing population growth rate.
+disease-induced mortality) can lead to cases where focusing on low quality patches is more effective in increasing population growth rate.
+Indeed, in these situations, we see that low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptibles, produce higher occupancy than high quality distributions.
+
+\emph{What does this model offer beyond the literature cited?
+WRT Hess 1996, the extension to explicit habitat patches offers insight into the roles of specific patches; even though connectivity is an important driving factor in this model, it also allows us to consider the role of other processes (e.g.
+extinction, explicity pathogen transmission in space).}
 
 Also discuss \cite{Sharp2011}, who show that management actions that increase carrying capacity can increase disease spread and destabilize population.
-And \cite{Hess1996} who finds that implementing a quarantine on centralized patches reduces pathogen spread.
 
@@
 
+\emph{As with the notes above, emphasize not just how this work fits with other literature, but also what it adds.
+WRT these other metapop disease modeling papers, may not want to equate increased patch quality with increased movement so easily and instead delve more into what including patch quality offers us.
+"This refelcts the the results of blah in that ..., but offers the additional insight that ..."}
+
 Spatial structure and disease spread represent two major factors influencing the dynamics of wildlife populations.
@@
 Failing to consider these interactions and their effect on the roles played by high and low quality habitat risks overlooking the importance of low-quality habitat for the persistence of wildlife populations and the importance of high-quality habitat for the persistence of wildlife disease.
+
+\emph{Again, use this last paragraph not just to summarize the results, but to emphasize the novelty of these insights.}
 
@@
 \includegraphics[height=5.5cm]{metapop(lattice).pdf}
-\caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infecteds persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
+\caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infectiouss persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
 \label{poutcome_lattice}
@@
 \includegraphics[height=7cm]{highlow(lattice).pdf}
-\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infected occupancy, and (c) mean total occupancy, measured relative to  a low variance metapopulation.
+\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy, measured relative to  a low variance metapopulation.
 Red solid lines show hiqh quality metapopulations, blue dashed show low quality, and black shows high variance.}
@@
 \includegraphics[height=5.5cm]{metapop(x0).pdf}
-\caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infecteds persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
+\caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infectiouss persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
 \label{poutcome_x0}
@@
 \includegraphics[height=7cm]{highlow(x0).pdf}
-\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infected occupancy, and (c) mean total occupancy, measured relative to  a low variance metapopulation.
+\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy, measured relative to  a low variance metapopulation.
 Red solid lines show hiqh quality metapopulations, blue dashed show low quality, and black shows high variance.}
@@
 \includegraphics[height=5.5cm]{metapop(delta).pdf}
-\caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infecteds persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
+\caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infectiouss persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
 \label{poutcome_delta}
@@
 \includegraphics[height=7cm]{highlow(delta).pdf}
-\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infected occupancy, and (c) mean total occupancy, measured relative to  a low variance metapopulation.
+\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy, measured relative to  a low variance metapopulation.
 Red solid lines show hiqh quality metapopulations, blue dashed show low quality, and black shows high variance.}

### 2014-06-30 Code/pubplots.R

@@ -10,9 +10,9 @@
 #   full = fully connected - all patches equally accessible from all other patches
 #   lattice = lattice structure - each patch has four accessible neighbors
 
-# conn <- "(full)"
+conn <- "(full)"
 # conn <- "(lattice)"
-conn <- "(x0)"
+# conn <- "(x0)"
 # conn <- "(delta)"
 
 
@@ -34,11 +34,13 @@ metapop.ext <- tapply(out$Sfin == 0 & out$Ifin == 0, list(out$longevity, out$var
 metapop.nd <- tapply(out$Sfin > 0 & out$Ifin == 0, list(out$longevity, out$variance), sum) / 100
 
 p.pan <- levelplot(metapop.pan, col.regions = brewer.pal(9, "Reds"), xlab = "Longevity", ylab = "Variance", 
-                   at = seq(0, 100, by = 12.5), colorkey = F)
+                   at = seq(0, 100, by = 12.5), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
 p.end <- levelplot(metapop.end, col.regions = brewer.pal(9, "Reds"), cuts = 8, xlab = "Longevity", ylab = "Variance", 
-                   at = seq(0, 100, by = 12.5), colorkey = F)
-p.ext <- levelplot(metapop.ext, col.regions = brewer.pal(9, "Reds"), cuts = 8, xlab = "Longevity", ylab = "Variance")
-p.nd <- levelplot(metapop.nd, col.regions = brewer.pal(9, "Reds"), cuts = 8, xlab = "Longevity", ylab = "Variance")
+                   at = seq(0, 100, by = 12.5), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9))
+p.ext <- levelplot(metapop.ext, col.regions = brewer.pal(9, "Reds"), cuts = 8, xlab = "Longevity", ylab = "Variance",
+                   scales = list(at = c(1, 3, 5, 7, 9))
+p.nd <- levelplot(metapop.nd, col.regions = brewer.pal(9, "Reds"), cuts = 8, xlab = "Longevity", ylab = "Variance",
+                  scales = list(at = c(1, 3, 5, 7, 9))
 
 key <- draw.colorkey(list(col = brewer.pal(9, "Reds"), 
                           at = seq(0, 100, by = 12.5),

### 2014-06-30 Code/simvis.r

@@ -9,7 +9,7 @@ parms<-data.frame("im"=0.5, #Controls how a patch imports colonists
                   "x"=1,
                   "es"=0.1,
                   "ei"=0.5,
-                  "delta"=0.5,
+                  "delta"=0.1,
                   "gamma0"=0.5,
                   "alpha"=0)
 
@@ -24,7 +24,7 @@ initial[51:100]<-"E"
 
 timesteps<-5000
 
-longevity <- 80
+longevity <- 180
 variance <- 0.2
 
 #===============================================================================

### 2014-07-08 Manuscript/manuscript.tex
Methods edits
Edits to methods section addressing comments from Colleen; trying to
clarify model structure, different parameterizations used, and the
different simulation studies performed.

@@
 \usepackage{hyperref}
+\usepackage{color}
 
@@
 This environmental persistence creates environmental pathogen reservoirs from which hosts can become infected without direct contact with an infectious individual.
-This additional transmission pathway can have important consequences for disease dynamics, with models showing that increased environmental persistence generally facilitates increased pathogen persistence and spread relative to direct transmission alone (\cite{Almberg2011}, \cite{Sharp2011}, \cite{Breban2009}.
+This additional transmission pathway can have important consequences for disease dynamics, with models showing that increased environmental persistence generally facilitates increased pathogen persistence and spread relative to direct transmission alone (\cite{Almberg2011}, \cite{Sharp2011}, \cite{Breban2009}).
 
@@
 \subsection{Model Structure}
-To address the above questions, we developed a theoretical stochastic patch occupancy model (SPOM) in which each patch can be in one of three possible states: occupied by susceptible hosts (S), occupied by infectious hosts (I), unoccupied by the host, but with an infectious pathogen reservoir, and unoccupied by both host and pathogen.
-State transitions are governed by host colonization, extinction, and infection rates, where a susceptible population can become infectious either through direct contact with infectious immigrants, or through a local pathogen reservoir.
+
+\begin{table}
+\caption{State transitions and their rates for patch $i$ in a metapopulation simulation.
+$S$ denotes occupied by  susceptible hosts, $I$ denotes occupied infectious hosts, and $\emptyset$ denotes unoccupied by the host (but potentially with pathogen reservoir with infection rate of $\gamma_i$).}
+\begin{tabular}{lr}
+State Transition & Rate \\
+\hline
+$S \rightarrow I$ & $\delta C_{Ii} + \gamma_i$\\
+$S \rightarrow \emptyset $ & $e_{Si}$\\
+$I \rightarrow \emptyset $ & $e_{Ii}$ \\
+$\emptyset \rightarrow S$ & $C_{Si}$\\
+$\emptyset \rightarrow I$ & $C_{Ii}$\\
+\label{transitions}
+\end{tabular}
+\end{table}
+
+To address the above questions, we developed a theoretical stochastic patch occupancy model (SPOM) in which each patch can be in one of three possible states: occupied by susceptible hosts (S), occupied by infectious hosts (I), and unoccupied by the host ($\emptyset$).
+State transitions are governed by host colonization, extinction, and infection rates, where a susceptible population can become infectious either through direct contact with infectious immigrants, or through a local pathogen reservoir (Table~\ref{transitions}).
 
@@
 where $e_X$ is the extinction rate of a patch of unit quality in state $X$ (here we assume that $e_I>e_S$ to account for disease-induced mortality), and $\alpha$ controls how extinction rate scales with patch quality.
-When an infectious population goes extinct, we assume that the infectious hosts leave behind an infectious pathogen reservoir on the patch.
+When an infectious population goes extinct, we assume that the hosts leave behind an infectious pathogen reservoir on the patch.
 
-Susceptible populations are infectious via direct contact with infectious colonists at rate $\delta C_{Ii}$, where $\delta$ is the transmission probability.
-In addition, environmental transmission can take place when a susceptible population occupies a patch with an environmental reservoir, and occurs at rate:
+Susceptible populations are infected via direct contact with infectious colonists at rate $\delta C_{Ii}$, where $\delta$ is the transmission probability.
+In addition, environmental transmission can take place when a susceptible population occupies a previously infected patch, and occurs at rate:
 \begin{equation}
@@
 \end{equation}
-where $\gamma_0$ is the initial infection rate of the pathogen reservoir, $t_{Ii}$ is the time since last infectious occupancy, and $r$ is the pathogen's decay rate in the environment.
+where $\gamma_0$ is the initial infection rate of the pathogen reservoir, $t_{Ii}$ is the time since last infectious occupancy ($t_{Ii} = \infty$, and thus $\gamma_i = 0$, if the patch has never been occupied by infectious hosts), and $r$ is the pathogen's decay rate in the environment.
 
 \begin{table}[h!]   
-\label{parameters}      
-\caption{Parameters of SPOM model and their interpretations.}
-\begin{tabular}{l p{8.5cm} r}
-Parameter & Interpretation &  Default Value\\
+\caption{Parameters of SPOM model, their meaning, and the values assigned under different parameterizations.  Empty cells indicate same value as the default parameterization.}
+\begin{tabular}{l p{5cm} rrrr}
+Parameter & Interpretation &  Default & Lattice & Equal Ext. & Weak Direct Inf.\\
 \hline
-$\xi_{im}$ & Scaling parameter for effect of target patch quality on immigration & 0.5\\
-$\xi_{em}$ & Scaling parameter for effect of target patch quality in emigration & 0.5\\
-$D$& Inverse of mean dispersal distance & 5\\
-$d_{ij}$ & Distance between patch $i$ and $j$ & 1 for all $i,j$\\
+$\xi_{im}$ & Scaling parameter for effect of target patch quality on immigration & 0.5 & -- & -- & --\\
+$\xi_{em}$ & Scaling parameter for effect of target patch quality in emigration & 0.5 & -- & -- & --\\
+$D$& Inverse of mean dispersal distance & 5 & 2 & -- & --\\
+$d_{ij}$ & Distance between patch $i$ and $j$ & 1 $\forall i \neq j$ & 1 $\forall i, j$ neighbors & -- & --\\
 \hline
-$e_S$ & Extinction rate of unit quality susceptible patch & 0.1\\
-$e_I$ & Extinction rate of unity quality infectious patch & 0.5\\
-$\alpha$ & Strength of environmental stochasticity & 1\\
+$e_S$ & Extinction rate of unit quality susceptible patch & 0.1 & -- & -- & --\\
+$e_I$ & Extinction rate of unity quality infectious patch & 0.5 & -- & -- & --\\
+$\alpha$ & Strength of environmental stochasticity & 1 & -- & 0 & --\\
 \hline
-$\delta$ & Probability of direct infection & 0.5\\
-$\gamma_0$ & Initial rate of infection from reservoir patch & 0.5
+$\delta$ & Probability of direct infection & 0.5 & -- & -- & 0.1\\
+$\gamma_0$ & Initial rate of infection from reservoir patch & 0.5 & -- & -- & --
 \end{tabular}
+\label{params}
 \end{table}
 
-\subsection{Model Implementation and Analysis}
+\subsection{Model Parameterization}
 
-The model outlined above was parameterized (Table 1) according to the following assumptions.
+The model outlined above was parameterized in four different ways to explore the consequences of different assumptions (Table~\ref{params}).
+The baseline model was parameterized  according to the following assumptions.
 (1) High quality patches maintain higher population sizes and therefore produce more colonists than low quality patches and have lower extinction rates than low quality patches ($\xi_{im}=0.5$, $\alpha=1$, \cite{Hanski2003}).
@@
 
-With the above parameters fixed, we explored the effects of pathogen environmental persistence and habitat heterogeneity by varying pathogen longevity (10 values ranging from 20 to 200 time steps, with values representing the half-life of the pathogen's infectivity), and the variance of the patch quality distribution (10 values ranging from 0.02 to 0.2).
-For each simulation (100 replicates for each of the 100 longevity-variance combinations), qualities for 100 patches were drawn from a uniform distribution with mean 1 and given variance.
-In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infectious.
+The baseline model above assumes that all patches are equally accessible from all other patches (i.e.
+are separated by distance 1), so to examine the effects of a more rigid spatial structure, we implemented a model with patches arranged in a square lattice such that each patch is only accessible from its four neighboring patches (Lattice, Table~\ref{params}).
+In addition, we explored a model with $\alpha = 0$ so that the extinction rate was constant with patch quality, simulating the effects of high environmental stochasticity (i.e.
+environmental stochasticity is strong enough that the population size of a patch does not affects its extinction rate)(Equal extinction, Table~\ref{params}).
+Lastly, we parameterized a model with $\delta = 0.1$ to reduce the relative importance of direct transmission and explore the disease dynamics when transmission is driven largely by the environmental reservoir (Weak direct infection, Table~\ref{params}).
+
+\subsection{Simulation studies}
+
+For each of the above parameterizations, we performed two different simulation experiments to explore the combined effects of pathogen environmental longevity and habitat quality distribution on disease dynamics and metapopulation occupancy.
+In the first of these, we investigated the effect of increasing habitat heterogeneity on the dynamics of pathogens with a range of environmental longevities.
+For these simulations, we explored 10 values of pathogen longevity ranging from 20 to 200 time steps, with values representing the half-life of the pathogen's infectivity, and 10 values for the variance of the patch quality distribution, ranging from 0.02 to 0.2.
+For each simulation (100 replicates for each of the 100 longevity-variance combinations), qualities for 100 patches were drawn from a uniform distribution with fixed mean 1 and given variance.
+In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infected.
 The state of the metapopulation was then tracked for 5000 time steps.
-To address the metapopulation-level questions above, we recorded the proportion of patches in each state at the end of each simulation and classified the simulation as host extinction, disease-free (no infectious populations after 5000 time steps), endemic (both susceptible and infectious populations present), or pandemic (only infectious populations).
-In addition, to address patch-level questions, for every simulation, we recorded the quality of each patch, the proportion of time spent in each state, and the number of transitions between each state.
+We recorded the proportion of patches in each state at the end of each simulation and classified the simulation as host extinction, disease-free (no infectious populations after 5000 time steps), endemic (both susceptible and infectious populations present), or pandemic (only infectious populations).
+In addition, to identify the roles played by individual patches of varying quality, for every simulation, we recorded the quality of each patch, the proportion of time spent in each state (i.e.
+occupied by susceptible hosts, occupied by infectious hosts, and unoccupied), and the number of transitions between each state.
 
-In the above simulations, the mean of the patch quality distribution was held constant, and thus the range was expanded symmetrically about one as the variance was increased.
-To further parse out the relative roles of high and low quality patches, we performed a series of simulations where we expanded the range of the quality distribution non-symmetrically to include high and low quality patches separately.
-For these simulations, at each value for longevity, we simulated 100 metapopulations for each of four patch quality distributions: low varaince, (range 0.75 to 1.25); intermediate variance, high quality (range 0.75 to 1.75); intermediate variance, low quality (range 0.25 to 1.25); and high variance (range 0.25 to 1.75).
+The above simulations held the mean of patch quality distribution constant, and explored the effect of increasing habitat heterogeneity around that mean (i.e.
+by expanding the quality distribution symmetrically about 1).
+To further parse out the relative influence of high and low quality patches separately, we performed a second series of simulations where we expanded the range of the uniform quality distribution to skew towards either high or low quality patches (i.e.
+by expanding the quality distribution non-symmetrically about 1).
+For these simulations, at each of the 10 longevities used above, we simulated 100 metapopulations for each of four patch quality distributions: low variance, (patch qualities ranging 0.75 to 1.25); intermediate variance, high quality (range 0.75 to 1.75); intermediate variance, low quality (range 0.25 to 1.25); and high variance (range 0.25 to 1.75).
 
-In addition, to explore the effects of the model assumptions and parameterization, we repeated the above analyses for three additional model parameterizations.
-The focal model assumes that all patches are equally accessible from all other patches, so to examine the effects of a more rigid spatial structure, we implemented a model with patches arranged in a lattice structure.
-In addition, we explored a model with $\alpha = 0$ so that the extinction rate was constant with patch quality, simulating the effecs of high environmental stochasticity (i.e.
-environmental stochasticity is strong enough that the population size of a patch does not affects its extinction rate).
-Lastly, we parameterized a model with $\delta = 0.1$ to reduce the relative importance of direct transmission and explore the disease dynamics when transmission is driven largely by the environmental reservoir.
-
-Continuous time stochastic simulations of the above model were implemented in the R language (citation) using the Gillespie algorithm (\cite{Gillespie1977}).
+Continuous time stochastic simulations of the above model were implemented in the R language (\cite{R2014}) using the Gillespie algorithm (\cite{Gillespie1977}).
 Code is available at \url{https://github.com/clint-leach/Metapop-Disease}.
@@
 
-The pathogen persisted in 70$\%$ of the 10000 simulations, spreading through the entire susceptible host population (i.e.
+\textcolor{red}{Similar to above edits to methods, need to better distinguish the different analyses and what they're testing.}
+
+The pathogen persisted in 70$\%$ of the 10000 simulations in the baseline model with symmetric patch quality distributions, spreading through the entire susceptible host population (i.e.
 a pandemic) in 37$\%$ of the simulations.
@@
 \includegraphics[height=5.5cm]{metapop(full).pdf}
-\caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infectiouss persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
+\caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infectious hosts persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
 \label{poutcome}
@@
 
+\textcolor{red}{Need to emphasize novelty of this work, especially compared to Hess, Gog, and Park.
+Discuss that movement in this model is not only heterogeneous, but also directional and what that means.
+Also consider reorganizing to put the more interesting elements at the front, and get rid of the less interesting parts (i.e.
+the first paragraph).
+As with notes for results and methods, need to be clear about which analyses support which claims, and specifically need to be careful about discussing the different patch quality distributions and distinguishing between the symmetric and asymmetric distributions.}
+
 The positive relationship between pathogen spread and environmental longevity demonstrated in the theoretical model above is well documented in models of empirical systems.
@@
 Similarly, \cite{Breban2009} show that environmental transmission of avian influenza virus allows it to persist in situations where direct transmission alone is insufficient.
-\emph{Other examples.}  Look into George et al (2013).
+\textcolor{red}{Other examples.}  Look into George et al (2013).
 
@@
 
-\emph{Emphasize own results relative to the contact network literature.
+\textcolor{red}{Emphasize own results relative to the contact network literature.
 Should try to avoid the "main model results suggest X, but sensitivity analysis models suggest Y" format and instead try to present a more complete picture/story.
@@
 
-\emph{High qualily patches also experience more infection events because they are able to support susceptibles long enough for them to get infected, whereas susceptibles on low quality patches are probably going to go extinct before they can become infected (though the effect persists when $\alpha = 0$ and is substantially diminished in the lattice model and when $\delta = 0.1$, so it seems that it is driven largely by connectivity and direct infection}
+\textcolor{red}{High qualily patches also experience more infection events because they are able to support susceptibles long enough for them to get infected, whereas susceptibles on low quality patches are probably going to go extinct before they can become infected (though the effect persists when $\alpha = 0$ and is substantially diminished in the lattice model and when $\delta = 0.1$, so it seems that it is driven largely by connectivity and direct infection}
 
@@
 Though this model provides evidence that high quality patches can serve as traps and superspreaders, and low quality patches can serve as susceptible refuges, the effects of these processes on total metapopulation occupancy depend on the strength and structure of the two transmission pathways.
-\emph{Need more here?}  In particular, these findings suggest the presence of a trade-off between facilitating metapopulation stability and inhibiting disease spread, with a pathogen's environmental longevity and ease of spread controlling the roles of different quality patches with respect to this trade-off.
+\textcolor{red}{Need more here?}  In particular, these findings suggest the presence of a trade-off between facilitating metapopulation stability and inhibiting disease spread, with a pathogen's environmental longevity and ease of spread controlling the roles of different quality patches with respect to this trade-off.
 When pathogen spread is limited or slow (e.g.
@@
 
-\emph{Maybe lay out the big picture in this paragraph: that there are essentially three different scenarios, very low pathogen spread, intermediate/endemic pathogen spread, and easy/pandemic pathogen spread and then go through each one and its consequences on the roles played by high and low quality habitat.}
+\textcolor{red}{Maybe lay out the big picture in this paragraph: that there are essentially three different scenarios, very low pathogen spread, intermediate/endemic pathogen spread, and easy/pandemic pathogen spread and then go through each one and its consequences on the roles played by high and low quality habitat.}
 
@@
 
-\emph{What does this model offer beyond the literature cited?
+\textcolor{red}{What does this model offer beyond the literature cited?
 WRT Hess 1996, the extension to explicit habitat patches offers insight into the roles of specific patches; even though connectivity is an important driving factor in this model, it also allows us to consider the role of other processes (e.g.
@@
 
-\emph{As with the notes above, emphasize not just how this work fits with other literature, but also what it adds.
+\textcolor{red}{As with the notes above, emphasize not just how this work fits with other literature, but also what it adds.
 WRT these other metapop disease modeling papers, may not want to equate increased patch quality with increased movement so easily and instead delve more into what including patch quality offers us.
@@
 
-\emph{Again, use this last paragraph not just to summarize the results, but to emphasize the novelty of these insights.}
+\textcolor{red}{Again, use this last paragraph not just to summarize the results, but to emphasize the novelty of these insights.}
 

### 2014-07-08 Code/pubplots.R

@@ -11,9 +11,9 @@
 #   lattice = lattice structure - each patch has four accessible neighbors
 
 conn <- "(full)"
-# conn <- "(lattice)"
-# conn <- "(x0)"
-# conn <- "(delta)"
+conn <- "(lattice)"
+conn <- "(x0)"
+conn <- "(delta)"
 
 
 # Metapopulation-level results
@@ -98,25 +98,25 @@ pdf(paste(getwd(), "/Manuscript", "/highlow", conn, ".pdf", sep=""), height=5, w
 par(mfrow = c(1, 3), bty = "l")
 
 S.occ <- tapply(highlow$S, list(highlow$longevity, highlow$treatment), mean)
-S.occ <- S.occ - S.occ[, 4]
+# S.occ <- S.occ - S.occ[, 4]
 
-plot(seq(20, 200, by = 20), S.occ[, 1], type = "b", lty = 1, ylim = c(-0.5, 0.5), pch = 19, col = "red", ylab = "S occupancy", xlab = "Longevity")
+plot(seq(20, 200, by = 20), S.occ[, 1], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", ylab = "S occupancy", xlab = "Longevity")
 lines(seq(20, 200, by = 20), S.occ[, 2], type = "b", lty = 2, pch = 19, col = "blue")
 lines(seq(20, 200, by = 20), S.occ[, 3], type = "b", lty = 3, pch = 19, col = "black")
 lines(seq(20, 200, by = 20), S.occ[, 4], type = "l", lty = 4, pch = 19, col = "grey")
 
 I.occ <- tapply(highlow$I, list(highlow$longevity, highlow$treatment), mean)
-I.occ <- I.occ - I.occ[, 4]
+# I.occ <- I.occ - I.occ[, 4]
 
-plot(seq(20, 200, by = 20), I.occ[, 1], type = "b", lty = 1, ylim = c(-0.5, 0.5), pch = 19, col = "red", ylab = "I occupancy", xlab = "Longevity")
+plot(seq(20, 200, by = 20), I.occ[, 1], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", ylab = "I occupancy", xlab = "Longevity")
 lines(seq(20, 200, by = 20), I.occ[, 2], type = "b", lty = 2, pch = 19, col = "blue")
[…]

### 2014-07-08 Manuscript/manuscript.tex
Results edits
Smaller edits following from edits to methods to try to better
distinguish between the different parameterizations and simulations.

@@
 To further parse out the relative influence of high and low quality patches separately, we performed a second series of simulations where we expanded the range of the uniform quality distribution to skew towards either high or low quality patches (i.e.
-by expanding the quality distribution non-symmetrically about 1).
+by expanding the quality distribution asymmetrically about 1).
 For these simulations, at each of the 10 longevities used above, we simulated 100 metapopulations for each of four patch quality distributions: low variance, (patch qualities ranging 0.75 to 1.25); intermediate variance, high quality (range 0.75 to 1.75); intermediate variance, low quality (range 0.25 to 1.25); and high variance (range 0.25 to 1.75).
@@
 
-The pathogen persisted in 70$\%$ of the 10000 simulations in the baseline model with symmetric patch quality distributions, spreading through the entire susceptible host population (i.e.
-a pandemic) in 37$\%$ of the simulations.
-Pathogen persistence increased substantially with longevity, with intermediate longevities favoring endemic disease, and high longevities facilitating pandemic dynamics (Fig~ \ref{poutcome}).
-The effect of longevity, however, depended on the variance of the patch quality distribution, with low variance metapopulations favoring endemic dynamics for a wider range of longevities (Fig~ \ref{poutcome}a), and high variance metapopulations lowering the longevity required for pandemic dynamics (Fig~ \ref{poutcome}b).
-The three alternative parameterizations (lattice structure, $\alpha = 0$, $\delta = 0.1$) produced qualitatively similar results, though pathogen spread in these models was generally more difficult (i.e.
-fewer pandemics), and the habitat quality variance exerted a relatively weaker influence on the epidemiological outcome (Figs~\ref{poutcome_lattice}, \ref{poutcome_x0}, \ref{poutcome_delta}).
+In the baseline parameterization, pathogen persistence increased substantially with longevity, with intermediate longevities favoring endemic disease, and high longevities facilitating pandemic dynamics (Fig~ \ref{poutcome}).
+The effect of longevity, however, depended on the shape of the patch quality distribution.
+In the simulations with patch quality distributions symmetric around 1, low variance metapopulations favored endemic dynamics for a wider range of longevities (Fig~ \ref{poutcome}a), and high variance metapopulations lowered the longevity required for pandemic dynamics (Fig~ \ref{poutcome}b).
+The three alternative parameterizations (models with lattice structure, equal extinction rates, and weak direct infection) produced qualitatively similar results, though pathogen spread in these models was generally more difficult (i.e.
+fewer pandemics), and the patch quality variance exerted a relatively weaker influence on the epidemiological outcome (Figs~\ref{poutcome_lattice}, \ref{poutcome_x0}, \ref{poutcome_delta}).
 
@@
 
-To separate the influence of high and low quality patches, we explored how susceptible and infectious patch occupancy changes when high and low quality patches are included separately in the patch quality distribution.
-Relative to a low-variance metapopulation (patch qualities roughly between 0.75 and 1.25), shifting the distribution toward low quality patches (distribution ranging from 0.25 to 1.25) increased mean susceptible occupancy, while shifting towards high quality patches (range from 0.75 to 1.75) decreased mean susceptible occupancy, with the largest effects observed at intermediate longevities (Fig~\ref{sens}(a)).
+When we explored the influence of high and low quality patches separately (i.e.
+in simulations that expanded the patch quality distribution asymmetrically), we found that, relative to a low-variance metapopulation (patch qualities roughly between 0.75 and 1.25), shifting the distribution to include low quality patches (distribution ranging from 0.25 to 1.25) increased mean susceptible occupancy, while shifting towards high quality patches (range from 0.75 to 1.75) decreased mean susceptible occupancy, with the largest differences observed at intermediate longevities (Fig~\ref{sens}(a)).
 Conversely, a shift towards low quality patches decreased infectious occupancy, while a shift towards high quality patches increased infectious occupancy (Fig~\ref{sens}(b)).
@@
 However, at high pathogen longevities, high quality patch distributions had a net positive effect on total occupancy relative to the other patch quality distributions (Fig~\ref{sens}(c)).
-Due to their lower extinction rates and high connectivity, high quality patches were able to maintain stable occupancy despite the widespread infection facilitated by high longevity.
-On the other hand, when high pathogen longevities led to pandemic dynamics, low quality distributions were unable to support infectious occupancy, resulting in an increase in host extinction events and a decrease in mean occupancy.
+Due to their lower extinction rates and high connectivity, high quality patches were able to maintain stable occupancy despite the widespread infection facilitated by high pathogen longevity.
+On the other hand, though low quality patch distributions increase total occupancy at intermediate longevities, when high pathogen longevities led to pandemic dynamics, low quality distributions were unable to support infectious occupancy, resulting in an increase in host extinction events and a decrease in mean total occupancy.
 The model in which the extinction rate does not scale with quality (i.e.
-$\alpha = 0$) produced very similar results (Fig~\ref{sens_x0}(c)).
-However, in simulations with weak direct infection ($\delta = 0.1$), the high quality patch distributions always produced higher occupancy than low quality patch distributions (Fig~\ref{sens_delta}(c)), while in the model with the lattice structure, low quality patch distributions actually produced higher occupancies than high quality distributions at high longevity (Fig~\ref{sens_lattice}(c)).
+$\alpha = 0$) produced very similar results, though with less pronounced differences between the different quality distributions (Fig~\ref{sens_x0}(c)).
+However, in simulations with weak direct infection ($\delta = 0.1$), the high quality patch distributions always produced higher occupancy than low quality patch distributions (Fig~\ref{sens_delta}(c)), while in the model with the lattice structure, low quality patch distributions actually produced higher occupancies than high quality distributions at high longevities (Fig~\ref{sens_lattice}(c)).
 
@@
 
-To understand the mechanisms behind these effects, we further examined within patch dynamics in high variance metapopulations (with qualities ranging from 0.25 to 1.75) that produced endemic dynamics.
+To understand the mechanisms behind these effects, we further examined within patch dynamics in the baseline model.
+To compare across the full range of patch qualities, and to ensure the opportunity for both suscepible and infectious occupancy, we limit this analysis to high variance metapopulations (with qualities ranging from 0.25 to 1.75) that produced endemic dynamics.
 Because of their lower extinction rate and higher recolonization rate, high quality patches supported more consistent occupancy than low quality patches (Fig~\ref{simvis}).
-However, as the epidemic progressed, high quality patches increasingly supported infectious occupants, while low quality patches supported infectiouss only briefly and infrequently (Fig~\ref{simvis}).
+However, as the epidemic progressed, high quality patches increasingly supported infectious occupants, while low quality patches supported infectious hosts only briefly and infrequently (Fig~\ref{simvis}).
 In addition, while high quality patches did support susceptible occupancy, they experienced a greater number of infection events (susceptible to infectious transitions) than low quality patches (Fig~\ref{infections}).
@@
 \includegraphics[height = 9cm]{simvis.png}
-\caption{Results from a single simulation with variance 0.2 and longevity 80.
+\caption{Results from a single representative simulation with variance 0.2 and longevity 80.
 The top panel shows total occupancy of susceptible (blue) and infectious (red) patches over time.

### 2014-07-10 Manuscript/manuscript.tex
Discussion edits
Initial discussion edits. Removed first paragraph about effect of
longevity.  Changes to network lit discussion.

@@
 
-The positive relationship between pathogen spread and environmental longevity demonstrated in the theoretical model above is well documented in models of empirical systems.
-Using a model parameterized for chronic wasting disease in mule deer, \cite{Almberg2011} find that increasing prion survival in the environment increases the resulting force of infection and peak prevalence, with the importance of reservoir transmission increasing through time.
-Similarly, \cite{Breban2009} show that environmental transmission of avian influenza virus allows it to persist in situations where direct transmission alone is insufficient.
-\textcolor{red}{Other examples.}  Look into George et al (2013).
+By modeling the spatial spread of an environmentally-transmitted pathogen, these simulations suggest that highly heterogeneous metapopulations are more prone to widespread pathogen spread than more homogeneous metapopulations (Fig~\ref{poutcome}).
+In particular, metapopulations with high patch quality variance require lower environmental longevities to experience pandemic dynamics (Fig~\ref{poutcome}).
+The mechanisms underlying this behavior can be explored by drawing parallels with models of pathogen spread through a contact network.
+In this context, our metapopulation can be viewed as a network, with patches as nodes connected through immigration (with the strength of the connection measured by connectivity).
+Many studies have shown that networks with high degree variance (where a node's degree is roughly equivalent to its total connectivity in our metapopulation model) are easily invaded by pathogens, which are then able to spread much more rapidly than in less heterogeneous networks (\cite{Pastor-Satorras2001}).
 
-By modeling the spatial spread of such an environmentally-transmitted pathogen, these simulations suggest that high variance metapopulations are more prone to widespread pathogen spread than low variance metapopulations (Fig~\ref{poutcome}).
-In particular, metapopulations with high patch quality variance require lower environmental longevities to experience pandemic dynamics (Fig~\ref{poutcome}).
-The mechanisms underlying this behavior can be understood by drawing parallels with models of pathogen spread through a contact network.
-In this context, our metapopulation can be viewed as a network with patches as nodes connected through immigration (with the strength of the connection measured by connectivity).
-Many studies have shown that networks with high degree variance (where a node's degree is roughly equivalent to its total connectivity in our metapopulation model) are easily invaded by pathogens, which are then able to spread much more rapidly than in less heterogeneous networks (\cite{Pastor-Satorras2001}).
-However, as the alternative parameterizations show, the overall effect of patch quality variance depends on the structure of the metapopulation, the resulting colonization and extinction rates, and the relative balance of the effects of low and high quality patches.
-In the focal model, the dynamics of high variance metapopulations were driven largely by the presence of high quality patches, whereas in the models with the lattice structure and with equal extinction rates across all patches, high and low quality patches were more balanced in their effects, thus canceling each other out as variance increased.
+However, by including environmental transmission, our models suggest that the overall effect of patch quality variance depends on the structure of the metapopulation, the resulting colonization and extinction rates, and the relative balance of the effects of low and high quality patches.
+In the baseline model, the increase in pathogen spread in high variance metapopulations results primarily from the presence of high quality patches and their ability to overwhelm low quality patches at high longevity through a combination of their relative stability, the high connectivity due to the uniform spatial structure, and the relative strength of direct infection.
+The assumptions tested in the other models all weaken these effects -- by constraining movement through the lattice structure; by limiting the influence of direct infection and thus the influence of colonists from infected high quality patches; and by reducing the stability of high quality patches relative to low -- thus preventing low quality patches from being overwhelmed by the high quality patches.
 
-\textcolor{red}{Emphasize own results relative to the contact network literature.
-Should try to avoid the "main model results suggest X, but sensitivity analysis models suggest Y" format and instead try to present a more complete picture/story.
-In the primary model, the increase in pathogen spread from high variance metapopulations likely results from the presence of high quality patches and their ability to overwhelm low quality patches at high longevity through a combination of their relative stability, the high connectivity due to the uniform spatial structure, and the relative strength of direct infection.
-The assumptions tested in the other models all weaken these effects -- by constraining movement through the lattice structure; by limiting the influence of direct infection and thus the influence of colonists from infected high quality patches; and by reducing the stability of high quality patches relative to low -- thus preventing low quality patches from being overwhelmed by the high quality patches.}
-
-The differences observed between the different parameterizations of the model highlight the fact that the effects of patch quality variance, and its interaction with longevity, result from the combined influence of both low and high quality patches in pathogen spread and metapopulation stability.
+The differences observed between the different parameterizations of the model highlight the fact that the effects of patch quality variance, and its interaction with longevity, result from the combined influence of both low and high quality patches on pathogen spread and metapopulation stability.
 Since a patch's connectivity is determined by its quality (Eqn~\ref{connectivity}), high quality patches function as highly connected hubs in the metapopulation.
@@
 As a result, low quality patches represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infectious occupancy (Fig~\ref{sens}).
-
-%However, the net effect of low and high quality patches on metapopulation occupancy depends on the relative strength of the different transmission pathways -- direct and environmental -- and on the structure of the metapopulation.
-When there is little pathogen spread (e.g.
-low longevity, or weak direct infection), high quality patch distributions support greater total occupancy than low quality patch distributions
-%
-%At intermediate longevities, the timescale of pathogen decay roughly matches the timescale of the colonization and extinction processes.
-In contrast, for low longevities, when the pathogen reservoir decays much faster than colonizations and extinctions occur, environmental transmission has little effect on disease dynamics and the effect of patch quality is relatively weak.
-Conversely, when the pathogen decays much slower than colonizations and extinctions occur, i.e.
-high longevity, the pathogen reservoir persists on nearly all patches and the effects of patch quality are washed out.
-In these cases, the role of patch quality lies primarily in its effect on extinction rates.
-Due to their lower extinction rates, high quality patches are able to maintain stable occupancy despite widespread (pandemic) infection, while low quality patches are unable to support infectious occupancy, resulting in an increase in host extinction events at the highest longevities.
+\textcolor{red}{Maybe also note that natural mortality on low quality patches is high enough that there's not enough time for susceptibles to to get infected.}
 

### 2014-07-10 Code/pubplots.R

@@ -27,23 +27,30 @@ library(RColorBrewer)
 
 metapop.S <- tapply(out$Sfin, list(out$longevity, out$variance), mean)
 metapop.I <- tapply(out$Ifin, list(out$longevity, out$variance), mean)
+metapop.occ <- tapply(out$Sfin + out$Ifin, list(out$longevity, out$variance), mean)
+maxI <- tapply(out$maxI, list(out$longevity, out$variance), mean)
+
+sd.S <- tapply(out$Sfin, list(out$longevity, out$variance), sd)
+sd.I <- tapply(out$Ifin, list(out$longevity, out$variance), sd)
+sd.maxI <- tapply(out$maxI, list(out$longevity, out$variance), sd)
 
 metapop.pan <- tapply(out$Sfin == 0 & out$Ifin > 0, list(out$longevity, out$variance), sum) / 100
 metapop.end <- tapply(out$Sfin > 0 & out$Ifin > 0, list(out$longevity, out$variance), sum) / 100
 metapop.ext <- tapply(out$Sfin == 0 & out$Ifin == 0, list(out$longevity, out$variance), sum) / 100
 metapop.nd <- tapply(out$Sfin > 0 & out$Ifin == 0, list(out$longevity, out$variance), sum) / 100
 
-p.pan <- levelplot(metapop.pan, col.regions = brewer.pal(9, "Reds"), xlab = "Longevity", ylab = "Variance", 
-                   at = seq(0, 100, by = 12.5), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
-p.end <- levelplot(metapop.end, col.regions = brewer.pal(9, "Reds"), cuts = 8, xlab = "Longevity", ylab = "Variance", 
-                   at = seq(0, 100, by = 12.5), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9))
-p.ext <- levelplot(metapop.ext, col.regions = brewer.pal(9, "Reds"), cuts = 8, xlab = "Longevity", ylab = "Variance",
-                   scales = list(at = c(1, 3, 5, 7, 9))
-p.nd <- levelplot(metapop.nd, col.regions = brewer.pal(9, "Reds"), cuts = 8, xlab = "Longevity", ylab = "Variance",
-                  scales = list(at = c(1, 3, 5, 7, 9))
+cols <- colorRampPalette(brewer.pal(9, "Reds"))(100)
+p.pan <- levelplot(metapop.pan, col.regions = cols, xlab = "Longevity", ylab = "Variance", 
[…]

### 2014-07-10 Code/simvis.r

@@ -24,7 +24,7 @@ initial[51:100]<-"E"
 
 timesteps<-5000
 
-longevity <- 180
+longevity <- 80
 variance <- 0.2
 
 #===============================================================================

### 2014-07-15 Manuscript/manuscript.tex
Results and discussion edits
Edits to results to switch presentation of metapopulation level results
from outcome to occupancy, and some initial edits to discussion to
reflect those changes.

@@
 
-In the baseline parameterization, pathogen persistence increased substantially with longevity, with intermediate longevities favoring endemic disease, and high longevities facilitating pandemic dynamics (Fig~ \ref{poutcome}).
-The effect of longevity, however, depended on the shape of the patch quality distribution.
-In the simulations with patch quality distributions symmetric around 1, low variance metapopulations favored endemic dynamics for a wider range of longevities (Fig~ \ref{poutcome}a), and high variance metapopulations lowered the longevity required for pandemic dynamics (Fig~ \ref{poutcome}b).
-The three alternative parameterizations (models with lattice structure, equal extinction rates, and weak direct infection) produced qualitatively similar results, though pathogen spread in these models was generally more difficult (i.e.
-fewer pandemics), and the patch quality variance exerted a relatively weaker influence on the epidemiological outcome (Figs~\ref{poutcome_lattice}, \ref{poutcome_x0}, \ref{poutcome_delta}).
+In the baseline parameterization, infectious occupancy increased substantially with longevity, while susceptible and total occupancy decreased (Fig~ \ref{poutcome}).
+Increasing the variance of the patch quality distribution around a mean of 1 had relatively little effect on average occupancy (Fig~\ref{poutcome}).
+However, though the mean showed little change, the variance in both susceptible and infected occupancy increased with the variance of the patch quality distribution, especially at intermediate longevities.
+This increase in variance is due to the greater variability in epidemiological outcome in high variance metapopulations, which generate both more pandemics and more disease die-outs than lower variance metapopulations.
+The three alternative parameterizations (models with lattice structure, equal extinction rates, and weak direct infection) produced qualitatively similar results, though pathogen spread in these models was generally more limited (Figs~\ref{poutcome_lattice}, \ref{poutcome_x0}, \ref{poutcome_delta}).
 
@@
 \includegraphics[height=5.5cm]{metapop(full).pdf}
-\caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infectious hosts persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
+\caption{Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
 \label{poutcome}
@@
 As with notes for results and methods, need to be clear about which analyses support which claims, and specifically need to be careful about discussing the different patch quality distributions and distinguishing between the symmetric and asymmetric distributions.}
+
+The distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of environmentally transmitted pathogens and the resulting patterns of host occupancy.
+These impacts are driven largely by the contrasting roles played by low and high quality habitat and their relative balance in the metapopulation.
+
+This work, together with previous studies of disease dynamics in a metapopulation (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}), suggests that there is a general trade-off between forces that facilitate metapopulation stability at the expense of disease spread, and forces that hinder disease spread at the expense of metapopulation stability.
+The position of these trade-offs, and their relative importance depend largely on the strength and structure of pathogen transmission.
+
+These previous studies have focused movement as the mechanism behind this trade-off.
+Shifting the patch quality distribution towards high quality patches in this model is in many ways analogous to increasing movement in the models of Hess, Gog, and Park.
+However, by habitat heterogeneity and metapopulation structure, the model presented here allows us to drill down below the metapopulation level and investigate the role of individual patches.
+In the Park model, movement is homogeneous, and increasing movement increases pathogen exposure equally across all patches, whereas in the model here, movement is heterogeneous and directional.
+In our model, we're actually holding per capita movement rates constant, and changing the structure and direction of that movement.
+
+Though beyond the scope of the current study, future work could more deeply investigate the effect of changing movement rates across different patch quality distributions.
 

### 2014-07-15 Code/pubplots.R

@@ -55,12 +55,28 @@ key <- draw.colorkey(list(col = cols,
                           height = 0.7), 
                      draw = F)
 
-pdf(paste(getwd(), "/Manuscript", "/metapop", conn, ".pdf", sep=""), height=5, width=10)
-
 grid.arrange(p.end, p.pan, key, ncol = 3, widths = c(0.45, 0.45, 0.1))
 
-dev.off()
+cols <- colorRampPalette(brewer.pal(9, "Reds"))(100)
 
+S <- levelplot(metapop.S, col.regions = cols, xlab = "Longevity", ylab = "Variance",
+               at = seq(0, 1, by = 0.01), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+I <- levelplot(metapop.I, col.regions = cols, xlab = "Longevity", ylab = "Variance",
+               at = seq(0, 1, by = 0.01), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+occ <- levelplot(metapop.occ, col.regions = cols, xlab = "Longevity", ylab = "Variance",
+               at = seq(0, 1, by = 0.01), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+
+key <- draw.colorkey(list(col = cols, 
+                          at = seq(0, 1, by = 0.01),
+                          labels = list(at = seq(0, 1, by = 0.25)),
+                          height = 0.7), 
+                     draw = F)
+
+pdf(paste(getwd(), "/Manuscript", "/metapop", conn, ".pdf", sep=""), height=5, width=14)
+
+grid.arrange(S, I, occ, key, ncol = 4, widths = c(0.3, 0.3, 0.3, 0.1))
+
+dev.off()
 #===============================================================================
 
 # Plotting logistic regression surface for prob. of pathogen persistence
@@ -129,6 +145,18 @@ lines(seq(20, 200, by = 20), occ[, 3], type = "b", lty = 3, pch = 19, col = "bla
 lines(seq(20, 200, by = 20), occ[, 4], type = "l", lty = 4, pch = 19, col = "grey")
 
 dev.off()
+
+par(mfrow = c(1, 2))
+
+plot(seq(20, 200, by = 20), maxI[, 1], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", ylab = "Max I", xlab = "Longevity")
[…]

### 2014-07-16 Manuscript/manuscript.tex
Reorganization of discussion
Some reorganization of discussion to put paragraphs on the roles of low
and high quality patches first.  Some new transitions and notes on
additional discussion needed.

@@
 
-\textcolor{red}{Need to emphasize novelty of this work, especially compared to Hess, Gog, and Park.
-Discuss that movement in this model is not only heterogeneous, but also directional and what that means.
-Also consider reorganizing to put the more interesting elements at the front, and get rid of the less interesting parts (i.e.
-the first paragraph).
-As with notes for results and methods, need to be clear about which analyses support which claims, and specifically need to be careful about discussing the different patch quality distributions and distinguishing between the symmetric and asymmetric distributions.}
+The distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of environmentally transmitted pathogens and the resulting patterns of host occupancy.
+These impacts are driven largely by the contrasting effects of low and high quality habitat on pathogen transmission, and their relative balance in the metapopulation.
+\textcolor{red}{Need more here?
+Possibly discussing some of the background colonization, extinction processes that are modified by patch quality, and how generally those interact with environmental transmission.}
 
-The distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of environmentally transmitted pathogens and the resulting patterns of host occupancy.
-These impacts are driven largely by the contrasting roles played by low and high quality habitat and their relative balance in the metapopulation.
-
-This work, together with previous studies of disease dynamics in a metapopulation (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}), suggests that there is a general trade-off between forces that facilitate metapopulation stability at the expense of disease spread, and forces that hinder disease spread at the expense of metapopulation stability.
-The position of these trade-offs, and their relative importance depend largely on the strength and structure of pathogen transmission.
-
-These previous studies have focused movement as the mechanism behind this trade-off.
-Shifting the patch quality distribution towards high quality patches in this model is in many ways analogous to increasing movement in the models of Hess, Gog, and Park.
-However, by habitat heterogeneity and metapopulation structure, the model presented here allows us to drill down below the metapopulation level and investigate the role of individual patches.
-In the Park model, movement is homogeneous, and increasing movement increases pathogen exposure equally across all patches, whereas in the model here, movement is heterogeneous and directional.
-In our model, we're actually holding per capita movement rates constant, and changing the structure and direction of that movement.
-
-Though beyond the scope of the current study, future work could more deeply investigate the effect of changing movement rates across different patch quality distributions.
-
-By modeling the spatial spread of an environmentally-transmitted pathogen, these simulations suggest that highly heterogeneous metapopulations are more prone to widespread pathogen spread than more homogeneous metapopulations (Fig~\ref{poutcome}).
-In particular, metapopulations with high patch quality variance require lower environmental longevities to experience pandemic dynamics (Fig~\ref{poutcome}).
-The mechanisms underlying this behavior can be explored by drawing parallels with models of pathogen spread through a contact network.
-In this context, our metapopulation can be viewed as a network, with patches as nodes connected through immigration (with the strength of the connection measured by connectivity).
-Many studies have shown that networks with high degree variance (where a node's degree is roughly equivalent to its total connectivity in our metapopulation model) are easily invaded by pathogens, which are then able to spread much more rapidly than in less heterogeneous networks (\cite{Pastor-Satorras2001}).
-
-However, by including environmental transmission, our models suggest that the overall effect of patch quality variance depends on the structure of the metapopulation, the resulting colonization and extinction rates, and the relative balance of the effects of low and high quality patches.
-In the baseline model, the increase in pathogen spread in high variance metapopulations results primarily from the presence of high quality patches and their ability to overwhelm low quality patches at high longevity through a combination of their relative stability, the high connectivity due to the uniform spatial structure, and the relative strength of direct infection.
-The assumptions tested in the other models all weaken these effects -- by constraining movement through the lattice structure; by limiting the influence of direct infection and thus the influence of colonists from infected high quality patches; and by reducing the stability of high quality patches relative to low -- thus preventing low quality patches from being overwhelmed by the high quality patches.
-
-The differences observed between the different parameterizations of the model highlight the fact that the effects of patch quality variance, and its interaction with longevity, result from the combined influence of both low and high quality patches on pathogen spread and metapopulation stability.
 Since a patch's connectivity is determined by its quality (Eqn~\ref{connectivity}), high quality patches function as highly connected hubs in the metapopulation.
@@
 
-\textcolor{red}{High qualily patches also experience more infection events because they are able to support susceptibles long enough for them to get infected, whereas susceptibles on low quality patches are probably going to go extinct before they can become infected (though the effect persists when $\alpha = 0$ and is substantially diminished in the lattice model and when $\delta = 0.1$, so it seems that it is driven largely by connectivity and direct infection}
-
 This trap effect, coupled with the relatively low extinction rate on high quality patches helps to create a stable platform from which the pathogen can spread through the rest of the metapopulation.
@@
 Once the pathogen infects high quality patches, the larger number of host colonists produced by these patches allow it to spread to the rest of the metapopulation relatively easily.
-This process helps to maintain infectious occupancy throughout the metapopulation, which in turn feeds back on high quality patches to ensure a steady stream of infectious colonists that help maintain the trap effect.
+This process helps to maintain infectious occupancy throughout the metapopulation, which in turn feeds back on high quality patches to ensure a steady stream of infectious colonists that help maintain the environmental reservoir and the trap effect.
 Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread and infectious occupancy in the metapopulation (Fig~\ref{sens}).
 
-And \cite{Hess1996} who finds that implementing a quarantine on centralized patches reduces pathogen spread.
+The importance of high quality habitat in maintaining pathogen spread suggests that high quality habitat may be a good target for control efforts, e.g.
+quarantine or vaccination.
+This echoes the insights of \cite{Hess1996} who simulated the spread of a directly transmitted pathogen through a metapopulation and found that implementing a quarantine on centralized patches reduced pathogen spread.
+This insight is even more important for environmentally transmitted pathogens.
 
-In contrast, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{sens}).
+In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{sens}).
 Even though individual low quality patches, due to their high extinction rates and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}), the presence of low quality patches increases overall susceptible occupancy relative to a lower variance metapopulation (Fig~\ref{sens}).
@@
 Though this model provides evidence that high quality patches can serve as traps and superspreaders, and low quality patches can serve as susceptible refuges, the effects of these processes on total metapopulation occupancy depend on the strength and structure of the two transmission pathways.
-\textcolor{red}{Need more here?}  In particular, these findings suggest the presence of a trade-off between facilitating metapopulation stability and inhibiting disease spread, with a pathogen's environmental longevity and ease of spread controlling the roles of different quality patches with respect to this trade-off.
+\textcolor{red}{Need more here?}  In particular, these findings suggest the presence of a trade-off between between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.
+The position of these trade-offs, and their relative importance depend largely on the strength and structure of the different avenues of pathogen transmission.
+Broadly, there are three different scenarios: very weak pathogen spread, intermediate/endemic pathogen spread, and easy/pandemic pathogen spread.
 When pathogen spread is limited or slow (e.g.
-due to low longevity, or weak direct transmission), high quality patches are a net benefit to the metapopulation, as their connectivity and stability facilitate greater occupancy.
-
-\textcolor{red}{Maybe lay out the big picture in this paragraph: that there are essentially three different scenarios, very low pathogen spread, intermediate/endemic pathogen spread, and easy/pandemic pathogen spread and then go through each one and its consequences on the roles played by high and low quality habitat.}
+due to low longevity, or weak direct transmission), high quality patches are a net benefit to the metapopulation, as their connectivity and stability facilitate greater occupancy with little additional pathogen spread.
 
@@
 
-\textcolor{red}{What does this model offer beyond the literature cited?
-WRT Hess 1996, the extension to explicit habitat patches offers insight into the roles of specific patches; even though connectivity is an important driving factor in this model, it also allows us to consider the role of other processes (e.g.
-extinction, explicity pathogen transmission in space).}
+Also discuss \cite{Sharp2011}, who show that management actions that increase carrying capacity can increase disease spread and destabilize population
 
-Also discuss \cite{Sharp2011}, who show that management actions that increase carrying capacity can increase disease spread and destabilize population.
-
-However, as pathogen longevity increases, the trade-off shifts and the addition of high quality patches is a net benefit to overall occupancy, reflecting the findings of \cite{Gog2002}, who find that under background infection from an alternative host, increasing movement rarely reduces occupancy, despite facilitating pathogen spread.
+As pathogen longevity increases, the trade-off shifts and the addition of high quality patches is a net benefit to overall occupancy, reflecting the findings of \cite{Gog2002}, who find that under background infection from an alternative host, increasing movement rarely reduces occupancy, despite facilitating pathogen spread.
 \cite{Park2012} further develops this work by noting that the magnitude of the detrimental effects of increased moevement on occupancy depends on the relative strength of different transmission pathways.
@@
 
+\textcolor{red}{What does this model offer beyond the literature cited?
+WRT Hess 1996, the extension to explicit habitat patches offers insight into the roles of specific patches; even though connectivity is an important driving factor in this model, it also allows us to consider the role of other processes (e.g.
+extinction, explicity pathogen transmission in space).}
+
 \textcolor{red}{As with the notes above, emphasize not just how this work fits with other literature, but also what it adds.
@@
 "This refelcts the the results of blah in that ..., but offers the additional insight that ..."}
+
+These previous studies have focused movement as the mechanism behind this trade-off.
+Shifting the patch quality distribution towards high quality patches in this model is in many ways analogous to increasing movement in the models of Hess, Gog, and Park.
+However, by habitat heterogeneity and metapopulation structure, the model presented here allows us to drill down below the metapopulation level and investigate the role of individual patches.
+In the Park model, movement is homogeneous, and increasing movement increases pathogen exposure equally across all patches, whereas in the model here, movement is heterogeneous and directional.
+In our model, we're actually holding per capita movement rates constant, and changing the structure and direction of that movement.
 

### 2014-07-22 Manuscript/manuscript.tex
Discussion edits
Restructuring and added discussion relating to Hess, Gog, and Park
papers.

@@
 Many, if not all, of the pathogens listed above affect spatially structured host populations (e.g.
-plague in prairie dog colonies, cite George et al 2013), and thus understanding how environmental transmission interacts with patterns of habitat quality is critical to managing disease in these systems.
-In this study then, we seek to explore how a pathogen's environmental longevity (how long it can persist and remain infections in the environment), and habitat heterogeneity (the variance in the distribution of patch quality in a metapopulation) interact to influence pathogen spread and host occupancy, with a specific interest in the roles played by high and low quality habitat.
+plague in prairie dog colonies, \cite{George2013}), and thus understanding how environmental transmission interacts with patterns of habitat quality is critical to managing disease in these systems.
+In this study then, we seek to explore how a pathogen's environmental longevity (how long it can persist and remain infections in the environment), and the distribution of habitat quality in a metapopulation interact to influence pathogen spread and host occupancy, with a specific interest in the roles played by high and low quality habitat.
 
@@
 \caption{State transitions and their rates for patch $i$ in a metapopulation simulation.
-$S$ denotes occupied by  susceptible hosts, $I$ denotes occupied infectious hosts, and $\emptyset$ denotes unoccupied by the host (but potentially with pathogen reservoir with infection rate of $\gamma_i$).}
+$S$ denotes occupied by  susceptible hosts, $I$ denotes occupied infectious hosts, and $\emptyset$ denotes unoccupied by the host.
+All patches are characterized by an environmental infection rate, $\gamma_i$, which is non-zero only if the patch has been previously occupied by infectious hosts.}
 \begin{tabular}{lr}
@@
 
-Susceptible populations are infected via direct contact with infectious colonists at rate $\delta C_{Ii}$, where $\delta$ is the transmission probability.
+Susceptible populations are infected via direct contact with infectious colonists at rate $\delta C_{Ii}$, where $\delta$ is the direct transmission probability.
 In addition, environmental transmission can take place when a susceptible population occupies a previously infected patch, and occurs at rate:
@@
 The baseline model above assumes that all patches are equally accessible from all other patches (i.e.
-are separated by distance 1), so to examine the effects of a more rigid spatial structure, we implemented a model with patches arranged in a square lattice such that each patch is only accessible from its four neighboring patches (Lattice, Table~\ref{params}).
-In addition, we explored a model with $\alpha = 0$ so that the extinction rate was constant with patch quality, simulating the effects of high environmental stochasticity (i.e.
-environmental stochasticity is strong enough that the population size of a patch does not affects its extinction rate)(Equal extinction, Table~\ref{params}).
+are separated by distance 1), so to examine the effects of a more rigid spatial structure, we also implemented a model with patches arranged in a square lattice such that each patch is only accessible from its four neighboring patches (Lattice, Table~\ref{params}).
+In addition, we explored a model with $\alpha = 0$ so that the extinction rate was constant with patch quality, simulating the effects of high environmental stochasticity, i.e.
+environmental stochasticity is strong enough that the population size of a patch does not affects its extinction rate (Equal extinction, Table~\ref{params}).
 Lastly, we parameterized a model with $\delta = 0.1$ to reduce the relative importance of direct transmission and explore the disease dynamics when transmission is driven largely by the environmental reservoir (Weak direct infection, Table~\ref{params}).
@@
 
-The above simulations held the mean of patch quality distribution constant, and explored the effect of increasing habitat heterogeneity around that mean (i.e.
+The above simulations held the mean of patch quality distribution constant and explored the effect of increasing habitat heterogeneity around that mean (i.e.
 by expanding the quality distribution symmetrically about 1).
@@
 
-\textcolor{red}{Similar to above edits to methods, need to better distinguish the different analyses and what they're testing.}
-
 In the baseline parameterization, infectious occupancy increased substantially with longevity, while susceptible and total occupancy decreased (Fig~ \ref{poutcome}).
@@
 However, though the mean showed little change, the variance in both susceptible and infected occupancy increased with the variance of the patch quality distribution, especially at intermediate longevities.
-This increase in variance is due to the greater variability in epidemiological outcome in high variance metapopulations, which generate both more pandemics and more disease die-outs than lower variance metapopulations.
+This increase in variance was due to the greater variability in epidemiological outcome in high variance metapopulations, which generated both more pandemics and more disease die-outs than lower variance metapopulations.
 The three alternative parameterizations (models with lattice structure, equal extinction rates, and weak direct infection) produced qualitatively similar results, though pathogen spread in these models was generally more limited (Figs~\ref{poutcome_lattice}, \ref{poutcome_x0}, \ref{poutcome_delta}).
@@
 
-However, at high pathogen longevities, high quality patch distributions had a net positive effect on total occupancy relative to the other patch quality distributions (Fig~\ref{sens}(c)).
+However, despite facilitating pathogen spread, at high pathogen longevities, high quality patch distributions had a net positive effect on total occupancy relative to the other patch quality distributions (Fig~\ref{sens}(c)).
 Due to their lower extinction rates and high connectivity, high quality patches were able to maintain stable occupancy despite the widespread infection facilitated by high pathogen longevity.
-On the other hand, though low quality patch distributions increase total occupancy at intermediate longevities, when high pathogen longevities led to pandemic dynamics, low quality distributions were unable to support infectious occupancy, resulting in an increase in host extinction events and a decrease in mean total occupancy.
-The model in which the extinction rate does not scale with quality (i.e.
+On the other hand, though low quality patch distributions inhibited pathogen spread at intermediate longevities, when high pathogen longevities led to pandemic dynamics, low quality distributions were unable to support infectious occupancy, resulting in an increase in host extinction events and a decrease in mean total occupancy.
+The model in which the extinction rate did not scale with quality (i.e.
 $\alpha = 0$) produced very similar results, though with less pronounced differences between the different quality distributions (Fig~\ref{sens_x0}(c)).
@@
 In addition, while high quality patches did support susceptible occupancy, they experienced a greater number of infection events (susceptible to infectious transitions) than low quality patches (Fig~\ref{infections}).
-This relationship became weaker as longevity increased, due to the fact that as longevity increases and the pathogen spreads more easily, there are fewer susceptibles remaining in the metapopulation to become infectious.
 A similar trend was observed across the other parameterizations, though the observed relationship between patch quality and number of infection events was shallower for the lattice and weak direct infection models (Figs~\ref{infections_lattice}, \ref{infections_x0}, \ref{infections_delta}).
@@
 
-The distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of environmentally transmitted pathogens and the resulting patterns of host occupancy.
-These impacts are driven largely by the contrasting effects of low and high quality habitat on pathogen transmission, and their relative balance in the metapopulation.
-\textcolor{red}{Need more here?
-Possibly discussing some of the background colonization, extinction processes that are modified by patch quality, and how generally those interact with environmental transmission.}
+This work demonstrates that the distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of environmentally transmitted pathogens and the resulting patterns of host occupancy.
+These impacts are driven largely by the contrasting effects of low and high quality habitat, and their relative balance in the metapopulation.
+These contrasting effects derive from differences in colonization and extinction rates, which in turn drive how low and high quality patches interact with and modulate pathogen transmission, either augmenting it in the case of high quality patches, or dampening it in the case of low quality habitat.
 
@@
 Attracting colonists helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).
-This infection pressure on high quality patches results from both transmission from the environmental reservoir and from direct transmission from infectious immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).
+This infection pressure on high quality patches results from both transmission from a persistent environmental reservoir and from direct transmission from infectious immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).
 
@@
 
-The importance of high quality habitat in maintaining pathogen spread suggests that high quality habitat may be a good target for control efforts, e.g.
-quarantine or vaccination.
-This echoes the insights of \cite{Hess1996} who simulated the spread of a directly transmitted pathogen through a metapopulation and found that implementing a quarantine on centralized patches reduced pathogen spread.
-This insight is even more important for environmentally transmitted pathogens.
-
 In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{sens}).
@@
 Because low quality patches are so infrequently colonized, at low to intermediate longevities, the environmental reservoir left behind by these infectious populations generally decays before it has the opportunity to infect new susceptible colonists.
-As a result, low quality patches represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infectious occupancy (Fig~\ref{sens}).
-\textcolor{red}{Maybe also note that natural mortality on low quality patches is high enough that there's not enough time for susceptibles to to get infected.}
+As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infectious occupancy (Fig~\ref{sens}).
 
-Though this model provides evidence that high quality patches can serve as traps and superspreaders, and low quality patches can serve as susceptible refuges, the effects of these processes on total metapopulation occupancy depend on the strength and structure of the two transmission pathways.
-\textcolor{red}{Need more here?}  In particular, these findings suggest the presence of a trade-off between between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.
-The position of these trade-offs, and their relative importance depend largely on the strength and structure of the different avenues of pathogen transmission.
-Broadly, there are three different scenarios: very weak pathogen spread, intermediate/endemic pathogen spread, and easy/pandemic pathogen spread.
-When pathogen spread is limited or slow (e.g.
+These findings, that low quality habitat inhibits disease spread while high quality habitat enhances it, suggest the presence of a trade-off between between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.
+This work thus expands on previous studies of disease spread in a metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).
+Shifting the patch quality distribution towards high quality patches in our model is in many ways analogous to increasing movement in the models of Hess, Gog, and Park, but by including habitat heterogeneity and metapopulation structure, we are able to investigate the role of individual patches and their contribution to total occupancy.
+
+These contributions, and the relative strength of the above trade-offs, depend largely on the strength and structure of the different avenues of pathogen transmission.
+Broadly, there are three different scenarios that emerge from this and the above studies: conditions that support very weak pathogen spread, conditions that facilitate intermediate/endemic pathogen spread, and conditions that produce easy/pandemic pathogen spread.
+When pathogen spread is weak (e.g.
 due to low longevity, or weak direct transmission), high quality patches are a net benefit to the metapopulation, as their connectivity and stability facilitate greater occupancy with little additional pathogen spread.
 
-However, in cases where the pathogen spreads more easily (i.e.
-at an endemic level), the addition of high quality patches functions similarly to increasing movement in the model developed by \cite{Hess1996}, increasing pathogen spread and reducing overall occupancy (Fig~\ref{sens}c).
-In these cases, where pathogen longevity is intermediate, or pathogen spread is limited by metapopulation structure, rather than serving as sources in the metapopulation, high quality patches become sinks.
+However, in cases where the pathogen is able to spread more easily (i.e.
+at an endemic level), the addition of high quality patches functions similarly to increasing movement in the model developed by \cite{Hess1996}: increasing pathogen spread and as a consequence reducing overall occupancy (Fig~\ref{sens}c).
+In these cases, where pathogen longevity is intermediate, or dispersal is limited by metapopulation structure (e.g.
+a lattice), rather than serving as sources in the metapopulation, high quality patches become sinks.
 More importantly, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).
@@
 disease-induced mortality) can lead to cases where focusing on low quality patches is more effective in increasing population growth rate.
-Indeed, in these situations, we see that low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptibles, produce higher occupancy than high quality distributions.
+Indeed, in these situations, our model suggests that low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can produce higher occupancy than high quality distributions.
 
-Also discuss \cite{Sharp2011}, who show that management actions that increase carrying capacity can increase disease spread and destabilize population
+As pathogen longevity increases and environmental reservoirs become more widespread, the trade-off shifts and the addition of high quality patches is again a net benefit to overall occupancy.
+In these situations, environmental reservoirs overwhelm the susceptible refuges on low quality patches, such that they remain contaminated even across infrequent colonization events.
+Under these conditions of widespread infection, high quality patches help to maintain metapopulation stability in the face of disease-induced mortality, thereby increasing total occupancy relative to more balanced or low quality patch distributions.
 
-As pathogen longevity increases, the trade-off shifts and the addition of high quality patches is a net benefit to overall occupancy, reflecting the findings of \cite{Gog2002}, who find that under background infection from an alternative host, increasing movement rarely reduces occupancy, despite facilitating pathogen spread.
-\cite{Park2012} further develops this work by noting that the magnitude of the detrimental effects of increased moevement on occupancy depends on the relative strength of different transmission pathways.
-Specifically, he finds that as environmental transmission becomes stronger (e.g.
-higher longevity) relative to direct transmission, then total occupancy increases monotonically with movement, i.e.
-movement is a net benefit to the metapopulation.
+These dynamics reflect the findings of \cite{Gog2002}, who found that under strong background infection from an alternative host, increasing movement increased total occupancy, despite facilitating pathogen spread.
+\cite{Park2012}, by adding environmental transmission to the mix, found more generally that in cases where background and environmental transmission were strong relative to direct transmission, increasing movement increased total occupancy monotonically, i.e.
+movement was a net benefit to the metapopulation.
 These studies, coupled with the results presented here, suggest that when there is a persistent source of infection throughout the metapopulation (e.g.
-when the pathogen is long-lived in the environment), factors that facilitate metapopulation stability (i.e.
+when the pathogen is long-lived in the environment and direct infection is relatively less important), factors that facilitate metapopulation stability (i.e.
 increased movement, high quality patches) are a greater benefit than factors that inhibit pathogen spread (i.e.
@@
 
-\textcolor{red}{What does this model offer beyond the literature cited?
-WRT Hess 1996, the extension to explicit habitat patches offers insight into the roles of specific patches; even though connectivity is an important driving factor in this model, it also allows us to consider the role of other processes (e.g.
-extinction, explicity pathogen transmission in space).}
-
-\textcolor{red}{As with the notes above, emphasize not just how this work fits with other literature, but also what it adds.
-WRT these other metapop disease modeling papers, may not want to equate increased patch quality with increased movement so easily and instead delve more into what including patch quality offers us.
-"This refelcts the the results of blah in that ..., but offers the additional insight that ..."}
-
-These previous studies have focused movement as the mechanism behind this trade-off.
-Shifting the patch quality distribution towards high quality patches in this model is in many ways analogous to increasing movement in the models of Hess, Gog, and Park.
-However, by habitat heterogeneity and metapopulation structure, the model presented here allows us to drill down below the metapopulation level and investigate the role of individual patches.
-In the Park model, movement is homogeneous, and increasing movement increases pathogen exposure equally across all patches, whereas in the model here, movement is heterogeneous and directional.
-In our model, we're actually holding per capita movement rates constant, and changing the structure and direction of that movement.
-
 Spatial structure and disease spread represent two major factors influencing the dynamics of wildlife populations.
-As we have seen, in cases where the pathogen is able to persist and remain infectious in the environment, these two factors can interact in complex and interesting ways.
+As we have seen, in cases where the pathogen is able to persist and remain infectious in the environment, these two factors can interact in complex ways.
 Failing to consider these interactions and their effect on the roles played by high and low quality habitat risks overlooking the importance of low-quality habitat for the persistence of wildlife populations and the importance of high-quality habitat for the persistence of wildlife disease.
-
-\textcolor{red}{Again, use this last paragraph not just to summarize the results, but to emphasize the novelty of these insights.}
 

### 2014-07-22 Analysis/highlow.analysis.r
Reorganize code
Separated publication plots into own functions to make it easier to
implement changes.; created new separate analysis scripts for each level
of analysis; separated code that actually runs into analysis folder.

@@ -8,6 +8,12 @@
 #   both = high quality variance (low.var+low+high)
 #=================================================================================
 
+conn <- "(full)"
+conn <- "(lattice)"
+conn <- "(x0)"
+conn <- "(delta)"
+
+
 highlow<-read.csv(paste(getwd(), "/Data/highlow", conn, ".csv", sep = ""), header=TRUE)
 
 highlow<-highlow[,-1]

### 2014-07-22 Analysis/metapop.analysis.r

@@ -1,0 +1,96 @@
+conn <- "(full)"
+conn <- "(lattice)"
+conn <- "(x0)"
+conn <- "(delta)"
+
+
+# Metapopulation-level results
+#===============================================================
+
+load(paste(getwd(), "/Data/out", conn, ".Rdata", sep = ""))
+
+library(gridExtra)
+library(lattice)
+library(RColorBrewer)
+
+metapop.S <- tapply(out$Sfin, list(out$longevity, out$variance), mean)
+metapop.I <- tapply(out$Ifin, list(out$longevity, out$variance), mean)
+metapop.occ <- tapply(out$Sfin + out$Ifin, list(out$longevity, out$variance), mean)
+maxI <- tapply(out$maxI, list(out$longevity, out$variance), mean)
+
+sd.S <- tapply(out$Sfin, list(out$longevity, out$variance), sd)
+sd.I <- tapply(out$Ifin, list(out$longevity, out$variance), sd)
+sd.maxI <- tapply(out$maxI, list(out$longevity, out$variance), sd)
+
+metapop.pan <- tapply(out$Sfin == 0 & out$Ifin > 0, list(out$longevity, out$variance), sum) / 100
+metapop.end <- tapply(out$Sfin > 0 & out$Ifin > 0, list(out$longevity, out$variance), sum) / 100
+metapop.ext <- tapply(out$Sfin == 0 & out$Ifin == 0, list(out$longevity, out$variance), sum) / 100
+metapop.nd <- tapply(out$Sfin > 0 & out$Ifin == 0, list(out$longevity, out$variance), sum) / 100
+
+cols <- colorRampPalette(brewer.pal(9, "Reds"))(100)
+p.pan <- levelplot(metapop.pan, col.regions = cols, xlab = "Longevity", ylab = "Variance", 
+                   at = seq(0, 100, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+p.end <- levelplot(metapop.end, col.regions = cols, xlab = "Longevity", ylab = "Variance", 
+                   at = seq(0, 100, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+p.ext <- levelplot(metapop.ext, col.regions = cols, xlab = "Longevity", ylab = "Variance",
+                   at = seq(0, 100, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
[…]

### 2014-07-22 Analysis/patch.analysis.r

@@ -1,0 +1,114 @@
+# Patch level results
+#========================================================
+
+conn <- "(full)"
+conn <- "(lattice)"
+conn <- "(x0)"
+conn <- "(delta)"
+
+load(paste(getwd(), "/Data/out", conn, ".Rdata", sep = ""))
+
+#Filter patch data to include only endemic cases
+patch<-out[which(out$Ifin>0 & out$Sfin > 0), ]
+
+#Filter patch data to include only high variance runs
+patch<-patch[which(abs(patch$variance - 0.2) < 0.01),]
+
+#Plotting proportion of time occupied by susceptibles and infecteds as a function of quality
+
+pdf(paste(dirname(getwd()), "/Manuscript", "/qualityraw", conn, ".pdf", sep=""), height=5, width=10)
+
+par(mfrow=c(1,2),cex=1)
+
+smoothScatter(patch$quality,patch$S,ylab="Proportion of time occupied by susceptibles",xlab="Patch Quality")
+lines(loess.smooth(patch$quality,patch$S),lwd=2,col="red")
+smoothScatter(patch$quality,patch$I,ylab="Proportion of time occupied by infecteds",xlab="Patch Quality")
+lines(loess.smooth(patch$quality,patch$I),lwd=2,col="red")
+
+dev.off()
+
+pdf(paste(dirname(getwd()), "/Manuscript", "/infevents", conn, ".pdf", sep=""), height=5, width=5)
+
+par(mfrow=c(1,1),cex=1)
+smoothScatter(patch$quality,patch$inf.events.tot,ylab="Number of infection events",xlab="Patch Quality")
+lines(loess.smooth(patch$quality,patch$inf.events.tot),lwd=3,col="red")
+
+dev.off()
+
+
+#===============================================================================
+# Figures to show how longevity affects role of quality
+
+# Plotting infection events against quality for range of longevities
+plot(seq(0.25, 1.75, by = 0.1), c(1:16), ylim = c(0, 10), type = "n", xlab = "Quality", ylab = "Infections")
+for(i in c(40, 60, 80, 100)){
+  lines(loess.smooth(patch$quality[patch$longevity == i], patch$inf.events.tot[patch$longevity == i]), type = "b", pch = as.character(i))  
+}
+
+
+# Showing simulation-level variability
+
+# Number of infection events
+library(scales)
+
[…]

### 2014-07-22 Analysis/pubplots.R

@@ -1,0 +1,24 @@
+
+
+models <- c("(full)", "(lattice)", "(x0)", "(delta)")
+
+source(paste(getwd(), "/Code/metapop.plot.r", sep = ""))
+source(paste(getwd(), "/Code/highlow.plot.r", sep = ""))
+source(paste(getwd(), "/Code/patch.plot.r", sep = ""))
+
+
+for(i in 1:length(models)){
+  metapop.plot(models[i])
+}
+
+for(i in 1:length(models)){
+  highlow.plot(models[i])
+}
+
+for(i in 1:length(models)){
+  patch.plot(models[i])
+}
+
+
+
+

### 2014-07-22 Code/highlow.plot.r

@@ -1,0 +1,30 @@
+highlow.plot <- function(conn){
+  
+  highlow<-read.csv(paste(getwd(), "/Data/highlow", conn, ".csv", sep = ""), header=TRUE)
+    
+  S.occ <- tapply(highlow$S, list(highlow$longevity, highlow$treatment), mean)
+  I.occ <- tapply(highlow$I, list(highlow$longevity, highlow$treatment), mean)
+  occ <- tapply(highlow$I + highlow$S, list(highlow$longevity, highlow$treatment), mean)
+  
+  pdf(paste(getwd(), "/Manuscript", "/highlow", conn, ".pdf", sep=""), height=5, width=10)
+  
+  par(mfrow = c(1, 3), bty = "l")
+  plot(seq(20, 200, by = 20), S.occ[, 1], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", ylab = "S occupancy", xlab = "Longevity")
+  lines(seq(20, 200, by = 20), S.occ[, 2], type = "b", lty = 2, pch = 19, col = "blue")
+  lines(seq(20, 200, by = 20), S.occ[, 3], type = "b", lty = 3, pch = 19, col = "black")
+  lines(seq(20, 200, by = 20), S.occ[, 4], type = "l", lty = 4, pch = 19, col = "grey")
+  
+  plot(seq(20, 200, by = 20), I.occ[, 1], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", ylab = "I occupancy", xlab = "Longevity")
+  lines(seq(20, 200, by = 20), I.occ[, 2], type = "b", lty = 2, pch = 19, col = "blue")
+  lines(seq(20, 200, by = 20), I.occ[, 3], type = "b", lty = 3, pch = 19, col = "black")
+  lines(seq(20, 200, by = 20), I.occ[, 4], type = "l", lty = 4, pch = 19, col = "grey")
+  
+  occ <- tapply(highlow$I + highlow$S, list(highlow$longevity, highlow$treatment), mean)
+  
+  plot(seq(20, 200, by = 20), occ[, 1], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", ylab = "Total occupancy", xlab = "Longevity")
+  lines(seq(20, 200, by = 20), occ[, 2], type = "b", lty = 2, pch = 19, col = "blue")
[…]

### 2014-07-22 Code/metapop.plot.r

@@ -1,0 +1,33 @@
+metapop.plot <- function(conn){
+  
+  library(gridExtra)
+  library(lattice)
+  library(RColorBrewer)
+  
+  load(paste(getwd(), "/Data/out", conn, ".Rdata", sep = ""))
+  
+  metapop.S <- tapply(out$Sfin, list(out$longevity, out$variance), mean)
+  metapop.I <- tapply(out$Ifin, list(out$longevity, out$variance), mean)
+  metapop.occ <- tapply(out$Sfin + out$Ifin, list(out$longevity, out$variance), mean)
+  
+  cols <- colorRampPalette(brewer.pal(9, "Reds"))(100)
+  
+  S <- levelplot(metapop.S, col.regions = cols, xlab = "Longevity", ylab = "Variance",
+                 at = seq(0, 1, by = 0.01), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+  I <- levelplot(metapop.I, col.regions = cols, xlab = "Longevity", ylab = "Variance",
+                 at = seq(0, 1, by = 0.01), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+  occ <- levelplot(metapop.occ, col.regions = cols, xlab = "Longevity", ylab = "Variance",
+                   at = seq(0, 1, by = 0.01), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+  
+  key <- draw.colorkey(list(col = cols, 
+                            at = seq(0, 1, by = 0.01),
+                            labels = list(at = seq(0, 1, by = 0.25)),
+                            height = 0.7), 
+                       draw = F)
+  
+  pdf(paste(getwd(), "/Manuscript", "/metapop", conn, ".pdf", sep=""), height=5, width=14)
+  
+  grid.arrange(S, I, occ, key, ncol = 4, widths = c(0.3, 0.3, 0.3, 0.1))
+  
+  dev.off()
+}

### 2014-07-22 Code/patch.plot.r

@@ -1,0 +1,28 @@
+patch.plot <- function(conn){
+  
+  library(scales)
+  
+  load(paste(getwd(), "/Data/out", conn, ".Rdata", sep = ""))
+  
+  #Filter patch data to include only endemic cases
+  patch<-out[which(out$Ifin>0 & out$Sfin > 0), ]
+  
+  #Filter patch data to include only high variance runs
+  patch<-patch[which(abs(patch$variance - 0.2) < 0.01),]
+  
+  reps <- unique(patch$repID)
+  
+  pdf(paste(getwd(), "/Manuscript", "/infevents", conn, ".pdf", sep=""), height=6, width=8)
+  
+  par(mfrow = c(1, 1))
+  
+  plot(patch$quality, patch$inf.events.early, type = "n", ylab = "Number of infection events", xlab = "Patch quality", bty = "l")
+  for(i in 1:length(reps)){
+    sim <- patch[patch$repID == reps[i], ]
+    try(lines(loess.smooth(sim$quality, sim$inf.events.tot), type = "l", col = alpha("gray", 0.3)))
+  }
+  
+  lines(loess.smooth(patch$quality, patch$inf.events.tot), lwd = 2, col = "black")
+ 
+  dev.off()
+}

### 2014-07-22 Code/pubplots.R

@@ -1,267 +1,0 @@
-
-# Code to load and format data (generated by shell and highlow.shell) and generate plots for publication.
-# Plots are generated in pdf format and saved in the adjacent manuscript folder to be incorporated into 
-# the latex file.
-
-# Note that loading the patch data can take up to a few minutes.
-
-
-#Choose the landscape structure to generate plots for:
-#   full = fully connected - all patches equally accessible from all other patches
-#   lattice = lattice structure - each patch has four accessible neighbors
-
-conn <- "(full)"
-conn <- "(lattice)"
-conn <- "(x0)"
-conn <- "(delta)"
-
-
-# Metapopulation-level results
-#===============================================================
-
-load(paste(getwd(), "/Data/out", conn, ".Rdata", sep = ""))
-
-library(gridExtra)
-library(lattice)
-library(RColorBrewer)
-
-metapop.S <- tapply(out$Sfin, list(out$longevity, out$variance), mean)
-metapop.I <- tapply(out$Ifin, list(out$longevity, out$variance), mean)
-metapop.occ <- tapply(out$Sfin + out$Ifin, list(out$longevity, out$variance), mean)
-maxI <- tapply(out$maxI, list(out$longevity, out$variance), mean)
-
-sd.S <- tapply(out$Sfin, list(out$longevity, out$variance), sd)
-sd.I <- tapply(out$Ifin, list(out$longevity, out$variance), sd)
-sd.maxI <- tapply(out$maxI, list(out$longevity, out$variance), sd)
-
-metapop.pan <- tapply(out$Sfin == 0 & out$Ifin > 0, list(out$longevity, out$variance), sum) / 100
-metapop.end <- tapply(out$Sfin > 0 & out$Ifin > 0, list(out$longevity, out$variance), sum) / 100
-metapop.ext <- tapply(out$Sfin == 0 & out$Ifin == 0, list(out$longevity, out$variance), sum) / 100
-metapop.nd <- tapply(out$Sfin > 0 & out$Ifin == 0, list(out$longevity, out$variance), sum) / 100
-
-cols <- colorRampPalette(brewer.pal(9, "Reds"))(100)
-p.pan <- levelplot(metapop.pan, col.regions = cols, xlab = "Longevity", ylab = "Variance", 
[…]

### 2014-07-23 Analysis/highlow.analysis.r
Correcting source file paths in analysis scripts
Set source paths to Code directory, and data loading paths to output
directory.

@@ -1,22 +1,23 @@
-#=================================================================================
-#
-#Code to analyze results from highlow scripts -- sensitivity analyses the remove
-#high or low quality patches.  The treatment names are as follows:
-#  low.var = missing highest and lowest 10 patches (replaced with 1)
-#   low = low quality patches added to mid 
-#   high = high quality patches added to mid 
-#   both = high quality variance (low.var+low+high)
-#=================================================================================
+# Code to analyze results from highlow scripts, in which the patch quality distribution
+# varies asymmetrically around 1.  The four distributions explored are:
+#   low var: patch qualities between 0.75 and 1.25
+#   intermediate var, high quality: 0.75 - 1.75
+#   intermediate var, low quality: 0.25 - 1.25
+#   high var: 0.25 - 1.75
 
-conn <- "(full)"
+# Choose which parameterization to explore
+conn <- "(full)"      
 conn <- "(lattice)"
 conn <- "(x0)"
 conn <- "(delta)"
 
-
-highlow<-read.csv(paste(getwd(), "/Data/highlow", conn, ".csv", sep = ""), header=TRUE)
+# Read in model output
+highlow<-read.csv(paste(getwd(), "/Output/highlow", conn, ".csv", sep = ""), header=TRUE)
 
 highlow<-highlow[,-1]
+
+#===============================================================================
+# Plotting mean occupancy as a function of longevity and treatment
 
 par(mfrow = c(1, 3), bty = "l")
 
@@ -44,10 +45,11 @@ lines(seq(20, 200, by = 20), occ[, 2], type = "b", lty = 2, pch = 19, col = "blu
 lines(seq(20, 200, by = 20), occ[, 3], type = "b", lty = 3, pch = 19, col = "black")
 lines(seq(20, 200, by = 20), occ[, 4], type = "l", lty = 4, pch = 19, col = "grey")
 
-
-#Outcome analysis
+#===============================================================================
+# Plotting proportion of different epidemiological outcomes
 
 par(mfrow = c(1, 4), bty = "l")
+
 pandemic <- tapply(highlow$S == 0 & highlow$I > 0, list(highlow$longevity, highlow$treatment), sum)
 
 plot(seq(20, 200, by = 20), pandemic[, 1], type = "b", lty = 1, ylim = c(0, 100), pch = 20, col = "red", xlab = "Longevity", ylab = "Pandemics")
[…]

### 2014-07-23 Analysis/highlow.shell.r

@@ -8,8 +8,8 @@ library(doSNOW)
 library(foreach)
 library(rlecuyer)
 
-source("highlow.runs.r")
-source("contSPOM(rates).r")
+source("Code/highlow.runs.r")
+source("Code/contSPOM(rates).r")
 
 #Set the number of replicates
 replicates <- 100

### 2014-07-23 Analysis/patch.analysis.r

@@ -1,22 +1,22 @@
-# Patch level results
-#========================================================
+# Patch level results -- scripts to analyze patch-level data from full simulation
+# model.
 
+# Select model parameterization to analyze
 conn <- "(full)"
 conn <- "(lattice)"
 conn <- "(x0)"
 conn <- "(delta)"
 
-load(paste(getwd(), "/Data/out", conn, ".Rdata", sep = ""))
+load(paste(getwd(), "/Output/out", conn, ".Rdata", sep = ""))
 
 #Filter patch data to include only endemic cases
 patch<-out[which(out$Ifin>0 & out$Sfin > 0), ]
 
 #Filter patch data to include only high variance runs
-patch<-patch[which(abs(patch$variance - 0.2) < 0.01),]
+patch<-patch[which(abs(patch$variance - 0.2) < 0.01), ]
 
-#Plotting proportion of time occupied by susceptibles and infecteds as a function of quality
-
-pdf(paste(dirname(getwd()), "/Manuscript", "/qualityraw", conn, ".pdf", sep=""), height=5, width=10)
+#===============================================================================
+#Plotting scatter cloud ofproportion of time occupied by susceptibles and infecteds as a function of quality
 
 par(mfrow=c(1,2),cex=1)
 
@@ -25,16 +25,11 @@ lines(loess.smooth(patch$quality,patch$S),lwd=2,col="red")
 smoothScatter(patch$quality,patch$I,ylab="Proportion of time occupied by infecteds",xlab="Patch Quality")
 lines(loess.smooth(patch$quality,patch$I),lwd=2,col="red")
 
-dev.off()
 
-pdf(paste(dirname(getwd()), "/Manuscript", "/infevents", conn, ".pdf", sep=""), height=5, width=5)
-
+# Plotting scatter cloud of number of infection events
 par(mfrow=c(1,1),cex=1)
 smoothScatter(patch$quality,patch$inf.events.tot,ylab="Number of infection events",xlab="Patch Quality")
 lines(loess.smooth(patch$quality,patch$inf.events.tot),lwd=3,col="red")
-
-dev.off()
-
 
 #===============================================================================
 # Figures to show how longevity affects role of quality
@@ -46,12 +41,12 @@ for(i in c(40, 60, 80, 100)){
 }
 
 
[…]

### 2014-07-23 Analysis/shell.r

@@ -8,8 +8,8 @@ library(foreach)
 library(doSNOW)
 library(rlecuyer)
 
-source("modelrun3.r")
-source("contSPOM(rates).r")
+source("Code/modelrun3.r")
+source("Code/contSPOM(rates).r")
 
 #Set the number of replicates
 replicates <- 100

### 2014-07-23 Analysis/simvis.r

@@ -1,5 +1,7 @@
 
-source("contSPOM(rates).r")
+# Code to generate visualization of a single simulation run
+
+source("Code/contSPOM(rates).r")
 
 # Defines fixed inputs for diseaseSPOM
 parms<-data.frame("im"=0.5, #Controls how a patch imports colonists

### 2014-07-23 Analysis/highlow.shell.r
Rename parameters to match manuscript
Rename scaling exponents to match notation used in manuscript.

@@ -21,16 +21,14 @@ longevity <- c(20, 100, 200)
 treatment <- expand.grid(rep(c("+high","+low","full","low.var"), replicates), longevity)
 
 #Defines fixed inputs for diseaseSPOM
-parms<-data.frame("im"=0.5, 
-                  "b"=0.5, 
-                  "D"=5, 
-                  "x"=1,
-                  "es"=0.1,
-                  "ei"=0.5,
-                  "delta"=0.5,
-                  "gamma0"=0.5,
-                  "xi"=1,
-                  "alpha"=0)
+parms<-data.frame("xi_im" = 0.5, 
+                  "xi_em" = 0.5,      
+                  "D" = 5, 
+                  "alpha" = 1,
+                  "es" = 0.1,
+                  "ei" = 0.5,
+                  "delta" = 0.5,
+                  "gamma0" = 0.5)
 
 # Connectivity matrix
 

### 2014-07-23 Analysis/shell.r

@@ -32,18 +32,15 @@ simID<-c(1:length(par.reps[,1]))
 par.reps<-cbind(par.reps,simID)
 
 
-
 #Defines fixed inputs for diseaseSPOM
-parms<-data.frame("im"=0.5, #Controls how a patch imports colonists
-                  "b"=0.5,  #Controls how a patch exports colonists      
-                  "D"=5, 
-                  "y"=10,
-                  "x"=1,
-                  "es"=0.1,
-                  "ei"=0.5,
-                  "delta"=0.5,
-                  "gamma0"=0.5,
-                  "alpha"=0)
+parms<-data.frame("xi_im" = 0.5, 
+                  "xi_em" = 0.5,      
+                  "D" = 5, 
+                  "alpha" = 1,
+                  "es" = 0.1,
+                  "ei" = 0.5,
+                  "delta" = 0.5,
+                  "gamma0" = 0.5)
   
 #Fully connected matrix
 distance<-matrix(1,nrow=100,ncol=100)

### 2014-07-23 Analysis/simvis.r

@@ -4,16 +4,14 @@
 source("Code/contSPOM(rates).r")
 
 # Defines fixed inputs for diseaseSPOM
-parms<-data.frame("im"=0.5, #Controls how a patch imports colonists
-                  "b"=0.5,  #Controls how a patch exports colonists      
-                  "D"=5, 
-                  "y"=10,
-                  "x"=1,
-                  "es"=0.1,
-                  "ei"=0.5,
-                  "delta"=0.1,
-                  "gamma0"=0.5,
-                  "alpha"=0)
+parms<-data.frame("xi_im" = 0.5, 
+                  "xi_em" = 0.5,      
+                  "D" = 5, 
+                  "alpha" = 1,
+                  "es" = 0.1,
+                  "ei" = 0.5,
+                  "delta" = 0.5,
+                  "gamma0" = 0.5)
 
 #Fully connected matrix
 distance<-matrix(1,nrow=100,ncol=100)

### 2014-07-23 Code/contSPOM(rates).r

@@ -1,202 +1,193 @@
-###############################################################################
-#Stochastic patch occupancy model (continuous time using Gillespie algorithm)
-###############################################################################
 
-#Same model as contSPOM.r with the following changes:
-  #Accepts a steady state threshold
-  #Able to store and output connectivity and colonization probabilities 
-  #Altered to replace patch level loop with vector operations
-
-
-diseaseSPOM<-function(distance, quality, initial, parms, r, timesteps,ss.threshold){
-
-#Inputs:
-#distance = nxn matrix of distances between patches
-#quality = nx1 vector of patch qualities
-#initial = nx1 vector of initial patch states
-#parms = dataframe of parameter values
-          #im = scaling parameter for effect of target patch quality in immigration
-          #b = scaling parameter for how patch quality affects the number of colonists produces
-          #D = inverse of mean dispersal distance
-          #y = minimum connectivity required for colonization prob to equal 0.5 = strength of allee effect
-          #e = extinction probability of unit patch
-          #x = how strongly extinction risk depends on patch quality
-          #theta = factor scaling extinction rate of infected patches
-          #delta = probability of direct infection
-          #gamma0 = probability of reservoir infection immediately following extinction of an infected patch
-          #r = decay rate of reservoirs
-          #alpha = probability of transition from reservoir to empty patch
-          
-im<-parms$im
-b<-parms$b
-D<-parms$D
-x<-parms$x
-es<-parms$es
-ei<-parms$ei
-delta<-parms$delta
-gamma0<-parms$gamma0
-alpha<-parms$alpha
-
+diseaseSPOM<-function(distance, quality, initial, parms, r, timesteps, ss.threshold){
+  # Runs stochastic patch occupancy model
+  #
+  # Args:
[…]

### 2014-07-23 Code/contSPOM(rates).r
Documentation of functions
Improved and updated documentation for model functions.

@@ -1,5 +1,5 @@
 
-diseaseSPOM<-function(distance, quality, initial, parms, r, timesteps, ss.threshold){
+diseaseSPOM <- function(distance, quality, initial, parms, r, timesteps, ss.threshold){
   # Runs stochastic patch occupancy model
   #
   # Args:
@@ -19,175 +19,158 @@ diseaseSPOM<-function(distance, quality, initial, parms, r, timesteps, ss.thresh
   #   timesteps = number of time steps to run
   #   ss.threshold = range within which occupancy can vary over 200 time steps to determine "steady state"
   #                  has been reached
+  #
+  # Returns:
+  #   matrix of dimension timesteps X (n+1) giving the state of each patch at each timestep
   
-  xi_im<-parms$xi_im
-  xi_em<-parms$xi_em
-  D<-parms$D
-  alpha<-parms$alpha
-  es<-parms$es
-  ei<-parms$ei
-  delta<-parms$delta
-  gamma0<-parms$gamma0
+  xi_im <- parms$xi_im
+  xi_em <- parms$xi_em
+  D <- parms$D
+  alpha <- parms$alpha
+  es <- parms$es
+  ei <- parms$ei
+  delta <- parms$delta
+  gamma0 <- parms$gamma0
     
-  n<-length(initial)
+  n <- length(initial)
   
-  #State matrix
-  state<-matrix(nrow=timesteps+1, ncol=n)
-  state[1,]<-initial
+  # State matrix
+  state <- matrix(nrow=timesteps+1, ncol=n)
+  state[1, ] <- initial
   
-  #Event times vector
-  event.times<-vector(mode="numeric",length=timesteps+1)
+  # Event times vector
+  event.times <- vector(mode="numeric", length=timesteps+1)
   
-  #Reservoir creation times
-  ti<-vector(mode="numeric", length=n)
+  # Reservoir creation times
+  ti <- vector(mode="numeric", length=n)
   
-  #Infectivity vector
-  gammas<-vector(mode="numeric",length=n)
+  # Infectivity vector
+  gammas <- vector(mode="numeric", length=n)
   
-  #Pairwise connectivity matrix
-  connectivity<-(quality^xi_im%*%t(quality^xi_em))*exp(-D*distance)
-  connectivity<-connectivity*distance
[…]

### 2014-07-23 Code/highlow.runs.r

@@ -1,9 +1,32 @@
-#############################################################################################
-#Code to run SPOM model with just high, just low, or all quality patches to evaluate the relative
-#contribution of each quality class to pandemic dynamics.
-#############################################################################################
 
 modelrun <- function(longevity, parms, distance, initial, timesteps, treatment){
+  # Takes in longevity and  type of patch quality distribution, runs the SPOM without infection
+  # until steady state, then introduces infection on a randomly selected occupied patch,
+  # calls SPOM model and generates output.
+  #
+  # Args:
+  #   longevity: value for pathogen longevity in environment, half-life of infectivity
+  #   parms: named numeric vector of parameter values
+  #   distance: nxn between patch distance matrix
+  #   initial: character vector giving initial state of each patch
+  #   timesteps: number of timesteps to run SPOM for
+  #   treatment: character value giving type of quality distribution (one of "low.var", "+high", "+low", "full")
+  #
+  # Returns:
+  #   1 x 11 dataframe giving processed results from simulation
+  #     treatment: patch quality distribution used
+  #     longevity: pathogen environmental longevity
+  #     S: final susceptible occupancy
+  #     I: final infectious occupancy
+  #     R: final reservoir occupancy
+  #     avgS: susceptible occupancy averaged over last 100 time steps
+  #     avgI: infectious occupancy averaged over last 100 time steps
+  #     avgR: reservoir occupancy averaged over last 100 time steps
+  #     maxI: maximum infectious occupancy
+  #     tmaxI: time to maximum infectious occupancy
+  #     quality0: quality of initially infected patch
+  
+  names(output) <- c("treatment", "longevity", "S", "I", "R", "avgS", "avgI", "avgR", "maxI", "tmaxI", "quality0")
   
   out.max <- 0.5 * sqrt(12 * 0.2) + 1
   out.min <- 2 - out.max
@@ -26,21 +49,21 @@ modelrun <- function(longevity, parms, distance, initial, timesteps, treatment){
   }
 
   
-  #Converts longevity (half-life) to decay rate
+  # Converts longevity (half-life) to decay rate
   r <- log(2) / longevity
   
-  #Runs simulation without infection until it reaches steady state
+  # Runs simulation without infection until it reaches steady state
[…]

### 2014-07-23 Code/modelrun3.r

@@ -1,19 +1,40 @@
-#############################################################################################
-#modelrun function that takes in longevity and variance values, runs the SPOM without infection
-#until steady state, then introduces infection on a randomly selected occupied patch and runs SPOM for
-#an additional 2000 time steps.  Collects the relevant occupancy data from diseaseSPOM output
-#and writes to csv file. Same as modelrun, but with number of replicates as an input and steady
-#state threshold.
 
-#Changes as of 01/2012
-  #Proportion of time in each state calculations corrected 12/20/11
-  #Number of state transitions for each patch added
-  #Unique simulation ID added to both metapop and patch-level data
-  #Patch quality drawn from uniform instead of gamma distribution
-  #Initial infected patch has to be occupied
-#############################################################################################
-
-modelrun<-function(longevity, variance, k, parms, distance, initial, timesteps){
+modelrun <- function(longevity, variance, k, parms, distance, initial, timesteps){
+  # Takes in longevity and variance values, runs the SPOM without infection
+  # until steady state, then introduces infection on a randomly selected occupied patch,
+  # calls SPOM model and generates output.
+  #
+  # Args:
+  #   longevity: value for pathogen longevity in environment, half-life of infectivity
+  #   variance: variance of patch quality distribution (uniform with mean 1)
+  #   k: replicate ID
+  #   parms: named numeric vector of parameter values
+  #   distance: nxn between patch distance matrix
+  #   initial: character vector giving initial state of each patch
+  #   timesteps: number of timesteps to run SPOM for
+  #
+  # Returns:
+  #   n x 19 dataframe giving patch-level results from simulation
+  #     quality: quality of patch
+  #     tinf: time to first infection
+  #     I: proportion of time patch infected
+  #     S: proportion of time patch susceptible
+  #     E: proportion of time patch empty
+  #     inf.events.early: number of infection events in first fifth of sim
+  #     inf.events.tot: total number of infection events
+  #     susc.col: number of susceptible colonization events
+  #     inf.col: number of infected colonization events
+  #     susc.ex: number of susceptible extinction events
+  #     inf.ex: number of infected extinction events
+  #     Sfin: final susceptible occupancy
+  #     Ifin: final infectious occupancy
+  #     Efin: final empty occupancy
+  #     maxI: maximum infectious occupancy
+  #     quality0: quality of initially infected patch
+  #     repID: replicate ID
+  #     longevity: pathogen environemental longevity
[…]

### 2014-07-23 Analysis/highlow.analysis.r
Rename results
Change "x0" results to "alpha0" to stay consistent with manuscript and
new parameter names.

@@ -8,7 +8,7 @@
 # Choose which parameterization to explore
 conn <- "(full)"      
 conn <- "(lattice)"
-conn <- "(x0)"
+conn <- "(alpha0)"
 conn <- "(delta)"
 
 # Read in model output

### 2014-07-23 Analysis/metapop.analysis.r

@@ -4,7 +4,7 @@
 # Choose model parameterization to explore
 conn <- "(full)"
 conn <- "(lattice)"
-conn <- "(x0)"
+conn <- "(alpha0)"
 conn <- "(delta)"
 
 load(paste(getwd(), "/Output/out", conn, ".Rdata", sep = ""))

### 2014-07-23 Analysis/patch.analysis.r

@@ -4,7 +4,7 @@
 # Select model parameterization to analyze
 conn <- "(full)"
 conn <- "(lattice)"
-conn <- "(x0)"
+conn <- "(alpha0)"
 conn <- "(delta)"
 
 load(paste(getwd(), "/Output/out", conn, ".Rdata", sep = ""))

### 2014-07-23 Analysis/pubplots.R

@@ -1,6 +1,6 @@
 # Script to generate all plots for manuscript
 
-models <- c("(full)", "(lattice)", "(x0)", "(delta)")
+models <- c("(full)", "(lattice)", "(alpha0)", "(delta)")
 
 source("Code/metapop.plot.r")
 source("Code/highlow.plot.r")

### 2014-07-23 Manuscript/manuscript.tex
Updates to figures
small changes to figures, and updating figure captions.

@@
 \includegraphics[height=7cm]{highlow(full).pdf}
-\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy, measured relative to  a low variance metapopulation.
-Red solid lines show hiqh quality metapopulations, blue dashed show low quality, and black shows high variance.}
+\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
+Red solid lines show hiqh quality metapopulations, blue dashed show low quality, black shows high variance, and gray shows low variance.}
 \label{sens}
@@
 \caption{The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
-Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed fit over all simulations.
-The thin black lines show the smoothed fits for longevities of 40 (top) and 140 (bottom).}
+Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
 \label{infections}
@@
 \includegraphics[height=5.5cm]{metapop(lattice).pdf}
-\caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infectiouss persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
+\caption{Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
 \label{poutcome_lattice}
@@
 \includegraphics[height=7cm]{highlow(lattice).pdf}
-\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy, measured relative to  a low variance metapopulation.
-Red solid lines show hiqh quality metapopulations, blue dashed show low quality, and black shows high variance.}
+\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
+Red solid lines show hiqh quality metapopulations, blue dashed show low quality, black shows high variance, and gray shows low variance.}
 \label{sens_lattice}
@@
 \caption{The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
-Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed fit over all simulations.
-The thin black lines show the smoothed fits for longevities of 40 (bottom) and 140 (top).}
+Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
 \label{infections_lattice}
@@
 \centering
-\includegraphics[height=5.5cm]{metapop(x0).pdf}
-\caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infectiouss persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
+\includegraphics[height=5.5cm]{metapop(alpha0).pdf}
+\caption{Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
 \label{poutcome_x0}
@@
 \centering
-\includegraphics[height=7cm]{highlow(x0).pdf}
-\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy, measured relative to  a low variance metapopulation.
-Red solid lines show hiqh quality metapopulations, blue dashed show low quality, and black shows high variance.}
+\includegraphics[height=7cm]{highlow(alpha0).pdf}
+\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
+Red solid lines show hiqh quality metapopulations, blue dashed show low quality, black shows high variance, and gray shows low variance.}
 \label{sens_x0}
@@
 \centering
-\includegraphics[height=7cm]{infevents(x0).pdf}
+\includegraphics[height=7cm]{infevents(alpha0).pdf}
 \caption{The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
-Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed fit over all simulations.
-The thin black lines show the smoothed fits for longevities of 40 (top) and 140 (bottom).}
+Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
 \label{infections_x0}
@@
 \includegraphics[height=5.5cm]{metapop(delta).pdf}
-\caption{Percent of simulations resulting in (a) endemic disease (both susceptibles and infectiouss persist) and (b) pandemic disease (no susceptibles persist), as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
+\caption{Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
 \label{poutcome_delta}
@@
 \includegraphics[height=7cm]{highlow(delta).pdf}
-\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy, measured relative to  a low variance metapopulation.
-Red solid lines show hiqh quality metapopulations, blue dashed show low quality, and black shows high variance.}
+\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
+Red solid lines show hiqh quality metapopulations, blue dashed show low quality, black shows high variance, and gray shows low variance.}
 \label{sens_delta}
@@
 \caption{The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
-Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed fit over all simulations.
-The thin black lines show the smoothed fits for longevities of 60 (bottom) and 120 (top).}
+Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
 \label{infections_delta}

### 2014-07-23 Analysis/simvis.r

@@ -81,7 +81,7 @@ levels(state$color) <- c("white", "red", "white", "blue")
 p1 <- ggplot(state, aes(x = Time, y = Patch)) + geom_tile(fill = state$color, alpha = 0.5) + 
               scale_fill_identity() + 
               scale_x_continuous(expand = c(0, 0)) +
-              scale_y_continuous(expand = c(0, 0)) +
+              scale_y_continuous(expression(Patch ~ Quality ~ symbol('\256')), expand = c(0, 0)) +
               theme_classic()
 
 
@@ -94,8 +94,11 @@ p2 <- ggplot(tseries, aes(x = Time, y = Occupancy, group = class, color = class,
               scale_x_continuous(labels = NULL, expand = c(0, 0)) +
               scale_y_continuous(expand = c(0, 0))
 
+png("Manuscript/simvis.png", width = 836, height = 506)
+
 grid.arrange(p2, p1, ncol = 1, heights = c(0.25, 0.75))
 
+dev.off()
 
 # MoMA version
 p1 + theme(line = element_blank(), text = element_blank())

### 2014-07-23 Code/highlow.plot.r

@@ -1,6 +1,6 @@
 highlow.plot <- function(conn){
   
-  highlow<-read.csv(paste(getwd(), "/Data/highlow", conn, ".csv", sep = ""), header=TRUE)
+  highlow<-read.csv(paste(getwd(), "/Output/highlow", conn, ".csv", sep = ""), header=TRUE)
     
   S.occ <- tapply(highlow$S, list(highlow$longevity, highlow$treatment), mean)
   I.occ <- tapply(highlow$I, list(highlow$longevity, highlow$treatment), mean)
@@ -11,19 +11,19 @@ highlow.plot <- function(conn){
   par(mfrow = c(1, 3), bty = "l")
   plot(seq(20, 200, by = 20), S.occ[, 1], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", ylab = "S occupancy", xlab = "Longevity")
   lines(seq(20, 200, by = 20), S.occ[, 2], type = "b", lty = 2, pch = 19, col = "blue")
-  lines(seq(20, 200, by = 20), S.occ[, 3], type = "b", lty = 3, pch = 19, col = "black")
+  lines(seq(20, 200, by = 20), S.occ[, 3], type = "l", lty = 3, pch = 19, col = "black")
   lines(seq(20, 200, by = 20), S.occ[, 4], type = "l", lty = 4, pch = 19, col = "grey")
   
   plot(seq(20, 200, by = 20), I.occ[, 1], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", ylab = "I occupancy", xlab = "Longevity")
   lines(seq(20, 200, by = 20), I.occ[, 2], type = "b", lty = 2, pch = 19, col = "blue")
-  lines(seq(20, 200, by = 20), I.occ[, 3], type = "b", lty = 3, pch = 19, col = "black")
+  lines(seq(20, 200, by = 20), I.occ[, 3], type = "l", lty = 3, pch = 19, col = "black")
   lines(seq(20, 200, by = 20), I.occ[, 4], type = "l", lty = 4, pch = 19, col = "grey")
   
   occ <- tapply(highlow$I + highlow$S, list(highlow$longevity, highlow$treatment), mean)
   
   plot(seq(20, 200, by = 20), occ[, 1], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", ylab = "Total occupancy", xlab = "Longevity")
[…]

### 2014-07-23 Code/metapop.plot.r

@@ -4,7 +4,7 @@ metapop.plot <- function(conn){
   library(lattice)
   library(RColorBrewer)
   
-  load(paste(getwd(), "/Data/out", conn, ".Rdata", sep = ""))
+  load(paste(getwd(), "/Output/out", conn, ".Rdata", sep = ""))
   
   metapop.S <- tapply(out$Sfin, list(out$longevity, out$variance), mean)
   metapop.I <- tapply(out$Ifin, list(out$longevity, out$variance), mean)

### 2014-07-23 Code/patch.plot.r

@@ -1,8 +1,8 @@
 patch.plot <- function(conn){
   
   library(scales)
-  
-  load(paste(getwd(), "/Data/out", conn, ".Rdata", sep = ""))
+
+  load(paste(getwd(), "/Output/out", conn, ".Rdata", sep = ""))
   
   #Filter patch data to include only endemic cases
   patch<-out[which(out$Ifin>0 & out$Sfin > 0), ]

### 2014-10-15 Manuscript/manuscript.tex
Edits and figure updates
Some edits from Colleen and changes to make figures more clear.

@@
 
-\title{Environmental pathogen reservoirs and habitat heterogeneity in a metapopulation}
+\title{The influence of habitat quality and environmental transmission on host-pathogen metapopulation dynamics}
 
@@
 
-\titlerunning{Pathogen Reservoirs}
+\titlerunning{Habitat quality and environmental transmission}
 
@@
 plague in prairie dog colonies, \cite{George2013}), and thus understanding how environmental transmission interacts with patterns of habitat quality is critical to managing disease in these systems.
-In this study then, we seek to explore how a pathogen's environmental longevity (how long it can persist and remain infections in the environment), and the distribution of habitat quality in a metapopulation interact to influence pathogen spread and host occupancy, with a specific interest in the roles played by high and low quality habitat.
+In this study then, we seek to explore how a pathogen's environmental longevity (how long it can persist and remain infectious in the environment), and the distribution of habitat quality in a metapopulation interact to influence pathogen spread and host occupancy, with a specific interest in the roles played by high and low quality habitat.
 
@@
 
-To address the above questions, we developed a theoretical stochastic patch occupancy model (SPOM) in which each patch can be in one of three possible states: occupied by susceptible hosts (S), occupied by infectious hosts (I), and unoccupied by the host ($\emptyset$).
+To address the above questions, we developed a stochastic patch occupancy model (SPOM) in which each patch can be in one of three possible states: occupied by susceptible hosts (S), occupied by infectious hosts (I), and unoccupied by the host ($\emptyset$).
 State transitions are governed by host colonization, extinction, and infection rates, where a susceptible population can become infectious either through direct contact with infectious immigrants, or through a local pathogen reservoir (Table~\ref{transitions}).
@@
 \end{equation}
-where $e_X$ is the extinction rate of a patch of unit quality in state $X$ (here we assume that $e_I>e_S$ to account for disease-induced mortality), and $\alpha$ controls how extinction rate scales with patch quality.
+where $e_X$ is the extinction rate of a patch of unit quality in state $X$ (here we assume that $e_I \geq e_S$ to account for disease-induced mortality), and $\alpha$ controls how extinction rate scales with patch quality.
 When an infectious population goes extinct, we assume that the hosts leave behind an infectious pathogen reservoir on the patch.
@@
 \begin{table}[h!]   
-\caption{Parameters of SPOM model, their meaning, and the values assigned under different parameterizations.  Empty cells indicate same value as the default parameterization.}
+\caption{Parameters of the SPOM model, their meaning, and the values assigned under different parameterizations.  Empty cells indicate the same value as the default parameterization.}
 \begin{tabular}{l p{5cm} rrrr}
@@
 $\xi_{im}$ & Scaling parameter for effect of target patch quality on immigration & 0.5 & -- & -- & --\\
-$\xi_{em}$ & Scaling parameter for effect of target patch quality in emigration & 0.5 & -- & -- & --\\
+$\xi_{em}$ & Scaling parameter for effect of target patch quality on emigration & 0.5 & -- & -- & --\\
 $D$& Inverse of mean dispersal distance & 5 & 2 & -- & --\\
@@
 In addition, we explored a model with $\alpha = 0$ so that the extinction rate was constant with patch quality, simulating the effects of high environmental stochasticity, i.e.
-environmental stochasticity is strong enough that the population size of a patch does not affects its extinction rate (Equal extinction, Table~\ref{params}).
+environmental stochasticity is strong enough that the population size of a patch does not affect its extinction rate (Equal extinction, Table~\ref{params}).
 Lastly, we parameterized a model with $\delta = 0.1$ to reduce the relative importance of direct transmission and explore the disease dynamics when transmission is driven largely by the environmental reservoir (Weak direct infection, Table~\ref{params}).
@@
 
-For each of the above parameterizations, we performed two different simulation experiments to explore the combined effects of pathogen environmental longevity and habitat quality distribution on disease dynamics and metapopulation occupancy.
+For each of the above parameterizations, we performed two different simulation experiments to explore the combined effects of pathogen longevity and habitat quality on disease dynamics and metapopulation occupancy.
 In the first of these, we investigated the effect of increasing habitat heterogeneity on the dynamics of pathogens with a range of environmental longevities.
@@
 
-The above simulations held the mean of patch quality distribution constant and explored the effect of increasing habitat heterogeneity around that mean (i.e.
+The above simulations held the mean of the patch quality distribution constant and explored the effect of increasing habitat heterogeneity around that mean (i.e.
 by expanding the quality distribution symmetrically about 1).
-To further parse out the relative influence of high and low quality patches separately, we performed a second series of simulations where we expanded the range of the uniform quality distribution to skew towards either high or low quality patches (i.e.
+To further parse out the relative influence of high and low quality patches separately, we performed a second series of simulations where we expanded the range of the uniform quality distribution to favor either high or low quality patches (i.e.
 by expanding the quality distribution asymmetrically about 1).
@@
 Since a patch's connectivity is determined by its quality (Eqn~\ref{connectivity}), high quality patches function as highly connected hubs in the metapopulation.
-Specifically, high quality patches both attract more colonists from other patches and produce more colonists that spread to other patches.
-These two properties have important implications for pathogen spread in the metapopulation.
 Attracting colonists helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).
-This infection pressure on high quality patches results from both transmission from a persistent environmental reservoir and from direct transmission from infectious immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).
+This increased infection pressure results from both transmission from a persistent environmental reservoir and from direct transmission from infectious immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).
 
@@
 the high extinction rate), together with the increased mortality of infectious populations, means that infectious populations do not persist long on low quality patches.
-Because low quality patches are so infrequently colonized, at low to intermediate longevities, the environmental reservoir left behind by these infectious populations generally decays before it has the opportunity to infect new susceptible colonists.
+Because low quality patches are so infrequently colonized, at relatively low to intermediate longevities, the environmental reservoir left behind by these infectious populations generally decays before it has the opportunity to infect new susceptible colonists.
 As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infectious occupancy (Fig~\ref{sens}).
@@
 In these cases, where pathogen longevity is intermediate, or dispersal is limited by metapopulation structure (e.g.
-a lattice), rather than serving as sources in the metapopulation, high quality patches become sinks.
+a lattice), rather than serving as sources that increase metapopulation occupancy, high quality patches become sinks.
 More importantly, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).
@@
 
+\section{Acknowledgements}
+
+CBL would like to thank members of the Webb Lab, past and present, for several rounds of useful feedback.
+This work was supported by USGS Cooperative Agreement 101485, and NSF Graduate Research Fellowship, No.
+DGE-1321845.
+
 \clearpage
@@
 \includegraphics[height=5.5cm]{metapop(lattice).pdf}
-\caption{Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
+\caption{Metapopulation with lattice structure.
+Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
 \label{poutcome_lattice}
@@
 \includegraphics[height=7cm]{highlow(lattice).pdf}
-\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
-Red solid lines show hiqh quality metapopulations, blue dashed show low quality, black shows high variance, and gray shows low variance.}
+\caption{Metapopulation with lattice structure.
+The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
+Red solid lines show hiqh quality metapopulations, blue dashed lines show low quality, black shows high variance, and gray shows low variance.}
 \label{sens_lattice}
@@
 \includegraphics[height=7cm]{infevents(lattice).pdf}
-\caption{The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
-Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
+\caption{Metapopulation with lattice structure.
+The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
+Light gray lines show the smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
 \label{infections_lattice}
@@
 \includegraphics[height=5.5cm]{metapop(alpha0).pdf}
-\caption{Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
+\caption{Extinction rate independent of quality.
+Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
 \label{poutcome_x0}
@@
 \includegraphics[height=7cm]{highlow(alpha0).pdf}
-\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
-Red solid lines show hiqh quality metapopulations, blue dashed show low quality, black shows high variance, and gray shows low variance.}
+\caption{Extinction rate independent of quality.
+The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
+Red solid lines show hiqh quality metapopulations, blue dashed lines show low quality, black shows high variance, and gray shows low variance.}
 \label{sens_x0}
@@
 \includegraphics[height=7cm]{infevents(alpha0).pdf}
-\caption{The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
-Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
+\caption{Extinction rate independent of quality.
+The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
+Light gray lines show the smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
 \label{infections_x0}
@@
 \includegraphics[height=5.5cm]{metapop(delta).pdf}
-\caption{Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
+\caption{Weak direct transmission.
+Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
 \label{poutcome_delta}
@@
 \includegraphics[height=7cm]{highlow(delta).pdf}
-\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
-Red solid lines show hiqh quality metapopulations, blue dashed show low quality, black shows high variance, and gray shows low variance.}
+\caption{Weak direct transmission.
+The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
+Red solid lines show hiqh quality metapopulations, blue dashed lines show low quality, black shows high variance, and gray shows low variance.}
 \label{sens_delta}
@@
 \includegraphics[height=7cm]{infevents(delta).pdf}
-\caption{The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
-Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
+\caption{Weak direct transmission.
+The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
+Light gray lines show the smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
 \label{infections_delta}

### 2014-10-15 Code/highlow.plot.r

@@ -9,22 +9,25 @@ highlow.plot <- function(conn){
   pdf(paste(getwd(), "/Manuscript", "/highlow", conn, ".pdf", sep=""), height=5, width=10)
   
   par(mfrow = c(1, 3), bty = "l")
-  plot(seq(20, 200, by = 20), S.occ[, 1], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", ylab = "S occupancy", xlab = "Longevity")
-  lines(seq(20, 200, by = 20), S.occ[, 2], type = "b", lty = 2, pch = 19, col = "blue")
-  lines(seq(20, 200, by = 20), S.occ[, 3], type = "l", lty = 3, pch = 19, col = "black")
-  lines(seq(20, 200, by = 20), S.occ[, 4], type = "l", lty = 4, pch = 19, col = "grey")
+  plot(seq(20, 200, by = 20), S.occ[, 1], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", 
+       ylab = "S occupancy", xlab = "Longevity", main = title("a", adj = 0))
+  lines(seq(20, 200, by = 20), S.occ[, 2], type = "b", lty = 1, pch = 19, col = "blue")
+  lines(seq(20, 200, by = 20), S.occ[, 3], type = "b", lty = 1, pch = 19, col = "black")
+  lines(seq(20, 200, by = 20), S.occ[, 4], type = "b", lty = 1, pch = 19, col = "grey")
   
-  plot(seq(20, 200, by = 20), I.occ[, 1], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", ylab = "I occupancy", xlab = "Longevity")
-  lines(seq(20, 200, by = 20), I.occ[, 2], type = "b", lty = 2, pch = 19, col = "blue")
-  lines(seq(20, 200, by = 20), I.occ[, 3], type = "l", lty = 3, pch = 19, col = "black")
-  lines(seq(20, 200, by = 20), I.occ[, 4], type = "l", lty = 4, pch = 19, col = "grey")
+  plot(seq(20, 200, by = 20), I.occ[, 1], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", 
+       ylab = "I occupancy", xlab = "Longevity", main = title("b", adj = 0))
+  lines(seq(20, 200, by = 20), I.occ[, 2], type = "b", lty = 1, pch = 19, col = "blue")
[…]

### 2014-10-15 Code/metapop.plot.r

@@ -13,11 +13,14 @@ metapop.plot <- function(conn){
   cols <- colorRampPalette(brewer.pal(9, "Reds"))(100)
   
   S <- levelplot(metapop.S, col.regions = cols, xlab = "Longevity", ylab = "Variance",
-                 at = seq(0, 1, by = 0.01), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+                 at = seq(0, 1, by = 0.01), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)),
+                 main = list("a", just = "left", x = 0.05))
   I <- levelplot(metapop.I, col.regions = cols, xlab = "Longevity", ylab = "Variance",
-                 at = seq(0, 1, by = 0.01), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+                 at = seq(0, 1, by = 0.01), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)),
+                 main = list("b", just = "left", x = 0.05))
   occ <- levelplot(metapop.occ, col.regions = cols, xlab = "Longevity", ylab = "Variance",
-                   at = seq(0, 1, by = 0.01), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+                   at = seq(0, 1, by = 0.01), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)),
+                   main = list("c", just = "left", x = 0.05))
   
   key <- draw.colorkey(list(col = cols, 
                             at = seq(0, 1, by = 0.01),

### 2014-10-16 Manuscript/manuscript.tex
Notational changes
Changing notation so that all parameters are greek letters.  Avoids
confusion between e and exponential, and between E and empty state; also
tweaked so gamma is a function of tau, the time since infection.

@@
 $S$ denotes occupied by  susceptible hosts, $I$ denotes occupied infectious hosts, and $\emptyset$ denotes unoccupied by the host.
-All patches are characterized by an environmental infection rate, $\gamma_i$, which is non-zero only if the patch has been previously occupied by infectious hosts.}
-\begin{tabular}{lr}
-State Transition & Rate \\
+All patches are characterized by an environmental infection rate, $\gamma(\tau_i)$, which is non-zero only if the patch has been previously occupied by infectious hosts.}
+\begin{tabular}{llr}
+State Transition & Process &  Rate \\
 \hline
-$S \rightarrow I$ & $\delta C_{Ii} + \gamma_i$\\
-$S \rightarrow \emptyset $ & $e_{Si}$\\
-$I \rightarrow \emptyset $ & $e_{Ii}$ \\
-$\emptyset \rightarrow S$ & $C_{Si}$\\
-$\emptyset \rightarrow I$ & $C_{Ii}$\\
+$S \rightarrow I$ & Infection (from contact or reservoir) & $\delta \lambda_{Ii} + \gamma(\tau_i)$\\
+$S \rightarrow \emptyset $ & Extinction (of susceptible) & $\epsilon_{Si}$\\
+$I \rightarrow \emptyset $ & Extinction (of infectious) &  $\epsilon_{Ii}$ \\
+$\emptyset \rightarrow S$ & Colonization (by susceptibles) & $\lambda_{Si}$\\
+$\emptyset \rightarrow I$ & Colonization (by infectious) & $\lambda_{Ii}$\\
 \label{transitions}
@@
 \begin{equation}
-C_{Xi}=A_i^{\xi_{im}} \sum_{i\neq j }\phi_jA_j^{\xi_{em}}e^{-D d_{ij}},
+\lambda_{Xi}=A_i^{\xi_{im}} \sum_{i\neq j }\phi_jA_j^{\xi_{em}}\exp(-D d_{ij}),
 \label{connectivity}
@@
 \begin{equation}
-E_{Xi}=\frac{e_X}{A_i^\alpha},
+\epsilon_{Xi}=\frac{\mu_X}{A_i^\alpha},
 \end{equation}
-where $e_X$ is the extinction rate of a patch of unit quality in state $X$ (here we assume that $e_I \geq e_S$ to account for disease-induced mortality), and $\alpha$ controls how extinction rate scales with patch quality.
+where $\mu_X$ is the extinction rate of a patch of unit quality in state $X$ (here we assume that $\mu_I \geq \mu_S$ to account for disease-induced mortality), and $\alpha$ controls how extinction rate scales with patch quality.
 When an infectious population goes extinct, we assume that the hosts leave behind an infectious pathogen reservoir on the patch.
 
-Susceptible populations are infected via direct contact with infectious colonists at rate $\delta C_{Ii}$, where $\delta$ is the direct transmission probability.
+Susceptible populations are infected via direct contact with infectious colonists at rate $\delta \lambda_{Ii}$, where $\delta$ is the direct transmission probability.
 In addition, environmental transmission can take place when a susceptible population occupies a previously infected patch, and occurs at rate:
 \begin{equation}
-\gamma_i=\gamma_0exp(-rt_{Ii}),
+\gamma(\tau_i)=\gamma_0exp(-r\tau_{i}),
 \end{equation}
-where $\gamma_0$ is the initial infection rate of the pathogen reservoir, $t_{Ii}$ is the time since last infectious occupancy ($t_{Ii} = \infty$, and thus $\gamma_i = 0$, if the patch has never been occupied by infectious hosts), and $r$ is the pathogen's decay rate in the environment.
+where $\gamma_0$ is the initial infection rate of the pathogen reservoir, $\tau_{i}$ is the time since last infectious occupancy ($\tau_{i} = \infty$, and thus $\gamma(\tau_i) = 0$, if the patch has never been occupied by infectious hosts), and $r$ is the pathogen's decay rate in the environment.
 
@@
 \hline
-$e_S$ & Extinction rate of unit quality susceptible patch & 0.1 & -- & -- & --\\
-$e_I$ & Extinction rate of unity quality infectious patch & 0.5 & -- & -- & --\\
+$\mu_S$ & Extinction rate of unit quality susceptible patch & 0.1 & -- & -- & --\\
+$\mu_I$ & Extinction rate of unity quality infectious patch & 0.5 & -- & -- & --\\
 $\alpha$ & Strength of environmental stochasticity & 1 & -- & 0 & --\\

### 2014-10-16 Analysis/highlow.analysis.r

@@ -278,7 +278,7 @@ p<-p+facet_grid(.~case)
 p+scale_colour_manual(values=c("control"="black","+high"="blue","+low"="red"),name="",breaks=c("+high","+low"))
 
 
-#==================================================================================================
+#===============================================================================
 
 #Digging into the role of initial quality
 
@@ -287,3 +287,18 @@ lines(loess.smooth(both$quality0,both$S))
 
 plot(both$quality0,both$I)
 lines(loess.smooth(both$quality0,both$I))
+
+#===============================================================================
+
+par(mfrow = c(1, 4))
+
+for(i in c("(full)", "(lattice)", "(alpha0)", "(delta)")){
+  
+  highlow<-read.csv(paste(getwd(), "/Output/highlow", i, ".csv", sep = ""), header=TRUE)
+  occ <- tapply(highlow$I + highlow$S, list(highlow$longevity, highlow$treatment), mean)
+  
+  plot(seq(20, 200, by = 20), occ[, 1] - occ[, 2], type = "b", pch = 19, main = i, bty = "l",
+       ylim = c(-0.2, 0.4), xlab = "longevity", ylab = "High quality occupancy - low quality occupancy")
+  abline(h = 0, col = "red")
+  
+}

### 2014-11-02 Manuscript/manuscript.tex
Results edits
Changes to results and figures; adding in high disease mortality;
highvlow figure; rearranging other figs and updating captions.

@@
 \hline
-$S \rightarrow I$ & Infection (from contact or reservoir) & $\delta \lambda_{Ii} + \gamma(\tau_i)$\\
-$S \rightarrow \emptyset $ & Extinction (of susceptible) & $\epsilon_{Si}$\\
-$I \rightarrow \emptyset $ & Extinction (of infectious) &  $\epsilon_{Ii}$ \\
-$\emptyset \rightarrow S$ & Colonization (by susceptibles) & $\lambda_{Si}$\\
-$\emptyset \rightarrow I$ & Colonization (by infectious) & $\lambda_{Ii}$\\
+$S \rightarrow I$ & Infection (from contact or reservoir) & $\delta C_{Ii} + \gamma(\tau_i)$\\
+$S \rightarrow \emptyset $ & Extinction (of susceptible) & $E_{Si}$\\
+$I \rightarrow \emptyset $ & Extinction (of infectious) &  $E_{Ii}$ \\
+$\emptyset \rightarrow S$ & Colonization (by susceptibles) & $C_{Si}$\\
+$\emptyset \rightarrow I$ & Colonization (by infectious) & $C_{Ii}$\\
 \label{transitions}
@@
 \begin{equation}
-\lambda_{Xi}=A_i^{\xi_{im}} \sum_{i\neq j }\phi_jA_j^{\xi_{em}}\exp(-D d_{ij}),
+C_{Xi}=A_i^{\xi_{im}} \sum_{i\neq j }\phi_jA_j^{\xi_{em}}\exp(-D d_{ij}),
 \label{connectivity}
@@
 \begin{equation}
-\epsilon_{Xi}=\frac{\mu_X}{A_i^\alpha},
+E_{Xi}=\frac{\mu_X}{A_i^\alpha},
 \end{equation}
@@
 \caption{Parameters of the SPOM model, their meaning, and the values assigned under different parameterizations.  Empty cells indicate the same value as the default parameterization.}
-\begin{tabular}{l p{5cm} rrrr}
-Parameter & Interpretation &  Default & Lattice & Equal Ext. & Weak Direct Inf.\\
+\begin{tabular}{l p{5cm} p{1.5cm} p{1.5cm} p{1.5cm} p{1.5cm} p{1.5cm}}
+Parameter & Interpretation &  Default & Local movement & High env. stoch. & Dominant reservoir trans. & High disease mortality \\
 \hline
-$\xi_{im}$ & Scaling parameter for effect of target patch quality on immigration & 0.5 & -- & -- & --\\
-$\xi_{em}$ & Scaling parameter for effect of target patch quality on emigration & 0.5 & -- & -- & --\\
-$D$& Inverse of mean dispersal distance & 5 & 2 & -- & --\\
-$d_{ij}$ & Distance between patch $i$ and $j$ & 1 $\forall i \neq j$ & 1 $\forall i, j$ neighbors & -- & --\\
+$\xi_{im}$ & Scaling parameter for effect of target patch quality on immigration & 0.5 & -- & -- & -- & --\\
+$\xi_{em}$ & Scaling parameter for effect of target patch quality on emigration & 0.5 & -- & -- & --& --\\
+$D$& Inverse of mean dispersal distance & 5 & 2 & -- & -- & --\\
+$d_{ij}$ & Distance between patch $i$ and $j$ & 1 $\forall i \neq j$ & 1 $\forall i, j$ neighbors & -- & -- & --\\
 \hline
-$\mu_S$ & Extinction rate of unit quality susceptible patch & 0.1 & -- & -- & --\\
-$\mu_I$ & Extinction rate of unity quality infectious patch & 0.5 & -- & -- & --\\
-$\alpha$ & Strength of environmental stochasticity & 1 & -- & 0 & --\\
+$\mu_S$ & Extinction rate of unit quality susceptible patch & 0.1 & -- & -- & -- & --\\
+$\mu_I$ & Extinction rate of unity quality infectious patch & 0.5 & -- & -- & -- & 1\\
+$\alpha$ & Strength of environmental stochasticity & 1 & -- & 0 & -- & --\\
 \hline
-$\delta$ & Probability of direct infection & 0.5 & -- & -- & 0.1\\
-$\gamma_0$ & Initial rate of infection from reservoir patch & 0.5 & -- & -- & --
+$\delta$ & Probability of direct infection & 0.5 & -- & -- & 0.1 & --\\
+$\gamma_0$ & Initial rate of infection from reservoir patch & 0.5 & -- & -- & -- & --
 \end{tabular}
@@
 
-The model outlined above was parameterized in four different ways to explore the consequences of different assumptions (Table~\ref{params}).
-The baseline model was parameterized  according to the following assumptions.
+The model outlined above was parameterized in five different ways to explore the consequences of different assumptions (Table~\ref{params}).
+The default model was parameterized  according to the following assumptions.
 (1) High quality patches maintain higher population sizes and therefore produce more colonists than low quality patches and have lower extinction rates than low quality patches ($\xi_{im}=0.5$, $\alpha=1$, \cite{Hanski2003}).
@@
 (3)  Connectivity among patches is determined entirely by quality ($d_{ij}=1$ for all $i \neq j$).
-In addition to these assumptions, parameters were chosen so that without infection, approximately 0.75 of the metapopulation was occupied ($e_S = 0.1$, $D=5$), and a range of epidemiological behaviors -- disease-free, endemic disease, and pandemic disease (i.e.
-all host populations infectious) -- was feasible, depending on the pathogen's environmental longevity ($e_I = 0.5$, $\delta = 0.5$, $\gamma_0 = 0.5$).
+In addition to these assumptions, parameters were chosen so that without infection, approximately 0.75 of the metapopulation was occupied ($\mu_S = 0.1$, $D=5$), and a range of epidemiological behaviors -- disease-free, endemic disease, and pandemic disease (i.e.
+all host populations infectious) -- was feasible, depending on the pathogen's environmental longevity ($\mu_I = 0.5$, $\delta = 0.5$, $\gamma_0 = 0.5$).
 
 The baseline model above assumes that all patches are equally accessible from all other patches (i.e.
-are separated by distance 1), so to examine the effects of a more rigid spatial structure, we also implemented a model with patches arranged in a square lattice such that each patch is only accessible from its four neighboring patches (Lattice, Table~\ref{params}).
+are separated by distance 1), so to examine the effects of a more rigid spatial structure, we also implemented a model with patches arranged in a square lattice such that each patch is only accessible from its four neighboring patches (Local movement, Table~\ref{params}).
 In addition, we explored a model with $\alpha = 0$ so that the extinction rate was constant with patch quality, simulating the effects of high environmental stochasticity, i.e.
-environmental stochasticity is strong enough that the population size of a patch does not affect its extinction rate (Equal extinction, Table~\ref{params}).
-Lastly, we parameterized a model with $\delta = 0.1$ to reduce the relative importance of direct transmission and explore the disease dynamics when transmission is driven largely by the environmental reservoir (Weak direct infection, Table~\ref{params}).
+environmental stochasticity is strong enough that the population size of a patch does not affect its extinction rate (High env.
+stoc., Table~\ref{params}).
+To explore the disease dynamics when transmission is driven largely by the environmental reservoir we parameterized a model with $\delta = 0.1$ to reduce the relative importance of direct transmission (Dominant reservoir trans., Table~\ref{params}).
+Lastly, to simulate a more virulent pathogen, we parameterized a model with $\mu_I = 1$ to substantially increase disease-induced mortality (High disease mortality, Table~\ref{params}).
 
@@
 In the first of these, we investigated the effect of increasing habitat heterogeneity on the dynamics of pathogens with a range of environmental longevities.
-For these simulations, we explored 10 values of pathogen longevity ranging from 20 to 200 time steps, with values representing the half-life of the pathogen's infectivity, and 10 values for the variance of the patch quality distribution, ranging from 0.02 to 0.2.
+For these simulations, we explored 10 values of pathogen longevity ranging from 0.1 to 3 times the expected residence time of a healthy population on a unit quality patch (i.e.
+10 units of time), with values representing the half-life of the pathogen's infectivity and 10 values for the variance of the patch quality distribution, ranging from 0.02 to 0.2.
 For each simulation (100 replicates for each of the 100 longevity-variance combinations), qualities for 100 patches were drawn from a uniform distribution with fixed mean 1 and given variance.
@@
 In the baseline parameterization, infectious occupancy increased substantially with longevity, while susceptible and total occupancy decreased (Fig~ \ref{poutcome}).
-Increasing the variance of the patch quality distribution around a mean of 1 had relatively little effect on average occupancy (Fig~\ref{poutcome}).
-However, though the mean showed little change, the variance in both susceptible and infected occupancy increased with the variance of the patch quality distribution, especially at intermediate longevities.
-This increase in variance was due to the greater variability in epidemiological outcome in high variance metapopulations, which generated both more pandemics and more disease die-outs than lower variance metapopulations.
-The three alternative parameterizations (models with lattice structure, equal extinction rates, and weak direct infection) produced qualitatively similar results, though pathogen spread in these models was generally more limited (Figs~\ref{poutcome_lattice}, \ref{poutcome_x0}, \ref{poutcome_delta}).
-
-\begin{figure}
-\centering
-\includegraphics[height=5.5cm]{metapop(full).pdf}
-\caption{Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
-\label{poutcome}
-\end{figure}
+However, increasing the variance of the patch quality distribution around a mean of 1 had relatively little effect on average occupancy (Fig~\ref{poutcome}).
+The four alternative parameterizations (models with local movement, high environmental stochasticity, dominant reservoir transmission, and high disease-induced mortality) produced qualitatively similar results, though pathogen spread in these models was generally more limited (Figs~\ref{poutcome_lattice}, \ref{poutcome_x0}, \ref{poutcome_delta}, \ref{poutcome_ei}).
 
 When we explored the influence of high and low quality patches separately (i.e.
-in simulations that expanded the patch quality distribution asymmetrically), we found that, relative to a low-variance metapopulation (patch qualities roughly between 0.75 and 1.25), shifting the distribution to include low quality patches (distribution ranging from 0.25 to 1.25) increased mean susceptible occupancy, while shifting towards high quality patches (range from 0.75 to 1.75) decreased mean susceptible occupancy, with the largest differences observed at intermediate longevities (Fig~\ref{sens}(a)).
-Conversely, a shift towards low quality patches decreased infectious occupancy, while a shift towards high quality patches increased infectious occupancy (Fig~\ref{sens}(b)).
-These general trends were consistent across the three other parameterizations (Figs~\ref{sens_lattice}, \ref{sens_x0}, \ref{sens_delta}).
+in simulations that expanded the patch quality distribution asymmetrically), we found that distributions that favor low quality habitat (distribution ranging from 0.25 to 1.25) increased mean susceptible occupancy, while distributions favoring high quality habitat (range from 0.75 to 1.75) decreased mean susceptible occupancy, with the largest differences observed at intermediate longevities (Fig~\ref{sens}(a)).
+Conversely, low quality habitat distributions decreased infectious occupancy, while high quality distributions increased infectious occupancy (Fig~\ref{sens}(b)).
+These general trends were consistent across the four other parameterizations (Figs~\ref{sens_lattice}, \ref{sens_x0}, \ref{sens_delta}, \ref{sens_ei}).
 
-However, despite facilitating pathogen spread, at high pathogen longevities, high quality patch distributions had a net positive effect on total occupancy relative to the other patch quality distributions (Fig~\ref{sens}(c)).
+However, despite facilitating pathogen spread, at high pathogen longevities, high quality patch distributions had a net positive effect on total occupancy relative to low quality distributions (Fig~\ref{sens}(c)).
 Due to their lower extinction rates and high connectivity, high quality patches were able to maintain stable occupancy despite the widespread infection facilitated by high pathogen longevity.
 On the other hand, though low quality patch distributions inhibited pathogen spread at intermediate longevities, when high pathogen longevities led to pandemic dynamics, low quality distributions were unable to support infectious occupancy, resulting in an increase in host extinction events and a decrease in mean total occupancy.
-The model in which the extinction rate did not scale with quality (i.e.
-$\alpha = 0$) produced very similar results, though with less pronounced differences between the different quality distributions (Fig~\ref{sens_x0}(c)).
-However, in simulations with weak direct infection ($\delta = 0.1$), the high quality patch distributions always produced higher occupancy than low quality patch distributions (Fig~\ref{sens_delta}(c)), while in the model with the lattice structure, low quality patch distributions actually produced higher occupancies than high quality distributions at high longevities (Fig~\ref{sens_lattice}(c)).
+
+The other model parameterizations produced a similar qualitative pattern, with the differences between total occupancy on high and low quality patches generally producing a "u" shape as a function of pathogen longevity (Fig~\ref{highvlow}).
+However, the five parameterizations differed in the magnitude and sign of this general relationship.
+High environmental stochasticity (i.e.
+$\alpha = 0$) produced very similar results to the default model, though with a larger positive effect of low quality habitat at intermediate longevity (Fig~\ref{highvlow}(c)).
+In simulations with reservoir transmission the dominant infection pathway ($\delta = 0.1$), the high quality patch distributions always produced occupancy higher than (or equal to ) the low quality patch distributions (Fig~\ref{highvlow}(d)), while in the models with local movement or high disease-induced mortality, low quality patch distributions actually produced higher occupancies than high quality distributions at high longevities (Fig~\ref{highvlow}(b, e)).
 
@@
 \includegraphics[height=7cm]{highlow(full).pdf}
-\caption{The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
-Red solid lines show hiqh quality metapopulations, blue dashed show low quality, black shows high variance, and gray shows low variance.}
+\caption{The effect of the presence of high and low quality patches on (a) median susceptible occupancy, (b) median infectious occupancy, and (c) median total occupancy.
+Red lines show hiqh quality metapopulations (patch qualities roughly between 0.75 and 1.75), blue lines show low quality metapopulations (patch qualities roughly between 0.25 and 1.25).}
 \label{sens}
@@
 
-To understand the mechanisms behind these effects, we further examined within patch dynamics in the baseline model.
+\begin{figure}
+\centering
+\includegraphics[height = 6cm]{highvlow.pdf}
+\caption{Differences between the median total occupancy of high quality patch distributions (quality between 0.75 and 1.75) and low quality patch distributions (quality between 0.25 to 1.25) across the five different model parameterizations.
+(a) default model, (b) local movement, (c) high environmental stochasticity, (d) dominant reservoir transmission, (e) high disease-induced mortality.}
+\label{highvlow}
+\end{figure}
+
+To understand the mechanisms behind these effects, we further examined within patch infection dynamics.
 To compare across the full range of patch qualities, and to ensure the opportunity for both suscepible and infectious occupancy, we limit this analysis to high variance metapopulations (with qualities ranging from 0.25 to 1.75) that produced endemic dynamics.
@@
 In addition, while high quality patches did support susceptible occupancy, they experienced a greater number of infection events (susceptible to infectious transitions) than low quality patches (Fig~\ref{infections}).
-A similar trend was observed across the other parameterizations, though the observed relationship between patch quality and number of infection events was shallower for the lattice and weak direct infection models (Figs~\ref{infections_lattice}, \ref{infections_x0}, \ref{infections_delta}).
+A similar trend was observed across the other parameterizations, though the observed relationship between patch quality and number of infection events was shallower for the lattice and weak direct infection models (Figs~\ref{infections_lattice}, \ref{infections_x0}, \ref{infections_delta}, \ref{infections_ei}).
 
@@
 \includegraphics[height = 9cm]{simvis.png}
-\caption{Results from a single representative simulation with variance 0.2 and longevity 80.
+\caption{Results from a single representative simulation with variance 0.2 and longevity of 0.3 times the expected residence time of a susceptible population on a unit quality patch.
 The top panel shows total occupancy of susceptible (blue) and infectious (red) patches over time.
@@
 
-\subsection{Sensitivity Analyses}
+\subsection{Occupancy results for symmetric quality distributions}
 
-\subsubsection{Lattice structure}
+\begin{figure}[h!]
+\centering
+\includegraphics[height=5.5cm]{metapop(full).pdf}
+\caption{Default model parameterization.
+Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay, measured in units of expected occupancy time of a susceptible population on a unit quality patch) and the variance of the patch quality distribution.}
+\label{poutcome}
+\end{figure}
 
@@
 \includegraphics[height=5.5cm]{metapop(lattice).pdf}
-\caption{Metapopulation with lattice structure.
-Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
+\caption{Metapopulation with lattice structure and local movement.
+Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay, measured in units of expected occupancy time of a susceptible population on a unit quality patch) and the variance of the patch quality distribution.}
 \label{poutcome_lattice}
@@
 
-\begin{figure}
+\begin{figure}[h!]
+\centering
+\includegraphics[height=5.5cm]{metapop(alpha0).pdf}
+\caption{High environmental stochasticity ($\alpha = 0$).
+Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay, measured in units of expected occupancy time of a susceptible population on a unit quality patch) and the variance of the patch quality distribution.}
+\label{poutcome_x0}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[height=5.5cm]{metapop(delta).pdf}
+\caption{Reservoir transmission dominant ($\delta = 0.1$).
+Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay, measured in units of expected occupancy time of a susceptible population on a unit quality patch) and the variance of the patch quality distribution.}
+\label{poutcome_delta}
+\end{figure}
+
+\begin{figure}[h!]
+\centering
+\includegraphics[height=5.5cm]{metapop(ei).pdf}
+\caption{High disease-induced mortality ($\mu_I = 1$).
+Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay, measured in units of expected occupancy time of a susceptible population on a unit quality patch) and the variance of the patch quality distribution.}
+\label{poutcome_ei}
+\end{figure}
+
+\clearpage
+
+\subsection{Occupancy results for asymmetric quality distributions}
+
+\begin{figure}[h!]
 \centering
 \includegraphics[height=7cm]{highlow(lattice).pdf}
-\caption{Metapopulation with lattice structure.
+\caption{Metapopulation with lattice structure and local movement.
 The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
-Red solid lines show hiqh quality metapopulations, blue dashed lines show low quality, black shows high variance, and gray shows low variance.}
+Red lines show high quality metapopulations (patch qualities roughly between 0.75 and 1.75), blue lines show low quality metapopulations (patch qualities roughly between 0.25 and 1.25).}
 \label{sens_lattice}
@@
 \centering
+\includegraphics[height=7cm]{highlow(alpha0).pdf}
+\caption{High environmental stochasticity ($\alpha = 0$).
+The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
+Red lines show high quality metapopulations (patch qualities roughly between 0.75 and 1.75), blue lines show low quality metapopulations (patch qualities roughly between 0.25 and 1.25).}
+\label{sens_x0}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[height=7cm]{highlow(delta).pdf}
+\caption{Reservoir transmission dominant ($\delta = 0.1$).
+The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
+Red lines show high quality metapopulations (patch qualities roughly between 0.75 and 1.75), blue lines show low quality metapopulations (patch qualities roughly between 0.25 and 1.25).}
+\label{sens_delta}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[height=7cm]{highlow(ei).pdf}
+\caption{High disease-induced mortality ($\mu_I = 1$).
+The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
+Red lines show high quality metapopulations (patch qualities roughly between 0.75 and 1.75), blue lines show low quality metapopulations (patch qualities roughly between 0.25 and 1.25).}
+\label{sens_ei}
+\end{figure}
+
+\clearpage
+
+\subsection{Patch-level infection events}
+
+\begin{figure}[h!]
+\centering
 \includegraphics[height=7cm]{infevents(lattice).pdf}
-\caption{Metapopulation with lattice structure.
+\caption{Metapopulation with lattice structure and local movement.
 The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
@@
 
-\clearpage
-
-\subsubsection{Extinction scaling, $\alpha = 0$}
-
-\begin{figure}[h!]
-\centering
-\includegraphics[height=5.5cm]{metapop(alpha0).pdf}
-\caption{Extinction rate independent of quality.
-Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
-\label{poutcome_x0}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[height=7cm]{highlow(alpha0).pdf}
-\caption{Extinction rate independent of quality.
-The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
-Red solid lines show hiqh quality metapopulations, blue dashed lines show low quality, black shows high variance, and gray shows low variance.}
-\label{sens_x0}
-\end{figure}
-
 \begin{figure}
@@
 \includegraphics[height=7cm]{infevents(alpha0).pdf}
-\caption{Extinction rate independent of quality.
+\caption{High environmental stochasticity ($\alpha = 0$).
 The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
@@
 
-\clearpage
-
-\subsubsection{Weak direct transmission, $\delta = 0.1$}
-
-\begin{figure}[h!]
+\begin{figure}
 \centering
-\includegraphics[height=5.5cm]{metapop(delta).pdf}
-\caption{Weak direct transmission.
-Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay) and the variance of the patch quality distribution.}
-\label{poutcome_delta}
+\includegraphics[height=7cm]{infevents(delta).pdf}
+\caption{Reservoir transmission dominant ($\delta = 0.1$).
+The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
+Light gray lines show the smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
+\label{infections_delta}
 \end{figure}
@@
 \centering
-\includegraphics[height=7cm]{highlow(delta).pdf}
-\caption{Weak direct transmission.
-The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
-Red solid lines show hiqh quality metapopulations, blue dashed lines show low quality, black shows high variance, and gray shows low variance.}
-\label{sens_delta}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[height=7cm]{infevents(delta).pdf}
-\caption{Weak direct transmission.
+\includegraphics[height=7cm]{infevents(ei).pdf}
+\caption{High disease-induced mortality ($\mu_I = 1$).
 The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
 Light gray lines show the smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
-\label{infections_delta}
+\label{infections_ei}
 \end{figure}

### 2014-11-02 Analysis/highlow.analysis.r

@@ -293,54 +293,63 @@ lines(loess.smooth(both$quality0,both$I))
 #===============================================================================
 
 par(mfrow = c(1, 5))
+sim <- c("(full)", "(lattice)", "(alpha0)", "(delta)", "(ei)")
+label <- c("a", "b", "c", "d","e")
 
-for(i in c("(full)", "(lattice)", "(alpha0)", "(delta)", "(ei)")){
+highlow<-read.csv(paste(getwd(), "/Output/highlow",sim[1], ".csv", sep = ""), header=TRUE)
+
+longevity <- sort(unique(highlow$longevity))
+
+S.occ <- tapply(highlow$S, list(highlow$longevity, highlow$treatment), median)
+I.occ <- tapply(highlow$I, list(highlow$longevity, highlow$treatment), median)
+occ <- tapply(highlow$I + highlow$S, list(highlow$longevity, highlow$treatment), median)
+
+plot(log10(longevity), occ[, 1] - occ[, 2], type = "b", pch = 19, bty = "l",
+    ylim = c(-0.4, 0.5), xlab = "log(longevity)", ylab = "High quality occupancy - low quality occupancy",
+    main = title(label[1], adj = 0),
+    cex.lab = 1.2)
+
+abline(h = 0, col = "red")
+
+for(i in 2:5){
   
-  highlow<-read.csv(paste(getwd(), "/Output/highlow", i, ".csv", sep = ""), header=TRUE)
+  highlow<-read.csv(paste(getwd(), "/Output/highlow",sim[i], ".csv", sep = ""), header=TRUE)
   
-  S.occ <- tapply(highlow$S, list(highlow$longevity, highlow$treatment), mean)
-  I.occ <- tapply(highlow$I, list(highlow$longevity, highlow$treatment), mean)
-  occ <- tapply(highlow$I + highlow$S, list(highlow$longevity, highlow$treatment), mean)
+  longevity <- sort(unique(highlow$longevity))
   
-  plot(log10(longevity), occ[, 1] - occ[, 2], type = "b", pch = 19, main = i, bty = "l",
-       ylim = c(-0.3, 0.5), xlab = "longevity", ylab = "High quality occupancy - low quality occupancy")
+  S.occ <- tapply(highlow$S, list(highlow$longevity, highlow$treatment), median)
+  I.occ <- tapply(highlow$I, list(highlow$longevity, highlow$treatment), median)
[…]

### 2014-11-02 Analysis/pubplots.R

@@ -1,6 +1,6 @@
 # Script to generate all plots for manuscript
 
-models <- c("(full)", "(lattice)", "(alpha0)", "(delta)")
+models <- c("(full)", "(lattice)", "(alpha0)", "(delta)", "(ei)")
 
 source("Code/metapop.plot.r")
 source("Code/highlow.plot.r")

### 2014-11-02 Code/highlow.plot.r

@@ -1,33 +1,29 @@
 highlow.plot <- function(conn){
   
   highlow<-read.csv(paste(getwd(), "/Output/highlow", conn, ".csv", sep = ""), header=TRUE)
+  
+  longevity <- sort(unique(highlow$longevity))
     
-  S.occ <- tapply(highlow$S, list(highlow$longevity, highlow$treatment), mean)
-  I.occ <- tapply(highlow$I, list(highlow$longevity, highlow$treatment), mean)
-  occ <- tapply(highlow$I + highlow$S, list(highlow$longevity, highlow$treatment), mean)
+  S.occ <- tapply(highlow$S, list(highlow$longevity, highlow$treatment), median)
+  I.occ <- tapply(highlow$I, list(highlow$longevity, highlow$treatment), median)
+  occ <- tapply(highlow$I + highlow$S, list(highlow$longevity, highlow$treatment), median)
   
   pdf(paste(getwd(), "/Manuscript", "/highlow", conn, ".pdf", sep=""), height=5, width=10)
   
   par(mfrow = c(1, 3), bty = "l")
-  plot(rownames(S.occ), S.occ[, 1], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", 
-       ylab = "S occupancy", xlab = "Longevity", main = title("a", adj = 0))
-  lines(rownames(S.occ), S.occ[, 2], type = "b", lty = 1, pch = 19, col = "blue")
-  lines(rownames(S.occ), S.occ[, 3], type = "b", lty = 1, pch = 19, col = "black")
-  lines(rownames(S.occ), S.occ[, 4], type = "b", lty = 1, pch = 19, col = "grey")
+  plot(log10(longevity), S.occ[, 1], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", 
+       ylab = "S occupancy", xlab = "log(longevity)", main = title("a", adj = 0))
+  lines(log10(longevity), S.occ[, 2], type = "b", lty = 1, pch = 19, col = "blue")
   
-  plot(rownames(I.occ), I.occ[, 1], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", 
-       ylab = "I occupancy", xlab = "Longevity", main = title("b", adj = 0))
-  lines(rownames(I.occ), I.occ[, 2], type = "b", lty = 1, pch = 19, col = "blue")
[…]

### 2014-11-02 Code/metapop.plot.r

@@ -6,20 +6,25 @@ metapop.plot <- function(conn){
   
   load(paste(getwd(), "/Output/out", conn, ".Rdata", sep = ""))
   
-  metapop.S <- tapply(out$Sfin, list(out$longevity, out$variance), mean)
-  metapop.I <- tapply(out$Ifin, list(out$longevity, out$variance), mean)
-  metapop.occ <- tapply(out$Sfin + out$Ifin, list(out$longevity, out$variance), mean)
+  labels <- signif(log10(longevity), 2)[c(1, 3, 5, 7, 9)]
+  
+  metapop.S <- tapply(out$Sfin, list(out$longevity, out$variance), median)
+  metapop.I <- tapply(out$Ifin, list(out$longevity, out$variance), median)
+  metapop.occ <- tapply(out$Sfin + out$Ifin, list(out$longevity, out$variance), median)
   
   cols <- colorRampPalette(brewer.pal(9, "Reds"))(100)
   
-  S <- levelplot(metapop.S, col.regions = cols, xlab = "Longevity", ylab = "Variance",
-                 at = seq(0, 1, by = 0.01), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)),
+  S <- levelplot(metapop.S, col.regions = cols, xlab = "log(longevity)", ylab = "Variance",
+                 at = seq(0, 1, by = 0.01), colorkey = F, 
+                 scales = list(at = c(1, 3, 5, 7, 9), x = list(labels = labels)),
                  main = list("a", just = "left", x = 0.05))
-  I <- levelplot(metapop.I, col.regions = cols, xlab = "Longevity", ylab = "Variance",
-                 at = seq(0, 1, by = 0.01), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)),
+  I <- levelplot(metapop.I, col.regions = cols, xlab = "log(longevity)", ylab = "Variance",
+                 at = seq(0, 1, by = 0.01), colorkey = F, 
+                 scales = list(at = c(1, 3, 5, 7, 9), x = list(labels = labels)),
                  main = list("b", just = "left", x = 0.05))
-  occ <- levelplot(metapop.occ, col.regions = cols, xlab = "Longevity", ylab = "Variance",
-                   at = seq(0, 1, by = 0.01), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)),
[…]

### 2014-11-02 Manuscript/manuscript.tex
Discussion edits
Edits to try to synthesize the insights from highvlow and the conditions
when high or low quality is better.

@@
 Since a patch's connectivity is determined by its quality (Eqn~\ref{connectivity}), high quality patches function as highly connected hubs in the metapopulation.
-Attracting colonists helps position high quality patches as the ecological "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).
+Attracting colonists helps position high quality patches as the infectious "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).
 This increased infection pressure results from both transmission from a persistent environmental reservoir and from direct transmission from infectious immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).
@@
 
-These findings, that low quality habitat inhibits disease spread while high quality habitat enhances it, suggest the presence of a trade-off between between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.
+Though there is a consistent trend in the effect of low and high quality habitat on pathogen spread, the net consequences of these effects for metapopulation occupancy vary depending on the structure and dynamics of the metapopulation.
+Specifically, these findings, that low quality habitat inhibits disease spread while high quality habitat enhances it, suggest the presence of a trade-off between between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.
 This work thus expands on previous studies of disease spread in a metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).
@@
 
-These contributions, and the relative strength of the above trade-offs, depend largely on the strength and structure of the different avenues of pathogen transmission.
-Broadly, there are three different scenarios that emerge from this and the above studies: conditions that support very weak pathogen spread, conditions that facilitate intermediate/endemic pathogen spread, and conditions that produce easy/pandemic pathogen spread.
+These contributions, and the relative strength of the above trade-offs, depend largely on the role of connectivity, how it is affected by patch quality, and the resulting strength and structure of the different avenues of pathogen transmission.
+Broadly, we see three different scenarios that roughly correspond to the three sections of the "u"-shaped relationships in Fig~\ref{highvlow}: conditions that support very weak pathogen spread, conditions that facilitate intermediate/endemic pathogen spread, and conditions that produce easy/pandemic pathogen spread.
 When pathogen spread is weak (e.g.
-due to low longevity, or weak direct transmission), high quality patches are a net benefit to the metapopulation, as their connectivity and stability facilitate greater occupancy with little additional pathogen spread.
+due to low longevity; the top left of the "u"), high quality patches are a net benefit to the metapopulation, as their connectivity and stability facilitate greater occupancy with little additional pathogen spread.
 
 However, in cases where the pathogen is able to spread more easily (i.e.
-at an endemic level), the addition of high quality patches functions similarly to increasing movement in the model developed by \cite{Hess1996}: increasing pathogen spread and as a consequence reducing overall occupancy (Fig~\ref{sens}c).
-In these cases, where pathogen longevity is intermediate, or dispersal is limited by metapopulation structure (e.g.
-a lattice), rather than serving as sources that increase metapopulation occupancy, high quality patches become sinks.
+at an endemic level; the trough of the "u"), the addition of high quality patches functions similarly to increasing movement in the model developed by \cite{Hess1996}: increasing pathogen spread and as a consequence reducing overall occupancy (Fig~\ref{sens}c).
+In these cases, the reduction in occupancy from the effect of high quality habitat on pathogen spread outweigh the positive consequences of increased metapopulation connectivity and stability.
+The shift in this trade-off can arise from conditions that change either side of that balance.
+High quality habitat distributions require lower pathogen longevities for pathogen spread to begin to take off, relative to low quality habitat distributions, and thus for intermediate longevities, high quality distributions experience considerably greater pathogen spread and low quality habitat thus produce higher occupancy.
+
+The trade-off can also switch in cases that disrupt the positive benefits of high quality habitat for metapopulation stability, i.e.
+with pathogens with high disease-induced mortality or where dispersal is limited by metapopulation structure.
+In these cases, rather than serving as sources that increase metapopulation occupancy, high quality patches become sinks.
 More importantly, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).
@@
 
-As pathogen longevity increases and environmental reservoirs become more widespread, the trade-off shifts and the addition of high quality patches is again a net benefit to overall occupancy.
+As pathogen longevity increases and environmental transmission becomes more important relative to direct transmission (i.e.
+the top right of the "u") the trade-off shifts back and the addition of high quality patches is again a net benefit to overall occupancy.
 In these situations, environmental reservoirs overwhelm the susceptible refuges on low quality patches, such that they remain contaminated even across infrequent colonization events.
-Under these conditions of widespread infection, high quality patches help to maintain metapopulation stability in the face of disease-induced mortality, thereby increasing total occupancy relative to more balanced or low quality patch distributions.
+Under these conditions of widespread environmental transmission, high quality patches help to maintain metapopulation stability in the face of disease-induced mortality, with little additional pathogen spread, thereby increasing total occupancy relative to more balanced or low quality patch distributions.
 
@@
 decreased movement, low quality patches).
+
+The different scenarios and parameterizations explored here represent a broad sample of the metapopulation dynamics that we might expect to observe in empirical systems.
+The consistent "u"-shaped trend in the relative benefit of low and high quality habitat as a function of pathogen longevity suggest that these trade-offs between metapopulation stability and pathogen spread are likely to be fairly general.
+By evaluating the relative importance of direct versus environmental transmission, along with the relative effect of habitat quality on connectivity and extinction, we can begin to diagnose where empirical systems (e.g.
+plague in prairie dogs, or white nose syndrome in bats) fall on this spectrum and the corresponding consequences for management.
 

### 2014-11-02 Analysis/simvis.r

@@ -24,7 +24,7 @@ initial[51:100]<-"E"
 
 timesteps<-5000
 
-longevity <- 1
+longevity <- 0.3
 variance <- 0.2
 
 #===============================================================================

### 2014-11-10 Manuscript/manuscript.tex
Plotting fixes and ms abstract.

@@
 \maketitle
+
+\begin{abstract} 
+Patterns of habitat quality are expected to play an important role in the dynamics of wildlife metapopulations, with the expectation that high quality habitat serves to stabilize and maintain occupancy.
+However, we might expect that the costs and benefits of different quality habitat might be modulated, or possibly even reversed, by the presence of an environmentally transmitted pathogen.
+We explore this possibility through the development of a stochastic patch occupancy model of host-pathogen dynamics, in which we can vary both the environmental longevity of the pathogen, and the distribution of habitat quality in the host metapopulation.
+\end{abstract}
 
@@
 
-Though there is a consistent trend in the effect of low and high quality habitat on pathogen spread, the net consequences of these effects for metapopulation occupancy vary depending on the structure and dynamics of the metapopulation.
-Specifically, these findings, that low quality habitat inhibits disease spread while high quality habitat enhances it, suggest the presence of a trade-off between between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.
+Though there is a consistent trend in the effect of low and high quality habitat on pathogen spread, the net consequences of these effects for total occupancy vary depending on the structure and dynamics of the metapopulation.
+Specifically, the findings that low quality habitat inhibits disease spread while high quality habitat enhances it suggest the presence of a trade-off between between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.
 This work thus expands on previous studies of disease spread in a metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).
@@
 at an endemic level; the trough of the "u"), the addition of high quality patches functions similarly to increasing movement in the model developed by \cite{Hess1996}: increasing pathogen spread and as a consequence reducing overall occupancy (Fig~\ref{sens}c).
-In these cases, the reduction in occupancy from the effect of high quality habitat on pathogen spread outweigh the positive consequences of increased metapopulation connectivity and stability.
+In these cases, the reduction in occupancy from the effect of high quality habitat on pathogen spread outweighs the positive consequences of increased metapopulation connectivity and stability.
 The shift in this trade-off can arise from conditions that change either side of that balance.
-High quality habitat distributions require lower pathogen longevities for pathogen spread to begin to take off, relative to low quality habitat distributions, and thus for intermediate longevities, high quality distributions experience considerably greater pathogen spread and low quality habitat thus produce higher occupancy.
+High quality habitat distributions require lower pathogen longevities for pathogen spread to begin to take off, relative to low quality habitat distributions.
+Thus, for intermediate longevities, high quality distributions experience considerably greater pathogen spread and lower occupancy than low quality distributions.
 
 The trade-off can also switch in cases that disrupt the positive benefits of high quality habitat for metapopulation stability, i.e.
-with pathogens with high disease-induced mortality or where dispersal is limited by metapopulation structure.
+pathogens with high disease-induced mortality or when dispersal is limited by metapopulation structure.
 In these cases, rather than serving as sources that increase metapopulation occupancy, high quality patches become sinks.
@@
 In these situations, environmental reservoirs overwhelm the susceptible refuges on low quality patches, such that they remain contaminated even across infrequent colonization events.
-Under these conditions of widespread environmental transmission, high quality patches help to maintain metapopulation stability in the face of disease-induced mortality, with little additional pathogen spread, thereby increasing total occupancy relative to more balanced or low quality patch distributions.
+Under these conditions of widespread environmental transmission, high quality patches help to maintain metapopulation stability in the face of disease-induced mortality, thereby increasing total occupancy relative to more balanced or low quality patch distributions.
 
@@
 The consistent "u"-shaped trend in the relative benefit of low and high quality habitat as a function of pathogen longevity suggest that these trade-offs between metapopulation stability and pathogen spread are likely to be fairly general.
-By evaluating the relative importance of direct versus environmental transmission, along with the relative effect of habitat quality on connectivity and extinction, we can begin to diagnose where empirical systems (e.g.
+By evaluating the relative importance of direct versus environmental transmission, along with the effect of habitat quality on connectivity and extinction, we can begin to diagnose where empirical systems (e.g.
 plague in prairie dogs, or white nose syndrome in bats) fall on this spectrum and the corresponding consequences for management.

### 2014-11-10 Analysis/highlow.analysis.r

@@ -304,7 +304,7 @@ S.occ <- tapply(highlow$S, list(highlow$longevity, highlow$treatment), median)
 I.occ <- tapply(highlow$I, list(highlow$longevity, highlow$treatment), median)
 occ <- tapply(highlow$I + highlow$S, list(highlow$longevity, highlow$treatment), median)
 
-plot(log10(longevity), occ[, 1] - occ[, 2], type = "b", pch = 19, bty = "l",
+plot(log10(longevity), occ[, "+high"] - occ[, "+low"], type = "b", pch = 19, bty = "l",
     ylim = c(-0.4, 0.5), xlab = "log(longevity)", ylab = "High quality occupancy - low quality occupancy",
     main = title(label[1], adj = 0),
     cex.lab = 1.2)
@@ -321,7 +321,7 @@ for(i in 2:5){
   I.occ <- tapply(highlow$I, list(highlow$longevity, highlow$treatment), median)
   occ <- tapply(highlow$I + highlow$S, list(highlow$longevity, highlow$treatment), median)
   
-  plot(log10(longevity), occ[, 1] - occ[, 2], type = "b", pch = 19, bty = "l",
+  plot(log10(longevity), occ[, "+high"] - occ[, "+low"], type = "b", pch = 19, bty = "l",
        ylim = c(-0.4, 0.5), xlab = "log(longevity)", ylab = "", yaxt = "n",
        main = title(label[i], adj = 0),
        cex.lab = 1.2)

### 2014-11-10 Analysis/simvis.r

@@ -24,7 +24,7 @@ initial[51:100]<-"E"
 
 timesteps<-5000
 
-longevity <- 0.3
+longevity <- 0.30
 variance <- 0.2
 
 #===============================================================================

### 2014-11-10 Code/highlow.plot.r

@@ -11,19 +11,19 @@ highlow.plot <- function(conn){
   pdf(paste(getwd(), "/Manuscript", "/highlow", conn, ".pdf", sep=""), height=5, width=10)
   
   par(mfrow = c(1, 3), bty = "l")
-  plot(log10(longevity), S.occ[, 1], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", 
+  plot(log10(longevity), S.occ[, "+high"], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", 
        ylab = "S occupancy", xlab = "log(longevity)", main = title("a", adj = 0))
-  lines(log10(longevity), S.occ[, 2], type = "b", lty = 1, pch = 19, col = "blue")
+  lines(log10(longevity), S.occ[, "+low"], type = "b", lty = 1, pch = 19, col = "blue")
   
-  plot(log10(longevity), I.occ[, 1], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", 
+  plot(log10(longevity), I.occ[, "+high"], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", 
        ylab = "I occupancy", xlab = "log(longevity)", main = title("b", adj = 0))
-  lines(log10(longevity), I.occ[, 2], type = "b", lty = 1, pch = 19, col = "blue")
+  lines(log10(longevity), I.occ[, "+low"], type = "b", lty = 1, pch = 19, col = "blue")
   
   occ <- tapply(highlow$I + highlow$S, list(highlow$longevity, highlow$treatment), mean)
   
-  plot(log10(longevity), occ[, 1], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red",
+  plot(log10(longevity), occ[, "+high"], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red",
        ylab = "Total occupancy", xlab = "log(longevity)", main = title("c", adj = 0))
-  lines(log10(longevity), occ[, 2], type = "b", lty = 1, pch = 19, col = "blue")
+  lines(log10(longevity), occ[, "+low"], type = "b", lty = 1, pch = 19, col = "blue")
   
   dev.off()
 }

### 2014-11-10 Code/metapop.plot.r

@@ -4,7 +4,7 @@ metapop.plot <- function(conn){
   library(lattice)
   library(RColorBrewer)
   
-  load(paste(getwd(), "/Output/out", conn, ".Rdata", sep = ""))
+  load(paste(getwd(), "/Output/out", conn, ".RData", sep = ""))
   
   labels <- signif(log10(longevity), 2)[c(1, 3, 5, 7, 9)]
   

### 2014-11-10 Code/patch.plot.r

@@ -2,7 +2,7 @@ patch.plot <- function(conn){
   
   library(scales)
 
-  load(paste(getwd(), "/Output/out", conn, ".Rdata", sep = ""))
+  load(paste(getwd(), "/Output/out", conn, ".RData", sep = ""))
   
   #Filter patch data to include only endemic cases
   patch<-out[which(out$Ifin>0 & out$Sfin > 0), ]

### 2014-11-10 Manuscript/manuscript.tex
Small edits: gov disclaimers, concise sentences.

@@
 %
-\documentclass{svjour3}
+\documentclass{article}
 
@@
 
-\title{The influence of habitat quality and environmental transmission on host-pathogen metapopulation dynamics}
+\title{Modelling the creation of ecological traps by environmentally persistent pathogens}
 
@@
 
-%\institute{}
-%\journalname{}
+\maketitle
 
-\titlerunning{Habitat quality and environmental transmission}
-
-\maketitle
+This draft manuscript is distributed solely for purposes of scientific peer review.
+Its content is deliberative and pre decisional, so it must not be disclosed or released by reviewers.
+Because the manuscript has not yet been approved for publication by the U.S.
+Geological Survey (USGS), it does not represent any official USGS finding or policy.
 
@@
 $S$ denotes occupied by  susceptible hosts, $I$ denotes occupied infectious hosts, and $\emptyset$ denotes unoccupied by the host.
-All patches are characterized by an environmental infection rate, $\gamma(\tau_i)$, which is non-zero only if the patch has been previously occupied by infectious hosts.}
+All patches are characterized by an environmental infection rate, $\gamma(\tau_i)$, where $\tau_i$ is the time since the patch was last occupied by infectious individuals.}
 \begin{tabular}{llr}
@@
 
-When we explored the influence of high and low quality patches separately (i.e.
-in simulations that expanded the patch quality distribution asymmetrically), we found that distributions that favor low quality habitat (distribution ranging from 0.25 to 1.25) increased mean susceptible occupancy, while distributions favoring high quality habitat (range from 0.75 to 1.75) decreased mean susceptible occupancy, with the largest differences observed at intermediate longevities (Fig~\ref{sens}(a)).
+We found that distributions with more low quality habitat (distribution ranging from 0.25 to 1.25) increased mean susceptible occupancy, while distributions with more high quality habitat (range from 0.75 to 1.75) decreased mean susceptible occupancy, with the largest differences observed at intermediate longevities (Fig~\ref{sens}(a)).
 Conversely, low quality habitat distributions decreased infectious occupancy, while high quality distributions increased infectious occupancy (Fig~\ref{sens}(b)).
@@
 
-Since a patch's connectivity is determined by its quality (Eqn~\ref{connectivity}), high quality patches function as highly connected hubs in the metapopulation.
-Attracting colonists helps position high quality patches as the infectious "traps" hypothesized above, wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).
+When a patch's connectivity is determined by its quality (Eqn~\ref{connectivity}), high quality patches function as highly connected hubs in the metapopulation.
+Attracting colonists helps position high quality patches as infectious "traps", wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).
 This increased infection pressure results from both transmission from a persistent environmental reservoir and from direct transmission from infectious immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).
 
-This trap effect, coupled with the relatively low extinction rate on high quality patches helps to create a stable platform from which the pathogen can spread through the rest of the metapopulation.
+The relatively low extinction rate on high quality patches can help to create a stable platform from which the pathogen can spread through the rest of the metapopulation.
 Indeed, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).
@@
 This phenomenon represents the other side of the hub role played by high quality patches in that low quality patches are relatively weakly connected to the metapopulation and are therefore infrequently colonized by infectious individuals.
-Moreover, the instability of low quality patches (i.e.
-the high extinction rate), together with the increased mortality of infectious populations, means that infectious populations do not persist long on low quality patches.
-Because low quality patches are so infrequently colonized, at relatively low to intermediate longevities, the environmental reservoir left behind by these infectious populations generally decays before it has the opportunity to infect new susceptible colonists.
+Moreover, the high extinction rate, together with the increased mortality of infectious populations, means that infectious populations do not persist long on low quality patches and the environmental reservoir left behind by these infectious populations may decay before it has the opportunity to infect new susceptible colonists.
 As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infectious occupancy (Fig~\ref{sens}).
@@
 DGE-1321845.
+
+Any mention of trade, product, or firm names is for descriptive purposes only and does not imply endorsement by the U.S. Government.
 

### 2014-11-10 Manuscript/manuscript.tex
Formatting, author addresses, Proc. B bib.

@@
-%
 \documentclass{article}
@@
 \usepackage{graphicx}
-\usepackage{natbib}
 \usepackage{tabularx}
@@
 \usepackage{color}
+\usepackage{authblk}
 
@@
 
-\author{Clinton B. Leach \and Paul C. Cross \and Colleen T. Webb}
+\author[1]{Clinton B. Leach$^*$}
+\author[1]{Colleen T. Webb}
+\affil[1]{Department of Biology, Graduate Degree Program in Ecology, Colorado State University, Fort Collins, CO, USA}
+\author[2]{Paul C. Cross}
+\affil[2]{Northern Rocky Mountain Science Center, US Geological Survey, Bozeman, Montana, USA}
+\affil[*]{Corresponding author: clint.leach@colostate.edu}
 
@@
 
+\begin{quote}
 This draft manuscript is distributed solely for purposes of scientific peer review.
@@
 Geological Survey (USGS), it does not represent any official USGS finding or policy.
+\end{quote}
 
@@
 
-\begin{table}
-\caption{State transitions and their rates for patch $i$ in a metapopulation simulation.
-$S$ denotes occupied by  susceptible hosts, $I$ denotes occupied infectious hosts, and $\emptyset$ denotes unoccupied by the host.
-All patches are characterized by an environmental infection rate, $\gamma(\tau_i)$, where $\tau_i$ is the time since the patch was last occupied by infectious individuals.}
-\begin{tabular}{llr}
-State Transition & Process &  Rate \\
-\hline
-$S \rightarrow I$ & Infection (from contact or reservoir) & $\delta C_{Ii} + \gamma(\tau_i)$\\
-$S \rightarrow \emptyset $ & Extinction (of susceptible) & $E_{Si}$\\
-$I \rightarrow \emptyset $ & Extinction (of infectious) &  $E_{Ii}$ \\
-$\emptyset \rightarrow S$ & Colonization (by susceptibles) & $C_{Si}$\\
-$\emptyset \rightarrow I$ & Colonization (by infectious) & $C_{Ii}$\\
-\label{transitions}
-\end{tabular}
-\end{table}
-
 To address the above questions, we developed a stochastic patch occupancy model (SPOM) in which each patch can be in one of three possible states: occupied by susceptible hosts (S), occupied by infectious hosts (I), and unoccupied by the host ($\emptyset$).
@@
 where $\gamma_0$ is the initial infection rate of the pathogen reservoir, $\tau_{i}$ is the time since last infectious occupancy ($\tau_{i} = \infty$, and thus $\gamma(\tau_i) = 0$, if the patch has never been occupied by infectious hosts), and $r$ is the pathogen's decay rate in the environment.
-
-\begin{table}[h!]   
-\caption{Parameters of the SPOM model, their meaning, and the values assigned under different parameterizations.  Empty cells indicate the same value as the default parameterization.}
-\begin{tabular}{l p{5cm} p{1.5cm} p{1.5cm} p{1.5cm} p{1.5cm} p{1.5cm}}
-Parameter & Interpretation &  Default & Local movement & High env. stoch. & Dominant reservoir trans. & High disease mortality \\
-\hline
-$\xi_{im}$ & Scaling parameter for effect of target patch quality on immigration & 0.5 & -- & -- & -- & --\\
-$\xi_{em}$ & Scaling parameter for effect of target patch quality on emigration & 0.5 & -- & -- & --& --\\
-$D$& Inverse of mean dispersal distance & 5 & 2 & -- & -- & --\\
-$d_{ij}$ & Distance between patch $i$ and $j$ & 1 $\forall i \neq j$ & 1 $\forall i, j$ neighbors & -- & -- & --\\
-\hline
-$\mu_S$ & Extinction rate of unit quality susceptible patch & 0.1 & -- & -- & -- & --\\
-$\mu_I$ & Extinction rate of unity quality infectious patch & 0.5 & -- & -- & -- & 1\\
-$\alpha$ & Strength of environmental stochasticity & 1 & -- & 0 & -- & --\\
-\hline
-$\delta$ & Probability of direct infection & 0.5 & -- & -- & 0.1 & --\\
-$\gamma_0$ & Initial rate of infection from reservoir patch & 0.5 & -- & -- & -- & --
-\end{tabular}
-\label{params}
-\end{table}
 
@@
 
-In the baseline parameterization, infectious occupancy increased substantially with longevity, while susceptible and total occupancy decreased (Fig~ \ref{poutcome}).
-However, increasing the variance of the patch quality distribution around a mean of 1 had relatively little effect on average occupancy (Fig~\ref{poutcome}).
-The four alternative parameterizations (models with local movement, high environmental stochasticity, dominant reservoir transmission, and high disease-induced mortality) produced qualitatively similar results, though pathogen spread in these models was generally more limited (Figs~\ref{poutcome_lattice}, \ref{poutcome_x0}, \ref{poutcome_delta}, \ref{poutcome_ei}).
+In the baseline parameterization, infectious occupancy increased substantially with longevity, while susceptible and total occupancy decreased (Supplemental).
+However, increasing the variance of the patch quality distribution around a mean of 1 had relatively little effect on average occupancy (Supplemental).
+The four alternative parameterizations (models with local movement, high environmental stochasticity, dominant reservoir transmission, and high disease-induced mortality) produced qualitatively similar results, though pathogen spread in these models was generally more limited (Supplemental).
 
@@
 Conversely, low quality habitat distributions decreased infectious occupancy, while high quality distributions increased infectious occupancy (Fig~\ref{sens}(b)).
-These general trends were consistent across the four other parameterizations (Figs~\ref{sens_lattice}, \ref{sens_x0}, \ref{sens_delta}, \ref{sens_ei}).
+These general trends were consistent across the four other parameterizations (Supplemental).
 
@@
 
-\begin{figure}
-\centering
-\includegraphics[height=7cm]{highlow(full).pdf}
-\caption{The effect of the presence of high and low quality patches on (a) median susceptible occupancy, (b) median infectious occupancy, and (c) median total occupancy.
-Red lines show hiqh quality metapopulations (patch qualities roughly between 0.75 and 1.75), blue lines show low quality metapopulations (patch qualities roughly between 0.25 and 1.25).}
-\label{sens}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[height = 6cm]{highvlow.pdf}
-\caption{Differences between the median total occupancy of high quality patch distributions (quality between 0.75 and 1.75) and low quality patch distributions (quality between 0.25 to 1.25) across the five different model parameterizations.
-(a) default model, (b) local movement, (c) high environmental stochasticity, (d) dominant reservoir transmission, (e) high disease-induced mortality.}
-\label{highvlow}
-\end{figure}
-
 To understand the mechanisms behind these effects, we further examined within patch infection dynamics.
@@
 In addition, while high quality patches did support susceptible occupancy, they experienced a greater number of infection events (susceptible to infectious transitions) than low quality patches (Fig~\ref{infections}).
-A similar trend was observed across the other parameterizations, though the observed relationship between patch quality and number of infection events was shallower for the lattice and weak direct infection models (Figs~\ref{infections_lattice}, \ref{infections_x0}, \ref{infections_delta}, \ref{infections_ei}).
-
-\begin{figure}
-\centering
-\includegraphics[height = 9cm]{simvis.png}
-\caption{Results from a single representative simulation with variance 0.2 and longevity of 0.3 times the expected residence time of a susceptible population on a unit quality patch.
-The top panel shows total occupancy of susceptible (blue) and infectious (red) patches over time.
-The bottom panel shows the state -- susceptible (blue), infectious (red), or unoccupied (white) -- of individual patches through time.
-Patches are stacked vertically with the lowest quality ($\sim$ 0.25) at the bottom and the highest quality ($\sim$ 1.75) at the top.}
-\label{simvis}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[height=7cm]{infevents(full).pdf}
-\caption{The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
-Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
-\label{infections}
-\end{figure}
+A similar trend was observed across the other parameterizations, though the observed relationship between patch quality and number of infection events was shallower for the lattice and weak direct infection models (Supplemental).
 
@@
 
-\section{Appendix}
+\bibliographystyle{vancouver}     
+\bibliography{metapop}   % name your BibTeX data base
 
-\subsection{Occupancy results for symmetric quality distributions}
+\clearpage
 
-\begin{figure}[h!]
-\centering
-\includegraphics[height=5.5cm]{metapop(full).pdf}
-\caption{Default model parameterization.
-Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay, measured in units of expected occupancy time of a susceptible population on a unit quality patch) and the variance of the patch quality distribution.}
-\label{poutcome}
+\begin{figure}
+\caption{The effect of the presence of high and low quality patches on (a) median susceptible occupancy, (b) median infectious occupancy, and (c) median total occupancy.
+Red lines show hiqh quality metapopulations (patch qualities roughly between 0.75 and 1.75), blue lines show low quality metapopulations (patch qualities roughly between 0.25 and 1.25).}
+\label{sens}
 \end{figure}
 
-\begin{figure}[h!]
+\begin{figure}
 \centering
-\includegraphics[height=5.5cm]{metapop(lattice).pdf}
-\caption{Metapopulation with lattice structure and local movement.
-Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay, measured in units of expected occupancy time of a susceptible population on a unit quality patch) and the variance of the patch quality distribution.}
-\label{poutcome_lattice}
+\caption{Differences between the median total occupancy of high quality patch distributions (quality between 0.75 and 1.75) and low quality patch distributions (quality between 0.25 to 1.25) across the five different model parameterizations.
+(a) default model, (b) local movement, (c) high environmental stochasticity, (d) dominant reservoir transmission, (e) high disease-induced mortality.}
+\label{highvlow}
 \end{figure}
 
-\begin{figure}[h!]
+\begin{figure}
 \centering
-\includegraphics[height=5.5cm]{metapop(alpha0).pdf}
-\caption{High environmental stochasticity ($\alpha = 0$).
-Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay, measured in units of expected occupancy time of a susceptible population on a unit quality patch) and the variance of the patch quality distribution.}
-\label{poutcome_x0}
+\caption{Results from a single representative simulation with variance 0.2 and longevity of 0.3 times the expected residence time of a susceptible population on a unit quality patch.
+The top panel shows total occupancy of susceptible (blue) and infectious (red) patches over time.
+The bottom panel shows the state -- susceptible (blue), infectious (red), or unoccupied (white) -- of individual patches through time.
+Patches are stacked vertically with the lowest quality ($\sim$ 0.25) at the bottom and the highest quality ($\sim$ 1.75) at the top.}
+\label{simvis}
 \end{figure}
 
-\begin{figure}[h!]
+\begin{figure}
 \centering
-\includegraphics[height=5.5cm]{metapop(delta).pdf}
-\caption{Reservoir transmission dominant ($\delta = 0.1$).
-Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay, measured in units of expected occupancy time of a susceptible population on a unit quality patch) and the variance of the patch quality distribution.}
-\label{poutcome_delta}
-\end{figure}
-
-\begin{figure}[h!]
-\centering
-\includegraphics[height=5.5cm]{metapop(ei).pdf}
-\caption{High disease-induced mortality ($\mu_I = 1$).
-Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay, measured in units of expected occupancy time of a susceptible population on a unit quality patch) and the variance of the patch quality distribution.}
-\label{poutcome_ei}
+\caption{The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
+Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
+\label{infections}
 \end{figure}
@@
 
-\subsection{Occupancy results for asymmetric quality distributions}
+\begin{table}
+\caption{State transitions and their rates for patch $i$ in a metapopulation simulation.
+$S$ denotes occupied by  susceptible hosts, $I$ denotes occupied infectious hosts, and $\emptyset$ denotes unoccupied by the host.
+All patches are characterized by an environmental infection rate, $\gamma(\tau_i)$, where $\tau_i$ is the time since the patch was last occupied by infectious individuals.}
+\begin{tabular}{llr}
+State Transition & Process &  Rate \\
+\hline
+$S \rightarrow I$ & Infection (from contact or reservoir) & $\delta C_{Ii} + \gamma(\tau_i)$\\
+$S \rightarrow \emptyset $ & Extinction (of susceptible) & $E_{Si}$\\
+$I \rightarrow \emptyset $ & Extinction (of infectious) &  $E_{Ii}$ \\
+$\emptyset \rightarrow S$ & Colonization (by susceptibles) & $C_{Si}$\\
+$\emptyset \rightarrow I$ & Colonization (by infectious) & $C_{Ii}$\\
+\label{transitions}
+\end{tabular}
+\end{table}
 
-\begin{figure}[h!]
-\centering
-\includegraphics[height=7cm]{highlow(lattice).pdf}
-\caption{Metapopulation with lattice structure and local movement.
-The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
-Red lines show high quality metapopulations (patch qualities roughly between 0.75 and 1.75), blue lines show low quality metapopulations (patch qualities roughly between 0.25 and 1.25).}
-\label{sens_lattice}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[height=7cm]{highlow(alpha0).pdf}
-\caption{High environmental stochasticity ($\alpha = 0$).
-The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
-Red lines show high quality metapopulations (patch qualities roughly between 0.75 and 1.75), blue lines show low quality metapopulations (patch qualities roughly between 0.25 and 1.25).}
-\label{sens_x0}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[height=7cm]{highlow(delta).pdf}
-\caption{Reservoir transmission dominant ($\delta = 0.1$).
-The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
-Red lines show high quality metapopulations (patch qualities roughly between 0.75 and 1.75), blue lines show low quality metapopulations (patch qualities roughly between 0.25 and 1.25).}
-\label{sens_delta}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[height=7cm]{highlow(ei).pdf}
-\caption{High disease-induced mortality ($\mu_I = 1$).
-The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
-Red lines show high quality metapopulations (patch qualities roughly between 0.75 and 1.75), blue lines show low quality metapopulations (patch qualities roughly between 0.25 and 1.25).}
-\label{sens_ei}
-\end{figure}
-
-\clearpage
-
-\subsection{Patch-level infection events}
-
-\begin{figure}[h!]
-\centering
-\includegraphics[height=7cm]{infevents(lattice).pdf}
-\caption{Metapopulation with lattice structure and local movement.
-The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
-Light gray lines show the smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
-\label{infections_lattice}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[height=7cm]{infevents(alpha0).pdf}
-\caption{High environmental stochasticity ($\alpha = 0$).
-The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
-Light gray lines show the smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
-\label{infections_x0}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[height=7cm]{infevents(delta).pdf}
-\caption{Reservoir transmission dominant ($\delta = 0.1$).
-The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
-Light gray lines show the smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
-\label{infections_delta}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[height=7cm]{infevents(ei).pdf}
-\caption{High disease-induced mortality ($\mu_I = 1$).
-The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
-Light gray lines show the smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
-\label{infections_ei}
-\end{figure}
-
-\clearpage
-
-\bibliographystyle{spbasic}     
-\bibliography{metapop}   % name your BibTeX data base
+\begin{table}[h!]   
+\caption{Parameters of the SPOM model, their meaning, and the values assigned under different parameterizations.  Empty cells indicate the same value as the default parameterization.}
+\begin{tabular}{l p{4cm} p{1.5cm} p{1.5cm} p{1.5cm} p{1.5cm} p{1.5cm}}
+Parameter & Interpretation &  Default & Local movement & High env. stoch. & Dominant reservoir trans. & High disease mortality \\
+\hline
+$\xi_{im}$ & Scaling parameter for effect of target patch quality on immigration & 0.5 & -- & -- & -- & --\\
+$\xi_{em}$ & Scaling parameter for effect of target patch quality on emigration & 0.5 & -- & -- & --& --\\
+$D$& Inverse of mean dispersal distance & 5 & 2 & -- & -- & --\\
+$d_{ij}$ & Distance between patch $i$ and $j$ & 1 $\forall i \neq j$ & 1 $\forall i, j$ neighbors & -- & -- & --\\
+\hline
+$\mu_S$ & Extinction rate of unit quality susceptible patch & 0.1 & -- & -- & -- & --\\
+$\mu_I$ & Extinction rate of unity quality infectious patch & 0.5 & -- & -- & -- & 1\\
+$\alpha$ & Strength of environmental stochasticity & 1 & -- & 0 & -- & --\\
+\hline
+$\delta$ & Probability of direct infection & 0.5 & -- & -- & 0.1 & --\\
+$\gamma_0$ & Initial rate of infection from reservoir patch & 0.5 & -- & -- & -- & --
+\end{tabular}
+\label{params}
+\end{table}
 

### 2014-11-10 Manuscript/supplementary.tex
Formatting, author addresses, Proc. B bib.

\documentclass{article} 
\usepackage{graphicx}

\title{Supplementary Figures}

\begin{document}

\maketitle

\section{Occupancy results for symmetric quality distributions}

\begin{figure}[h!]
\centering
\includegraphics[height=5.5cm]{figure/metapop(full).pdf}
\caption{Default model parameterization.  Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay, measured in units of expected occupancy time of a susceptible population on a unit quality patch) and the variance of the patch quality distribution.}
\label{poutcome}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[height=5.5cm]{figure/metapop(lattice).pdf}
\caption{Metapopulation with lattice structure and local movement.  Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay, measured in units of expected occupancy time of a susceptible population on a unit quality patch) and the variance of the patch quality distribution.}
\label{poutcome_lattice}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[height=5.5cm]{figure/metapop(alpha0).pdf}
\caption{High environmental stochasticity ($\alpha = 0$).  Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay, measured in units of expected occupancy time of a susceptible population on a unit quality patch) and the variance of the patch quality distribution.}
\label{poutcome_x0}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[height=5.5cm]{figure/metapop(delta).pdf}
\caption{Reservoir transmission dominant ($\delta = 0.1$).  Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay, measured in units of expected occupancy time of a susceptible population on a unit quality patch) and the variance of the patch quality distribution.}
\label{poutcome_delta}
\end{figure}

\begin{figure}[h!]
\centering
\includegraphics[height=5.5cm]{figure/metapop(ei).pdf}
\caption{High disease-induced mortality ($\mu_I = 1$).  Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay, measured in units of expected occupancy time of a susceptible population on a unit quality patch) and the variance of the patch quality distribution.}
\label{poutcome_ei}
\end{figure}

\clearpage

\section{Occupancy results for asymmetric quality distributions}

\begin{figure}[h!]
\centering
\includegraphics[height=7cm]{figure/highlow(lattice).pdf}
\caption{Metapopulation with lattice structure and local movement.  The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.  Red lines show high quality metapopulations (patch qualities roughly between 0.75 and 1.75), blue lines show low quality metapopulations (patch qualities roughly between 0.25 and 1.25).}
\label{sens_lattice}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=6cm]{figure/highlow(alpha0).pdf}
\caption{High environmental stochasticity ($\alpha = 0$).  The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.  Red lines show high quality metapopulations (patch qualities roughly between 0.75 and 1.75), blue lines show low quality metapopulations (patch qualities roughly between 0.25 and 1.25).}
\label{sens_x0}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=6cm]{figure/highlow(delta).pdf}
\caption{Reservoir transmission dominant ($\delta = 0.1$).  The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.  Red lines show high quality metapopulations (patch qualities roughly between 0.75 and 1.75), blue lines show low quality metapopulations (patch qualities roughly between 0.25 and 1.25).}
\label{sens_delta}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=6cm]{figure/highlow(ei).pdf}
\caption{High disease-induced mortality ($\mu_I = 1$).  The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.  Red lines show high quality metapopulations (patch qualities roughly between 0.75 and 1.75), blue lines show low quality metapopulations (patch qualities roughly between 0.25 and 1.25).}
\label{sens_ei}
\end{figure}

\clearpage

\section{Patch-level infection events}

\begin{figure}[h!]
\centering
\includegraphics[height=7cm]{figure/infevents(lattice).pdf}
\caption{Metapopulation with lattice structure and local movement.  The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.  Light gray lines show the smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
\label{infections_lattice}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=7cm]{figure/infevents(alpha0).pdf}
\caption{High environmental stochasticity ($\alpha = 0$).  The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.  Light gray lines show the smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
\label{infections_x0}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=7cm]{figure/infevents(delta).pdf}
\caption{Reservoir transmission dominant ($\delta = 0.1$).  The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.  Light gray lines show the smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
\label{infections_delta}
\end{figure}

\begin{figure}
\centering
\includegraphics[height=7cm]{figure/infevents(ei).pdf}
\caption{High disease-induced mortality ($\mu_I = 1$).  The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.  Light gray lines show the smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
\label{infections_ei}
\end{figure}

\end{document}

### 2014-11-19 Manuscript/trapdisc.Rmd
Reparameterization of disease-induced mortality.

---
title: "Discussion on traps"
author: "Clint Leach"
date: "11/12/2014"
output: pdf_document
---

# How do we define a trap?

This question comes with two follow-up questions:

* How do we measure/detect a trap in the simulation model?
* How do the assumptions of the model affect our interpretation of the above measurements?

Given these questions, it seems there are three ways we can define a trap.

## At the individual level

Here, a trap is preferred habitat that increases individual-level disease risk and mortality.

* In the model, we can interpret the increase in infection events with patch quality as an indicator that high quality patches are traps, as colonization of high quality habitat increases disease risk.
* What does this mean about mortality and how do our assumptions about disease-induced mortality actually translate to the individual level?

The Hanski model assumes that extinction rate of patch $i$ is given by $E_i = \frac{e_X}{A_i^\alpha}$, which derives from the assumption that patch extinction is driven by environmental stochasticity, and thus that larger populations on higher quality habitat are less prone to extinction.

In the model, we assume that $e_I > e_S$, i.e. that infected populations have higher extinction rates.  By Hanski's assumptions then, this means that we expect infected populations to be smaller than healthy populations and more susceptible to environmental stochasticity.  Following this, a better way to parameterize extinction might be as:

$$
E_{Ii} = \frac{e}{(\nu A_i)^{\alpha}}
$$

where $\nu$ is now the factor by which population size is reduced by infection (which, assuming $\alpha = 1$ as we did in most simulations is the same as just choosing $e_I > e_S$).

However, if we take these as our underlying assumptions, this should also imply that the emigration rate, the number of colonists produced by a patch, should also change with infection status, such that

$$
C_{Ii} = A_i^{\xi_{im}} \sum \phi_j (\nu A_j) ^ {\xi_{em}} \exp ({-Dd_{ij}})
$$

I wouldn't expect this to have a major effect on the results, but would likely reduce the "superspreader" effect that we see from high quality patches.

## At the patch level

A trap is a preferred habitat that cannot maintain a population on its own (i.e. a sink, $\lambda < 1$).  I think this is the most traditional definition in the literature, but is difficult to fit into the occupancy model framework, as we don't have population dynamics at the patch level.

## At the metapopulation level

A trap is preferred habitat that reduces overall metapopulation occupancy.  In addition to infection events, we've also been loosely using this definition in the paper (i.e. that high quality habitat is a trap when high quality habitat distributions produce lower occupancy than low quality distributions).  I'm not sure if it makes sense though to define a trap by a metapopulation-scale effect, as "trap" really refers to the dynamics within the patch.

That said, if we parameterize disease induced mortality in terms of population size above, instead of looking at occupancy in the model results, we can instead look at total population size, which might make more sense when considering whether high or low quality habitat is better for the metapopulation.

### 2014-11-19 Analysis/highlow.shell.r

@@ -27,7 +27,7 @@ parms<-data.frame("xi_im" = 0.5,
                   "D" = 5, 
                   "alpha" = 1,
                   "es" = 0.1,
-                  "ei" = 0.5,
+                  "nu" = 0.2,
                   "delta" = 0.5,
                   "gamma0" = 0.5)
 

### 2014-11-19 Analysis/shell.r

@@ -40,7 +40,7 @@ parms<-data.frame("xi_im" = 0.5,
                   "D" = 5, 
                   "alpha" = 1,
                   "es" = 0.1,
-                  "ei" = 0.5,
+                  "nu" = 0.2,
                   "delta" = 0.5,
                   "gamma0" = 0.5)
   
@@ -72,7 +72,8 @@ getDoParWorkers()
 
 #Looping through each parameter combo and calling modelrun
 system.time(out<-foreach(i = 1:length(par.reps[,1]),.verbose=TRUE,.combine="rbind") %dopar% 
-  modelrun(par.reps[i,1],par.reps[i,2],par.reps[i,3],parms,distance,initial,timesteps))
+  modelrun(par.reps[i,1],par.reps[i,2],par.reps[i,3],parms,distance,initial,timesteps)
+  )
 
 stopCluster(w)
 

### 2014-11-19 Analysis/simvis.r

@@ -9,7 +9,7 @@ parms<-data.frame("xi_im" = 0.5,
                   "D" = 5, 
                   "alpha" = 1,
                   "es" = 0.1,
-                  "ei" = 0.5,
+                  "nu" = 0.2,
                   "delta" = 0.5,
                   "gamma0" = 0.5)
 

### 2014-11-19 Code/contSPOM(rates).r

@@ -11,7 +11,7 @@ diseaseSPOM <- function(distance, quality, initial, parms, r, timesteps, ss.thre
   #     xi_em = scaling parameter for how patch quality affects the number of colonists produced
   #     D = inverse of mean dispersal distance
   #     es = extinction probability of unit susceptible patch
-  #     ei = extinction probability of unit infectious patch
+  #     nu = factor by which population size is reduced by infection
   #     alpha = how strongly extinction risk depends on patch quality
   #     delta = probability of direct infection
   #     gamma0 = probability of reservoir infection immediately following extinction of an infected patch
@@ -28,7 +28,7 @@ diseaseSPOM <- function(distance, quality, initial, parms, r, timesteps, ss.thre
   D <- parms$D
   alpha <- parms$alpha
   es <- parms$es
-  ei <- parms$ei
+  nu <- parms$nu
   delta <- parms$delta
   gamma0 <- parms$gamma0
     
@@ -53,7 +53,7 @@ diseaseSPOM <- function(distance, quality, initial, parms, r, timesteps, ss.thre
   
   #Extinction vectors
   extinctionS <- es / quality ^ alpha
-  extinctionI <- ei / quality ^ alpha
+  extinctionI <- es / (nu * quality) ^ alpha
   
   # Generates occupancy vector for S and I
   occ.S <- vector(mode="numeric", length=timesteps)
@@ -74,7 +74,7 @@ diseaseSPOM <- function(distance, quality, initial, parms, r, timesteps, ss.thre
     
     # Calculates connectivities to susceptible and infected patches
     connectivityS <- connectivity %*% (state[t, ] == "S")
-    connectivityI <- connectivity %*% (state[t ,] == "I")
+    connectivityI <- connectivity %*% (nu ^ xi_em * (state[t ,] == "I"))
     
     # Updates infectivity of reservoirs 
     gammas[which(ti > 0)] <- gamma0 * exp(-r * (event.times[t] - ti[which(ti > 0)]))

### 2014-11-19 Code/highlow.runs.r

@@ -19,9 +19,8 @@ modelrun <- function(longevity, parms, distance, initial, timesteps, treatment){
   #     S: final susceptible occupancy
   #     I: final infectious occupancy
   #     R: final reservoir occupancy
-  #     avgS: susceptible occupancy averaged over last 100 time steps
-  #     avgI: infectious occupancy averaged over last 100 time steps
-  #     avgR: reservoir occupancy averaged over last 100 time steps
+  #     S.pop: mean effective susceptible population size (sum(A_i * p_i(S)))
+  #     I.pop: mean effective infectious population size (sum(nu * A_i * p_i(I)))
   #     maxI: maximum infectious occupancy
   #     tmaxI: time to maximum infectious occupancy
   #     quality0: quality of initially infected patch
@@ -67,8 +66,10 @@ modelrun <- function(longevity, parms, distance, initial, timesteps, treatment){
   output <- diseaseSPOM(distance, quality, initial.inf, parms, r, timesteps, 0.05)
   output <- na.omit(output)
   
+  runtime <- dim(output)[1]
+  
   # Processing output
-  state <- output[, -1]
+  state <- output[(runtime - 499):runtime, -1]
   susc <- state == "S"
   inf <- state == "I"
   empty <- state == "E" | state == "R"
@@ -77,12 +78,15 @@ modelrun <- function(longevity, parms, distance, initial, timesteps, treatment){
   I <- rowSums(inf) / 100
   E <- rowSums(empty) / 100
   
+  # Effective population size (assuming pop proportional to quality)
+  S.pop <- (colSums(susc) / (dim(state)[1])) %*% quality
+  I.pop <- (colSums(inf) / (dim(state)[1])) %*% (parms$nu * quality)
+  
   # Generating and collecting output
   output <- data.frame(treatment, longevity, S[length(S)], I[length(I)], E[length(E)],
-                           mean(S[(length(S)-100):length(S)]), mean(I[(length(I)-100):length(I)]), mean(E[(length(I)-100):length(I)]),
-                           max(I), output[which.max(I),1], initial.quality)
+                           S.pop, I.pop, max(I), output[which.max(I),1], initial.quality)
   
-  names(output) <- c("treatment", "longevity", "S", "I", "R", "avgS", "avgI", "avgR", "maxI", "tmaxI", "quality0")
+  names(output) <- c("treatment", "longevity", "S", "I", "R", "S.pop", "I.pop", "maxI", "tmaxI", "quality0")
   
   return(output)
   

### 2014-11-19 Code/modelrun3.r

@@ -17,11 +17,10 @@ modelrun <- function(longevity, variance, k, parms, distance, initial, timesteps
   #   n x 19 dataframe giving patch-level results from simulation
   #     quality: quality of patch
   #     tinf: time to first infection
-  #     I: proportion of time patch infected
-  #     S: proportion of time patch susceptible
-  #     E: proportion of time patch empty
-  #     inf.events.early: number of infection events in first fifth of sim
-  #     inf.events.tot: total number of infection events
+  #     I: proportion of time patch infected over last 500 time steps
+  #     S: proportion of time patch susceptible over last 500 time steps
+  #     E: proportion of time patch empty over last 500 time steps
+  #     inf.events: number of infection events
   #     susc.col: number of susceptible colonization events
   #     inf.col: number of infected colonization events
   #     susc.ex: number of susceptible extinction events
@@ -34,7 +33,6 @@ modelrun <- function(longevity, variance, k, parms, distance, initial, timesteps
   #     repID: replicate ID
   #     longevity: pathogen environemental longevity
   #     variance: variance of patch quality distribution
-  
   
 	#Generates quality vector with desired variance
   max <- 0.5 * sqrt(12 * variance) + 1
@@ -61,8 +59,9 @@ modelrun <- function(longevity, variance, k, parms, distance, initial, timesteps
 	output <- na.omit(output)
   
   # Processing output
+  runtime <- dim(output)[1]
   
-  state <- output[, -1]
+  state <- output[(runtime - 499):runtime, -1]
   susc <- state == "S"
   inf <- state == "I"
   empty <- state == "E" | state == "R"
@@ -71,26 +70,25 @@ modelrun <- function(longevity, variance, k, parms, distance, initial, timesteps
   I <- rowSums(inf) / 100
   E <- rowSums(empty) / 100
   
-  inf.events <- susc[-(dim(output)[1]), ] * inf[-1, ]
-  susc.col <- empty[-(dim(output)[1]), ] * susc[-1, ]
-  inf.col <- empty[-(dim(output)[1]), ] * inf[-1, ]
-  susc.ex <- susc[-(dim(output)[1]), ] * empty[-1, ]
-  inf.ex <- inf[-(dim(output)[1]), ] * empty[-1, ]
+  inf.events <- susc[-(dim(state)[1]), ] * inf[-1, ]
[…]

### 2014-11-20 Manuscript/manuscript.tex
Udpated parameterization in manuscript.

@@
 Patterns of habitat quality are expected to play an important role in the dynamics of wildlife metapopulations, with the expectation that high quality habitat serves to stabilize and maintain occupancy.
-However, we might expect that the costs and benefits of different quality habitat might be modulated, or possibly even reversed, by the presence of an environmentally transmitted pathogen.
-We explore this possibility through the development of a stochastic patch occupancy model of host-pathogen dynamics, in which we can vary both the environmental longevity of the pathogen, and the distribution of habitat quality in the host metapopulation.
+However, we might expect that the costs and benefits of different quality habitat might be modulated, or possibly even reversed, by the presence of an environmentally persistent pathogen.
+In such cases, the presence of environmental pathogen reservoirs on high quality habitat may lead to the creation of ecological traps, wherein host individuals preferentially colonize high quality habitat, but are then exposed to increased infection risk.
+We explore this possibility through the development of a stochastic patch occupancy model in which we can vary both the environmental longevity of the pathogen, and the distribution of habitat quality in the host metapopulation.
+This suggests that for sufficiently long-lived environmental reservoirs, high quality habitat can serve as an ecological trap, and can even be a net detriment to the host metapopulation relative to low quality habitat.
+This inversion of the relative roles of high and low quality habitat highlights the importance of considering the interaction between spatial structure and pathogen transmission when managing wildlife populations exposed to an environmentally persistent pathogen.
 \end{abstract}
@@
 
-Building on the framework developed by \cite{Hanski2003}, the rate at which patch \emph{i}, with quality $A_i$, is colonized by individuals in state $X \in (S,I)$ is given by its connectivity:
-\begin{equation}
-C_{Xi}=A_i^{\xi_{im}} \sum_{i\neq j }\phi_jA_j^{\xi_{em}}\exp(-D d_{ij}),
+We assume that patch population size is proportional to patch quality ($A_i$), and that disease-induced mortality reduces the size of infectious populations by a factor of $\nu$.
+Building on the framework developed by \cite{Hanski2003} then,the rates at which patch \emph{i}, with quality $A_i$, is colonized by susceptible and infectious individuals, respectively, is given by its connectivity:
+\begin{align}
+C_{Si} &= A_i^{\xi_{im}} \sum_{i\neq j }\phi(S_j)A_j^{\xi_{em}}\exp(-D d_{ij}),\\
+C_{Ii} &= A_i^{\xi_{im}} \sum_{i\neq j }\phi(I_j)(\nu A_j) ^{\xi_{em}}\exp(-D d_{ij}),
 \label{connectivity}
-\end{equation}
-where $\phi_j$ is an indicator function that is $1$ if patch $j$ is in state $X$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality; $D$ is the inverse of mean dispersal distance; and $d_{ij}$ is the distance between patches $i$ and $j$.
+\end{align}
+where $\phi(X_j)$ is an indicator function that is $1$ if patch $j$ is in state $X$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality (i.e.
+population size); $D$ is the inverse of mean dispersal distance; and $d_{ij}$ is the distance between patches $i$ and $j$.
 Essentially, $C_{Xi}$ sums the colonization effort from all patches in state $X$ to patch $i$.
 
-The extinction rate of patch $i$ in state $X$ is given by:
-\begin{equation}
-E_{Xi}=\frac{\mu_X}{A_i^\alpha},
-\end{equation}
-where $\mu_X$ is the extinction rate of a patch of unit quality in state $X$ (here we assume that $\mu_I \geq \mu_S$ to account for disease-induced mortality), and $\alpha$ controls how extinction rate scales with patch quality.
+The state-dependent extinction rates of populations on patch $i$ are then given by:
+\begin{align}
+E_{Si} & =\frac{\mu}{A_i^\alpha},\\
+E_{Ii} & = \frac{\mu}{(\nu A_i) ^ \alpha}
+\end{align}
+where $\mu$ is the extinction rate of a patch of unit quality, and $\alpha$ controls how extinction rate scales with population size.
 When an infectious population goes extinct, we assume that the hosts leave behind an infectious pathogen reservoir on the patch.
 
-Susceptible populations are infected via direct contact with infectious colonists at rate $\delta \lambda_{Ii}$, where $\delta$ is the direct transmission probability.
+Susceptible populations are infected via direct contact with infectious colonists at rate $\delta C_{Ii}$, where $\delta$ is the direct transmission probability.
 In addition, environmental transmission can take place when a susceptible population occupies a previously infected patch, and occurs at rate:
@@
 The default model was parameterized  according to the following assumptions.
-(1) High quality patches maintain higher population sizes and therefore produce more colonists than low quality patches and have lower extinction rates than low quality patches ($\xi_{im}=0.5$, $\alpha=1$, \cite{Hanski2003}).
+(1) The rate of emigration from a patch increases with population size, while the extinction rate of a patch decreases with populations size ($\xi_{im}=0.5$, $\alpha=1$, \cite{Hanski2003}).
 (2) High quality patches attract more colonists than low quality patches ($\xi_{em}=0.5$, \cite{Hanski2003}).
 (3)  Connectivity among patches is determined entirely by quality ($d_{ij}=1$ for all $i \neq j$).
-In addition to these assumptions, parameters were chosen so that without infection, approximately 0.75 of the metapopulation was occupied ($\mu_S = 0.1$, $D=5$), and a range of epidemiological behaviors -- disease-free, endemic disease, and pandemic disease (i.e.
-all host populations infectious) -- was feasible, depending on the pathogen's environmental longevity ($\mu_I = 0.5$, $\delta = 0.5$, $\gamma_0 = 0.5$).
+In addition to these assumptions, parameters were chosen so that without infection, approximately 0.75 of the metapopulation was occupied ($\mu = 0.1$, $D=5$), and a range of epidemiological behaviors -- disease-free, endemic disease, and pandemic disease (i.e.
+all host populations infectious) -- was feasible, depending on the pathogen's environmental longevity ($\nu = 0.2$, $\delta = 0.5$, $\gamma_0 = 0.5$).
 
@@
 To explore the disease dynamics when transmission is driven largely by the environmental reservoir we parameterized a model with $\delta = 0.1$ to reduce the relative importance of direct transmission (Dominant reservoir trans., Table~\ref{params}).
-Lastly, to simulate a more virulent pathogen, we parameterized a model with $\mu_I = 1$ to substantially increase disease-induced mortality (High disease mortality, Table~\ref{params}).
+Lastly, to simulate a more virulent pathogen, we parameterized a model with $\nu = 0.1$ to substantially increase disease-induced mortality (High disease mortality, Table~\ref{params}).
 
@@
 For these simulations, we explored 10 values of pathogen longevity ranging from 0.1 to 3 times the expected residence time of a healthy population on a unit quality patch (i.e.
-10 units of time), with values representing the half-life of the pathogen's infectivity and 10 values for the variance of the patch quality distribution, ranging from 0.02 to 0.2.
+10 units of time), with values representing the half-life of the pathogen's infectivity.
+We also took 10 values for the variance of the patch quality distribution, ranging from 0.02 to 0.2.
 For each simulation (100 replicates for each of the 100 longevity-variance combinations), qualities for 100 patches were drawn from a uniform distribution with fixed mean 1 and given variance.
@@
 The state of the metapopulation was then tracked for 5000 time steps.
-We recorded the proportion of patches in each state at the end of each simulation and classified the simulation as host extinction, disease-free (no infectious populations after 5000 time steps), endemic (both susceptible and infectious populations present), or pandemic (only infectious populations).
+We recorded the proportion of patches in each state at the end of each simulation and also computed the mean effective population size of susceptible and infectious hosts over the last 500 time steps of the simulation (assuming an effective population size of $A_i$ for susceptible populations and $\nu A_i$ for infectious populations).
 In addition, to identify the roles played by individual patches of varying quality, for every simulation, we recorded the quality of each patch, the proportion of time spent in each state (i.e.
@@
 
+%\begin{table}[h!]   
+%\caption{Parameters of the SPOM model, their meaning, and the values assigned under different parameterizations.  Empty cells indicate the same value as the default parameterization.}
+%\begin{tabular}{p{1cm} p{4cm} p{1.5cm} p{1.5cm} p{1.5cm} p{1.5cm} p{1.5cm}}
+%Param. & Interpretation &  Default & Local movement & High env. stoch. & Dominant reservoir trans. & High virulence \\
+%\hline
+%$\xi_{im}$ & Effect of patch quality on immigration & 0.5 & -- & -- & -- & --\\
+%$\xi_{em}$ & Effect of population size on emigration & 0.5 & -- & -- & --& --\\
+%$D$& Inverse of mean dispersal distance & 5 & 2 & -- & -- & --\\
+%$d_{ij}$ & Distance between patch $i$ and $j$ & 1 $\forall i \neq j$ & 1 $\forall i, j$ neighbors & -- & -- & --\\
+%\hline
+%$\mu$ & Extinction rate of unit quality patch & 0.1 & -- & -- & -- & --\\
+%$\nu$ & Disease-induced reduction in population size & 0.2 & -- & -- & -- & 1\\
+%$\alpha$ & Strength of environmental stochasticity & 1 & -- & 0 & -- & --\\
+%\hline
+%$\delta$ & Probability of direct infection & 0.5 & -- & -- & 0.1 & --\\
+%$\gamma_0$ & Initial rate of infection from reservoir patch & 0.5 & -- & -- & -- & --
+%\end{tabular}
+%\label{params}
+%\end{table}
+
 \begin{table}[h!]   
-\caption{Parameters of the SPOM model, their meaning, and the values assigned under different parameterizations.  Empty cells indicate the same value as the default parameterization.}
-\begin{tabular}{l p{4cm} p{1.5cm} p{1.5cm} p{1.5cm} p{1.5cm} p{1.5cm}}
-Parameter & Interpretation &  Default & Local movement & High env. stoch. & Dominant reservoir trans. & High disease mortality \\
+\caption{Parameters of the SPOM model, their meaning, and the values assigned under alternative parameterizations.
+Empty cells indicate that the default values were used across all parameterizations.}
+\begin{tabular}{p{1cm} p{4cm} p{2cm} p{3cm}}
+Param. & Interpretation &  Default & Alternative \\
 \hline
-$\xi_{im}$ & Scaling parameter for effect of target patch quality on immigration & 0.5 & -- & -- & -- & --\\
-$\xi_{em}$ & Scaling parameter for effect of target patch quality on emigration & 0.5 & -- & -- & --& --\\
-$D$& Inverse of mean dispersal distance & 5 & 2 & -- & -- & --\\
-$d_{ij}$ & Distance between patch $i$ and $j$ & 1 $\forall i \neq j$ & 1 $\forall i, j$ neighbors & -- & -- & --\\
+$\xi_{im}$ & Effect of patch quality on immigration & 0.5 & -- \\
+$\xi_{em}$ & Effect of population size on emigration & 0.5 & -- \\
+$D$& Inverse of mean dispersal distance & 5 & 2 (local movement)\\
+$d_{ij}$ & Distance between patch $i$ and $j$ & 1 $\forall i \neq j$ & 1 $\forall i, j$ neighbors (local movement)\\
 \hline
-$\mu_S$ & Extinction rate of unit quality susceptible patch & 0.1 & -- & -- & -- & --\\
-$\mu_I$ & Extinction rate of unity quality infectious patch & 0.5 & -- & -- & -- & 1\\
-$\alpha$ & Strength of environmental stochasticity & 1 & -- & 0 & -- & --\\
+$\mu$ & Extinction rate of unit quality patch & 0.1 & -- \\
+$\nu$ & Disease-induced reduction in population size & 0.2 & 0.1 (high virulence)\\
+$\alpha$ & Strength of environmental stochasticity & 1 & 0 (high stochasticity)\\
 \hline
-$\delta$ & Probability of direct infection & 0.5 & -- & -- & 0.1 & --\\
-$\gamma_0$ & Initial rate of infection from reservoir patch & 0.5 & -- & -- & -- & --
+$\delta$ & Probability of direct infection & 0.5 & 0.1 (dominant env. transmission)\\
+$\gamma_0$ & Initial rate of infection from reservoir patch & 0.5 & -- 
 \end{tabular}

### 2015-03-31 Manuscript/manuscript.tex
Updates to methods

@@
 
-\subsection{Model Structure}
+\subsection{Stochastic patch occupancy model}
 
@@
 \end{align}
-where $\phi(X_j)$ is an indicator function that is $1$ if patch $j$ is in state $X$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality (i.e.
-population size); $D$ is the inverse of mean dispersal distance; and $d_{ij}$ is the distance between patches $i$ and $j$.
-Essentially, $C_{Xi}$ sums the colonization effort from all patches in state $X$ to patch $i$.
+where $\phi(X_j)$ is an indicator function that is $1$ if patch $j$ is in state $X$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch population size; $D$ is the inverse of mean dispersal distance; and $d_{ij}$ is the distance between patches $i$ and $j$.
+Essentially, $C_{Xi}$ sums the colonization effort (or propagule pressure) from all patches in state $X$ to focal patch $i$.
 
@@
 
-\subsection{Model Parameterization}
+\subsection{Model parameterization}
 
-The model outlined above was parameterized in five different ways to explore the consequences of different assumptions (Table~\ref{params}).
-The default model was parameterized  according to the following assumptions.
+The model was parameterized  according to the following assumptions.
 (1) The rate of emigration from a patch increases with population size, while the extinction rate of a patch decreases with populations size ($\xi_{im}=0.5$, $\alpha=1$, \cite{Hanski2003}).
-(2) High quality patches attract more colonists than low quality patches ($\xi_{em}=0.5$, \cite{Hanski2003}).
+(2) Migrants preferentially select high quality habitat, that is, high quality patches attract more colonists than low quality patches ($\xi_{em}=0.5$, \cite{Hanski2003}).
 (3)  Connectivity among patches is determined entirely by quality ($d_{ij}=1$ for all $i \neq j$).
-In addition to these assumptions, parameters were chosen so that without infection, approximately 0.75 of the metapopulation was occupied ($\mu = 0.1$, $D=5$), and a range of epidemiological behaviors -- disease-free, endemic disease, and pandemic disease (i.e.
-all host populations infectious) -- was feasible, depending on the pathogen's environmental longevity ($\nu = 0.2$, $\delta = 0.5$, $\gamma_0 = 0.5$).
+In addition to these assumptions, parameters were chosen so that without infection, approximately 0.75 of the metapopulation was occupied ($\mu = 0.1$, $D=5$).
+To explore the dynamics of a range of possible pathogens, disease induced mortality ($\nu$), probability of direct infection ($\delta$), and environmental longevity were varied over a range of possible values.
 
 The baseline model above assumes that all patches are equally accessible from all other patches (i.e.
-are separated by distance 1), so to examine the effects of a more rigid spatial structure, we also implemented a model with patches arranged in a square lattice such that each patch is only accessible from its four neighboring patches (Local movement, Table~\ref{params}).
-In addition, we explored a model with $\alpha = 0$ so that the extinction rate was constant with patch quality, simulating the effects of high environmental stochasticity, i.e.
-environmental stochasticity is strong enough that the population size of a patch does not affect its extinction rate (High env.
-stoc., Table~\ref{params}).
-To explore the disease dynamics when transmission is driven largely by the environmental reservoir we parameterized a model with $\delta = 0.1$ to reduce the relative importance of direct transmission (Dominant reservoir trans., Table~\ref{params}).
-Lastly, to simulate a more virulent pathogen, we parameterized a model with $\nu = 0.1$ to substantially increase disease-induced mortality (High disease mortality, Table~\ref{params}).
+are separated by distance 1).
+To examine the effects of a more rigid spatial structure, we also implemented a model with patches arranged in a square lattice such that each patch is only accessible from its four neighboring patches (in which case, average migration distance was adjusted to $D = 2$ to maintain roughly equivalent occupancy).
 
@@
 
-For each of the above parameterizations, we performed two different simulation experiments to explore the combined effects of pathogen longevity and habitat quality on disease dynamics and metapopulation occupancy.
-In the first of these, we investigated the effect of increasing habitat heterogeneity on the dynamics of pathogens with a range of environmental longevities.
-For these simulations, we explored 10 values of pathogen longevity ranging from 0.1 to 3 times the expected residence time of a healthy population on a unit quality patch (i.e.
-10 units of time), with values representing the half-life of the pathogen's infectivity.
-We also took 10 values for the variance of the patch quality distribution, ranging from 0.02 to 0.2.
-For each simulation (100 replicates for each of the 100 longevity-variance combinations), qualities for 100 patches were drawn from a uniform distribution with fixed mean 1 and given variance.
+To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with either high or low quality habitat distributions (with patch qualities ranging from 0.76 to 1.77, and from 0.23 to 1.24, respectively).
+Specifically, we implemented a full-factorial design in which disease-induced mortality ($\nu$) and probability of direct infection ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).
+In addition, pathogen longevity, representing the half-life of the pathogen's infectivity in the environment, was varied over 3 values (ranging from 0.3 to 3 times the expected residence time of a healthy population on a unit quality patch, i.e., 10 units of time).
+
+For each simulation -- 100 replicates for each combination of disease (determined by $\nu$, $\delta$ and longevity), habitat quality distribution (high or low), and landscape structure (fully connected or lattice), qualities for 100 patches were drawn from a uniform distribution on the given habitat quality range (above).
 In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infected.
@@
 We recorded the proportion of patches in each state at the end of each simulation and also computed the mean effective population size of susceptible and infectious hosts over the last 500 time steps of the simulation (assuming an effective population size of $A_i$ for susceptible populations and $\nu A_i$ for infectious populations).
-In addition, to identify the roles played by individual patches of varying quality, for every simulation, we recorded the quality of each patch, the proportion of time spent in each state (i.e.
-occupied by susceptible hosts, occupied by infectious hosts, and unoccupied), and the number of transitions between each state.
 
-The above simulations held the mean of the patch quality distribution constant and explored the effect of increasing habitat heterogeneity around that mean (i.e.
-by expanding the quality distribution symmetrically about 1).
-To further parse out the relative influence of high and low quality patches separately, we performed a second series of simulations where we expanded the range of the uniform quality distribution to favor either high or low quality patches (i.e.
-by expanding the quality distribution asymmetrically about 1).
-For these simulations, at each of the 10 longevities used above, we simulated 100 metapopulations for each of four patch quality distributions: low variance, (patch qualities ranging 0.75 to 1.25); intermediate variance, high quality (range 0.75 to 1.75); intermediate variance, low quality (range 0.25 to 1.25); and high variance (range 0.25 to 1.75).
+To more carefully explore the potential for high quality habitat to serve as infectious traps, we performed an additional simulation experiment in which we varied the effect of habitat quality on immigration rate by choosing different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating migrants that select all patches with equal probability).
+To better isolate the effects of habitat preference and its interaction with pathogen longevity, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), and the patch quality distribution (with patches evenly spaced over the full range of qualities explored in the simulation above.
 
@@
 \begin{table}[h!]   
-\caption{Parameters of the SPOM model, their meaning, and the values assigned under alternative parameterizations.
-Empty cells indicate that the default values were used across all parameterizations.}
-\begin{tabular}{p{1cm} p{4cm} p{2cm} p{3cm}}
-Param. & Interpretation &  Default & Alternative \\
+\caption{Parameters of the SPOM model, their meaning, and the values, or range of values, assigned.}
+\begin{tabular}{p{1cm} p{6cm} r}
+Param. & Interpretation &  Value(s) \\
 \hline
-$\xi_{im}$ & Effect of patch quality on immigration & 0.5 & -- \\
-$\xi_{em}$ & Effect of population size on emigration & 0.5 & -- \\
-$D$& Inverse of mean dispersal distance & 5 & 2 (local movement)\\
-$d_{ij}$ & Distance between patch $i$ and $j$ & 1 $\forall i \neq j$ & 1 $\forall i, j$ neighbors (local movement)\\
+$\xi_{im}$ & Effect of patch quality on immigration (preference for high quality habitat) & 0, 0.5 \\
+$\xi_{em}$ & Effect of population size on emigration & 0.5  \\
+$D$& Inverse of mean dispersal distance & 2 (lattice), 5 (full) \\
+$d_{ij}$ & Distance between patch $i$ and $j$ & \parbox[t]{4cm}{\raggedleft $1 \forall i, j$ neighbors (lattice)\\ $1 \forall  i\neq j$ (full)}\\
 \hline
-$\mu$ & Extinction rate of unit quality patch & 0.1 & -- \\
-$\nu$ & Disease-induced reduction in population size & 0.2 & 0.1 (high virulence)\\
-$\alpha$ & Strength of environmental stochasticity & 1 & 0 (high stochasticity)\\
+$\mu$ & Extinction rate of unit quality patch & 0.1 \\
+$\nu$ & Disease-induced reduction in population size & 0.1 - 1 \\
+$\alpha$ & Strength of environmental stochasticity & 1 \\
 \hline
-$\delta$ & Probability of direct infection & 0.5 & 0.1 (dominant env. transmission)\\
-$\gamma_0$ & Initial rate of infection from reservoir patch & 0.5 & -- 
+$\delta$ & Probability of direct infection & 0 - 0.9 \\
+$\gamma_0$ & Initial rate of infection from reservoir patch & 0.5
 \end{tabular}

### 2015-04-06 Manuscript/manuscript.tex
Small methods edits/additions.

@@
 To address the above questions, we developed a stochastic patch occupancy model (SPOM) in which each patch can be in one of three possible states: occupied by susceptible hosts (S), occupied by infectious hosts (I), and unoccupied by the host ($\emptyset$).
-State transitions are governed by host colonization, extinction, and infection rates, where a susceptible population can become infectious either through direct contact with infectious immigrants, or through a local pathogen reservoir (Table~\ref{transitions}).
+State transitions are governed by host colonization, extinction, and infection rates, where a susceptible population can become infectious either through direct contact with infectious immigrants, or through a local environmental pathogen reservoir (Table~\ref{transitions}).
 
@@
 \end{equation}
-where $\gamma_0$ is the initial infection rate of the pathogen reservoir, $\tau_{i}$ is the time since last infectious occupancy ($\tau_{i} = \infty$, and thus $\gamma(\tau_i) = 0$, if the patch has never been occupied by infectious hosts), and $r$ is the pathogen's decay rate in the environment.
+where $\gamma_0$ is the initial infection rate of the pathogen reservoir, $\tau_{i}$ is the time since last infectious occupancy ($\tau_{i} = \infty$, and thus $\gamma(\tau_i) = 0$, if the patch has never been occupied by infectious hosts), and $r$ is the pathogen's decay rate in the environment (determined by the pathogen's environmental longevity).
 
@@
 
-The baseline model above assumes that all patches are equally accessible from all other patches (i.e.
+The parameterization above assumes that all patches are equally accessible from all other patches (i.e.
 are separated by distance 1).
@@
 To more carefully explore the potential for high quality habitat to serve as infectious traps, we performed an additional simulation experiment in which we varied the effect of habitat quality on immigration rate by choosing different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating migrants that select all patches with equal probability).
-To better isolate the effects of habitat preference and its interaction with pathogen longevity, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), and the patch quality distribution (with patches evenly spaced over the full range of qualities explored in the simulation above.
+To better isolate the effects of habitat preference and its interaction with pathogen longevity in these simulations, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), landscape structure (fully connected), and the patch quality distribution (with patches evenly spaced over the full range of qualities explored in the simulation above).
+Here again we performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_im$) and recorded the proportion of time spent occupied by infectious hosts for each patch (as a measure of disease risk) and the final total effective population size of the host.
 

### 2015-04-09 Manuscript/manuscript.tex
Figure 1 results

@@
 
-In the baseline parameterization, infectious occupancy increased substantially with longevity, while susceptible and total occupancy decreased (Supplemental).
-However, increasing the variance of the patch quality distribution around a mean of 1 had relatively little effect on average occupancy (Supplemental).
-The four alternative parameterizations (models with local movement, high environmental stochasticity, dominant reservoir transmission, and high disease-induced mortality) produced qualitatively similar results, though pathogen spread in these models was generally more limited (Supplemental).
+A wide range of disease dynamics were observed over the range of pathogens and habitat quality distributions explored.  
+Epidemic outcome was determined largely by disease induced mortality, with low disease-induced mortality (i.e.
+high $\nu$) facilitating pandemic dynamics (in which no susceptible populations persist), and very high disease-induced mortality (i.e.
+low $\nu$) facilitating disease-free or endemic dynamics (Fig~\ref{endemic}).
+Pathogens with higher environmental longevities were generally able to persist and spread more easily, even with high disease-induced morality (Fig~\ref{endemic}(c, f)).  
+The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics, with higher quality distributions (with qualities ranging roughly from 0.75 to 1.75) allowing pandemic spread for a greater range of pathogens (Fig~\ref{endemic}, a-c vs.
+d-f).
 
@@
 \begin{figure}
-\caption{The effect of the presence of high and low quality patches on (a) median susceptible occupancy, (b) median infectious occupancy, and (c) median total occupancy.
-Red lines show hiqh quality metapopulations (patch qualities roughly between 0.75 and 1.75), blue lines show low quality metapopulations (patch qualities roughly between 0.25 and 1.25).}
-\label{sens}
+\caption{The proportion of simulations resulting in endemic disease dynamics (persistence of both susceptible and infectious hosts) as a function of disease-induced mortality ($\nu$), and the probability of direct transmission ($\delta$).
+Panel columns show low (0.3, a, d), medium (1, b, e), and high (3.2, c, f) pathogen longevities, while rows show high quality (ranging from 0.76 to 1.77, a - c) and low quality (ranging from 0.23 to 1.24, d - f) patch quality distributions.
+Darker shading corresponds to more frequent endemic dynamics.
+The white region above the endemic region generally corresponds to pandemic dynamics (no susceptible host populations persist), while the region below the endemic region corresponds to disease-free dynamics.}
+\label{endemic}
 \end{figure}

### 2015-05-26 Manuscript/manuscript.tex
Updated results and figure captions

@@
 
-We found that distributions with more low quality habitat (distribution ranging from 0.25 to 1.25) increased mean susceptible occupancy, while distributions with more high quality habitat (range from 0.75 to 1.75) decreased mean susceptible occupancy, with the largest differences observed at intermediate longevities (Fig~\ref{sens}(a)).
-Conversely, low quality habitat distributions decreased infectious occupancy, while high quality distributions increased infectious occupancy (Fig~\ref{sens}(b)).
-These general trends were consistent across the four other parameterizations (Supplemental).
+However, despite facilitating pathogen spread, for pathogens with either low or high disease-induced mortalities, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions (Fig~\ref{highvlow}).
+In the case of pathogens that were unable to persist due to high disease-induced mortality and weak transmission, high quality habitat distributions supported greater occupancy than low quality distributions due to the greater stability and connectivity afforded by high quality habitat.
+Similarly, when low disease-induced mortality and high transmission (from either environmental or direct transmission) allowed a pathogen to spread widely regardless of habitat quality, high quality distributions again supported larger host populations than low quality distributions.
 
-However, despite facilitating pathogen spread, at high pathogen longevities, high quality patch distributions had a net positive effect on total occupancy relative to low quality distributions (Fig~\ref{sens}(c)).
-Due to their lower extinction rates and high connectivity, high quality patches were able to maintain stable occupancy despite the widespread infection facilitated by high pathogen longevity.
-On the other hand, though low quality patch distributions inhibited pathogen spread at intermediate longevities, when high pathogen longevities led to pandemic dynamics, low quality distributions were unable to support infectious occupancy, resulting in an increase in host extinction events and a decrease in mean total occupancy.
+For pathogens falling between these extremes (i.e. with intermediate disease-induced mortality), the habitat quality distribution had a substantial effect on pathogen spread.  
+In particular, for a given disease, low quality habitat distributions slowed pathogen spread relative to high quality distributions and as a result supported larger host population sizes (Fig ~\ref{highvlow}, \ref{popsizes}).
+For a given pathogen in this range (e.g.
+with $\nu = 0.2$ and $\delta = 0.3$), this relative advantage of low quality habitat increased with increasing environmental longevity (Fig~\ref{popsizes}), largely due to the relative increase in the persistence and spread of environmentally long-lived pathogens in high quality habitat distributions.
 
-The other model parameterizations produced a similar qualitative pattern, with the differences between total occupancy on high and low quality patches generally producing a "u" shape as a function of pathogen longevity (Fig~\ref{highvlow}).
-However, the five parameterizations differed in the magnitude and sign of this general relationship.
-High environmental stochasticity (i.e.
-$\alpha = 0$) produced very similar results to the default model, though with a larger positive effect of low quality habitat at intermediate longevity (Fig~\ref{highvlow}(c)).
-In simulations with reservoir transmission the dominant infection pathway ($\delta = 0.1$), the high quality patch distributions always produced occupancy higher than (or equal to ) the low quality patch distributions (Fig~\ref{highvlow}(d)), while in the models with local movement or high disease-induced mortality, low quality patch distributions actually produced higher occupancies than high quality distributions at high longevities (Fig~\ref{highvlow}(b, e)).
-
-To understand the mechanisms behind these effects, we further examined within patch infection dynamics.
-To compare across the full range of patch qualities, and to ensure the opportunity for both suscepible and infectious occupancy, we limit this analysis to high variance metapopulations (with qualities ranging from 0.25 to 1.75) that produced endemic dynamics.
+To better understand the mechanisms behind these effects, we further examined within patch infection dynamics for a pathogen in this intermediate range (again with $\nu = 0.2$ and $\delta = 0.3$).  
 Because of their lower extinction rate and higher recolonization rate, high quality patches supported more consistent occupancy than low quality patches (Fig~\ref{simvis}).  
-However, as the epidemic progressed, high quality patches increasingly supported infectious occupants, while low quality patches supported infectious hosts only briefly and infrequently (Fig~\ref{simvis}).
-In addition, while high quality patches did support susceptible occupancy, they experienced a greater number of infection events (susceptible to infectious transitions) than low quality patches (Fig~\ref{infections}).
-A similar trend was observed across the other parameterizations, though the observed relationship between patch quality and number of infection events was shallower for the lattice and weak direct infection models (Supplemental).
+However, disease risk, measured by the probability of infection, increased with patch quality, an effect that became more pronounced as pathogen longevity increased (Fig~\ref{pinfection}).
+Indeed, as an epidemic progressed, high quality patches increasingly supported infectious occupants, while low quality patches supported infectious hosts only briefly and infrequently (Fig~\ref{simvis}).
+Thus, from the perspective of a migrant individual, high quality habitat posed increased disease risk, with the preferential selection of high quality habitat (i.e.
+$\xi_{im} = 0.5$) increasing this disease risk further (Fig~\ref{pinfection}, a-c vs.
+d-f).
+Moreover, preferential selection of high quality habitat had dramatic consequences for the stability of the host metapopulation when the pathogen is long-lived in the environment, increasing extinction risk substantially (Fig~\ref{pext}).
 
@@
 \caption{The proportion of simulations resulting in endemic disease dynamics (persistence of both susceptible and infectious hosts) as a function of disease-induced mortality ($\nu$), and the probability of direct transmission ($\delta$).
-Panel columns show low (0.3, a, d), medium (1, b, e), and high (3.2, c, f) pathogen longevities, while rows show high quality (ranging from 0.76 to 1.77, a - c) and low quality (ranging from 0.23 to 1.24, d - f) patch quality distributions.
+Panel columns show low (0.3, a, d), medium (1.0, b, e), and high (3.2, c, f) pathogen longevities, while rows show high quality (ranging from 0.76 to 1.77, a - c) and low quality (ranging from 0.23 to 1.24, d - f) patch quality distributions.
 Darker shading corresponds to more frequent endemic dynamics.
@@
 \centering
-\caption{Differences between the median total occupancy of high quality patch distributions (quality between 0.75 and 1.75) and low quality patch distributions (quality between 0.25 to 1.25) across the five different model parameterizations.
-(a) default model, (b) local movement, (c) high environmental stochasticity, (d) dominant reservoir transmission, (e) high disease-induced mortality.}
+\caption{Differences between the median total effective population size of hosts on high quality patch distributions (quality between 0.76 and 1.77) and low quality patch distributions (quality between 0.23 to 1.24) as a function of disease-induced mortality ($\nu$) and the probability of direct transmission ($\delta$).
+Reds indicate higher median host population sizes in high quality patch distributions, while blues indicate higher median host population sizes in low quality patch distributions.
+Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.}
 \label{highvlow}
@@
 \centering
-\caption{Results from a single representative simulation with variance 0.2 and longevity of 0.3 times the expected residence time of a susceptible population on a unit quality patch.
-The top panel shows total occupancy of susceptible (blue) and infectious (red) patches over time.
+\caption{Boxplots showing the range, over 100 replicate simulations, of susceptible (a), infectious (b), and total (c) host population sizes for high quality (red) and low quality (blue) habitat distributions.
+Disease-induced mortality ($\nu$) and direct transmission rate ($\delta$) are fixed at values of 0.2 and 0.3, respectively.
+Colored points show medians, while vertical lines indicate the outer-quartile ranges, with gray points indicating outliers (determined as points outside 1.5 times the inner-quartile range).}
+\label{popsizes}
+\end{figure}
+
+\begin{figure}
+\centering
+\caption{Results from a single representative simulation with habitat quality uniformly distributed between 0.23 and 1.77, and a pathogen with a disease induced mortality of $\nu = 0.2$, a direct transmission rate of $\delta = 0.3$, and an envioronmental longevity equal to the expected residence time of a susceptible population on a unit quality patch (10 units of time here).
+The top panel shows total proportion of patches occupied by susceptible (blue) and infectious (red) hosts over time.
 The bottom panel shows the state -- susceptible (blue), infectious (red), or unoccupied (white) -- of individual patches through time.
-Patches are stacked vertically with the lowest quality ($\sim$ 0.25) at the bottom and the highest quality ($\sim$ 1.75) at the top.}
+Patches are stacked vertically with the lowest quality at the bottom and the highest quality at the top.}
 \label{simvis}
@@
 \centering
-\caption{The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
-Light gray lines show smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
-\label{infections}
+\caption{Boxplots showing the probability of infection (i.e.
+proportion of time occupied by infectious hosts) as a function of patch quality.
+Panel columns show low (0.3, a, d), medium (1.0, b, e), and high (3.2, c, f) pathogen longevities, while rows show the effect of preferential selection of high quality habitat (a-c, $\xi_{im} = 0$,  no preference; d-f, $\xi_{im} = 0.5$, preference for high quality).
+Black points show the median over 100 replicate simulations, while vertical lines show the outer-quartile range, and gray points indicate outliers.
+Disease induced mortality and probability of direct infection fixed at $\nu = 0.2$ and $\delta = 0.3$.}
+\label{pinfection}
 \end{figure}
 
+\begin{figure}
+\centering
+\caption{Probability of host extinction as a function of pathogen environmental longevity and habitat preference, with black showing no habitat preference ($\xi_{im} = 0$) and red showing preference for high quality habitat ($\xi_{im} = 0.5$).
+Disease induced mortality and probability of direct infection fixed at $\nu = 0.2$ and $\delta = 0.3$.}
+\label{pext}
+\end{figure}
 \clearpage

### 2015-05-26 Analysis/simvis.r

@@ -6,7 +6,7 @@ source("Code/contSPOM(rates).r")
 # Defines fixed inputs for diseaseSPOM
 parms<-data.frame("xi_im" = 0.5, 
                   "xi_em" = 0.5,      
-                  "D" = 10, 
+                  "D" = 5, 
                   "alpha" = 1,
                   "es" = 0.1,
                   "nu" = 0.2,

### 2015-05-26 Manuscript/manuscript.tex
Edits to discussion of roles of low/high quality

@@
 This work demonstrates that the distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of environmentally transmitted pathogens and the resulting patterns of host occupancy.
-These impacts are driven largely by the contrasting effects of low and high quality habitat, and their relative balance in the metapopulation.
-These contrasting effects derive from differences in colonization and extinction rates, which in turn drive how low and high quality patches interact with and modulate pathogen transmission, either augmenting it in the case of high quality patches, or dampening it in the case of low quality habitat.
+These impacts are driven largely by differences in the colonization and extinction rates of high and low quality habitat, which in turn drive how low and high quality patches interact with and modulate pathogen transmission, either augmenting it in the case of high quality patches, or dampening it in the case of low quality habitat.
 
-When a patch's connectivity is determined by its quality (Eqn~\ref{connectivity}), high quality patches function as highly connected hubs in the metapopulation.
-Attracting colonists helps position high quality patches as infectious "traps", wherein susceptible hosts repeatedly colonize high quality patches and subsequently become infected (Fig~\ref{infections}).
-This increased infection pressure results from both transmission from a persistent environmental reservoir and from direct transmission from infectious immigrants attracted to the high quality patch, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).
+Because high quality habitat supports larger host population sizes, host populations on high quality habitat are less susceptible to environmental stochasticity and thus experience fewer extinction events.
+Due to this additional stability, high quality habitat is more consistently occupied, by both susceptible and infectious hosts, leading to the increase in infection probability with patch quality.
+From the perspective of a migrant susceptible host then, selection of high quality habitat leads to increased disease risk relative to other habitats, either because the patch is likely already occupied by infected hosts, or because susceptible hosts are likely to become infected (either through the environmental reservoir or direct contact).
+Thus, when high quality habitat is preferentially selected by migrants (i.e.
+$\xi_em > 0$), high quality habitats form ecological traps, that is, preferentially selected habitat that reduces fitness relative to other habitats \cite{Robertson2006}).
+Moreover, this preferential selection of high quality habitat further exacerbates the increase in disease risk with habitat quality (Fig~\ref{pinfection}) by increasing the risk of direct infection from infectious migrants also selecting high quality habitat, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).
 
-The relatively low extinction rate on high quality patches can help to create a stable platform from which the pathogen can spread through the rest of the metapopulation.
+This relatively low extinction rate on high quality patches, and the resulting increased probability of infection, helps then to create a relatively stable platform from which the pathogen can spread through the rest of the metapopulation.
 Indeed, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).  
-Once the pathogen infects high quality patches, the larger number of host colonists produced by these patches allow it to spread to the rest of the metapopulation relatively easily.
-This process helps to maintain infectious occupancy throughout the metapopulation, which in turn feeds back on high quality patches to ensure a steady stream of infectious colonists that help maintain the environmental reservoir and the trap effect.
-Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread and infectious occupancy in the metapopulation (Fig~\ref{sens}).
+This helps to maintain infectious occupancy throughout the metapopulation, which in turn feeds back on high quality patches to ensure a steady stream of infectious colonists that help maintain the environmental reservoir and the trap effect.
+Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread, even for more virulent pathogens (Fig~\ref{endemic}, \ref{popsizes}).
 
-In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{sens}).
-Even though individual low quality patches, due to their high extinction rates and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}), the presence of low quality patches increases overall susceptible occupancy relative to a lower variance metapopulation (Fig~\ref{sens}).
-This phenomenon represents the other side of the hub role played by high quality patches in that low quality patches are relatively weakly connected to the metapopulation and are therefore infrequently colonized by infectious individuals.
-Moreover, the high extinction rate, together with the increased mortality of infectious populations, means that infectious populations do not persist long on low quality patches and the environmental reservoir left behind by these infectious populations may decay before it has the opportunity to infect new susceptible colonists.
-As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infectious occupancy (Fig~\ref{sens}).
+In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{endemic}, \ref{popsizes}). 
+Even though individual low quality patches, due to their high extinction rates and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}), the presence of low quality patches facilitates susceptible persistence for a greater range of pathogens (Fig~\ref{endemic}).
+This phenomenon represents the other side of the role played by high quality patches in that low quality patches are relatively unstable (i.e.
+low population sizes and high extinction rates) and thus do not support infectious populations for long.
+Since they are infrequently colonized (when $\xi_em > 0$), the environmental reservoir left behind by these infectious populations likely decays before it has the opportunity to infect new susceptible colonists.
+As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infectious occupancy (Fig~\ref{endemic}, \ref{popsizes}).
 
-Though there is a consistent trend in the effect of low and high quality habitat on pathogen spread, the net consequences of these effects for total occupancy vary depending on the structure and dynamics of the metapopulation.
+Though there is a consistent trend in the effect of low and high quality habitat on pathogen spread, the net consequences of these effects for total occupancy vary depending on the traits of the pathogen.
 Specifically, the findings that low quality habitat inhibits disease spread while high quality habitat enhances it suggest the presence of a trade-off between between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.

### 2015-05-27 Manuscript/manuscript.tex
More discussion edits (relative benefit of high v. low)

@@
 
+Blah blah blah....something about lattice looking similar (supp info).
+
 \section{Discussion}
@@
 Thus, when high quality habitat is preferentially selected by migrants (i.e.
-$\xi_em > 0$), high quality habitats form ecological traps, that is, preferentially selected habitat that reduces fitness relative to other habitats \cite{Robertson2006}).
+$\xi_{em} > 0$), high quality habitats form ecological traps, that is, preferentially selected habitat that reduces fitness relative to other habitats \cite{Robertson2006}).
 Moreover, this preferential selection of high quality habitat further exacerbates the increase in disease risk with habitat quality (Fig~\ref{pinfection}) by increasing the risk of direct infection from infectious migrants also selecting high quality habitat, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).
@@
 low population sizes and high extinction rates) and thus do not support infectious populations for long.
-Since they are infrequently colonized (when $\xi_em > 0$), the environmental reservoir left behind by these infectious populations likely decays before it has the opportunity to infect new susceptible colonists.
+Since they are infrequently colonized (when $\xi_{em} > 0$), the environmental reservoir left behind by these infectious populations likely decays before it has the opportunity to infect new susceptible colonists.
 As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infectious occupancy (Fig~\ref{endemic}, \ref{popsizes}).
 
-Though there is a consistent trend in the effect of low and high quality habitat on pathogen spread, the net consequences of these effects for total occupancy vary depending on the traits of the pathogen.
-Specifically, the findings that low quality habitat inhibits disease spread while high quality habitat enhances it suggest the presence of a trade-off between between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.
-This work thus expands on previous studies of disease spread in a metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).
-Shifting the patch quality distribution towards high quality patches in our model is in many ways analogous to increasing movement in the models of Hess, Gog, and Park, but by including habitat heterogeneity and metapopulation structure, we are able to investigate the role of individual patches and their contribution to total occupancy.
+These findings that low quality habitat inhibits disease spread while high quality habitat enhances it suggest the presence of a trade-off between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.
+In this light, this work expands on previous studies of disease spread in a metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).
+Shifting the patch quality distribution towards high quality habitat in our model is  conceptually similar to increasing movement in these models, and as in these models, the net consequences of this trade-off for the host metapopulation depend on the traits of the pathogen.
+Broadly, we see three different scenarios: pathogen traits produce very weak pathogen spread and high quality habitat is beneficial; pathogen traits facilitate intermediate/endemic pathogen spread and low quality habitat is beneficial; and pathogen traits produce easy/pandemic pathogen spread and high quality habitat is again beneficial.
 
-These contributions, and the relative strength of the above trade-offs, depend largely on the role of connectivity, how it is affected by patch quality, and the resulting strength and structure of the different avenues of pathogen transmission.
-Broadly, we see three different scenarios that roughly correspond to the three sections of the "u"-shaped relationships in Fig~\ref{highvlow}: conditions that support very weak pathogen spread, conditions that facilitate intermediate/endemic pathogen spread, and conditions that produce easy/pandemic pathogen spread.
-When pathogen spread is weak (e.g.
-due to low longevity; the top left of the "u"), high quality patches are a net benefit to the metapopulation, as their connectivity and stability facilitate greater occupancy with little additional pathogen spread.
+When pathogen spread is weak, high quality patches are a net benefit to the metapopulation, as their connectivity and stability facilitate greater occupancy with little additional pathogen spread.
 
 However, in cases where the pathogen is able to spread more easily (i.e.
-at an endemic level; the trough of the "u"), the addition of high quality patches functions similarly to increasing movement in the model developed by \cite{Hess1996}: increasing pathogen spread and as a consequence reducing overall occupancy (Fig~\ref{sens}c).
+at an endemic level), the addition of high quality patches functions similarly to increasing movement in the model developed by \cite{Hess1996}: increasing pathogen spread and as a consequence reducing overall occupancy (Fig~\ref{sens}c).
 In these cases, the reduction in occupancy from the effect of high quality habitat on pathogen spread outweighs the positive consequences of increased metapopulation connectivity and stability.

### 2015-05-27 Manuscript/manuscript.tex
More discussion edits.

@@
 Due to this additional stability, high quality habitat is more consistently occupied, by both susceptible and infectious hosts, leading to the increase in infection probability with patch quality.
-From the perspective of a migrant susceptible host then, selection of high quality habitat leads to increased disease risk relative to other habitats, either because the patch is likely already occupied by infected hosts, or because susceptible hosts are likely to become infected (either through the environmental reservoir or direct contact).
+From the perspective of a migrant susceptible host then, selection of high quality habitat leads to increased disease risk relative to other habitats, either because the patch is likely already occupied by infectious hosts, or because susceptible hosts are likely to become infected (either through the environmental reservoir or direct contact).
 Thus, when high quality habitat is preferentially selected by migrants (i.e.
@@
 
-When pathogen spread is weak, high quality patches are a net benefit to the metapopulation, as their connectivity and stability facilitate greater occupancy with little additional pathogen spread.
+In the first of these scenarios, when pathogens have high disease-induced mortality and weak transmission (bottom left of panels in Fig~\ref{highvlow}), pathogen spread is so limited that habitat quality has little effect on epidemic dynamics.
+As a result, high quality patches are a net benefit to the host metapopulation relative to low quality habitat, because their connectivity and stability facilitate larger total population sizes with little additional pathogen spread.
 
-However, in cases where the pathogen is able to spread more easily (i.e.
-at an endemic level), the addition of high quality patches functions similarly to increasing movement in the model developed by \cite{Hess1996}: increasing pathogen spread and as a consequence reducing overall occupancy (Fig~\ref{sens}c).
-In these cases, the reduction in occupancy from the effect of high quality habitat on pathogen spread outweighs the positive consequences of increased metapopulation connectivity and stability.
-The shift in this trade-off can arise from conditions that change either side of that balance.
+Similarly, in the third scenario, when the pathogen spreads easily as a result of low disease-induced mortality and high transmission (upper portions of panels in Fig~\ref{highvlow}), high quality habitat is again a net benefit to the total host population size.
+Under these conditions, habitat quality again has a limited effect on epidemic dynamics, as the pathogen spreads easily enough to overwhelm any metapopulation structure.
+Thus, the benefit of high quality habitat derives primarily from its ability to maintain metapopulation stability in the face of widespread disease-induced mortality.
+
+However, in the second scenario, where the pathogen's traits facilitate intermediate spread (i.e.
+at an endemic level), the distribution of habitat quality exerts a large influence on epidemic dynamics.
+In these cases, the presence of high quality habitat functions similarly to increasing movement in the model developed by Hess (\cite{Hess1996}): increasing pathogen spread and as a consequence reducing overall host population size (Fig~\ref{highvlow}, \ref{popsizes}).
+Here, the consequences of increased pathogen spread outweigh the positive consequences of increased metapopulation connectivity and stability.  
 High quality habitat distributions require lower pathogen longevities for pathogen spread to begin to take off, relative to low quality habitat distributions.
@@
 
-The trade-off can also switch in cases that disrupt the positive benefits of high quality habitat for metapopulation stability, i.e.
-pathogens with high disease-induced mortality or when dispersal is limited by metapopulation structure.
 In these cases, rather than serving as sources that increase metapopulation occupancy, high quality patches become sinks.
@@
 Indeed, in these situations, our model suggests that low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can produce higher occupancy than high quality distributions.
-
-As pathogen longevity increases and environmental transmission becomes more important relative to direct transmission (i.e.
-the top right of the "u") the trade-off shifts back and the addition of high quality patches is again a net benefit to overall occupancy.
-In these situations, environmental reservoirs overwhelm the susceptible refuges on low quality patches, such that they remain contaminated even across infrequent colonization events.
-Under these conditions of widespread environmental transmission, high quality patches help to maintain metapopulation stability in the face of disease-induced mortality, thereby increasing total occupancy relative to more balanced or low quality patch distributions.
 

### 2015-06-01 Manuscript/manuscript.tex
Reorganization of discussion

@@
 
-This work demonstrates that the distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of environmentally transmitted pathogens and the resulting patterns of host occupancy.
-These impacts are driven largely by differences in the colonization and extinction rates of high and low quality habitat, which in turn drive how low and high quality patches interact with and modulate pathogen transmission, either augmenting it in the case of high quality patches, or dampening it in the case of low quality habitat.
+\subsection*{Introduction and outline of 3 regions}
+
+This work demonstrates that for many pathogens, the distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of environmentally transmitted pathogens and the resulting patterns of host occupancy.
+These impacts are driven largely by the interaction between pathogen traits (i.e.
+disease-induced mortality, transmission rates) and the colonization and extinction rates of high and low quality habitat.
+
+The general findings that low quality habitat can inhibit disease spread while high quality habitat can enhance it suggest the presence of a trade-off between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.
+In this light, this work expands on previous studies of disease spread in a metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).
+Shifting the patch quality distribution towards high quality habitat in our model is  conceptually similar to increasing movement in these models, and as in these models, the net consequences of this trade-off for the host metapopulation depend on the traits of the pathogen.
+Broadly, we see three different scenarios: pathogen traits produce very weak pathogen spread and high quality habitat is a net benefit; pathogen traits facilitate intermediate/endemic pathogen spread and low quality habitat is a net benefit; and pathogen traits produce easy/pandemic pathogen spread and high quality habitat is again a net benefit.
+
+\subsection*{Pathogens for which high quality beneficial}
+
+In the first of these scenarios, when pathogens have high disease-induced mortality and weak transmission (bottom left of panels in Fig~\ref{highvlow}), pathogen spread is so limited that habitat quality has little effect on epidemic dynamics.
+As a result, high quality patches are a net benefit to the host metapopulation relative to low quality habitat, because their connectivity and stability facilitate larger total population sizes with little additional pathogen spread.
+
+Similarly, in the third scenario, when the pathogen spreads easily as a result of low disease-induced mortality and high transmission (upper portions of panels in Fig~\ref{highvlow}), high quality habitat is again a net benefit to the total host population size.
+Under these conditions, habitat quality again has a limited effect on epidemic dynamics, as the pathogen spreads easily enough to overwhelm any metapopulation structure.
+Here then, the benefit of high quality habitat derives primarily from its ability to maintain metapopulation stability in the face of widespread disease-induced mortality.
+
+\subsection*{Pathogens for which high quality detrimental}
+
+However, in the second scenario, where the pathogen's traits facilitate intermediate spread (i.e.
+at an endemic level), the distribution of habitat quality exerts a large influence on epidemic dynamics.
+Here, the consequences of increased pathogen spread outweigh the positive consequences of increased metapopulation connectivity and stability.
+
+%In these cases, the presence of high quality habitat functions similarly to increasing movement in the model developed by Hess (\cite{Hess1996}): increasing pathogen spread and as a consequence reducing overall host population size (Fig~\ref{highvlow}, \ref{popsizes}).
+%High quality habitat distributions require lower pathogen longevities for pathogen spread to begin to take off, relative to low quality habitat distributions.
+Thus, for intermediate longevities, high quality distributions experience considerably greater pathogen spread and lower occupancy than low quality distributions.
+
+\subsubsection*{Role of high quality}
 
@@
 
+\subsubsection*{Role of low quality}
+
 In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{endemic}, \ref{popsizes}). 
@@
 
-These findings that low quality habitat inhibits disease spread while high quality habitat enhances it suggest the presence of a trade-off between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.
-In this light, this work expands on previous studies of disease spread in a metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).
-Shifting the patch quality distribution towards high quality habitat in our model is  conceptually similar to increasing movement in these models, and as in these models, the net consequences of this trade-off for the host metapopulation depend on the traits of the pathogen.
-Broadly, we see three different scenarios: pathogen traits produce very weak pathogen spread and high quality habitat is beneficial; pathogen traits facilitate intermediate/endemic pathogen spread and low quality habitat is beneficial; and pathogen traits produce easy/pandemic pathogen spread and high quality habitat is again beneficial.
-
-In the first of these scenarios, when pathogens have high disease-induced mortality and weak transmission (bottom left of panels in Fig~\ref{highvlow}), pathogen spread is so limited that habitat quality has little effect on epidemic dynamics.
-As a result, high quality patches are a net benefit to the host metapopulation relative to low quality habitat, because their connectivity and stability facilitate larger total population sizes with little additional pathogen spread.
-
-Similarly, in the third scenario, when the pathogen spreads easily as a result of low disease-induced mortality and high transmission (upper portions of panels in Fig~\ref{highvlow}), high quality habitat is again a net benefit to the total host population size.
-Under these conditions, habitat quality again has a limited effect on epidemic dynamics, as the pathogen spreads easily enough to overwhelm any metapopulation structure.
-Thus, the benefit of high quality habitat derives primarily from its ability to maintain metapopulation stability in the face of widespread disease-induced mortality.
-
-However, in the second scenario, where the pathogen's traits facilitate intermediate spread (i.e.
-at an endemic level), the distribution of habitat quality exerts a large influence on epidemic dynamics.
-In these cases, the presence of high quality habitat functions similarly to increasing movement in the model developed by Hess (\cite{Hess1996}): increasing pathogen spread and as a consequence reducing overall host population size (Fig~\ref{highvlow}, \ref{popsizes}).
-Here, the consequences of increased pathogen spread outweigh the positive consequences of increased metapopulation connectivity and stability.  
-High quality habitat distributions require lower pathogen longevities for pathogen spread to begin to take off, relative to low quality habitat distributions.
-Thus, for intermediate longevities, high quality distributions experience considerably greater pathogen spread and lower occupancy than low quality distributions.
+\subsubsection*{Overall consequences}
 
@@
 Indeed, in these situations, our model suggests that low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can produce higher occupancy than high quality distributions.
+
+\subsubsection*{Conclusions}
 

### 2015-06-01 Manuscript/manuscript.tex
More discussion edits

@@
 However, in the second scenario, where the pathogen's traits facilitate intermediate spread (i.e.
-at an endemic level), the distribution of habitat quality exerts a large influence on epidemic dynamics.
-Here, the consequences of increased pathogen spread outweigh the positive consequences of increased metapopulation connectivity and stability.
+at an endemic level), the distribution of habitat quality exerts a large influence on epidemic dynamics, with consequences for the overall host population size.
+In particular, the ability of these pathogens to persist and spread is influenced substantially by the colonization and extinction rates of high and low quality habitat, and the contrasting roles that these habitat types play.
 
-%In these cases, the presence of high quality habitat functions similarly to increasing movement in the model developed by Hess (\cite{Hess1996}): increasing pathogen spread and as a consequence reducing overall host population size (Fig~\ref{highvlow}, \ref{popsizes}).
 %High quality habitat distributions require lower pathogen longevities for pathogen spread to begin to take off, relative to low quality habitat distributions.
@@
 
-\subsubsection*{Role of high quality}
+\subsubsection*{Role of high quality habitat}
 
-Because high quality habitat supports larger host population sizes, host populations on high quality habitat are less susceptible to environmental stochasticity and thus experience fewer extinction events.
-Due to this additional stability, high quality habitat is more consistently occupied, by both susceptible and infectious hosts, leading to the increase in infection probability with patch quality.
-From the perspective of a migrant susceptible host then, selection of high quality habitat leads to increased disease risk relative to other habitats, either because the patch is likely already occupied by infectious hosts, or because susceptible hosts are likely to become infected (either through the environmental reservoir or direct contact).
-Thus, when high quality habitat is preferentially selected by migrants (i.e.
-$\xi_{em} > 0$), high quality habitats form ecological traps, that is, preferentially selected habitat that reduces fitness relative to other habitats \cite{Robertson2006}).
-Moreover, this preferential selection of high quality habitat further exacerbates the increase in disease risk with habitat quality (Fig~\ref{pinfection}) by increasing the risk of direct infection from infectious migrants also selecting high quality habitat, reflecting results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).
+Because high quality habitat supports larger host population sizes, populations on high quality habitat are less susceptible to environmental stochasticity and thus experience less frequent extinction events.
+Due to this additional stability, high quality habitat is more consistently occupied, by both susceptible and infectious hosts, leading to an increase in infection probability (i.e.
+proportion of time occupied by infectious hosts) with patch quality (Fig~\ref{pinfection}).
+From the perspective of a migrant susceptible host then, selection of high quality habitat leads to increased disease risk relative to other habitats, either because the patch is likely already occupied by infectious hosts, or because the local susceptible hosts are likely to soon become infected (either through the presences of an environmental reservoir or direct contact).
+Thus, if high quality habitat is preferentially selected by migrants (i.e.
+$\xi_{em} > 0$), high quality habitat becomes an ecological trap, that is, preferentially selected habitat that reduces fitness relative to other habitats \cite{Robertson2006}).
+Moreover, this preferential selection of high quality habitat further exacerbates the increase in disease risk with habitat quality (i.e., the fitness costs of high quality habitat, Fig~\ref{pinfection}) by increasing the risk of direct transmission from infectious migrants also selecting high quality habitat.
+This echoes results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).
 
-This relatively low extinction rate on high quality patches, and the resulting increased probability of infection, helps then to create a relatively stable platform from which the pathogen can spread through the rest of the metapopulation.
-Indeed, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).  
-This helps to maintain infectious occupancy throughout the metapopulation, which in turn feeds back on high quality patches to ensure a steady stream of infectious colonists that help maintain the environmental reservoir and the trap effect.
+The relatively low extinction rate on high quality patches, and the resulting increased probability of infection, helps to create a relatively stable platform from which the pathogen can spread through the rest of the metapopulation.
+Indeed, for pathogens with traits in this intermediate range, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).  
+This helps to maintain infectious occupancy throughout the metapopulation, which, when coupled with preferential selection of high quality habitat, feeds back on high quality patches to ensure a steady stream of infectious colonists that help maintain the environmental reservoir and the trap effect.
 Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread, even for more virulent pathogens (Fig~\ref{endemic}, \ref{popsizes}).
+The consequences of the concomitant widespread disease-induced mortality outweigh the positive consequences of increased metapopulation connectivity and stability, leading to a net decrease in total host population size on high quality habitat distributions (Fig~\ref{highvlow}, \ref{popsizes}).
 
-\subsubsection*{Role of low quality}
+\subsubsection*{Role of low quality habitat}
 
@@
 low population sizes and high extinction rates) and thus do not support infectious populations for long.
-Since they are infrequently colonized (when $\xi_{em} > 0$), the environmental reservoir left behind by these infectious populations likely decays before it has the opportunity to infect new susceptible colonists.
+Since they are infrequently colonized (when high quality habitat is preferentially selected, $\xi_{em} > 0$), the environmental reservoir left behind by these infectious populations likely decays before it has the opportunity to infect new susceptible colonists.
 As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infectious occupancy (Fig~\ref{endemic}, \ref{popsizes}).
@@
 
-In these cases, rather than serving as sources that increase metapopulation occupancy, high quality patches become sinks.
+Thus, for pathogens with traits that facilitate this intermediate level of spread, we see a reversal of the relative importance of high and low quality habitat for overall metapopulation stability.
+Here, rather than serving as sources that increase host population size, we see that high quality patches become sinks.  
 More importantly, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).  
-From a management perspective, it is critical to understand this shift, as previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable (\cite{Strasser2010}).
+From a management perspective, it is critical to understand this shift, as previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable for the host (\cite{Strasser2010}).
 Generally, this means that managers should focus on maintaining high quality habitat, but Strasser et al.
 (2010) show that stochastic disturbance (e.g.
-disease-induced mortality) can lead to cases where focusing on low quality patches is more effective in increasing population growth rate.
-Indeed, in these situations, our model suggests that low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can produce higher occupancy than high quality distributions.
+disease-induced mortality) can lead to cases where focusing on low quality habitat is more effective in increasing population growth rate.
+Indeed, in these situations, our model suggests that low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can maintain larger host populations than high quality distributions.
 
-\subsubsection*{Conclusions}
+\subsection*{Conclusions}
 
-These dynamics reflect the findings of \cite{Gog2002}, who found that under strong background infection from an alternative host, increasing movement increased total occupancy, despite facilitating pathogen spread.
-\cite{Park2012}, by adding environmental transmission to the mix, found more generally that in cases where background and environmental transmission were strong relative to direct transmission, increasing movement increased total occupancy monotonically, i.e.
-movement was a net benefit to the metapopulation.
+\subsubsection*{Discussion of Hess, Gog, Park and role of longevity}
+
+This partitioning of pathogen dynamics into three regions based on whether high quality habitat is a net benefit or detriment to the host metapopulation echoes the work of (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).
+In particular, for a given pathogen, Hess (\cite{Hess1996}) similarly identified three regions of disease dynamics which determined whether increased movement had a positive or negative effect on host occupancy.
+In cases where the host movement rate was either low enough that the pathogen was unable to persist or high enough to cause a pandemic, Hess found that further increasing movement increased host occupancy, while at intermediate levels that resulted in endemic disease, increasing movement decreased host occupancy.
+These three scenarios, distinguished by the host movement rate, map to the the three scenarios we outline here based on pathogen traits (disease-induced mortality and direct and environmental transmission).
+The conceptual similarity of these results suggests more generally that in metapopulations facing intermediate (i.e.
+endemic) pathogen spread, improving factors that increase host metapopulation stability and connectivity (e.g.
+increased movement, high quality habitat) can actually decrease total host population size as a result of increased pathogen spread.
+
+However, we find that as the environmental longevity of the pathogen increases, high quality habitat is a net benefit for a larger range of potential pathogens (Fig~\ref{highvlow}, red region grows from panels a.
+to c.).
+These dynamics reflect the findings of \cite{Gog2002}, who found that under strong background infection from an alternative host, increasing movement always increased total occupancy, despite facilitating pathogen spread.
+Park (\cite{Park2012}) expanded on this by adding environmental transmission to the mix and found that in cases where background and environmental transmission were strong relative to direct transmission, increasing movement was a net benefit to the host metapopulation.
 These studies, coupled with the results presented here, suggest that when there is a persistent source of infection throughout the metapopulation (e.g.
-when the pathogen is long-lived in the environment and direct infection is relatively less important), factors that facilitate metapopulation stability (i.e.
-increased movement, high quality patches) are a greater benefit than factors that inhibit pathogen spread (i.e.
+when the pathogen is long-lived in the environment), factors that facilitate metapopulation stability (i.e.
+increased movement, high quality patches) are generally a greater benefit than factors that inhibit pathogen spread (i.e.
 decreased movement, low quality patches).
 
-The different scenarios and parameterizations explored here represent a broad sample of the metapopulation dynamics that we might expect to observe in empirical systems.
-The consistent "u"-shaped trend in the relative benefit of low and high quality habitat as a function of pathogen longevity suggest that these trade-offs between metapopulation stability and pathogen spread are likely to be fairly general.
+The different scenarios and pathogens explored here represent a broad sample of the metapopulation dynamics that we might expect to observe in empirical systems.
+The consistent patterns in the relative benefit of low and high quality habitat as a function of pathogen traits suggest that these trade-offs between metapopulation stability and pathogen spread are likely to be fairly general.
 By evaluating the relative importance of direct versus environmental transmission, along with the effect of habitat quality on connectivity and extinction, we can begin to diagnose where empirical systems (e.g.
 plague in prairie dogs, or white nose syndrome in bats) fall on this spectrum and the corresponding consequences for management.
-
-Spatial structure and disease spread represent two major factors influencing the dynamics of wildlife populations.
-As we have seen, in cases where the pathogen is able to persist and remain infectious in the environment, these two factors can interact in complex ways.
-Failing to consider these interactions and their effect on the roles played by high and low quality habitat risks overlooking the importance of low-quality habitat for the persistence of wildlife populations and the importance of high-quality habitat for the persistence of wildlife disease.
+As we have seen, failing to consider the characteristics of a pathogen and their effect on the roles played by high and low quality habitat risks overlooking the importance of low-quality habitat for the persistence of wildlife populations and the importance of high-quality habitat for the persistence of wildlife disease.
 

### 2015-06-01 Manuscript/manuscript.tex
Clarity edits and supp lattice figures

@@
 \begin{abstract} 
-Patterns of habitat quality are expected to play an important role in the dynamics of wildlife metapopulations, with the expectation that high quality habitat serves to stabilize and maintain occupancy.
+The distribution of habitat quality is expected to play an important role in the dynamics of wildlife metapopulations, with the expectation that high quality habitat serves to stabilize and maintain occupancy.
 However, we might expect that the costs and benefits of different quality habitat might be modulated, or possibly even reversed, by the presence of an environmentally persistent pathogen.
@@
 Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\cite{Dragon1995}), plague (\cite{Eisen2008}), and white nose syndrome (\cite{Lindner2011}), among others.
-This environmental persistence creates environmental pathogen reservoirs from which hosts can become infected without direct contact with an infectious individual.
+This environmental persistence creates environmental pathogen reservoirs from which susceptible hosts can become infected without direct contact with an infectious individual.
 This additional transmission pathway can have important consequences for disease dynamics, with models showing that increased environmental persistence generally facilitates increased pathogen persistence and spread relative to direct transmission alone (\cite{Almberg2011}, \cite{Sharp2011}, \cite{Breban2009}).
@@
 Indeed, the quality of a habitat patch can affect its extinction and colonization rates, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).
-These processes in turn may then influence where environmental pathogen reservoirs get established and how they affect disease dynamics and host occupancy throughout the metapopulation.
+These processes in turn may influence how a pathogen spreads, where environmental pathogen reservoirs get established, and the resulting effects on host occupancy and population size.
 
-Specifically, we expect that high quality habitat patches, which support greater host density and traffic than lower quality habitat, might be more likely to form pathogen reservoirs.
+Specifically, we expect that high quality habitat patches, which support greater host density and traffic than lower quality habitat, might be more likely to form pathogen reservoirs and support consistently infectious populations.
 As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Almberg2011}).
@@
 plague in prairie dog colonies, \cite{George2013}), and thus understanding how environmental transmission interacts with patterns of habitat quality is critical to managing disease in these systems.
-In this study then, we seek to explore how a pathogen's environmental longevity (how long it can persist and remain infectious in the environment), and the distribution of habitat quality in a metapopulation interact to influence pathogen spread and host occupancy, with a specific interest in the roles played by high and low quality habitat.
+In this study then, we seek to explore how a pathogen's traits, including its environmental longevity (how long it can persist and remain infectious in the environment), interact with the distribution of habitat quality to influence pathogen spread and host occupancy, with a specific interest in the roles played by high and low quality habitat.
 
@@
 We assume that patch population size is proportional to patch quality ($A_i$), and that disease-induced mortality reduces the size of infectious populations by a factor of $\nu$.
-Building on the framework developed by \cite{Hanski2003} then,the rates at which patch \emph{i}, with quality $A_i$, is colonized by susceptible and infectious individuals, respectively, is given by its connectivity:
+Building on the framework developed by \cite{Hanski2003} then, the rates at which patch \emph{i}, with quality $A_i$, is colonized by susceptible and infectious individuals, respectively, is given by its connectivity:
 \begin{align}
@@
 \end{align}
-where $\phi(X_j)$ is an indicator function that is $1$ if patch $j$ is in state $X$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch population size; $D$ is the inverse of mean dispersal distance; and $d_{ij}$ is the distance between patches $i$ and $j$.
+where $\phi(X_j)$ is an indicator function that is $1$ if patch $j$ is in state $X$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality; $D$ is the inverse of mean dispersal distance; and $d_{ij}$ is the distance between patches $i$ and $j$.
 Essentially, $C_{Xi}$ sums the colonization effort (or propagule pressure) from all patches in state $X$ to focal patch $i$.
@@
 
-The model was parameterized  according to the following assumptions.
+The model was parameterized  according to the following assumptions (Table~\ref{params}).
 (1) The rate of emigration from a patch increases with population size, while the extinction rate of a patch decreases with populations size ($\xi_{im}=0.5$, $\alpha=1$, \cite{Hanski2003}).
-(2) Migrants preferentially select high quality habitat, that is, high quality patches attract more colonists than low quality patches ($\xi_{em}=0.5$, \cite{Hanski2003}).
+(2) Migrants preferentially select high quality habitat, that is, high quality patches attract more colonists than low quality patches ($\xi_{em}=0.5$, \cite{Hanski2003}, though we relax this below).
 (3)  Connectivity among patches is determined entirely by quality ($d_{ij}=1$ for all $i \neq j$).
@@
 
-For each simulation -- 100 replicates for each combination of disease (determined by $\nu$, $\delta$ and longevity), habitat quality distribution (high or low), and landscape structure (fully connected or lattice), qualities for 100 patches were drawn from a uniform distribution on the given habitat quality range (above).
+For each simulation -- 100 replicates for each combination of disease (determined by $\nu$, $\delta$, and longevity), habitat quality distribution (high or low), and landscape structure (fully connected or lattice), qualities for 100 patches were drawn from a uniform distribution on the given habitat quality range (above).
 In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infected.
@@
 To more carefully explore the potential for high quality habitat to serve as infectious traps, we performed an additional simulation experiment in which we varied the effect of habitat quality on immigration rate by choosing different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating migrants that select all patches with equal probability).
-To better isolate the effects of habitat preference and its interaction with pathogen longevity in these simulations, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), landscape structure (fully connected), and the patch quality distribution (with patches evenly spaced over the full range of qualities explored in the simulation above).
-Here again we performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_im$) and recorded the proportion of time spent occupied by infectious hosts for each patch (as a measure of disease risk) and the final total effective population size of the host.
+To isolate the effects of habitat preference and its interaction with pathogen longevity in these simulations, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), landscape structure (fully connected), and the patch quality distribution (with patches evenly spaced over the full range of qualities explored in the simulation above).
+Here again we performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by infectious hosts (as a measure of disease risk) and the final total effective host population size.
 
@@
 Epidemic outcome was determined largely by disease induced mortality, with low disease-induced mortality (i.e.
-high $\nu$) facilitating pandemic dynamics (in which no susceptible populations persist), and very high disease-induced mortality (i.e.
+high $\nu$) facilitating pandemic dynamics (in which no susceptible host populations persist), and very high disease-induced mortality (i.e.
 low $\nu$) facilitating disease-free or endemic dynamics (Fig~\ref{endemic}).
 Pathogens with higher environmental longevities were generally able to persist and spread more easily, even with high disease-induced morality (Fig~\ref{endemic}(c, f)).  
-The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics, with higher quality distributions (with qualities ranging roughly from 0.75 to 1.75) allowing pandemic spread for a greater range of pathogens (Fig~\ref{endemic}, a-c vs.
+The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics, with higher quality distributions (with qualities ranging roughly from 0.75 to 1.75) facilitating pandemic spread for a greater range of pathogens (Fig~\ref{endemic}, a-c vs.
 d-f).
+Metapopulations with a lattice structure exhibited qualitatively similar patterns, but with generally more limited pathogen spread and larger regions of endemic dynamics (Supplement).
 
@@
 In particular, for a given disease, low quality habitat distributions slowed pathogen spread relative to high quality distributions and as a result supported larger host population sizes (Fig ~\ref{highvlow}, \ref{popsizes}).
-For a given pathogen in this range (e.g.
-with $\nu = 0.2$ and $\delta = 0.3$), this relative advantage of low quality habitat increased with increasing environmental longevity (Fig~\ref{popsizes}), largely due to the relative increase in the persistence and spread of environmentally long-lived pathogens in high quality habitat distributions.
+For a given pathogen in this range (here with $\nu = 0.2$ and $\delta = 0.3$), the relative advantage of low quality habitat increased with increasing environmental longevity (Fig~\ref{popsizes}), largely due to the more pronounced effect of increased longevity on pathogen persistence and spread in high quality habitat distributions.
+Again, metapopulations with a lattice structure displayed similar patterns, but required lower disease-induced mortality to see the benefit of low quality habitat (Supplement).
 
@@
 Because of their lower extinction rate and higher recolonization rate, high quality patches supported more consistent occupancy than low quality patches (Fig~\ref{simvis}).  
-However, disease risk, measured by the probability of infection, increased with patch quality, an effect that became more pronounced as pathogen longevity increased (Fig~\ref{pinfection}).
+However, disease risk, measured by the proportion of time a patch is occupied by infectious hosts, increased with patch quality, an effect that became more pronounced as pathogen longevity increased (Fig~\ref{pinfection}).
 Indeed, as an epidemic progressed, high quality patches increasingly supported infectious occupants, while low quality patches supported infectious hosts only briefly and infrequently (Fig~\ref{simvis}).
-Thus, from the perspective of a migrant individual, high quality habitat posed increased disease risk, with the preferential selection of high quality habitat (i.e.
+Thus, from the perspective of a migrant individual, colonizing high quality habitat posed increased disease risk, with the preferential selection of high quality habitat (i.e.
 $\xi_{im} = 0.5$) increasing this disease risk further (Fig~\ref{pinfection}, a-c vs.
 d-f).
-Moreover, preferential selection of high quality habitat had dramatic consequences for the stability of the host metapopulation when the pathogen is long-lived in the environment, increasing extinction risk substantially (Fig~\ref{pext}).
-
-Blah blah blah....something about lattice looking similar (supp info).
+Moreover, preferential selection of high quality habitat had dramatic consequences for the stability of the host metapopulation when the pathogen is long-lived in the environment, increasing extinction risk substantially compared to random habitat selection (i.e.
+$\xi_{em} = 0$, Fig~\ref{pext}).
 

### 2015-06-01 Analysis/pubplots.R

@@ -6,9 +6,10 @@ library(RColorBrewer)
 library(ggplot2)
 library(ggthemes)
 
-###
+#===============================================================================
 ### Figure 1
-###
+
+pdf("Manuscript/figure/figure_1.pdf", width = 10, height = 6)
 
 load(paste(getwd(), "/Output/pathgrid.RData", sep = ""))
 
@@ -39,12 +40,15 @@ key <- draw.colorkey(list(col = cols,
 grid.arrange(arrangeGrob(low.l, med.l, high.l, low.h, med.h, high.h, ncol = 3), 
              key, ncol = 2, widths = c(0.9, 0.1))
 
+dev.off()
 
-###
+#===============================================================================
 ### Figure 2
-###
 
 rm(list = ls())
+
+pdf("Manuscript/figure/figure_2.pdf", width = 10, height = 5)
+
 load(paste(getwd(), "/Output/pathgrid.RData", sep = ""))
 
 dat <- tapply(out$S.pop + out$I.pop, list(out$longevity, out$treatment, out$delta, out$nu), median)
@@ -66,11 +70,14 @@ key <- draw.colorkey(list(col = cols,
 
 grid.arrange(low, med, high, key, ncol = 4, widths = c(0.3, 0.3, 0.3, 0.1))
 
-###
+dev.off()
+
+#===============================================================================
 ### Figure 3
-###
 
 rm(list = ls())
+
+pdf("Manuscript/figure/figure_3.pdf", width = 10, height = 5)
 load(paste(getwd(), "/Output/pathgrid.RData", sep = ""))
 sub <- out[abs(out$delta - 0.3) < 0.001 & out$nu == 0.2, ]
 
@@ -92,18 +99,20 @@ p.tot <- p.tot + geom_tufteboxplot(aes(colour = factor(treatment)), outlier.colo
 
 grid.arrange(p.S, p.I, p.tot, ncol = 3)
 
+dev.off()
 
-###
+#===============================================================================
 ### Figure 4
-###
 
 # See code in simvis.r
 
-###
+#===============================================================================
 ### Figure 5
-###
 
 rm(list = ls())
+
+pdf("Manuscript/figure/figure_5.pdf", width = 12, height = 6)
+
 load(paste(getwd(), "/Output/trap.RData", sep = ""))
 
 quality <- unique(out$quality)
[…]

### 2015-06-01 Analysis/simvis.r

@@ -3,6 +3,8 @@
 
 source("Code/contSPOM(rates).r")
 
+set.seed(123)
+
 # Defines fixed inputs for diseaseSPOM
 parms<-data.frame("xi_im" = 0.5, 
                   "xi_em" = 0.5,      
@@ -95,7 +97,7 @@ p2 <- ggplot(tseries, aes(x = Time, y = Occupancy, group = class, color = class,
               scale_x_continuous(labels = NULL, expand = c(0, 0)) +
               scale_y_continuous(expand = c(0, 0))
 
-png("Manuscript/simvis.png", width = 836, height = 506)
+pdf("Manuscript/figure/figure_4.pdf", width = 12, height = 8)
 
 grid.arrange(p2, p1, ncol = 1, heights = c(0.25, 0.75))
 

### 2015-06-02 Manuscript/supplementary.tex
Updated supplement

@@
 
-\title{Supplementary Figures}
+\title{Supplementary figures}
 
@@
 
-\section{Occupancy results for symmetric quality distributions}
+This supplement contains results from simulations on a metapopulation with patches arranged in a lattice structure (i.e.
+each habitat patch is accessible only from its four nearest neighbors).
+Supplemental figures 1 and 2 are analogous to figures 1 and 2 from the main text.
 
-\begin{figure}[h!]
-\centering
-\includegraphics[height=5.5cm]{figure/metapop(full).pdf}
-\caption{Default model parameterization.
-Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay, measured in units of expected occupancy time of a susceptible population on a unit quality patch) and the variance of the patch quality distribution.}
-\label{poutcome}
-\end{figure}
-
-\begin{figure}[h!]
-\centering
-\includegraphics[height=5.5cm]{figure/metapop(lattice).pdf}
-\caption{Metapopulation with lattice structure and local movement.
-Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay, measured in units of expected occupancy time of a susceptible population on a unit quality patch) and the variance of the patch quality distribution.}
-\label{poutcome_lattice}
-\end{figure}
-
-\begin{figure}[h!]
-\centering
-\includegraphics[height=5.5cm]{figure/metapop(alpha0).pdf}
-\caption{High environmental stochasticity ($\alpha = 0$).
-Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay, measured in units of expected occupancy time of a susceptible population on a unit quality patch) and the variance of the patch quality distribution.}
-\label{poutcome_x0}
-\end{figure}
-
-\begin{figure}[h!]
-\centering
-\includegraphics[height=5.5cm]{figure/metapop(delta).pdf}
-\caption{Reservoir transmission dominant ($\delta = 0.1$).
-Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay, measured in units of expected occupancy time of a susceptible population on a unit quality patch) and the variance of the patch quality distribution.}
-\label{poutcome_delta}
-\end{figure}
-
-\begin{figure}[h!]
-\centering
-\includegraphics[height=5.5cm]{figure/metapop(ei).pdf}
-\caption{High disease-induced mortality ($\mu_I = 1$).
-Mean occupancy of (a) susceptible patches, (b) infectious patches, and (c) both, as a function of the pathogen's environmental longevity (half-life of infectivity decay, measured in units of expected occupancy time of a susceptible population on a unit quality patch) and the variance of the patch quality distribution.}
-\label{poutcome_ei}
-\end{figure}
-
-\clearpage
-
-\section{Occupancy results for asymmetric quality distributions}
-
-\begin{figure}[h!]
-\centering
-\includegraphics[height=7cm]{figure/highlow(lattice).pdf}
-\caption{Metapopulation with lattice structure and local movement.
-The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
-Red lines show high quality metapopulations (patch qualities roughly between 0.75 and 1.75), blue lines show low quality metapopulations (patch qualities roughly between 0.25 and 1.25).}
-\label{sens_lattice}
+\begin{figure}
+\caption{The proportion of simulations resulting in endemic disease dynamics (persistence of both susceptible and infectious hosts) as a function of disease-induced mortality ($\nu$), and the probability of direct transmission ($\delta$).
+Panel columns show low (0.3, a, d), medium (1.0, b, e), and high (3.2, c, f) pathogen longevities, while rows show high quality (ranging from 0.76 to 1.77, a - c) and low quality (ranging from 0.23 to 1.24, d - f) patch quality distributions.
+Darker shading corresponds to more frequent endemic dynamics.
+The white region above the endemic region generally corresponds to pandemic dynamics (no susceptible host populations persist), while the region below the endemic region corresponds to disease-free dynamics.}
+\label{endemic}
 \end{figure}
@@
 \centering
-\includegraphics[height=6cm]{figure/highlow(alpha0).pdf}
-\caption{High environmental stochasticity ($\alpha = 0$).
-The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
-Red lines show high quality metapopulations (patch qualities roughly between 0.75 and 1.75), blue lines show low quality metapopulations (patch qualities roughly between 0.25 and 1.25).}
-\label{sens_x0}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[height=6cm]{figure/highlow(delta).pdf}
-\caption{Reservoir transmission dominant ($\delta = 0.1$).
-The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
-Red lines show high quality metapopulations (patch qualities roughly between 0.75 and 1.75), blue lines show low quality metapopulations (patch qualities roughly between 0.25 and 1.25).}
-\label{sens_delta}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[height=6cm]{figure/highlow(ei).pdf}
-\caption{High disease-induced mortality ($\mu_I = 1$).
-The effect of the presence of high and low quality patches on (a) mean susceptible occupancy, (b) mean infectious occupancy, and (c) mean total occupancy.
-Red lines show high quality metapopulations (patch qualities roughly between 0.75 and 1.75), blue lines show low quality metapopulations (patch qualities roughly between 0.25 and 1.25).}
-\label{sens_ei}
-\end{figure}
-
-\clearpage
-
-\section{Patch-level infection events}
-
-\begin{figure}[h!]
-\centering
-\includegraphics[height=7cm]{figure/infevents(lattice).pdf}
-\caption{Metapopulation with lattice structure and local movement.
-The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
-Light gray lines show the smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
-\label{infections_lattice}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[height=7cm]{figure/infevents(alpha0).pdf}
-\caption{High environmental stochasticity ($\alpha = 0$).
-The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
-Light gray lines show the smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
-\label{infections_x0}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[height=7cm]{figure/infevents(delta).pdf}
-\caption{Reservoir transmission dominant ($\delta = 0.1$).
-The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
-Light gray lines show the smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
-\label{infections_delta}
-\end{figure}
-
-\begin{figure}
-\centering
-\includegraphics[height=7cm]{figure/infevents(ei).pdf}
-\caption{High disease-induced mortality ($\mu_I = 1$).
-The number of infection events ($S$ to $I$ transitions) for individual patches in endemic simulations.
-Light gray lines show the smoothed fit for single simulations, while the thick black line shows the smoothed trend over all simulations.}
-\label{infections_ei}
+\caption{Differences between the median total effective population size of hosts on high quality patch distributions (quality between 0.76 and 1.77) and low quality patch distributions (quality between 0.23 to 1.24) as a function of disease-induced mortality ($\nu$) and the probability of direct transmission ($\delta$).
+Reds indicate higher median host population sizes in high quality patch distributions, while blues indicate higher median host population sizes in low quality patch distributions.
+Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.}
+\label{highvlow}
 \end{figure}

### 2015-06-02 Manuscript/manuscript.tex
More discussion edits

@@
 
-\subsection*{Introduction and outline of 3 regions}
-
-This work demonstrates that for many pathogens, the distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of environmentally transmitted pathogens and the resulting patterns of host occupancy.
+This work demonstrates that for many pathogens, the distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of environmentally transmitted pathogens and the resulting size and stability of the host population.
 These impacts are driven largely by the interaction between pathogen traits (i.e.
@@
 In this light, this work expands on previous studies of disease spread in a metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).
-Shifting the patch quality distribution towards high quality habitat in our model is  conceptually similar to increasing movement in these models, and as in these models, the net consequences of this trade-off for the host metapopulation depend on the traits of the pathogen.
-Broadly, we see three different scenarios: pathogen traits produce very weak pathogen spread and high quality habitat is a net benefit; pathogen traits facilitate intermediate/endemic pathogen spread and low quality habitat is a net benefit; and pathogen traits produce easy/pandemic pathogen spread and high quality habitat is again a net benefit.
+Shifting the patch quality distribution towards high quality habitat in our model is  conceptually similar to increasing movement in these models.
+And, as in these models, the net consequences of this trade-off for the host metapopulation depend on the traits of the pathogen.
+Broadly, we can divide the observed dynamics into three different scenarios: pathogen traits produce very weak pathogen spread and high quality habitat is a net benefit; pathogen traits facilitate intermediate/endemic pathogen spread and low quality habitat is a net benefit; and pathogen traits produce easy/pandemic pathogen spread and high quality habitat is again a net benefit.
 
-\subsection*{Pathogens for which high quality beneficial}
+\subsection*{Pathogens for which high quality habitat increases host population}
 
-In the first of these scenarios, when pathogens have high disease-induced mortality and weak transmission (bottom left of panels in Fig~\ref{highvlow}), pathogen spread is so limited that habitat quality has little effect on epidemic dynamics.
+In the first of these scenarios, when pathogens have high disease-induced mortality and weak transmission (bottom left of panels in Figs~\ref{endemic}, \ref{highvlow}), pathogen spread is so limited that habitat quality has little effect on epidemic dynamics.
 As a result, high quality patches are a net benefit to the host metapopulation relative to low quality habitat, because their connectivity and stability facilitate larger total population sizes with little additional pathogen spread.
 
-Similarly, in the third scenario, when the pathogen spreads easily as a result of low disease-induced mortality and high transmission (upper portions of panels in Fig~\ref{highvlow}), high quality habitat is again a net benefit to the total host population size.
+Similarly, in the third scenario, when the pathogen spreads easily as a result of low disease-induced mortality and high transmission (upper portions of panels in Figs~\ref{endemic}, \ref{highvlow}), high quality habitat is again a net benefit to the total host population size.
 Under these conditions, habitat quality again has a limited effect on epidemic dynamics, as the pathogen spreads easily enough to overwhelm any metapopulation structure.
-Here then, the benefit of high quality habitat derives primarily from its ability to maintain metapopulation stability in the face of widespread disease-induced mortality.
+Here then, the benefit of high quality habitat relative to low quality derives primarily from its ability to maintain metapopulation stability in the face of widespread disease-induced mortality.
 
-\subsection*{Pathogens for which high quality detrimental}
+\subsection*{Pathogens for which high quality habitat decreases host population}
 
@@
 
-%High quality habitat distributions require lower pathogen longevities for pathogen spread to begin to take off, relative to low quality habitat distributions.
-Thus, for intermediate longevities, high quality distributions experience considerably greater pathogen spread and lower occupancy than low quality distributions.
+Because high quality habitat supports larger host population sizes, host populations on high quality habitat are less susceptible to environmental stochasticity and thus experience less frequent extinction events.
+Due to this additional stability, high quality habitat is better able to both maintain a consistent pool of susceptible hosts to which infection can spread, and support infectious host populations under disease-induced mortality.
+As a result, the probability of infection (i.e. proportion of time occupied by infectious hosts) increases with patch quality (Fig~\ref{pinfection}).
+From the perspective of a migrant susceptible host then, selection of high quality habitat leads to increased disease risk relative to other habitats, either because the patch is likely already occupied by infectious hosts, or because the local susceptible hosts are likely to soon become infected (either through the presence of an environmental reservoir or direct contact).
 
-\subsubsection*{Role of high quality habitat}
-
-Because high quality habitat supports larger host population sizes, populations on high quality habitat are less susceptible to environmental stochasticity and thus experience less frequent extinction events.
-Due to this additional stability, high quality habitat is more consistently occupied, by both susceptible and infectious hosts, leading to an increase in infection probability (i.e.
-proportion of time occupied by infectious hosts) with patch quality (Fig~\ref{pinfection}).
-From the perspective of a migrant susceptible host then, selection of high quality habitat leads to increased disease risk relative to other habitats, either because the patch is likely already occupied by infectious hosts, or because the local susceptible hosts are likely to soon become infected (either through the presences of an environmental reservoir or direct contact).
 Thus, if high quality habitat is preferentially selected by migrants (i.e.
-$\xi_{em} > 0$), high quality habitat becomes an ecological trap, that is, preferentially selected habitat that reduces fitness relative to other habitats \cite{Robertson2006}).
-Moreover, this preferential selection of high quality habitat further exacerbates the increase in disease risk with habitat quality (i.e., the fitness costs of high quality habitat, Fig~\ref{pinfection}) by increasing the risk of direct transmission from infectious migrants also selecting high quality habitat.
+$\xi_{em} > 0$), it becomes an ecological trap, that is, preferentially selected habitat that reduces fitness relative to other habitats (\cite{Robertson2006}).
+This fitness cost (i.e. the disease risk) becomes more severe as pathogen longevity increases, reflecting the more prominent role of environmental transmission on high quality habitat.
+Moreover, the preferential selection of high quality habitat further exacerbates the increase in disease risk with habitat quality (Fig~\ref{pinfection}) by increasing the risk of direct transmission from infectious migrants also selecting high quality habitat.
 This echoes results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).
@@
 Indeed, for pathogens with traits in this intermediate range, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).  
-This helps to maintain infectious occupancy throughout the metapopulation, which, when coupled with preferential selection of high quality habitat, feeds back on high quality patches to ensure a steady stream of infectious colonists that help maintain the environmental reservoir and the trap effect.
+This helps to maintain infectious occupancy throughout the metapopulation, which, when coupled with preferential selection of high quality habitat, feeds back on high quality patches to ensure a steady stream of infectious colonists.
+These infectious colonists then help to maintain the environmental reservoir and the trap effect.  
 Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread, even for more virulent pathogens (Fig~\ref{endemic}, \ref{popsizes}).
 The consequences of the concomitant widespread disease-induced mortality outweigh the positive consequences of increased metapopulation connectivity and stability, leading to a net decrease in total host population size on high quality habitat distributions (Fig~\ref{highvlow}, \ref{popsizes}).
-
-\subsubsection*{Role of low quality habitat}
 
@@
 Since they are infrequently colonized (when high quality habitat is preferentially selected, $\xi_{em} > 0$), the environmental reservoir left behind by these infectious populations likely decays before it has the opportunity to infect new susceptible colonists.
-As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and thereby reducing infectious occupancy (Fig~\ref{endemic}, \ref{popsizes}).
-
-\subsubsection*{Overall consequences}
+As a result, low quality patches effectively represent a dead-end the pathogen, reducing the number of patches from which it can spread, and thereby reducing total infectious occupancy and maintaining larger host population sizes (Fig~\ref{endemic}, \ref{popsizes}).
 
@@
 More importantly, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).  
+Indeed, we see that the combination of the trap effect (i.e.
+host preference for high quality habitat) along with an environmentally long-lived pathogen can have severe consequences for metapopulation stability, substantially increasing the risk of host extinction (Fig~\ref{pext}).
 From a management perspective, it is critical to understand this shift, as previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable for the host (\cite{Strasser2010}).
@@
 disease-induced mortality) can lead to cases where focusing on low quality habitat is more effective in increasing population growth rate.
-Indeed, in these situations, our model suggests that low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can maintain larger host populations than high quality distributions.
+Similarly, our model suggests that low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can maintain larger total host populations than high quality distributions.
 
 \subsection*{Conclusions}
-
-\subsubsection*{Discussion of Hess, Gog, Park and role of longevity}
 
@@
 In cases where the host movement rate was either low enough that the pathogen was unable to persist or high enough to cause a pandemic, Hess found that further increasing movement increased host occupancy, while at intermediate levels that resulted in endemic disease, increasing movement decreased host occupancy.
-These three scenarios, distinguished by the host movement rate, map to the the three scenarios we outline here based on pathogen traits (disease-induced mortality and direct and environmental transmission).
+These three scenarios, distinguished in \cite{Hess1996} by the host movement rate, map to the the three scenarios we outline here based on pathogen traits (disease-induced mortality and direct and environmental transmission).
 The conceptual similarity of these results suggests more generally that in metapopulations facing intermediate (i.e.
-endemic) pathogen spread, improving factors that increase host metapopulation stability and connectivity (e.g.
+endemic) pathogen spread, factors that increase host metapopulation stability and connectivity (e.g.
 increased movement, high quality habitat) can actually decrease total host population size as a result of increased pathogen spread.
 
-However, we find that as the environmental longevity of the pathogen increases, high quality habitat is a net benefit for a larger range of potential pathogens (Fig~\ref{highvlow}, red region grows from panels a.
+However, we find that as the environmental longevity of the pathogen increases, the region in which we observe these effects becomes smaller, that is, high quality habitat is a net benefit for a larger range of potential pathogens (Fig~\ref{highvlow}, red region grows from panels a.
 to c.).
 These dynamics reflect the findings of \cite{Gog2002}, who found that under strong background infection from an alternative host, increasing movement always increased total occupancy, despite facilitating pathogen spread.
-Park (\cite{Park2012}) expanded on this by adding environmental transmission to the mix and found that in cases where background and environmental transmission were strong relative to direct transmission, increasing movement was a net benefit to the host metapopulation.
+Park (\cite{Park2012}) expanded on this by adding environmental transmission to the mix and found that in cases where background and environmental transmission were strong relative to direct transmission, increasing movement was again a net benefit to the host metapopulation.
 These studies, coupled with the results presented here, suggest that when there is a persistent source of infection throughout the metapopulation (e.g.

### 2015-06-04 Manuscript/manuscript.tex
Small edits and exploratory plots

@@
 This environmental persistence creates environmental pathogen reservoirs from which susceptible hosts can become infected without direct contact with an infectious individual.
-This additional transmission pathway can have important consequences for disease dynamics, with models showing that increased environmental persistence generally facilitates increased pathogen persistence and spread relative to direct transmission alone (\cite{Almberg2011}, \cite{Sharp2011}, \cite{Breban2009}).
+This additional transmission pathway can have important consequences for disease dynamics, with models showing that increased environmental longevity generally facilitates increased pathogen persistence and spread relative to direct transmission alone (\cite{Almberg2011}, \cite{Sharp2011}, \cite{Breban2009}).
 
@@
 Many, if not all, of the pathogens listed above affect spatially structured host populations (e.g.
-plague in prairie dog colonies, \cite{George2013}), and thus understanding how environmental transmission interacts with patterns of habitat quality is critical to managing disease in these systems.
+plague in prairie dog colonies, \cite{George2013}), and thus understanding how pathogen transmission interacts with patterns of habitat quality is critical to managing disease in these systems.
 In this study then, we seek to explore how a pathogen's traits, including its environmental longevity (how long it can persist and remain infectious in the environment), interact with the distribution of habitat quality to influence pathogen spread and host occupancy, with a specific interest in the roles played by high and low quality habitat.
@@
 In addition to these assumptions, parameters were chosen so that without infection, approximately 0.75 of the metapopulation was occupied ($\mu = 0.1$, $D=5$).
-To explore the dynamics of a range of possible pathogens, disease induced mortality ($\nu$), probability of direct infection ($\delta$), and environmental longevity were varied over a range of possible values.
+To explore the dynamics of a range of possible pathogens, disease induced mortality ($\nu$), probability of direct infection ($\delta$), and environmental longevity (representing the half-life of the pathogen's infectivity in the environment) were varied over a range of possible values.
 
@@
 Specifically, we implemented a full-factorial design in which disease-induced mortality ($\nu$) and probability of direct infection ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).
-In addition, pathogen longevity, representing the half-life of the pathogen's infectivity in the environment, was varied over 3 values (ranging from 0.3 to 3 times the expected residence time of a healthy population on a unit quality patch, i.e., 10 units of time).
+In addition, pathogen longevity was varied over 3 values (ranging from 0.3 to 3 times the expected residence time of a healthy population on a unit quality patch, i.e., 10 units of time).
 
@@
 
-To more carefully explore the potential for high quality habitat to serve as infectious traps, we performed an additional simulation experiment in which we varied the effect of habitat quality on immigration rate by choosing different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating migrants that select all patches with equal probability).
+To more carefully explore the potential for high quality habitat to serve as infectious traps, we performed an additional simulation experiment in which we varied the effect of habitat quality on immigration rate by choosing different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating that migrants select all patches with equal probability).
 To isolate the effects of habitat preference and its interaction with pathogen longevity in these simulations, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), landscape structure (fully connected), and the patch quality distribution (with patches evenly spaced over the full range of qualities explored in the simulation above).
@@
 A wide range of disease dynamics were observed over the range of pathogens and habitat quality distributions explored.  
-Epidemic outcome was determined largely by disease induced mortality, with low disease-induced mortality (i.e.
+Epidemic outcome was influenced most strongly by disease induced mortality, with low disease-induced mortality (i.e.
 high $\nu$) facilitating pandemic dynamics (in which no susceptible host populations persist), and very high disease-induced mortality (i.e.
@@
 
-However, despite facilitating pathogen spread, for pathogens with either low or high disease-induced mortalities, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions (Fig~\ref{highvlow}).
-In the case of pathogens that were unable to persist due to high disease-induced mortality and weak transmission, high quality habitat distributions supported greater occupancy than low quality distributions due to the greater stability and connectivity afforded by high quality habitat.
+Despite facilitating pathogen spread, for pathogens with either low or high disease-induced mortalities, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions (Fig~\ref{highvlow}).
+In the case of pathogens that were unable to persist due to high disease-induced mortality and weak transmission, high quality habitat distributions supported larger host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat.
 Similarly, when low disease-induced mortality and high transmission (from either environmental or direct transmission) allowed a pathogen to spread widely regardless of habitat quality, high quality distributions again supported larger host populations than low quality distributions.
@@
 For pathogens falling between these extremes (i.e. with intermediate disease-induced mortality), the habitat quality distribution had a substantial effect on pathogen spread.  
-In particular, for a given disease, low quality habitat distributions slowed pathogen spread relative to high quality distributions and as a result supported larger host population sizes (Fig ~\ref{highvlow}, \ref{popsizes}).
-For a given pathogen in this range (here with $\nu = 0.2$ and $\delta = 0.3$), the relative advantage of low quality habitat increased with increasing environmental longevity (Fig~\ref{popsizes}), largely due to the more pronounced effect of increased longevity on pathogen persistence and spread in high quality habitat distributions.
+In particular, for these pathgoens, low quality habitat distributions slowed pathogen spread relative to high quality distributions and as a result supported larger host population sizes (Fig ~\ref{highvlow}, \ref{popsizes}).
+For a given pathogen in this range (here with $\nu = 0.2$ and $\delta = 0.3$), the relative advantage of low quality habitat increased with increasing environmental longevity (Fig~\ref{popsizes}), largely due to the more pronounced effect of increased longevity on pathogen persistence and spread in high quality habitat distributions (note that this is not a general trend and in some cases is even reversed).
 Again, metapopulations with a lattice structure displayed similar patterns, but required lower disease-induced mortality to see the benefit of low quality habitat (Supplement).
@@
 Because of their lower extinction rate and higher recolonization rate, high quality patches supported more consistent occupancy than low quality patches (Fig~\ref{simvis}).  
-However, disease risk, measured by the proportion of time a patch is occupied by infectious hosts, increased with patch quality, an effect that became more pronounced as pathogen longevity increased (Fig~\ref{pinfection}).
+However, disease risk, measured by the proportion of time a patch spent occupied by infectious hosts, increased with patch quality, an effect that became more pronounced as pathogen longevity increased (Fig~\ref{pinfection}).
 Indeed, as an epidemic progressed, high quality patches increasingly supported infectious occupants, while low quality patches supported infectious hosts only briefly and infrequently (Fig~\ref{simvis}).

### 2015-06-04 Analysis/pathgrid.analysis.r

@@ -240,3 +240,29 @@ p.tot <- p.tot + geom_tufteboxplot(aes(colour = factor(treatment)), outlier.colo
   scale_y_continuous(limits = c(0, 120), expand = c(0, 0.1))
 
 grid.arrange(p.S, p.I, p.tot, ncol = 3)
+
+#===============================================================================
+# Plots showing the effect of longevity across all delta and nu
+
+library(scales)
+
+diff <- dat[, 2, , ] - dat[, 1, , ]
+signdiff <- sign(diff)
+absdiff <- abs(diff)
+
+par(mfcol = c(10, 10), mar = c(0, 0, 0, 0))
+for(delta in 1:10){
+  for(nu in 10:1){
+    plot(log10(longevity), diff[, delta, nu], type = "l", axes = F, xlab = "", ylab = "")
+    rect(-10, 0, 10, 100, col = alpha("blue", alpha = 0.5))
+    rect(-10, 0, 10, -100, col = alpha("red", alpha = 0.5))
+  }
+}
+
+par(mfcol = c(10, 10), mar = c(0, 0, 0, 0))
+for(delta in 1:10){
+  for(nu in 10:1){
+    barplot(absdiff[, delta, nu], col = ifelse(signdiff[, delta, nu] > 0, "blue", "red"), 
+            axes = F, xlab = "", ylab = "", axisnames = F)
+  }
+}

### 2015-06-08 Manuscript/manuscript.tex
Exploratory plots and additional edits

@@
 Pathogens with higher environmental longevities were generally able to persist and spread more easily, even with high disease-induced morality (Fig~\ref{endemic}(c, f)).  
-The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics, with higher quality distributions (with qualities ranging roughly from 0.75 to 1.75) facilitating pandemic spread for a greater range of pathogens (Fig~\ref{endemic}, a-c vs.
+The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics, with higher quality distributions (with qualities ranging roughly from 0.75 to 1.75) facilitating pandemic spread for a larger range of pathogens (Fig~\ref{endemic}, a-c vs.
 d-f).
@@
 
-Despite facilitating pathogen spread, for pathogens with either low or high disease-induced mortalities, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions (Fig~\ref{highvlow}).
-In the case of pathogens that were unable to persist due to high disease-induced mortality and weak transmission, high quality habitat distributions supported larger host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat.
-Similarly, when low disease-induced mortality and high transmission (from either environmental or direct transmission) allowed a pathogen to spread widely regardless of habitat quality, high quality distributions again supported larger host populations than low quality distributions.
+Despite generally facilitating pathogen spread, for a range of pathogens, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions (red shaded regions in Fig~\ref{highvlow}).
+In the case of pathogens that were unable to persist due to high disease-induced mortality and weak transmission, high quality habitat distributions supported larger host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat (bottom left of panels in Fig~\ref{highvlow}).
+Similarly, when low disease-induced mortality and high transmission (from either environmental or direct transmission) allowed a pathogen to spread widely (i.e.
+at a pandemic level) in both habitat quality distributions, high quality distributions again supported larger host populations than low quality distributions (top portion of panels in Fig~\ref{highvlow}).
 
-For pathogens falling between these extremes (i.e. with intermediate disease-induced mortality), the habitat quality distribution had a substantial effect on pathogen spread.  
-In particular, for these pathgoens, low quality habitat distributions slowed pathogen spread relative to high quality distributions and as a result supported larger host population sizes (Fig ~\ref{highvlow}, \ref{popsizes}).
-For a given pathogen in this range (here with $\nu = 0.2$ and $\delta = 0.3$), the relative advantage of low quality habitat increased with increasing environmental longevity (Fig~\ref{popsizes}), largely due to the more pronounced effect of increased longevity on pathogen persistence and spread in high quality habitat distributions (note that this is not a general trend and in some cases is even reversed).
-Again, metapopulations with a lattice structure displayed similar patterns, but required lower disease-induced mortality to see the benefit of low quality habitat (Supplement).
+However, for pathogens falling between these extremes (e.g.
+with intermediate disease-induced mortality), low quality habitat was a net benefit to the host, supporting larger total population sizes than high quality habitat (blue regions of Fig ~\ref{highvlow}).
+The spread of these pathogens was strongly influenced by the habitat quality distribution, with low quality habitat limiting disease spread and better maintaining susceptible occupancy relative to high quality habitat (Fig~\ref{popsizes}, Supplemental).
+In these cases, the more widespread disease-induced mortality faced by hosts on high quality habitat distributions results in smaller total population sizes than on low quality habitat.
+As pathogen longevity increases, this range of pathogens for which low quality habitat is beneficial to the host population becomes smaller, with this benefit apparent only for the most virulent pathogens.
+Again, metapopulations with a lattice structure displayed similar patterns (Supplement).
 
@@
 
-This work demonstrates that for many pathogens, the distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of environmentally transmitted pathogens and the resulting size and stability of the host population.
+This work demonstrates that the distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of a variety of environmentally transmitted pathogens and the resulting size and stability of the host population.
 These impacts are driven largely by the interaction between pathogen traits (i.e.
-disease-induced mortality, transmission rates) and the colonization and extinction rates of high and low quality habitat.
+disease-induced mortality, environmental and direct transmission rates) and the colonization and extinction rates of high and low quality habitat.
 
-The general findings that low quality habitat can inhibit disease spread while high quality habitat can enhance it suggest the presence of a trade-off between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.
-In this light, this work expands on previous studies of disease spread in a metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).
+The findings that low quality habitat can inhibit disease spread while high quality habitat can enhance it suggest the presence of a trade-off between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.
+In this light, this work expands on previous studies of disease spread in a homogeneous metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).
 Shifting the patch quality distribution towards high quality habitat in our model is  conceptually similar to increasing movement in these models.
@@
 Under these conditions, habitat quality again has a limited effect on epidemic dynamics, as the pathogen spreads easily enough to overwhelm any metapopulation structure.
-Here then, the benefit of high quality habitat relative to low quality derives primarily from its ability to maintain metapopulation stability in the face of widespread disease-induced mortality.
+Here then, the benefit of high quality habitat relative to low quality derives primarily from its ability to better maintain metapopulation stability in the face of widespread disease-induced mortality.
 
@@
 at an endemic level), the distribution of habitat quality exerts a large influence on epidemic dynamics, with consequences for the overall host population size.
-In particular, the ability of these pathogens to persist and spread is influenced substantially by the colonization and extinction rates of high and low quality habitat, and the contrasting roles that these habitat types play.
+In particular, the ability of these pathogens to persist and spread is influenced substantially by the colonization and extinction rates of high and low quality habitat, and the contrasting roles that these habitat types play in the metapopulation.
 
@@
 
-Thus, if high quality habitat is preferentially selected by migrants (i.e.
+Thus, if high quality habitat is preferred by migrants (i.e.
 $\xi_{em} > 0$), it becomes an ecological trap, that is, preferentially selected habitat that reduces fitness relative to other habitats (\cite{Robertson2006}).
@@
 Indeed, for pathogens with traits in this intermediate range, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).  
-This helps to maintain infectious occupancy throughout the metapopulation, which, when coupled with preferential selection of high quality habitat, feeds back on high quality patches to ensure a steady stream of infectious colonists.
+This helps to maintain infectious occupancy throughout the metapopulation, which, when coupled with preference for high quality habitat, feeds back on high quality patches to ensure a steady stream of infectious colonists.
 These infectious colonists then help to maintain the environmental reservoir and the trap effect.  
@@
 In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{endemic}, \ref{popsizes}). 
-Even though individual low quality patches, due to their high extinction rates and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}), the presence of low quality patches facilitates susceptible persistence for a greater range of pathogens (Fig~\ref{endemic}).
+Even though individual low quality patches, due to their high extinction rates and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}), the presence of low quality habitat facilitates susceptible persistence for a greater range of pathogens (Fig~\ref{endemic}).
 This phenomenon represents the other side of the role played by high quality patches in that low quality patches are relatively unstable (i.e.
@@
 Since they are infrequently colonized (when high quality habitat is preferentially selected, $\xi_{em} > 0$), the environmental reservoir left behind by these infectious populations likely decays before it has the opportunity to infect new susceptible colonists.
-As a result, low quality patches effectively represent a dead-end the pathogen, reducing the number of patches from which it can spread, and thereby reducing total infectious occupancy and maintaining larger host population sizes (Fig~\ref{endemic}, \ref{popsizes}).
+As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and allowing susceptible hosts to more easily persist and avoid infection.
+By preventing the widespread disease-induced mortality observed in high quality habitat distributions, low quality distributions are then able to maintain larger host population sizes (Fig~\ref{endemic}, \ref{popsizes}).
 
-Thus, for pathogens with traits that facilitate this intermediate level of spread, we see a reversal of the relative importance of high and low quality habitat for overall metapopulation stability.
+Thus, for pathogens with traits that facilitate this intermediate level of spread, we see a reversal of the relative importance of high and low quality habitat for overall host metapopulation stability.
 Here, rather than serving as sources that increase host population size, we see that high quality patches become sinks.  
@@
 host preference for high quality habitat) along with an environmentally long-lived pathogen can have severe consequences for metapopulation stability, substantially increasing the risk of host extinction (Fig~\ref{pext}).
-From a management perspective, it is critical to understand this shift, as previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable for the host (\cite{Strasser2010}).
+From a management perspective, it is critical to understand this shift in the effect of habitat quality, as previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable for the host (\cite{Strasser2010}).
 Generally, this means that managers should focus on maintaining high quality habitat, but Strasser et al.
@@
 disease-induced mortality) can lead to cases where focusing on low quality habitat is more effective in increasing population growth rate.
-Similarly, our model suggests that low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can maintain larger total host populations than high quality distributions.
+Similarly, our model suggests that in some cases, low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can maintain larger total host populations than high quality distributions.
 
@@
 The conceptual similarity of these results suggests more generally that in metapopulations facing intermediate (i.e.
-endemic) pathogen spread, factors that increase host metapopulation stability and connectivity (e.g.
+endemic) pathogen spread, factors that would normally increase host stability and connectivity in the absence of disease (e.g.
 increased movement, high quality habitat) can actually decrease total host population size as a result of increased pathogen spread.
 
-However, we find that as the environmental longevity of the pathogen increases, the region in which we observe these effects becomes smaller, that is, high quality habitat is a net benefit for a larger range of potential pathogens (Fig~\ref{highvlow}, red region grows from panels a.
+However, we find that as the environmental longevity of the pathogen increases, the region in which we observe this reversal becomes smaller, that is, high quality habitat is a net benefit for a larger range of potential pathogens (Fig~\ref{highvlow}, red region grows from panels a.
 to c.).
+Environmentally long-lived pathogens are able to spread widely for a larger range of conditions, largely overwhelming the ability of low quality habitat to provide refuges for susceptible hosts. 
 These dynamics reflect the findings of \cite{Gog2002}, who found that under strong background infection from an alternative host, increasing movement always increased total occupancy, despite facilitating pathogen spread.
@@
 These studies, coupled with the results presented here, suggest that when there is a persistent source of infection throughout the metapopulation (e.g.
-when the pathogen is long-lived in the environment), factors that facilitate metapopulation stability (i.e.
-increased movement, high quality patches) are generally a greater benefit than factors that inhibit pathogen spread (i.e.
-decreased movement, low quality patches).
+from long-lived environmental reservoirs), factors that facilitate metapopulation stability (e.g.
+increased movement, high quality habitat) are generally a greater benefit than factors that inhibit pathogen spread (e.g.
+decreased movement, low quality habitat).
 

### 2015-06-08 Analysis/pathgrid.analysis.r

@@ -143,7 +143,7 @@ grid.arrange(arrangeGrob(low, med, high, low.lattice, med.lattice, high.lattice,
 
 
 #===============================================================================
-# Plots of probability of pathogen persistence
+# Plots of differences in probability of pathogen persistence
 
 load(paste(getwd(), "/Output/pathgrid.RData", sep = ""))
 
@@ -178,6 +178,40 @@ key <- draw.colorkey(list(col = cols,
                      draw = F)
 
 grid.arrange(arrangeGrob(low, med, high, low.lattice, med.lattice, high.lattice, ncol = 3), 
+             key, ncol = 2, widths = c(0.9, 0.1))
+
+
+#===============================================================================
+# Plots of probability of pathogen persistence in high and low quality
+
+load(paste(getwd(), "/Output/pathgrid.RData", sep = ""))
+
+dat <- tapply(out$maxI > 0, list(out$longevity, out$treatment, out$delta, out$nu), sum)
+
+cols <- colorRampPalette(brewer.pal(11, "Greys"))(100)
+
+low.h <- levelplot(dat[1, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
+                 at = seq(0, 100, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+med.h <- levelplot(dat[2, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
+                 at = seq(0, 100, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+high.h <- levelplot(dat[3, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
+                  at = seq(0, 100, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+
+
+low.l <- levelplot(dat[1, 2, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
+                   at = seq(0, 100, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+med.l <- levelplot(dat[2, 2, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
+                   at = seq(0, 100, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+high.l <- levelplot(dat[3, 2, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
[…]

### 2015-06-08 Analysis/pubplots.R

@@ -79,7 +79,7 @@ rm(list = ls())
 
 pdf("Manuscript/figure/figure_3.pdf", width = 10, height = 5)
 load(paste(getwd(), "/Output/pathgrid.RData", sep = ""))
-sub <- out[abs(out$delta - 0.3) < 0.001 & out$nu == 0.2, ]
+sub <- out[abs(out$delta - 0.3) < 0.001 & abs(out$nu - 0.2) < 0.001, ]
 
 p.S <- ggplot(sub, aes(x = factor(log10(longevity)), y = S.pop)) + theme_classic()
 p.S <- p.S + xlab("log(longevity)") + ylab("S population") + theme(legend.position = "none")

### 2015-06-08 Manuscript/manuscript.tex
More descriptive captions

@@
 \centering
-\caption{Differences between the median total effective population size of hosts on high quality patch distributions (quality between 0.76 and 1.77) and low quality patch distributions (quality between 0.23 to 1.24) as a function of disease-induced mortality ($\nu$) and the probability of direct transmission ($\delta$).
-Reds indicate higher median host population sizes in high quality patch distributions, while blues indicate higher median host population sizes in low quality patch distributions.
+\caption{Differences between the median total effective population size of hosts on low quality patch distributions (quality between 0.23 to 1.24) and high quality patch distributions (quality between 0.76 and 1.77) as a function of disease-induced mortality ($\nu$) and the probability of direct transmission ($\delta$).
+Reds indicate that high quality habitat distributions support higher median host population sizes than low quality habitat, primarily corresponding to pathogens with either very limited or very extensive spread.
+Conversely, blues indicate higher median host population sizes on low quality habitat distributions, corresponding to pathogens with intermediate levels of spread.
 Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.}
@@
 Disease-induced mortality ($\nu$) and direct transmission rate ($\delta$) are fixed at values of 0.2 and 0.3, respectively.
-Colored points show medians, while vertical lines indicate the outer-quartile ranges, with gray points indicating outliers (determined as points outside 1.5 times the inner-quartile range).}
+Colored points show medians, while vertical lines indicate the outer-quartile ranges, with gray points indicating outliers (determined as points outside 1.5 times the inner-quartile range).
+At low longevities, this pathogen is largely unable to persist and high quality habitat supports larger populations of susceptible hosts.
+However, low quality habitat is better able to maintain these susceptible hosts as the pathogen's longevity increases, leading to larger total population sizes.}
 \label{popsizes}
@@
 The bottom panel shows the state -- susceptible (blue), infectious (red), or unoccupied (white) -- of individual patches through time.
-Patches are stacked vertically with the lowest quality at the bottom and the highest quality at the top.}
+Patches are stacked vertically with the lowest quality at the bottom and the highest quality at the top.
+Note the overall more consistent occupancy on high quality habitat and the only very brief spurts of infectious occupancy on low quality habitat.}
 \label{simvis}
@@
 Panel columns show low (0.3, a, d), medium (1.0, b, e), and high (3.2, c, f) pathogen longevities, while rows show the effect of preferential selection of high quality habitat (a-c, $\xi_{im} = 0$,  no preference; d-f, $\xi_{im} = 0.5$, preference for high quality).
+The bottom row demonstrates the existence of an ecological trap on high quality habitat resulting from the fitness costs of increased disease risk.
 Black points show the median over 100 replicate simulations, while vertical lines show the outer-quartile range, and gray points indicate outliers.
@@
 \caption{Probability of host extinction as a function of pathogen environmental longevity and habitat preference, with black showing no habitat preference ($\xi_{im} = 0$) and red showing preference for high quality habitat ($\xi_{im} = 0.5$).
-Disease induced mortality and probability of direct infection fixed at $\nu = 0.2$ and $\delta = 0.3$.}
+Disease induced mortality and probability of direct infection fixed at $\nu = 0.2$ and $\delta = 0.3$.
+The combination of preference for high quality habitat and an environmentally long-lived pathogen serve to substantially increase extinction risk.}
 \label{pext}

### 2015-06-08 Analysis/pathgrid.analysis.r

@@ -103,6 +103,36 @@ key <- draw.colorkey(list(col = cols,
 
 grid.arrange(arrangeGrob(low.h, med.h, high.h, low.l, med.l, high.l, ncol = 3), 
              key, ncol = 2, widths = c(0.9, 0.1))
+
+
+#===============================================================================
+# Plots of differences in susceptible and infectious population size
+
+cols <- colorRampPalette(brewer.pal(11, "RdBu"))(180)
+
+low.S <- levelplot(dat.S[1, 2, , ] - dat.S[1, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
+                 at = seq(-90, 90, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+med.S <- levelplot(dat.S[2, 2, , ] - dat.S[2, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
+                 at = seq(-90, 90, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+high.S <- levelplot(dat.S[3, 2, , ] - dat.S[3, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
+                  at = seq(-90, 90, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+
+low.I <- levelplot(dat.I[1, 2, , ] - dat.I[1, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
+                         at = seq(-90, 90, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+med.I <- levelplot(dat.I[2, 2, , ] - dat.I[2, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
+                         at = seq(-90, 90, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+high.I <- levelplot(dat.I[3, 2, , ] - dat.I[3, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
+                          at = seq(-90, 90, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+
+key <- draw.colorkey(list(col = cols, 
+                          at = seq(-90, 90, by = 1),
+                          labels = list(at = seq(-75, 75, by = 25)),
+                          height = 0.7), 
+                     draw = F)
+
[…]

### 2015-06-08 Analysis/pubplots.R

@@ -78,6 +78,7 @@ dev.off()
 rm(list = ls())
 
 pdf("Manuscript/figure/figure_3.pdf", width = 10, height = 5)
+
 load(paste(getwd(), "/Output/pathgrid.RData", sep = ""))
 sub <- out[abs(out$delta - 0.3) < 0.001 & abs(out$nu - 0.2) < 0.001, ]
 

### 2015-06-09 Manuscript/manuscript.tex
Title and abstract

@@
 
-\title{Modelling the creation of ecological traps by environmentally persistent pathogens}
+\title{The role of habitat quality in host metapopulations exposed to environmentally persistent pathogens}
 
@@
 In such cases, the presence of environmental pathogen reservoirs on high quality habitat may lead to the creation of ecological traps, wherein host individuals preferentially colonize high quality habitat, but are then exposed to increased infection risk.
-We explore this possibility through the development of a stochastic patch occupancy model in which we can vary both the environmental longevity of the pathogen, and the distribution of habitat quality in the host metapopulation.
-This suggests that for sufficiently long-lived environmental reservoirs, high quality habitat can serve as an ecological trap, and can even be a net detriment to the host metapopulation relative to low quality habitat.
+We explore this possibility through the development of a stochastic patch occupancy model in which we can vary both the traits of the pathogen (virulence, transmission, environmental longevity) and the distribution of habitat quality in the host metapopulation.
+This model suggests that for certain pathogens, high quality habitat can serve as an ecological trap, and can even be a net detriment to the host metapopulation relative to low quality habitat.
 This inversion of the relative roles of high and low quality habitat highlights the importance of considering the interaction between spatial structure and pathogen transmission when managing wildlife populations exposed to an environmentally persistent pathogen.
@@
 To isolate the effects of habitat preference and its interaction with pathogen longevity in these simulations, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), landscape structure (fully connected), and the patch quality distribution (with patches evenly spaced over the full range of qualities explored in the simulation above).
-Here again we performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by infectious hosts (as a measure of disease risk) and the final total effective host population size.
+Here again we performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by infectious hosts (as a measure of disease risk and the corresponding fitness costs) and the final total effective host population size.
 

### 2015-06-09 Manuscript/manuscript.tex
Clarity edits

@@
 
-We assume that patch population size is proportional to patch quality ($A_i$), and that disease-induced mortality reduces the size of infectious populations by a factor of $\nu$.
+We assume that the host population size supported by a patch is proportional to its quality ($A_i$), and that disease-induced mortality reduces the size of infectious populations by a factor of $\nu$.
 Building on the framework developed by \cite{Hanski2003} then, the rates at which patch \emph{i}, with quality $A_i$, is colonized by susceptible and infectious individuals, respectively, is given by its connectivity:
@@
 
-To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with either high or low quality habitat distributions (with patch qualities ranging from 0.76 to 1.77, and from 0.23 to 1.24, respectively).
-Specifically, we implemented a full-factorial design in which disease-induced mortality ($\nu$) and probability of direct infection ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).
+To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with patch quality distributions shifted towards either high or low quality habitat.
+These two distributions were chosen to have equal variance and an overlapping core of patches around unit quality to roughly simulate alternative management actions that favor expansion to include either additional high or low quality habitat (with qualities ranging from 0.76 to 1.77, and from 0.23 to 1.24 respectively).
+We then implemented a full-factorial design on each quality distribution in which the pathogen's disease-induced mortality ($\nu$) and probability of direct infection ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).
 In addition, pathogen longevity was varied over 3 values (ranging from 0.3 to 3 times the expected residence time of a healthy population on a unit quality patch, i.e., 10 units of time).

### 2015-06-09 Manuscript/manuscript.tex
Adding figures for ease of transfer

@@
 \begin{figure}
+\includegraphics[scale=0.5]{figure/figure_1.pdf}
 \caption{The proportion of simulations resulting in endemic disease dynamics (persistence of both susceptible and infectious hosts) as a function of disease-induced mortality ($\nu$), and the probability of direct transmission ($\delta$).
@@
 \begin{figure}
+\includegraphics[scale=0.5]{figure/figure_2.pdf}
 \centering
@@
 \begin{figure}
+\includegraphics[scale=0.5]{figure/figure_3.pdf}
 \centering
@@
 At low longevities, this pathogen is largely unable to persist and high quality habitat supports larger populations of susceptible hosts.
-However, low quality habitat is better able to maintain these susceptible hosts as the pathogen's longevity increases, leading to larger total population sizes.}
+However, low qualiimplying that patch quality does not influence individual host fitness ty habitat is better able to maintain these susceptible hosts as the pathogen's longevity increases, leading to larger total population sizes.}
 \label{popsizes}
@@
 \begin{figure}
+\includegraphics[scale=0.4]{figure/figure_4.pdf}
 \centering
@@
 \begin{figure}
+\includegraphics[scale=0.4]{figure/figure_5.pdf}
 \centering
@@
 \begin{figure}
+\includegraphics[scale=1]{figure/figure_6.pdf}
 \centering

### 2015-06-09 Code/highlow.plot.r
Delete old pub plots functions

@@ -1,29 +1,0 @@
-highlow.plot <- function(conn){
-  
-  highlow<-read.csv(paste(getwd(), "/Output/highlow", conn, ".csv", sep = ""), header=TRUE)
-  
-  longevity <- sort(unique(highlow$longevity))
-    
-  S.occ <- tapply(highlow$S, list(highlow$longevity, highlow$treatment), median)
-  I.occ <- tapply(highlow$I, list(highlow$longevity, highlow$treatment), median)
-  occ <- tapply(highlow$I + highlow$S, list(highlow$longevity, highlow$treatment), median)
-  
-  pdf(paste(getwd(), "/Manuscript", "/highlow", conn, ".pdf", sep=""), height=5, width=10)
-  
-  par(mfrow = c(1, 3), bty = "l")
-  plot(log10(longevity), S.occ[, "+high"], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", 
-       ylab = "S occupancy", xlab = "log(longevity)", main = title("a", adj = 0))
-  lines(log10(longevity), S.occ[, "+low"], type = "b", lty = 1, pch = 19, col = "blue")
-  
-  plot(log10(longevity), I.occ[, "+high"], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red", 
-       ylab = "I occupancy", xlab = "log(longevity)", main = title("b", adj = 0))
-  lines(log10(longevity), I.occ[, "+low"], type = "b", lty = 1, pch = 19, col = "blue")
-  
-  occ <- tapply(highlow$I + highlow$S, list(highlow$longevity, highlow$treatment), mean)
-  
-  plot(log10(longevity), occ[, "+high"], type = "b", lty = 1, ylim = c(0, 1), pch = 19, col = "red",
-       ylab = "Total occupancy", xlab = "log(longevity)", main = title("c", adj = 0))
-  lines(log10(longevity), occ[, "+low"], type = "b", lty = 1, pch = 19, col = "blue")
-  
-  dev.off()
-}

### 2015-06-09 Code/metapop.plot.r

@@ -1,41 +1,0 @@
-metapop.plot <- function(conn){
-  
-  library(gridExtra)
-  library(lattice)
-  library(RColorBrewer)
-  
-  load(paste(getwd(), "/Output/out", conn, ".RData", sep = ""))
-  
-  labels <- signif(log10(longevity), 2)[c(1, 3, 5, 7, 9)]
-  
-  metapop.S <- tapply(out$Sfin, list(out$longevity, out$variance), median)
-  metapop.I <- tapply(out$Ifin, list(out$longevity, out$variance), median)
-  metapop.occ <- tapply(out$Sfin + out$Ifin, list(out$longevity, out$variance), median)
-  
-  cols <- colorRampPalette(brewer.pal(9, "Reds"))(100)
-  
-  S <- levelplot(metapop.S, col.regions = cols, xlab = "log(longevity)", ylab = "Variance",
-                 at = seq(0, 1, by = 0.01), colorkey = F, 
-                 scales = list(at = c(1, 3, 5, 7, 9), x = list(labels = labels)),
-                 main = list("a", just = "left", x = 0.05))
-  I <- levelplot(metapop.I, col.regions = cols, xlab = "log(longevity)", ylab = "Variance",
-                 at = seq(0, 1, by = 0.01), colorkey = F, 
-                 scales = list(at = c(1, 3, 5, 7, 9), x = list(labels = labels)),
-                 main = list("b", just = "left", x = 0.05))
-  occ <- levelplot(metapop.occ, col.regions = cols, xlab = "log(longevity)", ylab = "Variance",
-                   at = seq(0, 1, by = 0.01), colorkey = F, 
-                   scales = list(at = c(1, 3, 5, 7, 9), x = list(labels = labels)),
-                   main = list("c", just = "left", x = 0.05))
-  
-  key <- draw.colorkey(list(col = cols, 
-                            at = seq(0, 1, by = 0.01),
-                            labels = list(at = seq(0, 1, by = 0.25)),
-                            height = 0.7), 
-                       draw = F)
-  
-  pdf(paste(getwd(), "/Manuscript", "/metapop", conn, ".pdf", sep=""), height=5, width=14)
-  
-  grid.arrange(S, I, occ, key, ncol = 4, widths = c(0.3, 0.3, 0.3, 0.1))
-  
-  dev.off()
-}

### 2015-06-09 Code/patch.plot.r

@@ -1,28 +1,0 @@
-patch.plot <- function(conn){
-  
-  library(scales)
-
-  load(paste(getwd(), "/Output/out", conn, ".RData", sep = ""))
-  
-  #Filter patch data to include only endemic cases
-  patch<-out[which(out$Ifin>0 & out$Sfin > 0), ]
-  
-  #Filter patch data to include only high variance runs
-  patch<-patch[which(abs(patch$variance - 0.2) < 0.01),]
-  
-  reps <- unique(patch$repID)
-  
-  pdf(paste(getwd(), "/Manuscript", "/infevents", conn, ".pdf", sep=""), height=6, width=8)
-  
-  par(mfrow = c(1, 1))
-  
-  plot(patch$quality, patch$inf.events.early, type = "n", ylab = "Number of infection events", xlab = "Patch quality", bty = "l")
-  for(i in 1:length(reps)){
-    sim <- patch[patch$repID == reps[i], ]
-    try(lines(loess.smooth(sim$quality, sim$inf.events.tot), type = "l", col = alpha("gray", 0.3)))
-  }
-  
-  lines(loess.smooth(patch$quality, patch$inf.events.tot), lwd = 2, col = "black")
- 
-  dev.off()
-}

### 2015-06-12 Manuscript/manuscript.tex
text edits, comments are embedded as (PCC: comment here). Didn't get to the discussion yet.

@@
 
-\title{The role of habitat quality in host metapopulations exposed to environmentally persistent pathogens}
+\title{The role of habitat quality in host metapopulations exposed to environmentally persistent pathogens.
+Or Identifying when environmentally persistent pathogens transform good habitat into ecological traps}
 
@@
 \author[2]{Paul C. Cross}
-\affil[2]{Northern Rocky Mountain Science Center, US Geological Survey, Bozeman, Montana, USA}
+\affil[2]{U.S. Geological Survey, Northern Rocky Mountain Science Center, Bozeman, Montana, USA}
 \affil[*]{Corresponding author: clint.leach@colostate.edu}
@@
 \begin{abstract} 
-The distribution of habitat quality is expected to play an important role in the dynamics of wildlife metapopulations, with the expectation that high quality habitat serves to stabilize and maintain occupancy.
-However, we might expect that the costs and benefits of different quality habitat might be modulated, or possibly even reversed, by the presence of an environmentally persistent pathogen.
-In such cases, the presence of environmental pathogen reservoirs on high quality habitat may lead to the creation of ecological traps, wherein host individuals preferentially colonize high quality habitat, but are then exposed to increased infection risk.
-We explore this possibility through the development of a stochastic patch occupancy model in which we can vary both the traits of the pathogen (virulence, transmission, environmental longevity) and the distribution of habitat quality in the host metapopulation.
-This model suggests that for certain pathogens, high quality habitat can serve as an ecological trap, and can even be a net detriment to the host metapopulation relative to low quality habitat.
+The distribution of habitat quality plays an important role in the dynamics of wildlife metapopulations, with the expectation that high quality habitat serves to stabilize and maintain occupancy (REF).
+However, the costs and benefits of different quality habitat might be modulated, or even reversed, by the presence of an environmentally persistent pathogen.
+The presence of environmental pathogen reservoirs on high quality habitat may lead to the creation of ecological traps, wherein host individuals preferentially colonize high quality habitat, but are then exposed to increased infection risk.
+We explored this possibility through the development of a stochastic patch occupancy model, where we varied the pathogen's virulence, transmission rate, and environmental persistence as well as the distribution of habitat quality in the host metapopulation.
+This model suggests that for certain (delete or explain what the certain refers to) pathogens, high quality habitat can serve as an ecological trap, and can be detrimental to host persistence relative to low quality habitat.
 This inversion of the relative roles of high and low quality habitat highlights the importance of considering the interaction between spatial structure and pathogen transmission when managing wildlife populations exposed to an environmentally persistent pathogen.
+
+PCC: abstract reads well. You could consider putting in a few more details. e.g. can you characterize the conditions/parameter space that creates the traps rather than just saying they may exist.
+
 \end{abstract}
@@
 Many wildlife diseases are caused by pathogens that can persist, and remain infectious, for long periods of time in the environment.
-Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\cite{Dragon1995}), plague (\cite{Eisen2008}), and white nose syndrome (\cite{Lindner2011}), among others.
+Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\cite{Dragon1995}), plague (\cite{Eisen2008}), and white nose syndrome (\cite{Lindner2011}), among others (PCC: may need latin names, could include chytrid?
+I think it is also potentially noteworthy that these pathogens also seem to be particularly notorious in wildlife....therefore this is a big deal).
 This environmental persistence creates environmental pathogen reservoirs from which susceptible hosts can become infected without direct contact with an infectious individual.
@@
 Indeed, the quality of a habitat patch can affect its extinction and colonization rates, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).
-These processes in turn may influence how a pathogen spreads, where environmental pathogen reservoirs get established, and the resulting effects on host occupancy and population size.
+These processes may influence how a pathogen spreads, where environmental pathogen reservoirs get established, and the resulting effects on host occupancy and population size.
 
-Specifically, we expect that high quality habitat patches, which support greater host density and traffic than lower quality habitat, might be more likely to form pathogen reservoirs and support consistently infectious populations.
-As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Almberg2011}).
+We expect that high quality habitat patches, which support greater host density and traffic than lower quality habitat, might be more likely to form pathogen reservoirs and support consistently infectious populations.
+As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Almberg2011}) (PCC: probably need a different reference for your first reference to "ecological trap").
 In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (\cite{Paull2012}).
-Similarly, we expect that low quality patches, which see relatively less host traffic, will be less likely to develop pathogen reservoirs and thus may serve as refuges on which susceptible hosts can escape infection.
+Similarly, we expect that low quality patches will be less likely to develop pathogen reservoirs and may serve as refuges where susceptible hosts can escape infection.
 
@@
 plague in prairie dog colonies, \cite{George2013}), and thus understanding how pathogen transmission interacts with patterns of habitat quality is critical to managing disease in these systems.
-In this study then, we seek to explore how a pathogen's traits, including its environmental longevity (how long it can persist and remain infectious in the environment), interact with the distribution of habitat quality to influence pathogen spread and host occupancy, with a specific interest in the roles played by high and low quality habitat.
+In this study, we explore how a pathogen's traits interact with the distribution of habitat quality to influence pathogen spread and host occupancy.
+We specifically focus on the relative importance of high and low quality habitats to host persistence in the presence of pathogens with different levels of environmental longevity.
 
@@
 \end{align}
-where $\phi(X_j)$ is an indicator function that is $1$ if patch $j$ is in state $X$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality; $D$ is the inverse of mean dispersal distance; and $d_{ij}$ is the distance between patches $i$ and $j$.
+
+PCC: not sure I understand why nothing is a function of time.
+
+where $\phi(X_j)$ is an indicator function that is $1$ if patch $j$ is in state $X$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality; $D$ is the inverse of mean dispersal distance; and $d_{ij}$ is the distance between patches $i$ and $j$  (PCC: not sure I would say d_ij is a distance here when 2 paragraphs down you set d_ij = 1).
 Essentially, $C_{Xi}$ sums the colonization effort (or propagule pressure) from all patches in state $X$ to focal patch $i$.
@@
 
-To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with patch quality distributions shifted towards either high or low quality habitat.
-These two distributions were chosen to have equal variance and an overlapping core of patches around unit quality to roughly simulate alternative management actions that favor expansion to include either additional high or low quality habitat (with qualities ranging from 0.76 to 1.77, and from 0.23 to 1.24 respectively).
+To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with patch quality distributions shifted towards either high or low quality habitat (PCC: while maintaining the same initial occupancy?).
+These two distributions were chosen to have equal variance and an overlapping core of patches around unit quality to roughly simulate alternative management actions that favor expansion (PCC: reword management actions that favor expansion) to include either additional high or low quality habitat (with qualities ranging from 0.76 to 1.77, and from 0.23 to 1.24 respectively).
 We then implemented a full-factorial design on each quality distribution in which the pathogen's disease-induced mortality ($\nu$) and probability of direct infection ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).
@@
 The state of the metapopulation was then tracked for 5000 time steps.
-We recorded the proportion of patches in each state at the end of each simulation and also computed the mean effective population size of susceptible and infectious hosts over the last 500 time steps of the simulation (assuming an effective population size of $A_i$ for susceptible populations and $\nu A_i$ for infectious populations).
+We recorded the proportion of patches in each state at the end of each simulation and also computed the mean effective population size of susceptible and infectious hosts over the last 500 time steps of the simulation (assuming an effective population size of $A_i$ for patches that are occupied and susceptible and $\nu A_i$ for occupied infectious patches).
 
@@
 To isolate the effects of habitat preference and its interaction with pathogen longevity in these simulations, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), landscape structure (fully connected), and the patch quality distribution (with patches evenly spaced over the full range of qualities explored in the simulation above).
-Here again we performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by infectious hosts (as a measure of disease risk and the corresponding fitness costs) and the final total effective host population size.
+We performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by infectious hosts (as a measure of disease risk and the corresponding fitness costs) and the final total effective host population size.
 
@@
 low $\nu$) facilitating disease-free or endemic dynamics (Fig~\ref{endemic}).
-Pathogens with higher environmental longevities were generally able to persist and spread more easily, even with high disease-induced morality (Fig~\ref{endemic}(c, f)).  
-The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics, with higher quality distributions (with qualities ranging roughly from 0.75 to 1.75) facilitating pandemic spread for a larger range of pathogens (Fig~\ref{endemic}, a-c vs.
+Pathogens with higher environmental longevities were generally able to persist (PCC: careful on word choice between persist and invade) and spread more easily, even with high disease-induced morality (Fig~\ref{endemic}(c, f)).
+The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics.
+Increasing overall habitat quality facilitated pandemic spread for a larger range of pathogens (Fig~\ref{endemic}, a-c vs.
 d-f).
@@
 
-Despite generally facilitating pathogen spread, for a range of pathogens, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions (red shaded regions in Fig~\ref{highvlow}).
-In the case of pathogens that were unable to persist due to high disease-induced mortality and weak transmission, high quality habitat distributions supported larger host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat (bottom left of panels in Fig~\ref{highvlow}).
+For a wide range of parameter values, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions despite generally facilitating pathogen spread (red shaded regions in Fig~\ref{highvlow}).
+(PCC: while true, one could also say: for the cases where the host and pathogen both persist there was a wide range of parameter values where low quality habitats supported more hosts.
+Essentially, I'm intersecting Fig.
+1 and 2, excluding the regions of the parameter space where the host or pathogen goes extinct (which is most of the red region in Fig 2).
+This is a very different sort of take home message.
+This is just a change of spin compared to how you phrase the paragraph.)
+In the case of pathogens that were unable to persist due to high disease-induced mortality and weak transmission, high quality habitat distributions supported larger host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat (bottom left corner of the panels in Fig~\ref{highvlow}).
 Similarly, when low disease-induced mortality and high transmission (from either environmental or direct transmission) allowed a pathogen to spread widely (i.e.
@@
 with intermediate disease-induced mortality), low quality habitat was a net benefit to the host, supporting larger total population sizes than high quality habitat (blue regions of Fig ~\ref{highvlow}).
+(PCC: Yes, may want to lead with this.)
 The spread of these pathogens was strongly influenced by the habitat quality distribution, with low quality habitat limiting disease spread and better maintaining susceptible occupancy relative to high quality habitat (Fig~\ref{popsizes}, Supplemental).
@@
 Darker shading corresponds to more frequent endemic dynamics.
-The white region above the endemic region generally corresponds to pandemic dynamics (no susceptible host populations persist), while the region below the endemic region corresponds to disease-free dynamics.}
+The white region above the endemic region generally corresponds to pandemic dynamics (no susceptible host populations persist), while the region below the endemic region corresponds to disease-free dynamics.
+PCC notes: need words and symbols on the axes.
+need to label the colorbar.
+I'd also draw some arrows across the rows and columns to show how pathogen longevity increases from left to right and habitat quality varies across rows.
+I'd also remove all the whitespace btwn plots so that the they are not super small in the paper.
+You could also directly label on the plots above and below the shading host vs.
+pathogen extinction.
+I would not refer to host extinction as "pandemics" as these are not synonyms.}
 \label{endemic}
@@
 At low longevities, this pathogen is largely unable to persist and high quality habitat supports larger populations of susceptible hosts.
-However, low qualiimplying that patch quality does not influence individual host fitness ty habitat is better able to maintain these susceptible hosts as the pathogen's longevity increases, leading to larger total population sizes.}
+However, low qualiimplying that patch quality does not influence individual host fitness ty habitat is better able to maintain these susceptible hosts as the pathogen's longevity increases, leading to larger total population sizes.
+PCC: took me a long time to understand this non-traditional boxplot.
+Personnally, I'd prefer a wide line for the inner quartiles and thin line for the outers.
+Then your eye is drawn towards and dwells on the most likely region rather than towards the unlikely region (the outer quartiles.)}
 \label{popsizes}
@@
 The bottom row demonstrates the existence of an ecological trap on high quality habitat resulting from the fitness costs of increased disease risk.
-Black points show the median over 100 replicate simulations, while vertical lines show the outer-quartile range, and gray points indicate outliers.
-Disease induced mortality and probability of direct infection fixed at $\nu = 0.2$ and $\delta = 0.3$.}
+(PCC: delete sentence goes in the main text, but it also isn't clear to me how you derive this conclusion from the figure) Black points show the median over 100 replicate simulations, while vertical lines show the outer-quartile range, and gray points indicate outliers.
+Disease induced mortality and probability of direct infection fixed at $\nu = 0.2$ and $\delta = 0.3$.
+PCC: you can scale y to be the same across all plots, which would allow you to drop your y axes on col 2 & 3, same goes for dropping the x axis on row 1 and squishing it together.
+Again, I would lable on the plot itself how longevity runs acroll the columns and rows indicate preference.
+If your main intent of the figure is to compare zero vs.
+high preference, then you should have those 2 cases on the same plot (maybe as two lines, dropping the boxplots) rather than making the reader make this connection across plots.
+The information in the boxplot seems to distract, rather than contribute, to your message here.
+}
 \label{pinfection}
@@
 Disease induced mortality and probability of direct infection fixed at $\nu = 0.2$ and $\delta = 0.3$.
-The combination of preference for high quality habitat and an environmentally long-lived pathogen serve to substantially increase extinction risk.}
+The combination of preference for high quality habitat and an environmentally long-lived pathogen serve to substantially increase extinction risk.
+(PCC: seems like you could say this in text w/o using a figure since you have only 6 points here and only 2 of them are different....i.e.
+The probability of host extinction was ten times higher when there was a preference for high quality patches and longer environmental persistence (0.3 vs.
+0.03))}
 \label{pext}

### 2015-06-15 Manuscript/manuscript.tex
a few more edits to the discussion

@@
 
-This work demonstrates that the distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of a variety of environmentally transmitted pathogens and the resulting size and stability of the host population.
-These impacts are driven largely by the interaction between pathogen traits (i.e.
-disease-induced mortality, environmental and direct transmission rates) and the colonization and extinction rates of high and low quality habitat.
+This work demonstrates that (PCC: distribution may often be confused with spatial distribution rather than quality) habitat quality in a metapopulation can have substantial impacts on the dynamics of a variety of environmentally transmitted pathogens and the resulting size and stability of the host population.
+(PCC: deleted, lots of keywords but no information).
+In many cases, high quality habitat can enhance the disease spread and impact of environmentally persistent pathogens.
+In addition, the habitats with many resources may become ecological traps that attract hosts, but are a net drain on the metapopulation due to the impacts of disease in those patches.
+This suggests a trade-off between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability (PCC: rather than saying "processes", elaborate on what specific process you mean.
+).
 
-The findings that low quality habitat can inhibit disease spread while high quality habitat can enhance it suggest the presence of a trade-off between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability.
-In this light, this work expands on previous studies of disease spread in a homogeneous metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).
+This work expands on previous studies of disease spread in a homogeneous metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).
 Shifting the patch quality distribution towards high quality habitat in our model is conceptually similar to increasing movement in these models.
 And, as in these models, the net consequences of this trade-off for the host metapopulation depend on the traits of the pathogen.
-Broadly, we can divide the observed dynamics into three different scenarios: pathogen traits produce very weak pathogen spread and high quality habitat is a net benefit; pathogen traits facilitate intermediate/endemic pathogen spread and low quality habitat is a net benefit; and pathogen traits produce easy/pandemic pathogen spread and high quality habitat is again a net benefit.
+Broadly, we can divide the observed dynamics into three different scenarios: limited pathogen spread, intermediate, and high pathogen endemicity (PCC: shouldn't there be a fourth scenario of pathogen drives the host extinct?).
+High quality habitats are a net benefit when pathogen spread is limited or is high enough that high and low quality patches are affected by disease.
+There is an intermediate scenario, however, where low quality habitats are more likely to remain uninfected and as a result boost the metapopulation size while high quality patches that are heavily infected tend to reduce the metapopulation size.
+(PCC Note: most things you label as pathogen traits are actually emergent properties of the host-pathogen-environment...e.g.
+virulence)  (PCC: my text here makes some of the stuff below redundant)
 
@@
 $\xi_{em} > 0$), it becomes an ecological trap, that is, preferentially selected habitat that reduces fitness relative to other habitats (\cite{Robertson2006}).
-This fitness cost (i.e. the disease risk) becomes more severe as pathogen longevity increases, reflecting the more prominent role of environmental transmission on high quality habitat.
+This fitness cost (i.e.
+the disease risk) becomes more severe as pathogen longevity increases, reflecting the more prominent role of environmental transmission on high quality habitat.
+(PCC: You're burying the lead.
+This stuff should go to the beginning of the discussion.
+)
 Moreover, the preferential selection of high quality habitat further exacerbates the increase in disease risk with habitat quality (Fig~\ref{pinfection}) by increasing the risk of direct transmission from infectious migrants also selecting high quality habitat.
@@
 Similarly, our model suggests that in some cases, low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can maintain larger total host populations than high quality distributions.
+(PCC: you could also relate to the early metapopulation work, which suggested how even unoccupied habitats were important to metapopulation persistence.
+Here we are layering on environmental pathogens as another reason why lower quality habitat and/or unoccupied habitats are important for conservation efforts.
+This also relates to migration-disease work, whereby migration can help reduce disease by moving the host off of contaminated sites (although there are also other mechanisms)
 
@@
 to c.).
+(PCC: would it be more accurate to describe the effect of envi.
+longevity as unimodal whereby at low longevities the pathogen works as though it were directly transmitted...and so there is a strong correlation between disease occupancy and host occupancy of a patch.
+At intermediate longevity, low quality patches are important as they are less likely to be occupied allowing time for decontamination.
+At high levels of longevity, patches are not unoccupied for long enough for decontamination to occur.
+If this is a fair representation, then we should avoid saying things like the 1st sentence as it implied a monotonic relationship.
+)
 Environmentally long-lived pathogens are able to spread widely for a larger range of conditions, largely overwhelming the ability of low quality habitat to provide refuges for susceptible hosts. 

### 2015-06-15 Manuscript/manuscript.tex
PCC comments and edits

--- 2015-06-12 Manuscript/manuscript.tex
@@
 
-where $\phi(X_j)$ is an indicator function that is $1$ if patch $j$ is in state $X$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality; $D$ is the inverse of mean dispersal distance; and $d_{ij}$ is the distance between patches $i$ and $j$  (PCC: not sure I would say d_ij is a distance here when 2 paragraphs down you set d_ij = 1).
+where $\phi(X_j)$ is an indicator function that is $1$ if patch $j$ is in state $X$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality; $D$ is the inverse of mean dispersal distance; and $d_{ij}$ is the distance between patches $i$ and $j$  (PCC: not sure I would say $d_{ij}$ is a distance here when 2 paragraphs down you set $d_ij = 1$).
 Essentially, $C_{Xi}$ sums the colonization effort (or propagule pressure) from all patches in state $X$ to focal patch $i$.
@@
 \begin{figure}
-\includegraphics[scale=0.5]{figure/figure_1.pdf}
+%\includegraphics[scale=0.5]{figure/figure_1.pdf}
 \caption{The proportion of simulations resulting in endemic disease dynamics (persistence of both susceptible and infectious hosts) as a function of disease-induced mortality ($\nu$), and the probability of direct transmission ($\delta$).
@@
 \begin{figure}
-\includegraphics[scale=0.5]{figure/figure_2.pdf}
+%\includegraphics[scale=0.5]{figure/figure_2.pdf}
 \centering
@@
 \begin{figure}
-\includegraphics[scale=0.5]{figure/figure_3.pdf}
+%\includegraphics[scale=0.5]{figure/figure_3.pdf}
 \centering
@@
 \begin{figure}
-\includegraphics[scale=0.4]{figure/figure_4.pdf}
+%\includegraphics[scale=0.4]{figure/figure_4.pdf}
 \centering
@@
 \begin{figure}
-\includegraphics[scale=0.4]{figure/figure_5.pdf}
+%\includegraphics[scale=0.4]{figure/figure_5.pdf}
 \centering
@@
 Disease induced mortality and probability of direct infection fixed at $\nu = 0.2$ and $\delta = 0.3$.
-PCC: you can scale y to be the same across all plots, which would allow you to drop your y axes on col 2 & 3, same goes for dropping the x axis on row 1 and squishing it together.
+PCC: you can scale y to be the same across all plots, which would allow you to drop your y axes on col 2 \& 3, same goes for dropping the x axis on row 1 and squishing it together.
 Again, I would lable on the plot itself how longevity runs acroll the columns and rows indicate preference.
@@
 \begin{figure}
-\includegraphics[scale=1]{figure/figure_6.pdf}
+%\includegraphics[scale=1]{figure/figure_6.pdf}
 \centering

### 2015-06-18 Manuscript/manuscript.tex
Edits to fig 1 and new exploratory figs.

@@
 \includegraphics[scale=0.5]{figure/figure_1.pdf}
-\caption{The proportion of simulations resulting in endemic disease dynamics (persistence of both susceptible and infectious hosts) as a function of disease-induced mortality ($\nu$), and the probability of direct transmission ($\delta$).
-Panel columns show low (0.3, a, d), medium (1.0, b, e), and high (3.2, c, f) pathogen longevities, while rows show high quality (ranging from 0.76 to 1.77, a - c) and low quality (ranging from 0.23 to 1.24, d - f) patch quality distributions.
+\caption{The proportion of simulations resulting in endemic disease dynamics (persistence of both susceptible and infectious hosts) as a function of infectious survival ($\nu$), and the probability of direct transmission ($\delta$).
+Panel columns show low (0.3, a, d), medium (1.0, b, e), and high (3.2, c, f) pathogen longevities, while rows show low quality (ranging from 0.23 to 1.24, a - c) and high quality (ranging from 0.76 to 1.77, d - f) patch quality distributions.
 Darker shading corresponds to more frequent endemic dynamics.
-The white region above the endemic region generally corresponds to pandemic dynamics (no susceptible host populations persist), while the region below the endemic region corresponds to disease-free dynamics.
-PCC notes: need words and symbols on the axes.
-need to label the colorbar.
-I'd also draw some arrows across the rows and columns to show how pathogen longevity increases from left to right and habitat quality varies across rows.
-I'd also remove all the whitespace btwn plots so that the they are not super small in the paper.
-You could also directly label on the plots above and below the shading host vs.
-pathogen extinction.
-I would not refer to host extinction as "pandemics" as these are not synonyms.}
+The white region above the endemic region generally corresponds to pandemic dynamics (no susceptible host populations persist), while the region below the endemic region corresponds to disease-free dynamics.}
 \label{endemic}

### 2015-06-18 Analysis/patch.analysis.r

@@ -214,9 +214,39 @@ p6 <- p6 + geom_tufteboxplot(outlier.colour = "grey80", position = position_dodg
 
 grid.arrange(p1, p3, p5, p2, p4, p6, nrow = 2)
 
+#===============================================================================
+# Multipanel figure comparing extinctions across longevity and xi_im
+
+ex <- tapply(out$inf.ex + out$susc.ex, list(out$xi_em, out$xi_im, out$longevity, out$quality), mean)
+
+par(mfrow = c(2, 3))
+
+plot(quality, ex[2, 1, 4, ], ylim = c(0, 7), ylab = "Extinctions", xlab = "Quality", pch = 20)
+plot(quality, ex[2, 1, 7, ], ylim = c(0, 7), ylab = "Extinctions", xlab = "Quality", pch = 20)
+plot(quality, ex[2, 1, 10, ], ylim = c(0, 7), ylab = "Extinctions", xlab = "Quality", pch = 20)
+lines(supsmu(quality, ex[2, 1, 4, ]))
+
+plot(quality, ex[2, 2, 4, ], ylim = c(0, 7), ylab = "Extinctions", xlab = "Quality", pch = 20)
+plot(quality, ex[2, 2, 7, ], ylim = c(0, 7), ylab = "Extinctions", xlab = "Quality", pch = 20)
+plot(quality, ex[2, 2, 10, ], ylim = c(0, 7), ylab = "Extinctions", xlab = "Quality", pch = 20)
+lines(supsmu(quality, ex[2, 2, 4, ]))
 
 #===============================================================================
-# Consequences of trap effect (with xi_em = 0.5)
+# Multipanel figure comparing S across longevity and xi_im
+
+S <- tapply(out$I, list(out$xi_em, out$xi_im, out$longevity, out$quality), median)
+
+par(mfrow = c(1, 3))
+
+plot(quality, S[2, 1, 4, ], ylim = c(0, 1), ylab = "Probability susceptible", xlab = "Quality", pch = 4)
+points(quality, S[2, 2, 4, ], col = "brown", pch = 4)
+plot(quality, S[2, 1, 7, ], ylim = c(0, 1), ylab = "Probability susceptible", xlab = "Quality", pch = 4)
+points(quality, S[2, 2, 7, ], col = "brown", pch = 4)
+plot(quality, S[2, 1, 10, ], ylim = c(0, 1), ylab = "Probability susceptible", xlab = "Quality", pch = 4)
[…]

### 2015-06-18 Analysis/pathgrid.analysis.r

@@ -140,35 +140,31 @@ load(paste(getwd(), "/Output/pathgrid.RData", sep = ""))
 
 dat <- tapply(out$S == 0 & out$I == 0, list(out$longevity, out$treatment, out$delta, out$nu), sum)
 
-cols <- colorRampPalette(brewer.pal(11, "RdBu"))(200)
+cols <- colorRampPalette(brewer.pal(9, "Greys"))(100)
 
-low <- levelplot(dat[1, 2, , ] - dat[1, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
-                 at = seq(-100, 100, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
-med <- levelplot(dat[2, 2, , ] - dat[2, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
-                 at = seq(-100, 100, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
-high <- levelplot(dat[3, 2, , ] - dat[3, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
-                  at = seq(-100, 100, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+low.h <- levelplot(dat[1, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
+                 at = seq(0, 100, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+med.h <- levelplot(dat[2, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
+                 at = seq(0, 100, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
+high.h <- levelplot(dat[3, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
+                  at = seq(0, 100, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
 
 
-load(paste(getwd(), "/Output/pathgrid(lattice).RData", sep = ""))
-
-dat.lat <- tapply(out$S == 0 & out$I == 0, list(out$longevity, out$treatment, out$delta, out$nu), sum)
-
-low.lattice <- levelplot(dat.lat[1, 2, , ] - dat.lat[1, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
[…]

### 2015-06-18 Analysis/pubplots.R

@@ -5,6 +5,7 @@ library(lattice)
 library(RColorBrewer)
 library(ggplot2)
 library(ggthemes)
+library(plyr)
 
 #===============================================================================
 ### Figure 1
@@ -13,32 +14,32 @@ pdf("Manuscript/figure/figure_1.pdf", width = 10, height = 6)
 
 load(paste(getwd(), "/Output/pathgrid.RData", sep = ""))
 
-dat <- tapply(out$I > 0 & out$S > 0, list(out$longevity, out$treatment, out$delta, out$nu), sum) / 100
+out$endemic <- out$S > 0 & out$I > 0
+dat <- ddply(out, c("longevity", "treatment", "delta", "nu"), summarise, sum = sum(endemic) / 100)
 
 cols <- colorRampPalette(brewer.pal(9, "Greys"))(100)
+labels <- c("d", "e", "f", "a", "b", "c")
 
-low.l <- levelplot(dat[1, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
-                   at = seq(0, 1, by = 0.01), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)), main = "a.")
-med.l <- levelplot(dat[2, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
-                   at = seq(0, 1, by = 0.01), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)), main = "b.")
-high.l <- levelplot(dat[3, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
-                    at = seq(0, 1, by = 0.01), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)), main = "c.")
+plot <- levelplot(sum ~ delta * nu | longevity * treatment, data = dat,
+          col.regions = cols,
+          at = seq(0, 1, by = 0.01),
+          xlab = list(label = expression("Probability of direct transmission"~(delta)), cex = 1),
+          ylab = list(label = expression("Infectious survival"~(nu)), cex = 1),
+          strip = F,
+          scales = list(alternating = F, cex = 1),
+          ylab.right = list(label = "Probability of endemic", cex = 1),
+          par.settings = list(layout.widths = list(axis.key.padding = 0, ylab.right = 3)),
+          panel=function(...){
+            panel.levelplot(...)
[…]

### 2015-06-18 Code/highlow.runs.r

@@ -1,7 +1,7 @@
 
 modelrun <- function(longevity, delta, nu, parms, distance, initial, timesteps, treatment){
   # Function that runs SPOM model for given pathogen parameters and habitat quality
-  # distribution and collects matapopulation-level simulation results.
+  # distribution and collects metapopulation-level simulation results.
   #
   # Takes in pathogen parameters and type of patch quality distribution, 
   # runs the SPOM without infection until steady state, then introduces infection

### 2015-06-18 Analysis/pubplots.R
Updates to fig 2 to match fig 1

@@ -48,28 +48,39 @@ dev.off()
 
 rm(list = ls())
 
-pdf("Manuscript/figure/figure_2.pdf", width = 10, height = 5)
+pdf("Manuscript/figure/figure_2.pdf", width = 10, height = 4)
 
 load(paste(getwd(), "/Output/pathgrid.RData", sep = ""))
 
-dat <- tapply(out$S.pop + out$I.pop, list(out$longevity, out$treatment, out$delta, out$nu), median)
+out$pop<- out$S.pop + out$I.pop
+
+high <- ddply(subset(out, treatment == "+high"), c("treatment", "longevity", "delta", "nu"), 
+              summarise, high = median(pop))
+low <- ddply(subset(out, treatment == "+low"), c("treatment", "longevity", "delta", "nu"), 
+              summarise, low = median(pop))
+joint <- join(high, low, by = c("longevity", "delta", "nu"))
+joint$diff <- joint$low - joint$high
 
 cols <- colorRampPalette(brewer.pal(11, "RdBu"))(140)
 
-low <- levelplot(dat[1, 2, , ] - dat[1, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
-                 at = seq(-70, 70, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)), main = "a.")
-med <- levelplot(dat[2, 2, , ] - dat[2, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
-                 at = seq(-70, 70, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)), main = "b.")
-high <- levelplot(dat[3, 2, , ] - dat[3, 1, , ], col.regions = cols, xlab = expression(delta), ylab = expression(nu),
-                  at = seq(-70, 70, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)), main = "c.")
+labels = c("a", "b", "c")
 
-key <- draw.colorkey(list(col = cols, 
-                          at = seq(-70, 70, by = 1),
-                          labels = list(at = seq(-50, 50, by = 25)),
-                          height = 0.7), 
-                     draw = F)
+plot <- levelplot(diff ~ delta * nu | longevity, data = joint,
+                  col.regions = cols,
+                  at = seq(-70, 70, by = 1),
+                  xlab = list(label = expression("Probability of direct transmission"~(delta)), cex = 1),
[…]

### 2015-06-19 Manuscript/manuscript.tex
Updates to fig 2 to match fig 1
Updates to fig 3

@@
 \begin{figure}
-%\includegraphics[scale=0.5]{figure/figure_2.pdf}
+\includegraphics[scale=0.5]{figure/figure_2.pdf}
 \centering
@@
 \begin{figure}
-%\includegraphics[scale=0.5]{figure/figure_3.pdf}
+\includegraphics[scale=0.5]{figure/figure_3.pdf}
 \centering
@@
 Disease-induced mortality ($\nu$) and direct transmission rate ($\delta$) are fixed at values of 0.2 and 0.3, respectively.
-Colored points show medians, while vertical lines indicate the outer-quartile ranges, with gray points indicating outliers (determined as points outside 1.5 times the inner-quartile range).
+Colored points show medians, while vertical lines indicate the inner-quartile ranges, with horizontal lines indicating the minimum and maximum.
 At low longevities, this pathogen is largely unable to persist and high quality habitat supports larger populations of susceptible hosts.
-However, low qualiimplying that patch quality does not influence individual host fitness ty habitat is better able to maintain these susceptible hosts as the pathogen's longevity increases, leading to larger total population sizes.
-PCC: took me a long time to understand this non-traditional boxplot.
-Personnally, I'd prefer a wide line for the inner quartiles and thin line for the outers.
-Then your eye is drawn towards and dwells on the most likely region rather than towards the unlikely region (the outer quartiles.)}
+However, low quality habitat is better able to maintain these susceptible hosts as the pathogen's longevity increases, leading to larger total population sizes.}
 \label{popsizes}
@@
 \begin{figure}
-%\includegraphics[scale=0.4]{figure/figure_4.pdf}
+\includegraphics[scale=0.4]{figure/figure_4.pdf}
 \centering

### 2015-06-19 Analysis/pubplots.R

@@ -94,23 +94,60 @@ pdf("Manuscript/figure/figure_3.pdf", width = 10, height = 5)
 load(paste(getwd(), "/Output/pathgrid.RData", sep = ""))
 sub <- out[abs(out$delta - 0.3) < 0.001 & abs(out$nu - 0.2) < 0.001, ]
 
-p.S <- ggplot(sub, aes(x = factor(log10(longevity)), y = S.pop)) + theme_classic()
-p.S <- p.S + xlab("log(longevity)") + ylab("S population") + theme(legend.position = "none")
-p.S <- p.S + geom_tufteboxplot(aes(colour = factor(treatment)), outlier.colour = "grey80", position = position_dodge(width = 0.4)) +
-  scale_y_continuous(limits = c(0, 120), expand = c(0, 0.1)) + ggtitle("a.")
+S <- ddply(sub, c("longevity", "treatment"), summarise, 
+           med = median(S.pop), 
+           lowq = quantile(S.pop, 0.25),
+           highq = quantile(S.pop, 0.75),
+           min = min(S.pop),
+           max = max(S.pop))
+
+I <- ddply(sub, c("longevity", "treatment"), summarise, 
+           med = median(I.pop), 
+           lowq = quantile(I.pop, 0.25),
+           highq = quantile(I.pop, 0.75),
+           min = min(I.pop),
+           max = max(I.pop))
+
+tot <- ddply(sub, c("longevity", "treatment"), summarise, 
+           med = median(S.pop + I.pop), 
+           lowq = quantile(S.pop + I.pop, 0.25),
+           highq = quantile(S.pop + I.pop, 0.75),
+           min = min(S.pop + I.pop),
+           max = max(S.pop + I.pop))
+
+pS <- ggplot(S, aes(x = factor(log10(longevity)), y = med))
+pS <- pS + geom_point(aes(colour = factor(treatment)), position = position_dodge(width = 0.4)) +
+             geom_point(aes(colour = factor(treatment), y = min), shape = 45, size = 5, position = position_dodge(width = 0.4)) +
+             geom_point(aes(colour = factor(treatment), y = max), shape = 45, size = 5, position = position_dodge(width = 0.4)) +
+             geom_errorbar(aes(colour = factor(treatment), ymin = lowq, ymax = highq), 
+                           position = position_dodge(width = 0.4), width = 0)
+pS <- pS + scale_y_continuous(limits = c(0, 125), expand = c(0, 0.1)) +
[…]

### 2015-06-19 Manuscript/manuscript.tex
Small intro and methods edits

@@
 
-Many wildlife diseases are caused by pathogens that can persist, and remain infectious, for long periods of time in the environment.
-Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\cite{Dragon1995}), plague (\cite{Eisen2008}), and white nose syndrome (\cite{Lindner2011}), among others (PCC: may need latin names, could include chytrid?
-I think it is also potentially noteworthy that these pathogens also seem to be particularly notorious in wildlife....therefore this is a big deal).
+Many prominent and problematic wildlife diseases are caused by pathogens that can persist, and remain infectious, for long periods of time in the environment.
+Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\emph{Bacillus anthracis}, \cite{Dragon1995}), plague (\emph{Yersinia pestis}, \cite{Eisen2008}), and white nose syndrome (\emph{Geomyces destructans}, \cite{Lindner2011}), among others.
 This environmental persistence creates environmental pathogen reservoirs from which susceptible hosts can become infected without direct contact with an infectious individual.
@@
 We expect that high quality habitat patches, which support greater host density and traffic than lower quality habitat, might be more likely to form pathogen reservoirs and support consistently infectious populations.
-As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Almberg2011}) (PCC: probably need a different reference for your first reference to "ecological trap").
+As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Robertson2006}, \cite{Almberg2011}).
 In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (\cite{Paull2012}).
@@
 \begin{align}
-C_{Si} &= A_i^{\xi_{im}} \sum_{i\neq j }\phi(S_j)A_j^{\xi_{em}}\exp(-D d_{ij}),\\
-C_{Ii} &= A_i^{\xi_{im}} \sum_{i\neq j }\phi(I_j)(\nu A_j) ^{\xi_{em}}\exp(-D d_{ij}),
+C_{Si}(t) &= A_i^{\xi_{im}} \sum_{i\neq j }\phi(S, t)_jA_j^{\xi_{em}}\exp(-D d_{ij}),\\
+C_{Ii}(t) &= A_i^{\xi_{im}} \sum_{i\neq j }\phi(I, t)_j(\nu A_j) ^{\xi_{em}}\exp(-D d_{ij}),
 \label{connectivity}
@@
 
-PCC: not sure I understand why nothing is a function of time.
-
-where $\phi(X_j)$ is an indicator function that is $1$ if patch $j$ is in state $X$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality; $D$ is the inverse of mean dispersal distance; and $d_{ij}$ is the distance between patches $i$ and $j$  (PCC: not sure I would say $d_{ij}$ is a distance here when 2 paragraphs down you set $d_ij = 1$).
+where $\phi(X, t)_j$ is an indicator function that is $1$ if patch $j$ is in state $X$ at time $t$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality; $D$ is the inverse of mean dispersal distance; and $d_{ij}$ is the effective distance between patches $i$ and $j$ .
 Essentially, $C_{Xi}$ sums the colonization effort (or propagule pressure) from all patches in state $X$ to focal patch $i$.
@@
 The parameterization above assumes that all patches are equally accessible from all other patches (i.e.
-are separated by distance 1).
+are separated by effective distance 1).
 To examine the effects of a more rigid spatial structure, we also implemented a model with patches arranged in a square lattice such that each patch is only accessible from its four neighboring patches (in which case, average migration distance was adjusted to $D = 2$ to maintain roughly equivalent occupancy).
@@
 
-To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with patch quality distributions shifted towards either high or low quality habitat (PCC: while maintaining the same initial occupancy?).
-These two distributions were chosen to have equal variance and an overlapping core of patches around unit quality to roughly simulate alternative management actions that favor expansion (PCC: reword management actions that favor expansion) to include either additional high or low quality habitat (with qualities ranging from 0.76 to 1.77, and from 0.23 to 1.24 respectively).
+To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with patch quality distributions shifted towards either high or low quality habitat.
+These two distributions were chosen to have equal variance and an overlapping core of patches around unit quality, with either the minimum or maximum shifted to simulate management actions that add either low or high quality habitat respectively (yielding distributions with qualities ranging from 0.23 to 1.24 and from 0.76 to 1.77).
 We then implemented a full-factorial design on each quality distribution in which the pathogen's disease-induced mortality ($\nu$) and probability of direct infection ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).
@@
 low $\nu$) facilitating disease-free or endemic dynamics (Fig~\ref{endemic}).
-Pathogens with higher environmental longevities were generally able to persist (PCC: careful on word choice between persist and invade) and spread more easily, even with high disease-induced morality (Fig~\ref{endemic}(c, f)).
+Pathogens with higher environmental longevities were generally able to invade and spread more easily, even with high disease-induced morality (Fig~\ref{endemic}(c, f)).  
 The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics.
@@
 This is just a change of spin compared to how you phrase the paragraph.)
-In the case of pathogens that were unable to persist due to high disease-induced mortality and weak transmission, high quality habitat distributions supported larger host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat (bottom left corner of the panels in Fig~\ref{highvlow}).
+In the case of pathogens that were unable to invade due to high disease-induced mortality and weak transmission, high quality habitat distributions supported larger host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat (bottom left corner of the panels in Fig~\ref{highvlow}).
 Similarly, when low disease-induced mortality and high transmission (from either environmental or direct transmission) allowed a pathogen to spread widely (i.e.
@@
 at an endemic level), the distribution of habitat quality exerts a large influence on epidemic dynamics, with consequences for the overall host population size.
-In particular, the ability of these pathogens to persist and spread is influenced substantially by the colonization and extinction rates of high and low quality habitat, and the contrasting roles that these habitat types play in the metapopulation.
+In particular, the ability of these pathogens to invade and spread is influenced substantially by the colonization and extinction rates of high and low quality habitat, and the contrasting roles that these habitat types play in the metapopulation.
 
@@
 In particular, for a given pathogen, Hess (\cite{Hess1996}) similarly identified three regions of disease dynamics which determined whether increased movement had a positive or negative effect on host occupancy.
-In cases where the host movement rate was either low enough that the pathogen was unable to persist or high enough to cause a pandemic, Hess found that further increasing movement increased host occupancy, while at intermediate levels that resulted in endemic disease, increasing movement decreased host occupancy.
+In cases where the host movement rate was either low enough that the pathogen was unable to invade or high enough to cause a pandemic, Hess found that further increasing movement increased host occupancy, while at intermediate levels that resulted in endemic disease, increasing movement decreased host occupancy.
 These three scenarios, distinguished in \cite{Hess1996} by the host movement rate, map to the the three scenarios we outline here based on pathogen traits (disease-induced mortality and direct and environmental transmission).
@@
 Colored points show medians, while vertical lines indicate the inner-quartile ranges, with horizontal lines indicating the minimum and maximum.
-At low longevities, this pathogen is largely unable to persist and high quality habitat supports larger populations of susceptible hosts.
+At low longevities, this pathogen is largely unable to invade and high quality habitat supports larger populations of susceptible hosts.
 However, low quality habitat is better able to maintain these susceptible hosts as the pathogen's longevity increases, leading to larger total population sizes.}

### 2015-06-30 Analysis/patch.analysis.r
Patch-level exploratory figures

@@ -232,11 +232,11 @@ plot(quality, ex[2, 2, 10, ], ylim = c(0, 7), ylab = "Extinctions", xlab = "Qual
 lines(supsmu(quality, ex[2, 2, 4, ]))
 
 #===============================================================================
-# Multipanel figure comparing S across longevity and xi_im
+# Multipanel figure comparing S and I across longevity and xi_im
 
-S <- tapply(out$I, list(out$xi_em, out$xi_im, out$longevity, out$quality), median)
+S <- tapply(out$S, list(out$xi_em, out$xi_im, out$longevity, out$quality), median)
 
-par(mfrow = c(1, 3))
+par(mfrow = c(2, 3))
 
 plot(quality, S[2, 1, 4, ], ylim = c(0, 1), ylab = "Probability susceptible", xlab = "Quality", pch = 4)
 points(quality, S[2, 2, 4, ], col = "brown", pch = 4)
@@ -245,6 +245,29 @@ points(quality, S[2, 2, 7, ], col = "brown", pch = 4)
 plot(quality, S[2, 1, 10, ], ylim = c(0, 1), ylab = "Probability susceptible", xlab = "Quality", pch = 4)
 points(quality, S[2, 2, 10, ], col = "brown", pch = 4)
 
+I <- tapply(out$I, list(out$xi_em, out$xi_im, out$longevity, out$quality), median)
+
+plot(quality, I[2, 1, 4, ], ylim = c(0, 1), ylab = "Probability infectious", xlab = "Quality", pch = 4)
+points(quality, I[2, 2, 4, ], col = "brown", pch = 4)
+plot(quality, I[2, 1, 7, ], ylim = c(0, 1), ylab = "Probability infectious", xlab = "Quality", pch = 4)
+points(quality, I[2, 2, 7, ], col = "brown", pch = 4)
+plot(quality, I[2, 1, 10, ], ylim = c(0, 1), ylab = "Probability infectious", xlab = "Quality", pch = 4)
+points(quality, I[2, 2, 10, ], col = "brown", pch = 4)
+
+par(mfrow = c(1, 3))
+
+occ <- tapply(out$I * 0.2 * out$quality + out$S * out$quality, list(out$xi_em, out$xi_im, out$longevity, out$quality), median)
+
+plot(quality, occ[2, 1, 4, ], ylim = c(0, 2), ylab = "Probability occupied", xlab = "Quality", pch = 4)
[…]

### 2015-06-30 Manuscript/manuscript.tex
Initial edits distinguishing metapop vs. ind level traps

@@
 Similarly, we expect that low quality patches will be less likely to develop pathogen reservoirs and may serve as refuges where susceptible hosts can escape infection.
+The combination of these effects may then lead to the creation of metapopulation-scale traps, where the inclusion of otherwise high quality habitat in the metapopulation actually reduces host population size relative to low quality habitat.
 
@@
 
-where $\phi(X, t)_j$ is an indicator function that is $1$ if patch $j$ is in state $X$ at time $t$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality; $D$ is the inverse of mean dispersal distance; and $d_{ij}$ is the effective distance between patches $i$ and $j$ .
+where $\phi(X, t)_j$ is an indicator function that is $1$ if patch $j$ is in state $X$ at time $t$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality; $D$ is the inverse of the host's mean dispersal distance; and $d_{ij}$ is the effective distance between patches $i$ and $j$ .
 Essentially, $C_{Xi}$ sums the colonization effort (or propagule pressure) from all patches in state $X$ to focal patch $i$.
@@
 (1) The rate of emigration from a patch increases with population size, while the extinction rate of a patch decreases with populations size ($\xi_{im}=0.5$, $\alpha=1$, \cite{Hanski2003}).  
-(2) Migrants preferentially select high quality habitat, that is, high quality patches attract more colonists than low quality patches ($\xi_{em}=0.5$, \cite{Hanski2003}, though we relax this below).
+(2) Migrants preferentially select high quality habitat, that is, the rate at which a patch is colonized scales with quality ($\xi_{em}=0.5$, \cite{Hanski2003}, though we relax this below).  
 (3)  Connectivity among patches is determined entirely by quality ($d_{ij}=1$ for all $i \neq j$).  
@@
 These two distributions were chosen to have equal variance and an overlapping core of patches around unit quality, with either the minimum or maximum shifted to simulate management actions that add either low or high quality habitat respectively (yielding distributions with qualities ranging from 0.23 to 1.24 and from 0.76 to 1.77).
-We then implemented a full-factorial design on each quality distribution in which the pathogen's disease-induced mortality ($\nu$) and probability of direct infection ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).
+We then implemented a full-factorial set of simulations on each quality distribution in which the pathogen's disease-induced mortality ($\nu$) and probability of direct infection ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).
 In addition, pathogen longevity was varied over 3 values (ranging from 0.3 to 3 times the expected residence time of a healthy population on a unit quality patch, i.e., 10 units of time).
@@
 
-To more carefully explore the potential for high quality habitat to serve as infectious traps, we performed an additional simulation experiment in which we varied the effect of habitat quality on immigration rate by choosing different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating that migrants select all patches with equal probability).
-To isolate the effects of habitat preference and its interaction with pathogen longevity in these simulations, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), landscape structure (fully connected), and the patch quality distribution (with patches evenly spaced over the full range of qualities explored in the simulation above).
-We performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by infectious hosts (as a measure of disease risk and the corresponding fitness costs) and the final total effective host population size.
+To more carefully explore the effects of habitat quality on patch-level dynamics, we performed an additional simulation experiment on metapopulations with patch qualities evenly spaced over the full range explored above (from 0.2 to 1.8).
+In these simulations, we varied longevity over the values used above and explored the consequences of host preference for high quality habitat (and the corresponding potential for ecological traps) by choosing two different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating that migrants select all patches with equal probability).
+To isolate the effects of habitat preference and its interaction with pathogen longevity, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), and landscape structure (fully connected).
+We performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by susceptible and infectious hosts, and the final total host population size.
 
@@
 For a wide range of parameter values, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions despite generally facilitating pathogen spread (red shaded regions in Fig~\ref{highvlow}).
-(PCC: while true, one could also say: for the cases where the host and pathogen both persist there was a wide range of parameter values where low quality habitats supported more hosts.
-Essentially, I'm intersecting Fig.
-1 and 2, excluding the regions of the parameter space where the host or pathogen goes extinct (which is most of the red region in Fig 2).
-This is a very different sort of take home message.
-This is just a change of spin compared to how you phrase the paragraph.)
 In the case of pathogens that were unable to invade due to high disease-induced mortality and weak transmission, high quality habitat distributions supported larger host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat (bottom left corner of the panels in Fig~\ref{highvlow}).
@@
 with intermediate disease-induced mortality), low quality habitat was a net benefit to the host, supporting larger total population sizes than high quality habitat (blue regions of Fig ~\ref{highvlow}).
-(PCC: Yes, may want to lead with this.)
 The spread of these pathogens was strongly influenced by the habitat quality distribution, with low quality habitat limiting disease spread and better maintaining susceptible occupancy relative to high quality habitat (Fig~\ref{popsizes}, Supplemental).
-In these cases, the more widespread disease-induced mortality faced by hosts on high quality habitat distributions results in smaller total population sizes than on low quality habitat.
-As pathogen longevity increases, this range of pathogens for which low quality habitat is beneficial to the host population becomes smaller, with this benefit apparent only for the most virulent pathogens.
+In these cases, high quality habitat serves as a sort of metapopulation-level trap, that is, otherwise beneficial habitat that is detrimental to the host population in the presence of disease.
+As pathogen longevity increases, the range of pathogens for which we observe this metapopulation-level trap becomes smaller, apparent only for the most virulent pathogens.
 Again, metapopulations with a lattice structure displayed similar patterns (Supplement).
 
-To better understand the mechanisms behind these effects, we further examined within patch infection dynamics for a pathogen in this intermediate range (again with $\nu = 0.2$ and $\delta = 0.3$).  
+To better understand the patch-level mechanisms behind these effects, we further examined within patch infection dynamics for a pathogen in this intermediate range (again with $\nu = 0.2$ and $\delta = 0.3$).
 Because of their lower extinction rate and higher recolonization rate, high quality patches supported more consistent occupancy than low quality patches (Fig~\ref{simvis}).  

### 2015-06-30 Manuscript/manuscript.tex
Fig 5 to supplement, remove fig 6

@@
 To better understand the patch-level mechanisms behind these effects, we further examined within patch infection dynamics for a pathogen in this intermediate range (again with $\nu = 0.2$ and $\delta = 0.3$).
-Because of their lower extinction rate and higher recolonization rate, high quality patches supported more consistent occupancy than low quality patches (Fig~\ref{simvis}).  
-However, disease risk, measured by the proportion of time a patch spent occupied by infectious hosts, increased with patch quality, an effect that became more pronounced as pathogen longevity increased (Fig~\ref{pinfection}).
+Because of their lower extinction rate and higher recolonization rate, high quality patches supported more consistent occupancy than low quality patches (Fig~\ref{simvis}, Supplement).  
+However, because of this stability, disease risk, measured by the proportion of time a patch spent occupied by infectious hosts, increased with patch quality (Supplement).
 Indeed, as an epidemic progressed, high quality patches increasingly supported infectious occupants, while low quality patches supported infectious hosts only briefly and infrequently (Fig~\ref{simvis}).
 Thus, from the perspective of a migrant individual, colonizing high quality habitat posed increased disease risk, with the preferential selection of high quality habitat (i.e.
-$\xi_{im} = 0.5$) increasing this disease risk further (Fig~\ref{pinfection}, a-c vs.
-d-f).
-Moreover, preferential selection of high quality habitat had dramatic consequences for the stability of the host metapopulation when the pathogen is long-lived in the environment, increasing extinction risk substantially compared to random habitat selection (i.e.
-$\xi_{em} = 0$, Fig~\ref{pext}).
+$\xi_{im} = 0.5$) increasing this disease risk further (Supplement).
+Moreover, assuming preferential selection of high quality habitat had dramatic consequences for the stability of the host metapopulation when the pathogen is long-lived in the environment, increasing the probability of extinction substantially compared to random habitat selection (0.29 vs.
+0.03).
 
@@
 
-\begin{figure}
-%\includegraphics[scale=0.4]{figure/figure_5.pdf}
-\centering
-\caption{Boxplots showing the probability of infection (i.e.
-proportion of time occupied by infectious hosts) as a function of patch quality.
-Panel columns show low (0.3, a, d), medium (1.0, b, e), and high (3.2, c, f) pathogen longevities, while rows show the effect of preferential selection of high quality habitat (a-c, $\xi_{im} = 0$,  no preference; d-f, $\xi_{im} = 0.5$, preference for high quality).
-The bottom row demonstrates the existence of an ecological trap on high quality habitat resulting from the fitness costs of increased disease risk.
-(PCC: delete sentence goes in the main text, but it also isn't clear to me how you derive this conclusion from the figure) Black points show the median over 100 replicate simulations, while vertical lines show the outer-quartile range, and gray points indicate outliers.
-Disease induced mortality and probability of direct infection fixed at $\nu = 0.2$ and $\delta = 0.3$.
-PCC: you can scale y to be the same across all plots, which would allow you to drop your y axes on col 2 \& 3, same goes for dropping the x axis on row 1 and squishing it together.
-Again, I would lable on the plot itself how longevity runs acroll the columns and rows indicate preference.
-If your main intent of the figure is to compare zero vs.
-high preference, then you should have those 2 cases on the same plot (maybe as two lines, dropping the boxplots) rather than making the reader make this connection across plots.
-The information in the boxplot seems to distract, rather than contribute, to your message here.
-}
-\label{pinfection}
-\end{figure}
-
-\begin{figure}
-%\includegraphics[scale=1]{figure/figure_6.pdf}
-\centering
-\caption{Probability of host extinction as a function of pathogen environmental longevity and habitat preference, with black showing no habitat preference ($\xi_{im} = 0$) and red showing preference for high quality habitat ($\xi_{im} = 0.5$).
-Disease induced mortality and probability of direct infection fixed at $\nu = 0.2$ and $\delta = 0.3$.
-The combination of preference for high quality habitat and an environmentally long-lived pathogen serve to substantially increase extinction risk.
-(PCC: seems like you could say this in text w/o using a figure since you have only 6 points here and only 2 of them are different....i.e.
-The probability of host extinction was ten times higher when there was a preference for high quality patches and longer environmental persistence (0.3 vs.
-0.03))}
-\label{pext}
-\end{figure}
 \clearpage

### 2015-07-01 Manuscript/manuscript.tex
Discussion edits
Largely trying to emphasize differences between two scales of "trap effect"

@@
 
-This work demonstrates that (PCC: distribution may often be confused with spatial distribution rather than quality) habitat quality in a metapopulation can have substantial impacts on the dynamics of a variety of environmentally transmitted pathogens and the resulting size and stability of the host population.
-(PCC: deleted, lots of keywords but no information).
-In many cases, high quality habitat can enhance the disease spread and impact of environmentally persistent pathogens.
-In addition, the habitats with many resources may become ecological traps that attract hosts, but are a net drain on the metapopulation due to the impacts of disease in those patches.
-This suggests a trade-off between processes that facilitate metapopulation stability at the expense of disease spread, and processes that hinder disease spread at the expense of metapopulation stability (PCC: rather than saying "processes", elaborate on what specific process you mean.
-).
+This work demonstrates that habitat quality in a metapopulation can have substantial impacts on the dynamics of a variety of environmentally transmitted pathogens and the resulting size and stability of the host population.
+In many cases, the presence of high quality habitat can enhance the spread and impact of environmentally persistent pathogens.  
+In addition, the patches with many resources may become metapopulation-scale ecological traps that are normally beneficial and attractive to hosts, but become a net drain on the metapopulation due to the impacts of disease.
+This suggests that the presence of a disease can lead to a trade-off between otherwise high quality habitat that facilitates metapopulation stability at the expense of increased disease spread, and low quality habitats that hinder disease spread at the expense of metapopulation stability.
 
-This work expands on previous studies of disease spread in a homogeneous metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).
+This expands on previous studies of disease spread in a homogeneous metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).
 Shifting the patch quality distribution towards high quality habitat in our model is conceptually similar to increasing movement in these models.
-And, as in these models, the net consequences of this trade-off for the host metapopulation depend on the traits of the pathogen.
-Broadly, we can divide the observed dynamics into three different scenarios: limited pathogen spread, intermediate, and high pathogen endemicity (PCC: shouldn't there be a fourth scenario of pathogen drives the host extinct?).
-High quality habitats are a net benefit when pathogen spread is limited or is high enough that high and low quality patches are affected by disease.
+And, as in these models, the net consequences of this trade-off for the host metapopulation depend on the characteristics of the disease.
+Broadly, we can divide the observed dynamics into three different scenarios: limited pathogen spread, intermediate, and pandemic spread (possibly leading to host extinction). 
+High quality habitats are a net benefit to the host population when pathogen spread is limited or is high enough that both high and low quality patches are affected by disease. 
+In these cases, the connectivity and stability of high quality patches facilitate larger total population sizes with little additional pathogen spread.
+
 There is an intermediate scenario, however, where low quality habitats are more likely to remain uninfected and as a result boost the metapopulation size while high quality patches that are heavily infected tend to reduce the metapopulation size.
-(PCC Note: most things you label as pathogen traits are actually emergent properties of the host-pathogen-environment...e.g.
-virulence)  (PCC: my text here makes some of the stuff below redundant)
+In this scenario, the ability of pathogens to invade and spread is influenced substantially by the colonization and extinction rates of high and low quality habitat, and the contrasting roles that these habitat types play in the metapopulation.
 
-\subsection*{Pathogens for which high quality habitat increases host population}
-
-In the first of these scenarios, when pathogens have high disease-induced mortality and weak transmission (bottom left of panels in Figs~\ref{endemic}, \ref{highvlow}), pathogen spread is so limited that habitat quality has little effect on epidemic dynamics.
-As a result, high quality patches are a net benefit to the host metapopulation relative to low quality habitat, because their connectivity and stability facilitate larger total population sizes with little additional pathogen spread.
-
-Similarly, in the third scenario, when the pathogen spreads easily as a result of low disease-induced mortality and high transmission (upper portions of panels in Figs~\ref{endemic}, \ref{highvlow}), high quality habitat is again a net benefit to the total host population size.
-Under these conditions, habitat quality again has a limited effect on epidemic dynamics, as the pathogen spreads easily enough to overwhelm any metapopulation structure.
-Here then, the benefit of high quality habitat relative to low quality derives primarily from its ability to better maintain metapopulation stability in the face of widespread disease-induced mortality.
-
-\subsection*{Pathogens for which high quality habitat decreases host population}
-
-However, in the second scenario, where the pathogen's traits facilitate intermediate spread (i.e.
-at an endemic level), the distribution of habitat quality exerts a large influence on epidemic dynamics, with consequences for the overall host population size.
-In particular, the ability of these pathogens to invade and spread is influenced substantially by the colonization and extinction rates of high and low quality habitat, and the contrasting roles that these habitat types play in the metapopulation.
-
-Because high quality habitat supports larger host population sizes, host populations on high quality habitat are less susceptible to environmental stochasticity and thus experience less frequent extinction events.
+Because high quality habitat supports larger host population sizes, these populations are less susceptible to environmental stochasticity and thus experience less frequent extinction events.
 Due to this additional stability, high quality habitat is better able to both maintain a consistent pool of susceptible hosts to which infection can spread, and support infectious host populations under disease-induced mortality.
-As a result, the probability of infection (i.e. proportion of time occupied by infectious hosts) increases with patch quality (Fig~\ref{pinfection}).
+As a result, the probability of infection (i.e. proportion of time occupied by infectious hosts) increases with patch quality (Supplement).
 From the perspective of a migrant susceptible host then, selection of high quality habitat leads to increased disease risk relative to other habitats, either because the patch is likely already occupied by infectious hosts, or because the local susceptible hosts are likely to soon become infected (either through the presence of an environmental reservoir or direct contact).
@@
 Thus, if high quality habitat is preferred by migrants (i.e.
-$\xi_{em} > 0$), it becomes an ecological trap, that is, preferentially selected habitat that reduces fitness relative to other habitats (\cite{Robertson2006}).
-This fitness cost (i.e.
-the disease risk) becomes more severe as pathogen longevity increases, reflecting the more prominent role of environmental transmission on high quality habitat.
-(PCC: You're burying the lead.
-This stuff should go to the beginning of the discussion.
-)
-Moreover, the preferential selection of high quality habitat further exacerbates the increase in disease risk with habitat quality (Fig~\ref{pinfection}) by increasing the risk of direct transmission from infectious migrants also selecting high quality habitat.
+$\xi_{em} > 0$), it forms an ecological trap at the individual scale, that is, preferentially selected habitat that reduces individual fitness relative to other habitats (\cite{Robertson2006}).
+Moreover, the preferential selection of high quality habitat further exacerbates the increase in disease risk with habitat quality (Supplement) by increasing the risk of direct transmission from infectious migrants also selecting high quality habitat.
 This echoes results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).
@@
 The relatively low extinction rate on high quality patches, and the resulting increased probability of infection, helps to create a relatively stable platform from which the pathogen can spread through the rest of the metapopulation.
-Indeed, for pathogens with traits in this intermediate range, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).  
-This helps to maintain infectious occupancy throughout the metapopulation, which, when coupled with preference for high quality habitat, feeds back on high quality patches to ensure a steady stream of infectious colonists.
-These infectious colonists then help to maintain the environmental reservoir and the trap effect.  
-Thus, through these two interacting mechanisms -- the trap and superspreader effects -- the presence of high quality patches serves to significantly increase pathogen spread, even for more virulent pathogens (Fig~\ref{endemic}, \ref{popsizes}).
-The consequences of the concomitant widespread disease-induced mortality outweigh the positive consequences of increased metapopulation connectivity and stability, leading to a net decrease in total host population size on high quality habitat distributions (Fig~\ref{highvlow}, \ref{popsizes}).
+Indeed, for pathogens with spread in this intermediate range, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).  
+This helps to maintain infectious occupancy throughout the metapopulation, which, when coupled with preference for high quality habitat, feeds back on high quality patches to ensure a steady stream of new infectious colonists.
+Thus, through these two interacting processes -- the ecological trap created by the preference of individuals for high quality habitat, and the superspreader role emerging from the increased connectivity and stability -- the presence of high quality patches serves to significantly increase pathogen spread, even for more virulent pathogens (Fig~\ref{endemic}, \ref{popsizes}).
+The consequences of the concomitant widespread disease-induced mortality outweigh the positive consequences of increased metapopulation connectivity and stability, leading to the creation of a metapopulation-scale trap and a net decrease in total host population size on high quality habitat distributions (Fig~\ref{highvlow}, \ref{popsizes}).
 
 In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{endemic}, \ref{popsizes}). 
-Even though individual low quality patches, due to their high extinction rates and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}), the presence of low quality habitat facilitates susceptible persistence for a greater range of pathogens (Fig~\ref{endemic}).
+Even though individual low quality patches, due to their high extinction and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}, Supplement), the presence of low quality habitat facilitates susceptible persistence for a greater range of pathogens (Fig~\ref{endemic}).
 This phenomenon represents the other side of the role played by high quality patches in that low quality patches are relatively unstable (i.e.
@@
 
-Thus, for pathogens with traits that facilitate this intermediate level of spread, we see a reversal of the relative importance of high and low quality habitat for overall host metapopulation stability.
+Thus, for pathogens with this intermediate level of spread, we see a reversal of the relative importance of high and low quality habitat for overall host metapopulation stability.
 Here, rather than serving as sources that increase host population size, we see that high quality patches become sinks.  
-More importantly, they become ecological traps, which can be even more detrimental to the overall population (\cite{Kristan2003}).  
-Indeed, we see that the combination of the trap effect (i.e.
-host preference for high quality habitat) along with an environmentally long-lived pathogen can have severe consequences for metapopulation stability, substantially increasing the risk of host extinction (Fig~\ref{pext}).
-From a management perspective, it is critical to understand this shift in the effect of habitat quality, as previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable for the host (\cite{Strasser2010}).
-Generally, this means that managers should focus on maintaining high quality habitat, but Strasser et al.
-(2010) show that stochastic disturbance (e.g.
+More importantly, they become ecological traps that expose migrant hosts to increased disease risk, which can be even more detrimental to the overall population (\cite{Kristan2003}).  
+Indeed, we see that the combination of host preference for high quality habitat, along with an environmentally long-lived pathogen can have severe consequences for metapopulation stability, substantially increasing the risk of host extinction.
+From a management perspective, it is critical to understand this shift in the effect of habitat quality in order to avoid falling into the metapopulation trap created by high quality habitat.
+Previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable for the host (i.e.
+high quality habitat), but Strasser et al.
+(\cite{Strasser2010}) show that stochastic disturbance (e.g.
 disease-induced mortality) can lead to cases where focusing on low quality habitat is more effective in increasing population growth rate.
-Similarly, our model suggests that in some cases, low quality patch distributions, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can maintain larger total host populations than high quality distributions.
+Similarly, our model suggests that in the presence of some pathogens, maintaining or including low quality habitat in the metapopulation, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can facilitate larger total host populations than focusing on otherwise high quality habitat.
 (PCC: you could also relate to the early metapopulation work, which suggested how even unoccupied habitats were important to metapopulation persistence.
 Here we are layering on environmental pathogens as another reason why lower quality habitat and/or unoccupied habitats are important for conservation efforts.
-This also relates to migration-disease work, whereby migration can help reduce disease by moving the host off of contaminated sites (although there are also other mechanisms)
+This also relates to migration-disease work, whereby migration can help reduce disease by moving the host off of contaminated sites (although there are also other mechanisms)).
 
@@
 
-However, we find that as the environmental longevity of the pathogen increases, the region in which we observe this reversal becomes smaller, that is, high quality habitat is a net benefit for a larger range of potential pathogens (Fig~\ref{highvlow}, red region grows from panels a.
+However, we find that as the environmental longevity of the pathogen increases, the region in which we observe this reversal becomes smaller, that is, high quality habitat is a metapopulation-level trap for a smaller range of potential pathogens (Fig~\ref{highvlow}, blue region shrinks from panels a.
 to c.).
@@
 The different scenarios and pathogens explored here represent a broad sample of the metapopulation dynamics that we might expect to observe in empirical systems.
-The consistent patterns in the relative benefit of low and high quality habitat as a function of pathogen traits suggest that these trade-offs between metapopulation stability and pathogen spread are likely to be fairly general.
+The consistent patterns in the relative benefit of low and high quality habitat suggest that these trade-offs between metapopulation stability and pathogen spread are likely to be fairly general.  
 By evaluating the relative importance of direct versus environmental transmission, along with the effect of habitat quality on connectivity and extinction, we can begin to diagnose where empirical systems (e.g.

### 2015-07-02 Manuscript/manuscript.tex
Language tweaks and added lit to discussion

@@
 To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with patch quality distributions shifted towards either high or low quality habitat.
-These two distributions were chosen to have equal variance and an overlapping core of patches around unit quality, with either the minimum or maximum shifted to simulate management actions that add either low or high quality habitat respectively (yielding distributions with qualities ranging from 0.23 to 1.24 and from 0.76 to 1.77).
+These two distributions were chosen to have equal variance and an overlapping core of patches around unit quality, with either the minimum or maximum shifted to simulate management actions that add either low or high quality habitat (yielding distributions with qualities ranging from 0.23 to 1.24 and from 0.76 to 1.77 respectively).
 We then implemented a full-factorial set of simulations on each quality distribution in which the pathogen's disease-induced mortality ($\nu$) and probability of direct infection ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).
@@
 
-To more carefully explore the effects of habitat quality on patch-level dynamics, we performed an additional simulation experiment on metapopulations with patch qualities evenly spaced over the full range explored above (from 0.2 to 1.8).
+To more carefully explore the role of habitat quality in within-patch dynamics, we performed an additional simulation experiment on metapopulations with patch qualities evenly spaced over the full range explored above (from 0.2 to 1.8).
 In these simulations, we varied longevity over the values used above and explored the consequences of host preference for high quality habitat (and the corresponding potential for ecological traps) by choosing two different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating that migrants select all patches with equal probability).
-To isolate the effects of habitat preference and its interaction with pathogen longevity, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), and landscape structure (fully connected).
+To isolate the effects of habitat preference and pathogen longevity, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), and landscape structure (fully connected).
 We performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by susceptible and infectious hosts, and the final total host population size.
@@
 The spread of these pathogens was strongly influenced by the habitat quality distribution, with low quality habitat limiting disease spread and better maintaining susceptible occupancy relative to high quality habitat (Fig~\ref{popsizes}, Supplemental).
-In these cases, high quality habitat serves as a sort of metapopulation-level trap, that is, otherwise beneficial habitat that is detrimental to the host population in the presence of disease.
-As pathogen longevity increases, the range of pathogens for which we observe this metapopulation-level trap becomes smaller, apparent only for the most virulent pathogens.
+In these cases, high quality habitat served as a sort of metapopulation-level trap, that is, otherwise beneficial habitat that is detrimental to the host population in the presence of disease.
+As pathogen longevity increased, the range of pathogens for which we observed this metapopulation-level trap became smaller, apparent only for the most virulent pathogens.
 Again, metapopulations with a lattice structure displayed similar patterns (Supplement).
 
-To better understand the patch-level mechanisms behind these effects, we further examined within patch infection dynamics for a pathogen in this intermediate range (again with $\nu = 0.2$ and $\delta = 0.3$).
-Because of their lower extinction rate and higher recolonization rate, high quality patches supported more consistent occupancy than low quality patches (Fig~\ref{simvis}, Supplement).  
+To better understand the patch-level mechanisms behind these effects, we further examined within patch infection dynamics for a pathogen in this intermediate range ($\nu = 0.2$ and $\delta = 0.3$).  
+Because of their lower extinction rate and higher recolonization rate, high quality patches were more consistently occupied than low quality patches (Fig~\ref{simvis}, Supplement).  
 However, because of this stability, disease risk, measured by the proportion of time a patch spent occupied by infectious hosts, increased with patch quality (Supplement).
@@
 
-This work demonstrates that habitat quality in a metapopulation can have substantial impacts on the dynamics of a variety of environmentally transmitted pathogens and the resulting size and stability of the host population.
+This work demonstrates that the habitat quality in a metapopulation can have substantial impacts on the dynamics of a variety of environmentally transmitted pathogens and the resulting size and stability of the host population.
 In many cases, the presence of high quality habitat can enhance the spread and impact of environmentally persistent pathogens.  
-In addition, the patches with many resources may become metapopulation-scale ecological traps that are normally beneficial and attractive to hosts, but become a net drain on the metapopulation due to the impacts of disease.
-This suggests that the presence of a disease can lead to a trade-off between otherwise high quality habitat that facilitates metapopulation stability at the expense of increased disease spread, and low quality habitats that hinder disease spread at the expense of metapopulation stability.
+In addition, these patches with many resources may become metapopulation-scale ecological traps that are normally beneficial and attractive to hosts, but become a net drain on the metapopulation due to the impacts of disease.
+This suggests that the presence of a disease can lead to a trade-off between otherwise high quality habitat that facilitates metapopulation stability at the expense of increased disease spread, and low quality habitat that hinders disease spread at the expense of metapopulation stability.
 
@@
 And, as in these models, the net consequences of this trade-off for the host metapopulation depend on the characteristics of the disease.
-Broadly, we can divide the observed dynamics into three different scenarios: limited pathogen spread, intermediate, and pandemic spread (possibly leading to host extinction). 
-High quality habitats are a net benefit to the host population when pathogen spread is limited or is high enough that both high and low quality patches are affected by disease. 
+Broadly, we can divide the observed dynamics into three different scenarios: limited, intermediate, and pandemic pathogen spread (possibly leading to host extinction). 
+High quality habitats are a net benefit to the host population when pathogen spread is either limited or is high enough that both high and low quality patches are affected by disease. 
 In these cases, the connectivity and stability of high quality patches facilitate larger total population sizes with little additional pathogen spread.
@@
 In this scenario, the ability of pathogens to invade and spread is influenced substantially by the colonization and extinction rates of high and low quality habitat, and the contrasting roles that these habitat types play in the metapopulation.
+
+\subsubsection*{Role of high quality habitat}
 
@@
 This helps to maintain infectious occupancy throughout the metapopulation, which, when coupled with preference for high quality habitat, feeds back on high quality patches to ensure a steady stream of new infectious colonists.
-Thus, through these two interacting processes -- the ecological trap created by the preference of individuals for high quality habitat, and the superspreader role emerging from the increased connectivity and stability -- the presence of high quality patches serves to significantly increase pathogen spread, even for more virulent pathogens (Fig~\ref{endemic}, \ref{popsizes}).
+Through these two interacting processes -- the ecological trap created by the preference of individuals for high quality habitat, and the superspreader role emerging from the increased connectivity and stability -- the presence of high quality patches serves to significantly increase pathogen spread, even for more virulent pathogens (Fig~\ref{endemic}, \ref{popsizes}).
 The consequences of the concomitant widespread disease-induced mortality outweigh the positive consequences of increased metapopulation connectivity and stability, leading to the creation of a metapopulation-scale trap and a net decrease in total host population size on high quality habitat distributions (Fig~\ref{highvlow}, \ref{popsizes}).
 
-In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy (Fig~\ref{endemic}, \ref{popsizes}). 
-Even though individual low quality patches, due to their high extinction and low colonization rates, are unable to support consistent occupancy (Fig~\ref{simvis}, Supplement), the presence of low quality habitat facilitates susceptible persistence for a greater range of pathogens (Fig~\ref{endemic}).
+\subsubsection*{Role of low quality habitat}
+In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy in the metapopulation (Fig~\ref{endemic}, \ref{popsizes}). 
+Even though individual low quality patches, due to their high extinction and low colonization rates, are only able to support transient occupancy (Fig~\ref{simvis}, Supplement), the presence of low quality habitat facilitates more widespread susceptible persistence for a greater range of pathogens (Fig~\ref{endemic}).
 This phenomenon represents the other side of the role played by high quality patches in that low quality patches are relatively unstable (i.e.
@@
 Since they are infrequently colonized (when high quality habitat is preferentially selected, $\xi_{em} > 0$), the environmental reservoir left behind by these infectious populations likely decays before it has the opportunity to infect new susceptible colonists.
-As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and allowing susceptible hosts to more easily persist and avoid infection.
-By preventing the widespread disease-induced mortality observed in high quality habitat distributions, low quality distributions are then able to maintain larger host population sizes (Fig~\ref{endemic}, \ref{popsizes}).
+As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and allowing susceptible hosts to more easily persist and avoid infection throughout the metapopulation.
+The opportunity for pathogen reservoir decay provided by the ephemeral occupancy of low quality habitat echoes the 'migratory escape' hypothesis (\cite{Loehle1995}), which suggests that annual migration allows hosts to vacate contaminated sites and return once they are clean.
+Taken with this literature, our results highlight the importance of transient/temporary habitat in mitigating host exposure to environmentally transmitted pathogens.
+
+\subsubsection*{Implications for management}
 
@@
 Indeed, we see that the combination of host preference for high quality habitat, along with an environmentally long-lived pathogen can have severe consequences for metapopulation stability, substantially increasing the risk of host extinction.
-From a management perspective, it is critical to understand this shift in the effect of habitat quality in order to avoid falling into the metapopulation trap created by high quality habitat.
+From a management perspective, it is critical to understand this shift in the effect of habitat quality in order to avoid falling into the metapopulation trap created by high quality habitat and its interaction with pathogen spread.
+
 Previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable for the host (i.e.
@@
 disease-induced mortality) can lead to cases where focusing on low quality habitat is more effective in increasing population growth rate.
-Similarly, our model suggests that in the presence of some pathogens, maintaining or including low quality habitat in the metapopulation, by hindering the spread of the pathogen and providing refuges for susceptible hosts, can facilitate larger total host populations than focusing on otherwise high quality habitat.
-(PCC: you could also relate to the early metapopulation work, which suggested how even unoccupied habitats were important to metapopulation persistence.
-Here we are layering on environmental pathogens as another reason why lower quality habitat and/or unoccupied habitats are important for conservation efforts.
-This also relates to migration-disease work, whereby migration can help reduce disease by moving the host off of contaminated sites (although there are also other mechanisms)).
+Ovaskainen and Hanski (\cite{Ovaskainen2003a}) similarly demonstrate that correlated extinctions reduce the contributions of well-connected habitat to metapopulation persistence and increase the contributions of more isolated habitat.
+Our model adds to this literature by demonstrating the potential for environmentally persistent pathogens to provide the kind of disturbance under which low quality habitat is beneficial.
 
@@
 to c.).
-(PCC: would it be more accurate to describe the effect of envi.
-longevity as unimodal whereby at low longevities the pathogen works as though it were directly transmitted...and so there is a strong correlation between disease occupancy and host occupancy of a patch.
-At intermediate longevity, low quality patches are important as they are less likely to be occupied allowing time for decontamination.
-At high levels of longevity, patches are not unoccupied for long enough for decontamination to occur.
-If this is a fair representation, then we should avoid saying things like the 1st sentence as it implied a monotonic relationship.
-)
 Environmentally long-lived pathogens are able to spread widely for a larger range of conditions, largely overwhelming the ability of low quality habitat to provide refuges for susceptible hosts. 

### 2015-07-02 Manuscript/supplementary.tex
Adding figures for ease of transfer
Updates to supp plots.

@@
 
-This supplement contains results from simulations on a metapopulation with patches arranged in a lattice structure (i.e.
+Supplemental figures 1 and 2 are analogous to figures 1 and 2 from the main text, but with habitat patches arranged in a lattice structure (i.e.
 each habitat patch is accessible only from its four nearest neighbors).
-Supplemental figures 1 and 2 are analogous to figures 1 and 2 from the main text.
+
+Supplemental figures 3 and 4 show within-patch occupancy for a given $\nu$ and $\delta$.
 
 \begin{figure}
+\includegraphics[scale=0.5]{figure/supplement_1.pdf}
 \caption{The proportion of simulations resulting in endemic disease dynamics (persistence of both susceptible and infectious hosts) as a function of disease-induced mortality ($\nu$), and the probability of direct transmission ($\delta$).
@@
 \begin{figure}
+\includegraphics[scale=0.5]{figure/supplement_2.pdf}
 \centering
@@
 
+\begin{figure}
+\includegraphics[scale=0.6]{figure/supplement_3}
+\centering
+\caption{Probability of susceptible occupancy (measured as the proportion of time spent occupied by susceptible hosts over the last 500 time steps) as a function of patch quality.
+Red points indicate host preference for high quality habitat ($\xi_{im} = 0.5$), while black points indicate no preference ($\xi_{im} = 0$).
+Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.
+Disease-induced mortality and probability of direct infection both fixed at $\nu = 0.2$ and $\delta = 0.3$, with fully connected landscape structure.}
+\end{figure}
+
+\begin{figure}
+\includegraphics[scale=0.6]{figure/supplement_4}
+\centering
+\caption{Probability of infectious occupancy (measured as the proportion of time spent occupied by infectious hosts over the last 500 time steps) as a function of patch quality.
+Red points indicate host preference for high quality habitat ($\xi_{im} = 0.5$), while black points indicate no preference ($\xi_{im} = 0$).
+Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.
+Disease-induced mortality and probability of direct infection both fixed at $\nu = 0.2$ and $\delta = 0.3$, with fully connected landscape structure.}
+\end{figure}
+
 \end{document}

### 2015-07-02 Analysis/pubplots.R

@@ -157,87 +157,6 @@ dev.off()
 # See code in simvis.r
 
 #===============================================================================
-### Figure 5
-
-rm(list = ls())
-
-pdf("Manuscript/figure/figure_5.pdf", width = 12, height = 6)
-
-load(paste(getwd(), "/Output/trap.RData", sep = ""))
-
-quality <- unique(out$quality)
-longevity <- unique(out$longevity)
-
-longs <- longevity[c(4, 7, 10)]
-
-labels <- vector(length = 100, mode = "character")
-labels[c(1, 50, 100)] <- c("0.2", "1", "1.8")
-
-sub <- subset(out, xi_em == 0.5 & xi_im == 0 & longevity == longs[1])
-p1 <- ggplot(sub, aes(x = as.factor(quality), y = I)) + theme_classic()
-p1 <- p1 + xlab("Quality") + ylab("Probability of infection") + ggtitle("a.")
-p1 <- p1 + geom_tufteboxplot(outlier.colour = "grey80", position = position_dodge(width = 1)) +
-  scale_y_continuous(expand = c(0, 0.01)) + scale_x_discrete(labels = labels) + theme(axis.ticks.x = element_blank())
-
-sub <- subset(out, xi_em == 0.5 & xi_im == 0.5 & longevity == longs[1])
-p2 <- ggplot(sub, aes(x = as.factor(quality), y = I)) + theme_classic()
-p2 <- p2 + xlab("Quality") + ylab("Probability of infection") + ggtitle("d.")
-p2 <- p2 + geom_tufteboxplot(outlier.colour = "grey80", position = position_dodge(width = 1)) +
-  scale_y_continuous(expand = c(0, 0.01)) + scale_x_discrete(labels = labels) + theme(axis.ticks.x = element_blank())
-
-
-sub <- subset(out, xi_em == 0.5 & xi_im == 0 & longevity == longs[2])
-p3 <- ggplot(sub, aes(x = as.factor(quality), y = I)) + theme_classic()
-p3 <- p3 + xlab("Quality") + ylab("Probability of infection") + ggtitle("b.") 
-p3 <- p3 + geom_tufteboxplot(outlier.colour = "grey80", position = position_dodge(width = 1)) +
-  scale_y_continuous(expand = c(0, 0.01)) + scale_x_discrete(labels = labels) + theme(axis.ticks.x = element_blank())
-
-sub <- subset(out, xi_em == 0.5 & xi_im == 0.5 & longevity == longs[2])
[…]

### 2015-07-08 Manuscript/manuscript.tex
A few text edits and comments.

@@
 
-\title{The role of habitat quality in host metapopulations exposed to environmentally persistent pathogens.
-Or Identifying when environmentally persistent pathogens transform good habitat into ecological traps}
+\title{When environmentally persistent pathogens transform good habitat into ecological traps}
 
@@
 \begin{abstract} 
-The distribution of habitat quality plays an important role in the dynamics of wildlife metapopulations, with the expectation that high quality habitat serves to stabilize and maintain occupancy (REF).
-However, the costs and benefits of different quality habitat might be modulated, or even reversed, by the presence of an environmentally persistent pathogen.
-The presence of environmental pathogen reservoirs on high quality habitat may lead to the creation of ecological traps, wherein host individuals preferentially colonize high quality habitat, but are then exposed to increased infection risk.
+The benefits of high quality habitat for host persistence may be modulated by the presence of an environmentally persistent pathogen.
+In some cases, the presence of environmental pathogen reservoirs on high quality habitat may lead to the creation of ecological traps, wherein host individuals preferentially colonize high quality habitat, but are then exposed to increased infection risk and disease-induced mortality.
 We explored this possibility through the development of a stochastic patch occupancy model, where we varied the pathogen's virulence, transmission rate, and environmental persistence as well as the distribution of habitat quality in the host metapopulation.
@@
 Indeed, the quality of a habitat patch can affect its extinction and colonization rates, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).
-These processes may influence how a pathogen spreads, where environmental pathogen reservoirs get established, and the resulting effects on host occupancy and population size.
+These processes will influence how a pathogen spreads, where environmental pathogen reservoirs get established, and the resulting effects on host occupancy and population size.
 
@@
 (3)  Connectivity among patches is determined entirely by quality ($d_{ij}=1$ for all $i \neq j$).  
-In addition to these assumptions, parameters were chosen so that without infection, approximately 0.75 of the metapopulation was occupied ($\mu = 0.1$, $D=5$).
-To explore the dynamics of a range of possible pathogens, disease induced mortality ($\nu$), probability of direct infection ($\delta$), and environmental longevity (representing the half-life of the pathogen's infectivity in the environment) were varied over a range of possible values.
+In addition to these assumptions, we chose parameters so that without infection, approximately 0.75 of the metapopulation was occupied ($\mu = 0.1$, $D=5$).
+We varied disease induced mortality ($\nu$), probability of direct infection ($\delta$), and environmental longevity to explore the dynamics of a range of possible pathogens.
 
@@
 To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with patch quality distributions shifted towards either high or low quality habitat.
-These two distributions were chosen to have equal variance and an overlapping core of patches around unit quality, with either the minimum or maximum shifted to simulate management actions that add either low or high quality habitat (yielding distributions with qualities ranging from 0.23 to 1.24 and from 0.76 to 1.77 respectively).
+These two distributions were chosen to have equal variance and an overlapping core of patches around unit quality, with either the minimum or maximum shifted to simulate regions with more, or less, high quality habitat (yielding distributions with qualities ranging from 0.23 to 1.24 and from 0.76 to 1.77 respectively).
 We then implemented a full-factorial set of simulations on each quality distribution in which the pathogen's disease-induced mortality ($\nu$) and probability of direct infection ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).
@@
 In these cases, high quality habitat served as a sort of metapopulation-level trap, that is, otherwise beneficial habitat that is detrimental to the host population in the presence of disease.
-As pathogen longevity increased, the range of pathogens for which we observed this metapopulation-level trap became smaller, apparent only for the most virulent pathogens.
+As pathogen longevity increased, the range of the parameter space where we observed this metapopulation-level trap became smaller, apparent only for the most virulent pathogens.
 Again, metapopulations with a lattice structure displayed similar patterns (Supplement).
@@
 This work demonstrates that the habitat quality in a metapopulation can have substantial impacts on the dynamics of a variety of environmentally transmitted pathogens and the resulting size and stability of the host population.
-In many cases, the presence of high quality habitat can enhance the spread and impact of environmentally persistent pathogens.  
+In some cases, the presence of high quality habitat can enhance the spread and impact of environmentally persistent pathogens.  
 In addition, these patches with many resources may become metapopulation-scale ecological traps that are normally beneficial and attractive to hosts, but become a net drain on the metapopulation due to the impacts of disease.
@@
 In this scenario, the ability of pathogens to invade and spread is influenced substantially by the colonization and extinction rates of high and low quality habitat, and the contrasting roles that these habitat types play in the metapopulation.
+
+PCC: nice summary here. Either in the beginning here or in the conclusions, you could more specifically relate these 3 regions of the parameter space to ongoing issues: 
+White-nose:  seems to be in the category of pandemic spread, probably not a trap scenario
+CWD: probably also "pandemic", but we're still on the early side of the epidemic, very long timescales
+Chytrid: ?
+Plague:  ?Colleen would know better?
+Anthrax: limited spread (or rather limited recurrance) so probably also not a trap scenario.
 
@@
 Through these two interacting processes -- the ecological trap created by the preference of individuals for high quality habitat, and the superspreader role emerging from the increased connectivity and stability -- the presence of high quality patches serves to significantly increase pathogen spread, even for more virulent pathogens (Fig~\ref{endemic}, \ref{popsizes}).
-The consequences of the concomitant widespread disease-induced mortality outweigh the positive consequences of increased metapopulation connectivity and stability, leading to the creation of a metapopulation-scale trap and a net decrease in total host population size on high quality habitat distributions (Fig~\ref{highvlow}, \ref{popsizes}).
+The consequences of widespread disease-induced mortality can outweigh the positive consequences of increased metapopulation connectivity and stability, leading to the creation of a metapopulation-scale trap and a net decrease in total host population size on high quality habitat distributions (Fig~\ref{highvlow}, \ref{popsizes}).
 
@@
 From a management perspective, it is critical to understand this shift in the effect of habitat quality in order to avoid falling into the metapopulation trap created by high quality habitat and its interaction with pathogen spread.
+PCC: this paragraph seems repetitive of earlier statements.
 
@@
 Ovaskainen and Hanski (\cite{Ovaskainen2003a}) similarly demonstrate that correlated extinctions reduce the contributions of well-connected habitat to metapopulation persistence and increase the contributions of more isolated habitat.
-Our model adds to this literature by demonstrating the potential for environmentally persistent pathogens to provide the kind of disturbance under which low quality habitat is beneficial.
+Our model adds to this literature by demonstrating the potential for environmentally persistent pathogens to provide the kind of disturbance under which low quality habitat is beneficial, and thus should be conserved.
 
@@
 Darker shading corresponds to more frequent endemic dynamics.
-The white region above the endemic region generally corresponds to pandemic dynamics (no susceptible host populations persist), while the region below the endemic region corresponds to disease-free dynamics.}
+The white region above the endemic region generally corresponds to pandemic dynamics (no susceptible host populations persist), while the region below the endemic region corresponds to disease-free dynamics.
+PCC: in this figure and in Fig 2 I'd put write the longevities in each sub-figure and remove the longevity label from the x-axis.
+I'd prefer to not refer to the persistence of susceptible patches as a "endemic".
+Maybe this figure should instead be a heatmap of the percentage of patches infected.
+This can then be contrasted with Fig.2 showing that only in the intermediate infection scenarios are there traps.
+This avoids the "endemic" v.
+pandemic issue}
 \label{endemic}
@@
 At low longevities, this pathogen is largely unable to invade and high quality habitat supports larger populations of susceptible hosts.
-However, low quality habitat is better able to maintain these susceptible hosts as the pathogen's longevity increases, leading to larger total population sizes.}
+However, low quality habitat is better able to maintain these susceptible hosts as the pathogen's longevity increases, leading to larger total population sizes.
+PCC: seems like you chose different longevities here for your low,med, high scenarios than in Fig.
+1&2.
+Not a big issue, but might point out that the low longevity here is really just a no-disease, completely healthy scenario}
 \label{popsizes}

### 2015-07-09 Manuscript/manuscript.tex
Redo figure 1 and longevity labels

@@
 \includegraphics[scale=0.5]{figure/figure_1.pdf}
-\caption{The proportion of simulations resulting in endemic disease dynamics (persistence of both susceptible and infectious hosts) as a function of infectious survival ($\nu$), and the probability of direct transmission ($\delta$).
+\caption{Median total susceptible population size as a function of infectious survival ($\nu$), and the probability of direct transmission ($\delta$).
 Panel columns show low (0.3, a, d), medium (1.0, b, e), and high (3.2, c, f) pathogen longevities, while rows show low quality (ranging from 0.23 to 1.24, a - c) and high quality (ranging from 0.76 to 1.77, d - f) patch quality distributions.
-Darker shading corresponds to more frequent endemic dynamics.
-The white region above the endemic region generally corresponds to pandemic dynamics (no susceptible host populations persist), while the region below the endemic region corresponds to disease-free dynamics.
-PCC: in this figure and in Fig 2 I'd put write the longevities in each sub-figure and remove the longevity label from the x-axis.
-I'd prefer to not refer to the persistence of susceptible patches as a "endemic".
-Maybe this figure should instead be a heatmap of the percentage of patches infected.
-This can then be contrasted with Fig.2 showing that only in the intermediate infection scenarios are there traps.
-This avoids the "endemic" v.
-pandemic issue}
+Darker shading corresponds to larger population sizes.}
 \label{endemic}
@@
 \centering
-\caption{Differences between the median total effective population size of hosts on low quality patch distributions (quality between 0.23 to 1.24) and high quality patch distributions (quality between 0.76 and 1.77) as a function of disease-induced mortality ($\nu$) and the probability of direct transmission ($\delta$).
+\caption{Differences between the median total effective population size of hosts on low quality patch distributions (quality between 0.23 to 1.24) and high quality patch distributions (quality between 0.76 and 1.77) as a function of infectious survival ($\nu$) and the probability of direct transmission ($\delta$).
 Reds indicate that high quality habitat distributions support higher median host population sizes than low quality habitat, primarily corresponding to pathogens with either very limited or very extensive spread.
@@
 Colored points show medians, while vertical lines indicate the inner-quartile ranges, with horizontal lines indicating the minimum and maximum.
-At low longevities, this pathogen is largely unable to invade and high quality habitat supports larger populations of susceptible hosts.
-However, low quality habitat is better able to maintain these susceptible hosts as the pathogen's longevity increases, leading to larger total population sizes.
-PCC: seems like you chose different longevities here for your low,med, high scenarios than in Fig.
-1&2.
-Not a big issue, but might point out that the low longevity here is really just a no-disease, completely healthy scenario}
+At low longevities, this pathogen is unable to invade and high quality habitat supports larger populations of susceptible hosts.
+However, low quality habitat is better able to maintain these susceptible hosts as the pathogen's longevity increases, leading to larger total population sizes.}
 \label{popsizes}

### 2015-07-09 Analysis/pubplots.R

@@ -14,32 +14,34 @@ pdf("Manuscript/figure/figure_1.pdf", width = 10, height = 6)
 
 load(paste(getwd(), "/Output/pathgrid.RData", sep = ""))
 
-out$endemic <- out$S > 0 & out$I > 0
-dat <- ddply(out, c("longevity", "treatment", "delta", "nu"), summarise, sum = sum(endemic) / 100)
+dat <- ddply(out, c("longevity", "treatment", "delta", "nu"), summarise, susc = median(S.pop))
 
-cols <- colorRampPalette(brewer.pal(9, "Greys"))(100)
+cols <- colorRampPalette(brewer.pal(9, "Greys"))(120)
 labels <- c("d", "e", "f", "a", "b", "c")
 
-plot <- levelplot(sum ~ delta * nu | longevity * treatment, data = dat,
+plot <- levelplot(susc ~ delta * nu | longevity * treatment, data = dat,
           col.regions = cols,
-          at = seq(0, 1, by = 0.01),
+          at = seq(0, 120, by = 1),
           xlab = list(label = expression("Probability of direct transmission"~(delta)), cex = 1),
           ylab = list(label = expression("Infectious survival"~(nu)), cex = 1),
           strip = F,
           scales = list(alternating = F, cex = 1),
-          ylab.right = list(label = "Probability of endemic", cex = 1),
+          ylab.right = list(label = "Susceptible population size", cex = 1),
           par.settings = list(layout.widths = list(axis.key.padding = 0, ylab.right = 3)),
           panel=function(...){
             panel.levelplot(...)
             panel.text(0, 1, labels[panel.number()], cex = 1)
           })
 
-col.lab <- textGrob(expression(Longevity ~ symbol("\256")))
 row.low <- textGrob("Low quality", rot = 90)
 row.high <- textGrob("High quality", rot = 90)
 
 grid.arrange(arrangeGrob(row.low, row.high, ncol = 1), plot, ncol = 2, 
-             widths = c(0.05, 0.95), sub = col.lab)
+             widths = c(0.05, 0.95))
+
+grid.text("Longevity = 0.32", x = unit(0.26, units = "npc"), y = unit(0.96, units = "npc"), draw = T)
+grid.text("Longevity = 1.0", x = unit(0.50, units = "npc"), y = unit(0.96, units = "npc"), draw = T)
+grid.text("Longevity = 3.2", x = unit(0.75, units = "npc"), y = unit(0.96, units = "npc"), draw = T)
 
 dev.off()
 
[…]

### 2015-07-15 Manuscript/manuscript.tex
Discussion edits
Fixing redundancy issues and adding discussion of actual diseases.

@@
 We explored this possibility through the development of a stochastic patch occupancy model, where we varied the pathogen's virulence, transmission rate, and environmental persistence as well as the distribution of habitat quality in the host metapopulation.
-This model suggests that for certain (delete or explain what the certain refers to) pathogens, high quality habitat can serve as an ecological trap, and can be detrimental to host persistence relative to low quality habitat.
+This model suggests that for pathogens with intermediate levels of spread, high quality habitat can serve as an ecological trap, and can be detrimental to host persistence relative to low quality habitat.
 This inversion of the relative roles of high and low quality habitat highlights the importance of considering the interaction between spatial structure and pathogen transmission when managing wildlife populations exposed to an environmentally persistent pathogen.
-
-PCC: abstract reads well. You could consider putting in a few more details. e.g. can you characterize the conditions/parameter space that creates the traps rather than just saying they may exist.
-
 \end{abstract}
@@
 Many prominent and problematic wildlife diseases are caused by pathogens that can persist, and remain infectious, for long periods of time in the environment.
-Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\emph{Bacillus anthracis}, \cite{Dragon1995}), plague (\emph{Yersinia pestis}, \cite{Eisen2008}), and white nose syndrome (\emph{Geomyces destructans}, \cite{Lindner2011}), among others.
+Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\emph{Bacillus anthracis}, \cite{Dragon1995}), plague (\emph{Yersinia pestis}, \cite{Eisen2008}), and white nose syndrome (\emph{Pseudogymnoascus destructans}, \cite{Reynolds2015}), among others.
 This environmental persistence creates environmental pathogen reservoirs from which susceptible hosts can become infected without direct contact with an infectious individual.
@@
 \begin{align}
-C_{Si}(t) &= A_i^{\xi_{im}} \sum_{i\neq j }\phi(S, t)_jA_j^{\xi_{em}}\exp(-D d_{ij}),\\
-C_{Ii}(t) &= A_i^{\xi_{im}} \sum_{i\neq j }\phi(I, t)_j(\nu A_j) ^{\xi_{em}}\exp(-D d_{ij}),
+C_{Si}(t) &= A_i^{\xi_{im}} \sum_{i\neq j }\phi_j(S, t)A_j^{\xi_{em}}\exp(-D d_{ij}),\\
+C_{Ii}(t) &= A_i^{\xi_{im}} \sum_{i\neq j }\phi_j(I, t)(\nu A_j) ^{\xi_{em}}\exp(-D d_{ij}),
 \label{connectivity}
@@
 
-where $\phi(X, t)_j$ is an indicator function that is $1$ if patch $j$ is in state $X$ at time $t$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality; $D$ is the inverse of the host's mean dispersal distance; and $d_{ij}$ is the effective distance between patches $i$ and $j$ .
+where $\phi_j(X, t)$ is an indicator function that is $1$ if patch $j$ is in state $X$ at time $t$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality; $D$ is the inverse of the host's mean dispersal distance; and $d_{ij}$ is the effective distance between patches $i$ and $j$ .
 Essentially, $C_{Xi}$ sums the colonization effort (or propagule pressure) from all patches in state $X$ to focal patch $i$.
@@
 
-PCC: nice summary here. Either in the beginning here or in the conclusions, you could more specifically relate these 3 regions of the parameter space to ongoing issues: 
-White-nose:  seems to be in the category of pandemic spread, probably not a trap scenario
-CWD: probably also "pandemic", but we're still on the early side of the epidemic, very long timescales
-Chytrid: ?
-Plague:  ?Colleen would know better?
-Anthrax: limited spread (or rather limited recurrance) so probably also not a trap scenario.
-
 \subsubsection*{Role of high quality habitat}
@@
 
-\subsubsection*{Implications for management}
-
-Thus, for pathogens with this intermediate level of spread, we see a reversal of the relative importance of high and low quality habitat for overall host metapopulation stability.
-Here, rather than serving as sources that increase host population size, we see that high quality patches become sinks.  
-More importantly, they become ecological traps that expose migrant hosts to increased disease risk, which can be even more detrimental to the overall population (\cite{Kristan2003}).  
-Indeed, we see that the combination of host preference for high quality habitat, along with an environmentally long-lived pathogen can have severe consequences for metapopulation stability, substantially increasing the risk of host extinction.
 From a management perspective, it is critical to understand this shift in the effect of habitat quality in order to avoid falling into the metapopulation trap created by high quality habitat and its interaction with pathogen spread.
-PCC: this paragraph seems repetitive of earlier statements.
-
 Previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable for the host (i.e.
@@
 
-The different scenarios and pathogens explored here represent a broad sample of the metapopulation dynamics that we might expect to observe in empirical systems.
-The consistent patterns in the relative benefit of low and high quality habitat suggest that these trade-offs between metapopulation stability and pathogen spread are likely to be fairly general.  
-By evaluating the relative importance of direct versus environmental transmission, along with the effect of habitat quality on connectivity and extinction, we can begin to diagnose where empirical systems (e.g.
-plague in prairie dogs, or white nose syndrome in bats) fall on this spectrum and the corresponding consequences for management.
-As we have seen, failing to consider the characteristics of a pathogen and their effect on the roles played by high and low quality habitat risks overlooking the importance of low-quality habitat for the persistence of wildlife populations and the importance of high-quality habitat for the persistence of wildlife disease.
+The three scenarios identified here provide a rough framework that we can use to identify the potential risk of metapopulation-level traps in wildlife populations currently affected by an environmentally persistent pathogen.
+For instance, high quality habitat seems unlikely to form metapopulation traps for bat populations affected by white nose syndrome due to the widespread infection both within and between hibernaculum and the scarcity of host refuges (\cite{Langwig2014}, \cite{ORegan2015}).
+Similarly, from these results, we would not expect the spread of chronic wasting disease to turn high quality habitat into traps for mule deer metapopulations, due again to widespread infection (albeit at low prevalence) across population units (\cite{Conner2004}) and the relative importance of local-scale transmission within winter ranges (\cite{Farnsworth2006}).
+However, we would expect to see high quality traps in the plague-prairie dog system, where the infection is widespread and disease-induced mortality is high, but uninfected colonies are still able to persist (\cite{Stapp2004}).
+Indeed, studies have suggested that large colonies are more likely to become infected (\cite{Snall2008}), and thus high quality habitat capable of supporting large population sizes may help facilitate plague persistence and spread.
+Thus, by considering the characteristics of a host-pathogen system and their effect on the roles played by high and low quality habitat, we can begin to diagnose into which of these three scenarios empirical systems fall and the corresponding consequences for management.
+As we have seen, failing to do so risks overlooking the importance of low-quality habitat for the persistence of wildlife populations and the importance of high-quality habitat for the persistence of wildlife disease.
 

### 2015-07-15 Analysis/patch.analysis.r
Changes to exploratory figure scripts

@@ -256,18 +256,14 @@ points(quality, I[2, 2, 10, ], col = "brown", pch = 4)
 
 par(mfrow = c(1, 3))
 
-occ <- tapply(out$I * 0.2 * out$quality + out$S * out$quality, list(out$xi_em, out$xi_im, out$longevity, out$quality), median)
+occ <- tapply(out$I + out$S, list(out$xi_em, out$xi_im, out$longevity, out$quality), median)
 
-plot(quality, occ[2, 1, 4, ], ylim = c(0, 2), ylab = "Probability occupied", xlab = "Quality", pch = 4)
+plot(quality, occ[2, 1, 4, ], ylim = c(0, 1), ylab = "Probability occupied", xlab = "Quality", pch = 4)
 points(quality, occ[2, 2, 4, ], col = "brown", pch = 4)
-abline(a = 0, b = 1)
-plot(quality, occ[2, 1, 7, ], ylim = c(0, 2), ylab = "Probability occupied", xlab = "Quality", pch = 4)
+plot(quality, occ[2, 1, 7, ], ylim = c(0, 1), ylab = "Probability occupied", xlab = "Quality", pch = 4)
 points(quality, occ[2, 2, 7, ], col = "brown", pch = 4)
-abline(a = 0, b = 1)
-plot(quality, occ[2, 1, 10, ], ylim = c(0, 2), ylab = "Probability occupied", xlab = "Quality", pch = 4)
+plot(quality, occ[2, 1, 10, ], ylim = c(0, 1), ylab = "Probability occupied", xlab = "Quality", pch = 4)
 points(quality, occ[2, 2, 10, ], col = "brown", pch = 4)
-abline(a = 0, b = 1)
-abline(a = 0, b = 0.2)
 #===============================================================================
 # Consequences of preference (with xi_em = 0.5)
 

### 2015-07-15 Analysis/pathgrid.analysis.r

@@ -245,7 +245,7 @@ grid.arrange(arrangeGrob(low.h, med.h, high.h, low.l, med.l, high.l, ncol = 3),
 
 load(paste(getwd(), "/Output/pathgrid.RData", sep = ""))
 
-dat <- tapply(out$maxI > 0, list(out$longevity, out$treatment, out$delta, out$nu), sum) / 100
+dat <- tapply(out$S > 0 & out$I > 0, list(out$longevity, out$treatment, out$delta, out$nu), sum) / 100
 
 cols <- colorRampPalette(brewer.pal(9, "Greys"))(100)
 

### 2015-07-15 Manuscript/manuscript.tex
Added sections required by Interface

@@
 
+\newpage
+
 \begin{abstract} 
@@
 \end{abstract}
+
+\subsubsection*{Keywords:}
+disease; metapopulation; habitat quality; environmental transmission; ecological trap
 
@@
 are separated by effective distance 1).
-To examine the effects of a more rigid spatial structure, we also implemented a model with patches arranged in a square lattice such that each patch is only accessible from its four neighboring patches (in which case, average migration distance was adjusted to $D = 2$ to maintain roughly equivalent occupancy).
+To examine the effects of a more rigid spatial structure, we also implemented a model with patches arranged in a square lattice such that each patch is only accessible from its four neighbouring patches (in which case, average migration distance was adjusted to $D = 2$ to maintain roughly equivalent occupancy).
 
@@
 From a management perspective, it is critical to understand this shift in the effect of habitat quality in order to avoid falling into the metapopulation trap created by high quality habitat and its interaction with pathogen spread.
-Previous theoretical work suggests that management should focus on maintaining patches where conditions are most favorable for the host (i.e.
+Previous theoretical work suggests that management should focus on maintaining patches where conditions are most favourable for the host (i.e.
 high quality habitat), but Strasser et al.
@@
 
-\section{Acknowledgements}
+\subsection*{Data accessibility}
+Simulation source code and generated data are available at \url{https://github.com/clint-leach/Metapop-Disease}, along with scripts used to generate figures.
 
-CBL would like to thank members of the Webb Lab, past and present, for several rounds of useful feedback.
-This work was supported by USGS Cooperative Agreement 101485, and NSF Graduate Research Fellowship, No.
-DGE-1321845.
+\subsection*{Competing interests}
+We have no competing interests.
 
+\subsection*{Author's contributions}
+CBL participated in the design of the study, wrote the code and carried out the simulations, and drafted the manuscript; CTW and PCC conceived of the study, participated in the design of the study, and helped draft the manuscript.
+All authors gave final approval for publication.
+
+\subsection*{Acknowledgements}
+
+CBL would like to thank members of the Webb Lab at Colorado State University, past and present, for several rounds of useful feedback.
+
+\subsection*{Funding}
+
+This work was supported by USGS Cooperative Agreement 101485, and NSF Graduate Research Fellowship, No. DGE-1321845.
+\\
+\\
 Any mention of trade, product, or firm names is for descriptive purposes only and does not imply endorsement by the U.S. Government.
@@
 
-%\begin{table}[h!]   
-%\caption{Parameters of the SPOM model, their meaning, and the values assigned under different parameterizations.  Empty cells indicate the same value as the default parameterization.}
-%\begin{tabular}{p{1cm} p{4cm} p{1.5cm} p{1.5cm} p{1.5cm} p{1.5cm} p{1.5cm}}
-%Param. & Interpretation &  Default & Local movement & High env. stoch. & Dominant reservoir trans. & High virulence \\
-%\hline
-%$\xi_{im}$ & Effect of patch quality on immigration & 0.5 & -- & -- & -- & --\\
-%$\xi_{em}$ & Effect of population size on emigration & 0.5 & -- & -- & --& --\\
-%$D$& Inverse of mean dispersal distance & 5 & 2 & -- & -- & --\\
-%$d_{ij}$ & Distance between patch $i$ and $j$ & 1 $\forall i \neq j$ & 1 $\forall i, j$ neighbors & -- & -- & --\\
-%\hline
-%$\mu$ & Extinction rate of unit quality patch & 0.1 & -- & -- & -- & --\\
-%$\nu$ & Disease-induced reduction in population size & 0.2 & -- & -- & -- & 1\\
-%$\alpha$ & Strength of environmental stochasticity & 1 & -- & 0 & -- & --\\
-%\hline
-%$\delta$ & Probability of direct infection & 0.5 & -- & -- & 0.1 & --\\
-%$\gamma_0$ & Initial rate of infection from reservoir patch & 0.5 & -- & -- & -- & --
-%\end{tabular}
-%\label{params}
-%\end{table}
-
 \begin{table}[h!]   

### 2015-07-15 Analysis/highlow.analysis.r
Removing outdated code
Delete code for analyzing results from previous simulations designs no longer in use.

@@ -1,355 +1,0 @@
-# Code to analyze results from highlow scripts, in which the patch quality distribution
-# varies asymmetrically around 1.  The four distributions explored are:
-#   low var: patch qualities between 0.75 and 1.25
-#   intermediate var, high quality: 0.75 - 1.75
-#   intermediate var, low quality: 0.25 - 1.25
-#   high var: 0.25 - 1.75
-
-# Choose which parameterization to explore
-conn <- "(full)"      
-conn <- "(lattice)"
-conn <- "(alpha0)"
-conn <- "(delta)"
-conn <- "(ei)"
-conn <- "(midei)"
-conn <- "(lowei)"
-
-# Read in model output
-highlow<-read.csv(paste(getwd(), "/Output/highlow", conn, ".csv", sep = ""), header=TRUE)
-
-highlow<-highlow[,-1]
-
-longevity <- sort(unique(highlow$longevity))
-
-#===============================================================================
-# Plotting mean occupancy as a function of longevity and treatment
-
-par(mfrow = c(1, 3), bty = "l")
-
-S.occ <- tapply(highlow$S.pop, list(highlow$longevity, highlow$treatment), median)
-
-plot(log10(longevity), S.occ[, "+high"], type = "b", lty = 1, ylim = c(0, 120), pch = 19, col = "red", ylab = "S occupancy", xlab = "Longevity")
-lines(log10(longevity), S.occ[, "+low"], type = "b", lty = 2, pch = 19, col = "blue")
-lines(log10(longevity), S.occ[, "full"], type = "b", lty = 3, pch = 19, col = "black")
-lines(log10(longevity), S.occ[, "low.var"], type = "l", lty = 4, pch = 19, col = "grey")
-
-I.occ <- tapply(highlow$I.pop, list(highlow$longevity, highlow$treatment), median)
-
-plot(log10(longevity), I.occ[, "+high"], type = "b", lty = 1, ylim = c(0, 120), pch = 19, col = "red", ylab = "I occupancy", xlab = "Longevity")
-lines(log10(longevity), I.occ[, "+low"], type = "b", lty = 2, pch = 19, col = "blue")
-lines(log10(longevity), I.occ[, "full"], type = "b", lty = 3, pch = 19, col = "black")
[…]

### 2015-07-15 Analysis/metapop.analysis.r

@@ -1,104 +1,0 @@
-# Scripts to analyze metapopulation level results from full simulation model
-# where the patch quality distribution expands symmetrically about 1.
-
-# Choose model parameterization to explore
-conn <- "(full)"
-conn <- "(lattice)"
-conn <- "(alpha0)"
-conn <- "(delta)"
-conn <- "(ei)"
-conn <- "(eimid)"
-
-load(paste(getwd(), "/Output/out", conn, ".RData", sep = ""))
-
-library(gridExtra)
-library(lattice)
-library(RColorBrewer)
-
-#===============================================================================
-# Level plots of mean occupancy as function of variance and longevity
-
-metapop.S <- tapply(out$S * out$quality, list(out$longevity, out$variance), sum) / 100
-metapop.I <- tapply(out$I * out$quality, list(out$longevity, out$variance), sum) / 100
-metapop.occ <- tapply(out$S * out$quality + out$I * out$quality, list(out$longevity, out$variance), sum) / 100
-maxI <- tapply(out$maxI, list(out$longevity, out$variance), mean)
-
-sd.S <- tapply(out$Sfin, list(out$longevity, out$variance), sd)
-sd.I <- tapply(out$Ifin, list(out$longevity, out$variance), sd)
-sd.maxI <- tapply(out$maxI, list(out$longevity, out$variance), sd)
-
-cols <- colorRampPalette(brewer.pal(9, "Reds"))(100)
-
-S <- levelplot(metapop.S, col.regions = cols, xlab = "Longevity", ylab = "Variance",
-               at = seq(0, 100, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
-I <- levelplot(metapop.I, col.regions = cols, xlab = "Longevity", ylab = "Variance",
-               at = seq(0, 100, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
-occ <- levelplot(metapop.occ, col.regions = cols, xlab = "Longevity", ylab = "Variance",
-                 at = seq(0, 100, by = 1), colorkey = F, scales = list(at = c(1, 3, 5, 7, 9)))
-
-key <- draw.colorkey(list(col = cols, 
-                          at = seq(0, 100, by = 1),
-                          labels = list(at = seq(0, 100, by = 25)),
-                          height = 0.7), 
-                     draw = F)
-
-
-grid.arrange(S, I, occ, key, ncol = 4, widths = c(0.3, 0.3, 0.3, 0.1))
-
[…]

### 2015-07-15 Analysis/patch.analysis.r

@@ -1,117 +1,4 @@
-# Patch level results -- scripts to analyze patch-level data from full simulation
-# model.
-
-# Select model parameterization to analyze
-conn <- "(full)"
-conn <- "(lattice)"
-conn <- "(alpha0)"
-conn <- "(delta)"
-
-load(paste(getwd(), "/Output/out", conn, ".Rdata", sep = ""))
-
-#Filter patch data to include only endemic cases
-patch<-out[which(out$Ifin>0 & out$Sfin > 0), ]
-
-#Filter patch data to include only high variance runs
-patch<-patch[which(abs(patch$variance - 0.2) < 0.01), ]
-
-#===============================================================================
-#Plotting scatter cloud ofproportion of time occupied by susceptibles and infecteds as a function of quality
-
-par(mfrow=c(1,2),cex=1)
-
-smoothScatter(patch$quality,patch$S,ylab="Proportion of time occupied by susceptibles",xlab="Patch Quality")
-lines(loess.smooth(patch$quality,patch$S),lwd=2,col="red")
-smoothScatter(patch$quality,patch$I,ylab="Proportion of time occupied by infecteds",xlab="Patch Quality")
-lines(loess.smooth(patch$quality,patch$I),lwd=2,col="red")
-
-
-# Plotting scatter cloud of number of infection events
-par(mfrow=c(1,1),cex=1)
-smoothScatter(patch$quality,patch$inf.events,ylab="Number of infection events",xlab="Patch Quality")
-lines(loess.smooth(patch$quality,patch$inf.events),lwd=3,col="red")
-
-#===============================================================================
-# Figures to show how longevity affects role of quality
-
-# Plotting infection events against quality for range of longevities
-plot(seq(0.25, 1.75, by = 0.1), c(1:16), ylim = c(0, 10), type = "n", xlab = "Quality", ylab = "Infections")
-for(i in unique(patch$longevity)){
-  lines(loess.smooth(patch$quality[patch$longevity == i], patch$inf.events[patch$longevity == i]), type = "b", pch = as.character(i))  
-}
-
-
-#===============================================================================
-# Showing simulation-level variability
-
-# Number of infection events
-
-library(scales)
-
-reps <- unique(patch$repID)
-
-par(mfrow = c(1, 1))
[…]

### 2015-07-16 Manuscript/manuscript.tex
Consistent vocabulary
Removing 'pandemic', 'endemic', 'pathogen traits', and using 'infectious survival' instead of 'mortality' for nu.

@@
 plague in prairie dog colonies, \cite{George2013}), and thus understanding how pathogen transmission interacts with patterns of habitat quality is critical to managing disease in these systems.
-In this study, we explore how a pathogen's traits interact with the distribution of habitat quality to influence pathogen spread and host occupancy.
+In this study, we explore how pathogen spread interacts with the distribution of habitat quality to influence host occupancy.
 We specifically focus on the relative importance of high and low quality habitats to host persistence in the presence of pathogens with different levels of environmental longevity.
@@
 To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with patch quality distributions shifted towards either high or low quality habitat.
-These two distributions were chosen to have equal variance and an overlapping core of patches around unit quality, with either the minimum or maximum shifted to simulate regions with more, or less, high quality habitat (yielding distributions with qualities ranging from 0.23 to 1.24 and from 0.76 to 1.77 respectively).
-We then implemented a full-factorial set of simulations on each quality distribution in which the pathogen's disease-induced mortality ($\nu$) and probability of direct infection ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).
+These two distributions were chosen to have equal variance and an overlapping core of patches around unit quality, with either the minimum or maximum shifted to simulate regions with more, or less, high quality habitat (yielding distributions with qualities ranging from 0.23 to 1.24 and from 0.76 to 1.77).
+We then implemented a full-factorial set of simulations on each quality distribution in which infectious survival ($\nu$) and probability of direct infection ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).
 In addition, pathogen longevity was varied over 3 values (ranging from 0.3 to 3 times the expected residence time of a healthy population on a unit quality patch, i.e., 10 units of time).
 
-For each simulation -- 100 replicates for each combination of disease (determined by $\nu$, $\delta$, and longevity), habitat quality distribution (high or low), and landscape structure (fully connected or lattice), qualities for 100 patches were drawn from a uniform distribution on the given habitat quality range (above).
+For each simulation -- 100 replicates for each combination of disease parameters ($\nu$, $\delta$, and longevity), habitat quality distribution (high or low), and landscape structure (fully connected or lattice), qualities for 100 patches were drawn from a uniform distribution on the given habitat quality range (above).
 In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infected.
@@
 In these simulations, we varied longevity over the values used above and explored the consequences of host preference for high quality habitat (and the corresponding potential for ecological traps) by choosing two different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating that migrants select all patches with equal probability).
-To isolate the effects of habitat preference and pathogen longevity, we fixed the direct transmission probability ($\delta = 0.3$), disease induced mortality ($\nu = 0.2$), and landscape structure (fully connected).
+To isolate the effects of habitat preference and pathogen longevity, we fixed the direct transmission probability ($\delta = 0.3$), infectious survival ($\nu = 0.2$), and landscape structure (fully connected).
 We performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by susceptible and infectious hosts, and the final total host population size.
@@
 
-A wide range of disease dynamics were observed over the range of pathogens and habitat quality distributions explored.  
-Epidemic outcome was influenced most strongly by disease induced mortality, with low disease-induced mortality (i.e.
-high $\nu$) facilitating pandemic dynamics (in which no susceptible host populations persist), and very high disease-induced mortality (i.e.
-low $\nu$) facilitating disease-free or endemic dynamics (Fig~\ref{endemic}).
+A wide range of disease dynamics were observed over the range of parameters and habitat quality distributions explored.  
+Epidemic outcome was influenced most strongly by infectious survival, with high infectious survival (i.e.
+low disease-induced mortality) facilitating widespread infection (in which no susceptible host populations persist), and very low infectious survival (i.e.
+high disease-induced mortality) facilitating disease-free dynamics or intermediate spread (Fig~\ref{endemic}).
 Pathogens with higher environmental longevities were generally able to invade and spread more easily, even with high disease-induced morality (Fig~\ref{endemic}(c, f)).  
 The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics.
-Increasing overall habitat quality facilitated pandemic spread for a larger range of pathogens (Fig~\ref{endemic}, a-c vs.
+Increasing overall habitat quality facilitated widespread infection for a larger range of parameters (Fig~\ref{endemic}, a-c vs.
 d-f).
-Metapopulations with a lattice structure exhibited qualitatively similar patterns, but with generally more limited pathogen spread and larger regions of endemic dynamics (Supplement).
+Metapopulations with a lattice structure exhibited qualitatively similar patterns, but with generally more limited pathogen spread and larger regions of susceptible maintenance (Supplement).
 
 For a wide range of parameter values, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions despite generally facilitating pathogen spread (red shaded regions in Fig~\ref{highvlow}).
-In the case of pathogens that were unable to invade due to high disease-induced mortality and weak transmission, high quality habitat distributions supported larger host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat (bottom left corner of the panels in Fig~\ref{highvlow}).
-Similarly, when low disease-induced mortality and high transmission (from either environmental or direct transmission) allowed a pathogen to spread widely (i.e.
+In the case of pathogens that were unable to invade due to low infectious survival and weak transmission, high quality habitat distributions supported larger host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat (bottom left corner of the panels in Fig~\ref{highvlow}).
+Similarly, when high infectious survival and strong transmission (from either environmental or direct transmission) allowed a pathogen to spread widely (i.e.
 at a pandemic level) in both habitat quality distributions, high quality distributions again supported larger host populations than low quality distributions (top portion of panels in Fig~\ref{highvlow}).
@@
 
-To better understand the patch-level mechanisms behind these effects, we further examined within patch infection dynamics for a pathogen in this intermediate range ($\nu = 0.2$ and $\delta = 0.3$).  
+To better understand the patch-level mechanisms behind these effects, we further examined within patch infection dynamics for parameters in this intermediate range ($\nu = 0.2$ and $\delta = 0.3$).  
 Because of their lower extinction rate and higher recolonization rate, high quality patches were more consistently occupied than low quality patches (Fig~\ref{simvis}, Supplement).  
@@
 And, as in these models, the net consequences of this trade-off for the host metapopulation depend on the characteristics of the disease.
-Broadly, we can divide the observed dynamics into three different scenarios: limited, intermediate, and pandemic pathogen spread (possibly leading to host extinction). 
+Broadly, we can divide the observed dynamics into three different scenarios: limited, intermediate, and widespread infection (possibly leading to host extinction). 
 High quality habitats are a net benefit to the host population when pathogen spread is either limited or is high enough that both high and low quality patches are affected by disease. 
@@
 \subsubsection*{Role of low quality habitat}
+
 In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy in the metapopulation (Fig~\ref{endemic}, \ref{popsizes}). 
@@
 Taken with this literature, our results highlight the importance of transient/temporary habitat in mitigating host exposure to environmentally transmitted pathogens.
+
+\subsubsection*{Consequences for management}
 
@@
 
-This partitioning of pathogen dynamics into three regions based on whether high quality habitat is a net benefit or detriment to the host metapopulation echoes the work of (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).
+This partitioning of pathogen dynamics into three regions based on whether high quality habitat is a net benefit or detriment to the host metapopulation echoes the work of \cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}.
 In particular, for a given pathogen, Hess (\cite{Hess1996}) similarly identified three regions of disease dynamics which determined whether increased movement had a positive or negative effect on host occupancy.
-In cases where the host movement rate was either low enough that the pathogen was unable to invade or high enough to cause a pandemic, Hess found that further increasing movement increased host occupancy, while at intermediate levels that resulted in endemic disease, increasing movement decreased host occupancy.
-These three scenarios, distinguished in \cite{Hess1996} by the host movement rate, map to the the three scenarios we outline here based on pathogen traits (disease-induced mortality and direct and environmental transmission).
-The conceptual similarity of these results suggests more generally that in metapopulations facing intermediate (i.e.
-endemic) pathogen spread, factors that would normally increase host stability and connectivity in the absence of disease (e.g.
+In cases where the host movement rate was either low enough that the pathogen was unable to invade or high enough to cause a widespread infection, Hess found that further increasing movement increased host occupancy, while at intermediate levels that resulted in moderate disease spread, increasing movement decreased host occupancy.
+These three scenarios, distinguished in \cite{Hess1996} by the host movement rate, map to the three scenarios we outline here based on disease parameters (infectious survival and direct and environmental transmission).
+The conceptual similarity of these results suggests more generally that in metapopulations facing intermediate pathogen spread, factors that would normally increase host stability and connectivity in the absence of disease (e.g.
 increased movement, high quality habitat) can actually decrease total host population size as a result of increased pathogen spread.
@@
 to c.).
-Environmentally long-lived pathogens are able to spread widely for a larger range of conditions, largely overwhelming the ability of low quality habitat to provide refuges for susceptible hosts. 
-These dynamics reflect the findings of \cite{Gog2002}, who found that under strong background infection from an alternative host, increasing movement always increased total occupancy, despite facilitating pathogen spread.
+Environmentally long-lived pathogens are able to spread widely under a larger range of conditions, largely overwhelming the ability of low quality habitat to provide refuges for susceptible hosts. 
+These dynamics reflect the findings of Gog \emph{et al.} (\cite{Gog2002}), who found that under strong background infection from an alternative host, increasing movement always increased total occupancy, despite facilitating pathogen spread.
 Park (\cite{Park2012}) expanded on this by adding environmental transmission to the mix and found that in cases where background and environmental transmission were strong relative to direct transmission, increasing movement was again a net benefit to the host metapopulation.
@@
 Similarly, from these results, we would not expect the spread of chronic wasting disease to turn high quality habitat into traps for mule deer metapopulations, due again to widespread infection (albeit at low prevalence) across population units (\cite{Conner2004}) and the relative importance of local-scale transmission within winter ranges (\cite{Farnsworth2006}).
-However, we would expect to see high quality traps in the plague-prairie dog system, where the infection is widespread and disease-induced mortality is high, but uninfected colonies are still able to persist (\cite{Stapp2004}).
-Indeed, studies have suggested that large colonies are more likely to become infected (\cite{Snall2008}), and thus high quality habitat capable of supporting large population sizes may help facilitate plague persistence and spread.
-Thus, by considering the characteristics of a host-pathogen system and their effect on the roles played by high and low quality habitat, we can begin to diagnose into which of these three scenarios empirical systems fall and the corresponding consequences for management.
+However, we might expect to see high quality traps in the plague-prairie dog system, where the infection is widespread and disease-induced mortality is high, but uninfected colonies are still able to persist (\cite{Stapp2004}).
+Indeed, studies have suggested that large colonies are more likely to become infected (\cite{Snall2008}), and thus high quality habitat capable of supporting large populations may help facilitate plague persistence and spread.
+Thus, by considering the characteristics of a host-pathogen system and their effect on the roles played by high and low quality habitat, we can begin to diagnose where empirical systems fall relative to these three scenarios and the corresponding consequences for management.
 As we have seen, failing to do so risks overlooking the importance of low-quality habitat for the persistence of wildlife populations and the importance of high-quality habitat for the persistence of wildlife disease.
@@
 Panel columns show low (0.3, a, d), medium (1.0, b, e), and high (3.2, c, f) pathogen longevities, while rows show low quality (ranging from 0.23 to 1.24, a - c) and high quality (ranging from 0.76 to 1.77, d - f) patch quality distributions.
-Darker shading corresponds to larger population sizes.}
+Darker shading corresponds to larger population sizes, while white indicates that no susceptible host populations persist (i.e.
+all host populations become infected).}
 \label{endemic}
@@
 \caption{Boxplots showing the range, over 100 replicate simulations, of susceptible (a), infectious (b), and total (c) host population sizes for high quality (red) and low quality (blue) habitat distributions.
-Disease-induced mortality ($\nu$) and direct transmission rate ($\delta$) are fixed at values of 0.2 and 0.3, respectively.
-Colored points show medians, while vertical lines indicate the inner-quartile ranges, with horizontal lines indicating the minimum and maximum.
+Infectious survival ($\nu$) and direct transmission rate ($\delta$) are fixed at values of 0.2 and 0.3, respectively.
+Coloured points show medians, while vertical lines indicate the inner-quartile ranges, with horizontal lines indicating the minimum and maximum.
 At low longevities, this pathogen is unable to invade and high quality habitat supports larger populations of susceptible hosts.
@@
 $\mu$ & Extinction rate of unit quality patch & 0.1 \\
-$\nu$ & Disease-induced reduction in population size & 0.1 - 1 \\
+$\nu$ & Infectious survival & 0.1 - 1 \\
 $\alpha$ & Strength of environmental stochasticity & 1 \\

### 2015-07-16 Manuscript/supplementary.tex
Updates to supp figures

@@
 \includegraphics[scale=0.5]{figure/supplement_1.pdf}
-\caption{The proportion of simulations resulting in endemic disease dynamics (persistence of both susceptible and infectious hosts) as a function of disease-induced mortality ($\nu$), and the probability of direct transmission ($\delta$).
-Panel columns show low (0.3, a, d), medium (1.0, b, e), and high (3.2, c, f) pathogen longevities, while rows show high quality (ranging from 0.76 to 1.77, a - c) and low quality (ranging from 0.23 to 1.24, d - f) patch quality distributions.
-Darker shading corresponds to more frequent endemic dynamics.
-The white region above the endemic region generally corresponds to pandemic dynamics (no susceptible host populations persist), while the region below the endemic region corresponds to disease-free dynamics.}
+\caption{Median total susceptible population size as a function of infectious survival ($\nu$), and the probability of direct transmission ($\delta$).
+Panel columns show low (0.3, a, d), medium (1.0, b, e), and high (3.2, c, f) pathogen longevities, while rows show low quality (ranging from 0.23 to 1.24, a - c) and high quality (ranging from 0.76 to 1.77, d - f) patch quality distributions.
+Darker shading corresponds to larger population sizes, while white indicates that no susceptible host populations persist (i.e.
+all host populations become infected).}
 \label{endemic}

### 2015-07-16 Analysis/pubplots.R

@@ -170,32 +170,35 @@ pdf("Manuscript/figure/supplement_1.pdf", width = 10, height = 6)
 
 load(paste(getwd(), "/Output/pathgrid(lattice).RData", sep = ""))
 
-out$endemic <- out$S > 0 & out$I > 0
-dat <- ddply(out, c("longevity", "treatment", "delta", "nu"), summarise, sum = sum(endemic) / 100)
+dat <- ddply(out, c("longevity", "treatment", "delta", "nu"), summarise, susc = median(S.pop))
 
-cols <- colorRampPalette(brewer.pal(9, "Greys"))(100)
+cols <- colorRampPalette(brewer.pal(9, "Greys"))(120)
 labels <- c("d", "e", "f", "a", "b", "c")
 
-plot <- levelplot(sum ~ delta * nu | longevity * treatment, data = dat,
+plot <- levelplot(susc ~ delta * nu | longevity * treatment, data = dat,
                   col.regions = cols,
-                  at = seq(0, 1, by = 0.01),
+                  at = seq(0, 120, by = 1),
                   xlab = list(label = expression("Probability of direct transmission"~(delta)), cex = 1),
                   ylab = list(label = expression("Infectious survival"~(nu)), cex = 1),
                   strip = F,
                   scales = list(alternating = F, cex = 1),
-                  ylab.right = list(label = "Probability of endemic", cex = 1),
+                  ylab.right = list(label = "Susceptible population size", cex = 1),
                   par.settings = list(layout.widths = list(axis.key.padding = 0, ylab.right = 3)),
                   panel=function(...){
                     panel.levelplot(...)
                     panel.text(0, 1, labels[panel.number()], cex = 1)
                   })
 
-col.lab <- textGrob(expression(Longevity ~ symbol("\256")))
 row.low <- textGrob("Low quality", rot = 90)
 row.high <- textGrob("High quality", rot = 90)
 
 grid.arrange(arrangeGrob(row.low, row.high, ncol = 1), plot, ncol = 2, 
-             widths = c(0.05, 0.95), sub = col.lab)
+             widths = c(0.05, 0.95))
+
+grid.text("Longevity = 0.32", x = unit(0.26, units = "npc"), y = unit(0.96, units = "npc"), draw = T)
[…]

### 2015-07-21 Manuscript/manuscript.tex
Making vocabulary more consistent and clear

@@
 To address the above questions, we developed a stochastic patch occupancy model (SPOM) in which each patch can be in one of three possible states: occupied by susceptible hosts (S), occupied by infectious hosts (I), and unoccupied by the host ($\emptyset$).
-State transitions are governed by host colonization, extinction, and infection rates, where a susceptible population can become infectious either through direct contact with infectious immigrants, or through a local environmental pathogen reservoir (Table~\ref{transitions}).
+State transitions are governed by host colonization, extinction, and transmission rates, where a susceptible population can become infectious either through direct contact with infectious immigrants, or through a local environmental pathogen reservoir (Table~\ref{transitions}).
 
-We assume that the host population size supported by a patch is proportional to its quality ($A_i$), and that disease-induced mortality reduces the size of infectious populations by a factor of $\nu$.
-Building on the framework developed by \cite{Hanski2003} then, the rates at which patch \emph{i}, with quality $A_i$, is colonized by susceptible and infectious individuals, respectively, is given by its connectivity:
+We assume that the host population size supported by a patch is proportional to its quality ($A_i$), and that a proportion, $\nu$, of hosts survive when a population is infected (i.e.
+the size of infectious populations is reduced by a factor of $\nu$).
+Building on the framework developed by \cite{Hanski2003} then, the rates at which patch \emph{i}, with quality (or equivalently population size) $A_i$, is colonized by susceptible and infectious individuals, respectively, is given by its connectivity:
 \begin{align}
@@
 \end{align}
-
-where $\phi_j(X, t)$ is an indicator function that is $1$ if patch $j$ is in state $X$ at time $t$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality; $D$ is the inverse of the host's mean dispersal distance; and $d_{ij}$ is the effective distance between patches $i$ and $j$ .
+where $\phi_j(X, t)$ is an indicator function that is $1$ if patch $j$ is in state $X$ at time $t$ and is $0$ otherwise; $\xi_{im}$ and $\xi_{em}$ control how rates of immigration and emigration, respectively, scale with patch quality/population size; $D$ is the inverse of the host's mean dispersal distance; and $d_{ij}$ is the effective distance between patches $i$ and $j$ .
 Essentially, $C_{Xi}$ sums the colonization effort (or propagule pressure) from all patches in state $X$ to focal patch $i$.
@@
 \end{equation}
-where $\gamma_0$ is the initial infection rate of the pathogen reservoir, $\tau_{i}$ is the time since last infectious occupancy ($\tau_{i} = \infty$, and thus $\gamma(\tau_i) = 0$, if the patch has never been occupied by infectious hosts), and $r$ is the pathogen's decay rate in the environment (determined by the pathogen's environmental longevity).
+where $\gamma_0$ is the initial transmission rate of the pathogen reservoir, $\tau_{i}$ is the time since last infectious occupancy ($\tau_{i} = \infty$, and thus $\gamma(\tau_i) = 0$, if the patch has never been occupied by infectious hosts), and $r$ is the pathogen's decay rate in the environment (determined by the pathogen's longevity, defined as its half-life in the environment).
 
@@
 The model was parameterized according to the following assumptions (Table~\ref{params}). 
-(1) The rate of emigration from a patch increases with population size, while the extinction rate of a patch decreases with populations size ($\xi_{im}=0.5$, $\alpha=1$, \cite{Hanski2003}).  
+(1) The rate of emigration from a patch increases with the local population size, while the extinction rate of a patch decreases with populations size ($\xi_{im}=0.5$, $\alpha=1$, \cite{Hanski2003}).
 (2) Migrants preferentially select high quality habitat, that is, the rate at which a patch is colonized scales with quality ($\xi_{em}=0.5$, \cite{Hanski2003}, though we relax this below).  
 (3)  Connectivity among patches is determined entirely by quality ($d_{ij}=1$ for all $i \neq j$).  
-In addition to these assumptions, we chose parameters so that without infection, approximately 0.75 of the metapopulation was occupied ($\mu = 0.1$, $D=5$).
-We varied disease induced mortality ($\nu$), probability of direct infection ($\delta$), and environmental longevity to explore the dynamics of a range of possible pathogens.
+In addition to these assumptions, we chose parameters so that without infection, approximately 0.75 of the patches were occupied ($\mu = 0.1$, $D=5$).
+We varied infectious survival ($\nu$), probability of direct transmission ($\delta$), and environmental longevity to explore the dynamics of a range of possible pathogens.
 
@@
 These two distributions were chosen to have equal variance and an overlapping core of patches around unit quality, with either the minimum or maximum shifted to simulate regions with more, or less, high quality habitat (yielding distributions with qualities ranging from 0.23 to 1.24 and from 0.76 to 1.77).
-We then implemented a full-factorial set of simulations on each quality distribution in which infectious survival ($\nu$) and probability of direct infection ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).
+We then implemented a full-factorial set of simulations on each quality distribution in which infectious survival ($\nu$) and probability of direct transmission ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).
 In addition, pathogen longevity was varied over 3 values (ranging from 0.3 to 3 times the expected residence time of a healthy population on a unit quality patch, i.e., 10 units of time).
@@
 The state of the metapopulation was then tracked for 5000 time steps.
-We recorded the proportion of patches in each state at the end of each simulation and also computed the mean effective population size of susceptible and infectious hosts over the last 500 time steps of the simulation (assuming an effective population size of $A_i$ for patches that are occupied and susceptible and $\nu A_i$ for occupied infectious patches).
+We recorded the proportion of patches in each state at the end of each simulation and also computed the mean total population size of susceptible and infectious hosts over the last 500 time steps of the simulation (summing the local population sizes for all occupied patches, assuming a local population size of $A_i$ for patches that are occupied and susceptible and $\nu A_i$ for occupied infectious patches).
 
@@
 high disease-induced mortality) facilitating disease-free dynamics or intermediate spread (Fig~\ref{endemic}).
-Pathogens with higher environmental longevities were generally able to invade and spread more easily, even with high disease-induced morality (Fig~\ref{endemic}(c, f)).  
+Pathogens with higher environmental longevities were generally able to invade and spread more easily, even with low infectious survival (Fig~\ref{endemic}(c, f)).  
 The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics.
@@
 For a wide range of parameter values, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions despite generally facilitating pathogen spread (red shaded regions in Fig~\ref{highvlow}).
-In the case of pathogens that were unable to invade due to low infectious survival and weak transmission, high quality habitat distributions supported larger host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat (bottom left corner of the panels in Fig~\ref{highvlow}).
+In the case of pathogens that were unable to invade due to low infectious survival and weak transmission, high quality habitat distributions supported larger total host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat (bottom left corner of the panels in Fig~\ref{highvlow}).
 Similarly, when high infectious survival and strong transmission (from either environmental or direct transmission) allowed a pathogen to spread widely (i.e.
-at a pandemic level) in both habitat quality distributions, high quality distributions again supported larger host populations than low quality distributions (top portion of panels in Fig~\ref{highvlow}).
+at a pandemic level) in both habitat quality distributions, high quality distributions again supported larger total host populations than low quality distributions (top portion of panels in Fig~\ref{highvlow}).
 
 However, for pathogens falling between these extremes (e.g.
-with intermediate disease-induced mortality), low quality habitat was a net benefit to the host, supporting larger total population sizes than high quality habitat (blue regions of Fig ~\ref{highvlow}).
+with intermediate infectious survival), low quality habitat was a net benefit to the host, supporting larger total population sizes than high quality habitat (blue regions of Fig ~\ref{highvlow}).
 The spread of these pathogens was strongly influenced by the habitat quality distribution, with low quality habitat limiting disease spread and better maintaining susceptible occupancy relative to high quality habitat (Fig~\ref{popsizes}, Supplemental).
-In these cases, high quality habitat served as a sort of metapopulation-level trap, that is, otherwise beneficial habitat that is detrimental to the host population in the presence of disease.
+In these cases, high quality habitat served as a sort of metapopulation-level trap, that is, otherwise beneficial habitat that is detrimental to the overall host population in the presence of disease.
 As pathogen longevity increased, the range of the parameter space where we observed this metapopulation-level trap became smaller, apparent only for the most virulent pathogens.
@@
 Broadly, we can divide the observed dynamics into three different scenarios: limited, intermediate, and widespread infection (possibly leading to host extinction). 
-High quality habitats are a net benefit to the host population when pathogen spread is either limited or is high enough that both high and low quality patches are affected by disease. 
+High quality habitats are a net benefit to the total host population when pathogen spread is either limited or is high enough that both high and low quality patches are affected by disease. 
 In these cases, the connectivity and stability of high quality patches facilitate larger total population sizes with little additional pathogen spread.
 
-There is an intermediate scenario, however, where low quality habitats are more likely to remain uninfected and as a result boost the metapopulation size while high quality patches that are heavily infected tend to reduce the metapopulation size.
+There is an intermediate scenario, however, where low quality habitats are more likely to remain uninfected and as a result boost the total population size while high quality patches that are heavily infected tend to reduce the total population size.
 In this scenario, the ability of pathogens to invade and spread is influenced substantially by the colonization and extinction rates of high and low quality habitat, and the contrasting roles that these habitat types play in the metapopulation.
@@
 
-Because high quality habitat supports larger host population sizes, these populations are less susceptible to environmental stochasticity and thus experience less frequent extinction events.
+Because high quality habitat supports larger local host population sizes, these populations are less susceptible to environmental stochasticity and thus experience less frequent extinction events.
 Due to this additional stability, high quality habitat is better able to both maintain a consistent pool of susceptible hosts to which infection can spread, and support infectious host populations under disease-induced mortality.
@@
 This phenomenon represents the other side of the role played by high quality patches in that low quality patches are relatively unstable (i.e.
-low population sizes and high extinction rates) and thus do not support infectious populations for long.
+small local population sizes and high extinction rates) and thus do not support infectious populations for long.
 Since they are infrequently colonized (when high quality habitat is preferentially selected, $\xi_{em} > 0$), the environmental reservoir left behind by these infectious populations likely decays before it has the opportunity to infect new susceptible colonists.
@@
 \centering
-\caption{Results from a single representative simulation with habitat quality uniformly distributed between 0.23 and 1.77, and a pathogen with a disease induced mortality of $\nu = 0.2$, a direct transmission rate of $\delta = 0.3$, and an envioronmental longevity equal to the expected residence time of a susceptible population on a unit quality patch (10 units of time here).
+\caption{Results from a single representative simulation with habitat quality uniformly distributed between 0.23 and 1.77, and a pathogen with an infectious survival of $\nu = 0.2$, a direct transmission rate of $\delta = 0.3$, and an envioronmental longevity equal to the expected residence time of a susceptible population on a unit quality patch (10 units of time here).
 The top panel shows total proportion of patches occupied by susceptible (blue) and infectious (red) hosts over time.
@@
 $S$ denotes occupied by  susceptible hosts, $I$ denotes occupied infectious hosts, and $\emptyset$ denotes unoccupied by the host.
-All patches are characterized by an environmental infection rate, $\gamma(\tau_i)$, where $\tau_i$ is the time since the patch was last occupied by infectious individuals.}
+All patches are characterized by an environmental transmission rate, $\gamma(\tau_i)$, where $\tau_i$ is the time since the patch was last occupied by infectious individuals.}
 \begin{tabular}{llr}
@@
 \hline
-$S \rightarrow I$ & Infection (from contact or reservoir) & $\delta C_{Ii} + \gamma(\tau_i)$\\
+$S \rightarrow I$ & Transmission (from contact or reservoir) & $\delta C_{Ii} + \gamma(\tau_i)$\\
 $S \rightarrow \emptyset $ & Extinction (of susceptible) & $E_{Si}$\\
@@
 \hline
-$\delta$ & Probability of direct infection & 0 - 0.9 \\
-$\gamma_0$ & Initial rate of infection from reservoir patch & 0.5
+$\delta$ & Probability of direct transmission & 0 - 0.9 \\
+$\gamma_0$ & Initial rate of transmission from reservoir patch & 0.5
 \end{tabular}

### 2015-07-21 Manuscript/supplementary.tex
Making vocabulary more consistent and clear

@@
 \centering
-\caption{Differences between the median total effective population size of hosts on high quality patch distributions (quality between 0.76 and 1.77) and low quality patch distributions (quality between 0.23 to 1.24) as a function of disease-induced mortality ($\nu$) and the probability of direct transmission ($\delta$).
+\caption{Differences between the median total effective population size of hosts on high quality patch distributions (quality between 0.76 and 1.77) and low quality patch distributions (quality between 0.23 to 1.24) as a function of infectious survival ($\nu$) and the probability of direct transmission ($\delta$).
 Reds indicate higher median host population sizes in high quality patch distributions, while blues indicate higher median host population sizes in low quality patch distributions.
@@
 Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.
-Disease-induced mortality and probability of direct infection both fixed at $\nu = 0.2$ and $\delta = 0.3$, with fully connected landscape structure.}
+Infectious survival and probability of direct infection both fixed at $\nu = 0.2$ and $\delta = 0.3$, with fully connected landscape structure.}
 \end{figure}

### 2015-07-21 Analysis/pubplots.R
Switching to eps figures

@@ -10,7 +10,7 @@ library(plyr)
 #===============================================================================
 ### Figure 1
 
-pdf("Manuscript/figure/figure_1.pdf", width = 10, height = 6)
+postscript("Manuscript/figure/figure_1.eps", width = 10, height = 6)
 
 load(paste(getwd(), "/Output/pathgrid.RData", sep = ""))
 
@@ -50,7 +50,7 @@ dev.off()
 
 rm(list = ls())
 
-pdf("Manuscript/figure/figure_2.pdf", width = 10, height = 4)
+postscript("Manuscript/figure/figure_2.eps", width = 10, height = 4)
 
 load(paste(getwd(), "/Output/pathgrid.RData", sep = ""))
 
@@ -94,7 +94,7 @@ dev.off()
 
 rm(list = ls())
 
-pdf("Manuscript/figure/figure_3.pdf", width = 10, height = 5)
+postscript("Manuscript/figure/figure_3.eps", width = 10, height = 5)
 
 load(paste(getwd(), "/Output/pathgrid.RData", sep = ""))
 sub <- out[abs(out$delta - 0.3) < 0.001 & abs(out$nu - 0.2) < 0.001, ]
@@ -166,7 +166,7 @@ dev.off()
 
 rm(list = ls())
 
-pdf("Manuscript/figure/supplement_1.pdf", width = 10, height = 6)
+postscript("Manuscript/figure/supplement_1.eps", width = 10, height = 6)
 
 load(paste(getwd(), "/Output/pathgrid(lattice).RData", sep = ""))
 
@@ -207,7 +207,7 @@ dev.off()
 
 rm(list = ls())
 
-pdf("Manuscript/figure/supplement_2.pdf", width = 10, height = 4)
+postscript("Manuscript/figure/supplement_2.eps", width = 10, height = 4)
 
 load(paste(getwd(), "/Output/pathgrid(lattice).RData", sep = ""))
 
@@ -252,7 +252,7 @@ dev.off()
 
 rm(list = ls())
 
-pdf("Manuscript/figure/supplement_3.pdf", width = 8, height = 4)
+postscript("Manuscript/figure/supplement_3.eps", width = 8, height = 4)
 
 load(paste(getwd(), "/Output/trap.RData", sep = ""))
 
@@ -283,7 +283,7 @@ dev.off()
 
 rm(list = ls())
 
-pdf("Manuscript/figure/supplement_4.pdf", width = 8, height = 4)
[…]

### 2015-07-21 Analysis/simvis.r

@@ -79,9 +79,9 @@ colnames(state) <- c(1:100)
 
 state <- melt(state)
 names(state) <- c("Time", "Patch", "color")
-levels(state$color) <- c("white", "red", "white", "blue")
+levels(state$color) <- c("white", "#ee2c2c", "white", "#1874cd")
 
-p1 <- ggplot(state, aes(x = Time, y = Patch)) + geom_tile(fill = state$color, alpha = 0.5) + 
+p1 <- ggplot(state, aes(x = Time, y = Patch)) + geom_tile(fill = state$color) + 
               scale_fill_identity() + 
               scale_x_continuous(expand = c(0, 0)) +
               scale_y_continuous(expression(Patch ~ Quality ~ symbol('\256')), expand = c(0, 0)) +
@@ -90,14 +90,14 @@ p1 <- ggplot(state, aes(x = Time, y = Patch)) + geom_tile(fill = state$color, al
 
 tseries <- data.frame(Time = rep(c(1:5001), 2), Occupancy = c(S, I), class = rep(c("S", "I"), each = 5001))
 
-p2 <- ggplot(tseries, aes(x = Time, y = Occupancy, group = class, color = class, alpha = 0.5)) + geom_line() + 
-              scale_color_manual(values = c("red", "blue")) +
+p2 <- ggplot(tseries, aes(x = Time, y = Occupancy, group = class, color = class)) + geom_line() + 
+              scale_color_manual(values = c("#ee2c2c", "#1874cd")) +
               guides(color = FALSE, alpha = FALSE) +
               theme_classic() + labs(x = NULL) + 
               scale_x_continuous(labels = NULL, expand = c(0, 0)) +
               scale_y_continuous(expand = c(0, 0))
 
-pdf("Manuscript/figure/figure_4.pdf", width = 12, height = 8)
+postscript("Manuscript/figure/figure_4.eps", width = 12, height = 8)
 
 grid.arrange(p2, p1, ncol = 1, heights = c(0.25, 0.75))
 

### 2015-07-23 Manuscript/manuscript.tex
Fonts and formatting

@@
 \usepackage{authblk}
+\usepackage[T1]{fontenc}
+\usepackage{pslatex}
 
@@
 
-Continuous time stochastic simulations of the above model were implemented in the R language (\cite{R2014}) using the Gillespie algorithm (\cite{Gillespie1977}).
-Code is available at \url{https://github.com/clint-leach/Metapop-Disease}.
+Continuous time stochastic simulations of the above model were implemented in the R language (\cite{R2014}) using the Gillespie algorithm (\cite{Gillespie1977}).  Code is available at \\
+\url{https://github.com/clint-leach/Metapop-Disease}.
 
@@
 \subsection*{Data accessibility}
-Simulation source code and generated data are available at \url{https://github.com/clint-leach/Metapop-Disease}, along with scripts used to generate figures.
+Simulation source code and generated data are available at \\  \url{https://github.com/clint-leach/Metapop-Disease}.
 
@@
 \begin{figure}
-\includegraphics[scale=0.5]{figure/figure_1.pdf}
+\includegraphics[angle=270, scale=0.5]{figure/figure_1.eps}
 \caption{Median total susceptible population size as a function of infectious survival ($\nu$), and the probability of direct transmission ($\delta$).
@@
 \begin{figure}
-\includegraphics[scale=0.5]{figure/figure_2.pdf}
+\includegraphics[angle=270, scale=0.5]{figure/figure_2.eps}
 \centering
@@
 \begin{figure}
-\includegraphics[scale=0.5]{figure/figure_3.pdf}
+\includegraphics[angle=270, scale=0.5]{figure/figure_3.eps}
 \centering
@@
 \begin{figure}
-\includegraphics[scale=0.4]{figure/figure_4.pdf}
+\includegraphics[angle=270, scale=0.5]{figure/figure_4.eps}
 \centering

### 2015-07-23 Analysis/patch.analysis.r
Model output and documentation

@@ -1,4 +1,7 @@
-# Patch level results -- scripts to analyze patch-level data 
+# Exploratory plots and anlyses of patch-level results generated by trap.shell.r
+# and stored in Output/trap.RData
+
+#===============================================================================
 
 load(paste(getwd(), "/Output/trap.RData", sep = ""))
 

### 2015-07-23 Analysis/pathgrid.analysis.r

@@ -1,3 +1,5 @@
+# Exploratory plots and anlysis of metapopulation-level resutls generated by 
+# pathgrid.shell.r and stored in Output/pathgrid.RData and Output/pathgrid(lattice).RData
 
 library(gridExtra)
 library(lattice)

### 2015-07-23 Analysis/pathgrid.shell.r
Distinct names for modelrun functions

@@ -82,6 +82,6 @@ getDoParWorkers()
 #===============================================================================
 #Looping through each parameter combo and calling modelrun
 out <- foreach(i = 1:dim(treatment)[1], .verbose=TRUE, .combine="rbind") %dopar% 
-  modelrun(treatment[i, 2], treatment[i, 3], treatment[i, 4], parms, distance, initial, timesteps, treatment[i, 1])
+  metapop.collect(treatment[i, 2], treatment[i, 3], treatment[i, 4], parms, distance, initial, timesteps, treatment[i, 1])
 
 stopCluster(w)

### 2015-07-23 Analysis/trap.shell.r

@@ -92,7 +92,7 @@ getDoParWorkers()
 out <- foreach(i = 1:length(par.reps[, 1]), .verbose=TRUE, .combine="rbind") %dopar% {
   parms["xi_im"] <- par.reps[i, 2]
   parms["xi_em"] <- par.reps[i, 3]
-  modelrun(par.reps[i, 1], range,par.reps[i, 4], parms, distance, initial, timesteps)
+  patch.collect(par.reps[i, 1], range,par.reps[i, 4], parms, distance, initial, timesteps)
   }
 
 stopCluster(w)

### 2015-07-23 Code/highlow.runs.r

@@ -1,5 +1,5 @@
 
-modelrun <- function(longevity, delta, nu, parms, distance, initial, timesteps, treatment){
+metapop.collect <- function(longevity, delta, nu, parms, distance, initial, timesteps, treatment){
   # Function that runs SPOM model for given pathogen parameters and habitat quality
   # distribution and collects metapopulation-level simulation results.
   #

### 2015-07-23 Code/modelrun3.r

@@ -1,5 +1,5 @@
 
-modelrun <- function(longevity, range, k, parms, distance, initial, timesteps){
+patch.collect <- function(longevity, range, k, parms, distance, initial, timesteps){
   # Function that runs the SPOM model for fixed pathogen parameters and 
   # habitat quality distribution and collects patch-level simulation results.
   #

### 2015-11-11 Manuscript/manuscript.tex
Clarification of longevity parameterization

@@
 \begin{equation}
-\gamma(\tau_i)=\gamma_0exp(-r\tau_{i}),
+\gamma(\tau_i)=\gamma_0\exp \left(-r \tau_{i}\right),
 \end{equation}
-where $\gamma_0$ is the initial transmission rate of the pathogen reservoir, $\tau_{i}$ is the time since last infectious occupancy ($\tau_{i} = \infty$, and thus $\gamma(\tau_i) = 0$, if the patch has never been occupied by infectious hosts), and $r$ is the pathogen's decay rate in the environment (determined by the pathogen's longevity, defined as its half-life in the environment).
+where $\gamma_0$ is the initial transmission rate of the pathogen reservoir, $\tau_{i}$ is the time since last infectious occupancy ($\tau_{i} = \infty$, and thus $\gamma(\tau_i) = 0$, if the patch has never been occupied by infectious hosts), and $r$ is the pathogen's decay rate in the environment, determined by the pathogen's longevity, $\beta$ (defined as its half-life in the environment relative to the expected residence time of a population on a unit quality patch):
+\begin{equation}
+r = \frac{\mu \log(2)}{\beta}.
+\end{equation}
 
@@
 The model was parameterized according to the following assumptions (Table~\ref{params}). 
-(1) The rate of emigration from a patch increases with the local population size, while the extinction rate of a patch decreases with populations size ($\xi_{im}=0.5$, $\alpha=1$, \cite{Hanski2003}).
-(2) Migrants preferentially select high quality habitat, that is, the rate at which a patch is colonized scales with quality ($\xi_{em}=0.5$, \cite{Hanski2003}, though we relax this below).  
+(1) The rate of emigration from a patch increases with the local population size, while the extinction rate of a patch decreases with populations size ($\xi_{em}=0.5$, $\alpha=1$, \cite{Hanski2003}).
+(2) Migrants preferentially select high quality habitat, that is, the rate at which a patch is colonized scales with quality ($\xi_{im}=0.5$, \cite{Hanski2003}, though we relax this below).  
 (3)  Connectivity among patches is determined entirely by quality ($d_{ij}=1$ for all $i \neq j$).  
@@
 We then implemented a full-factorial set of simulations on each quality distribution in which infectious survival ($\nu$) and probability of direct transmission ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).
-In addition, pathogen longevity was varied over 3 values (ranging from 0.3 to 3 times the expected residence time of a healthy population on a unit quality patch, i.e., 10 units of time).
+In addition, pathogen longevity ($\beta$) was varied over 3 values (ranging from 0.32 to 3.2).
 
-For each simulation -- 100 replicates for each combination of disease parameters ($\nu$, $\delta$, and longevity), habitat quality distribution (high or low), and landscape structure (fully connected or lattice), qualities for 100 patches were drawn from a uniform distribution on the given habitat quality range (above).
+For each simulation -- 100 replicates for each combination of disease parameters ($\nu$, $\delta$, and $\beta$), habitat quality distribution (high or low), and landscape structure (fully connected or lattice), qualities for 100 patches were drawn from a uniform distribution on the given habitat quality range (above).
 In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infected.
@@
 $\delta$ & Probability of direct transmission & 0 - 0.9 \\
-$\gamma_0$ & Initial rate of transmission from reservoir patch & 0.5
+$\gamma_0$ & Initial rate of transmission from reservoir patch & 0.5\\
+$\beta$ & Pathogen longevity (half-life in environment, relative to $1/\mu$) & 0.3 - 3
  \end{tabular}

### 2015-11-11 Analysis/pubplots.R

@@ -98,6 +98,7 @@ postscript("Manuscript/figure/figure_3.eps", width = 10, height = 5)
 
 load(paste(getwd(), "/Output/pathgrid.RData", sep = ""))
 sub <- out[abs(out$delta - 0.3) < 0.001 & abs(out$nu - 0.2) < 0.001, ]
+sub$longevity <- round(sub$longevity, 1)
 
 S <- ddply(sub, c("longevity", "treatment"), summarise, 
            med = median(S.pop), 
@@ -120,36 +121,36 @@ tot <- ddply(sub, c("longevity", "treatment"), summarise,
            min = min(S.pop + I.pop),
            max = max(S.pop + I.pop))
 
-pS <- ggplot(S, aes(x = factor(log10(longevity)), y = med))
+pS <- ggplot(S, aes(x = factor(longevity), y = med))
 pS <- pS + geom_point(aes(colour = factor(treatment)), position = position_dodge(width = 0.4)) +
              geom_point(aes(colour = factor(treatment), y = min), shape = 45, size = 5, position = position_dodge(width = 0.4)) +
              geom_point(aes(colour = factor(treatment), y = max), shape = 45, size = 5, position = position_dodge(width = 0.4)) +
              geom_errorbar(aes(colour = factor(treatment), ymin = lowq, ymax = highq), 
                            position = position_dodge(width = 0.4), width = 0)
 pS <- pS + scale_y_continuous(limits = c(0, 125), expand = c(0, 0.1)) +
-             ylab("S population") + xlab("log(longevity)") + ggtitle("a") +
+             ylab("S population") + xlab("longevity") + ggtitle("a") +
              theme_classic() + theme(legend.position = "none")
 
 
-pI <- ggplot(I, aes(x = factor(log10(longevity)), y = med))
+pI <- ggplot(I, aes(x = factor(longevity), y = med))
 pI <- pI + geom_point(aes(colour = factor(treatment)), position = position_dodge(width = 0.4)) +
            geom_point(aes(colour = factor(treatment), y = min), shape = 45, size = 5, position = position_dodge(width = 0.4)) +
            geom_point(aes(colour = factor(treatment), y = max), shape = 45, size = 5, position = position_dodge(width = 0.4)) +
            geom_errorbar(aes(colour = factor(treatment), ymin = lowq, ymax = highq), 
                          position = position_dodge(width = 0.4), width = 0)
 pI <- pI + scale_y_continuous(limits = c(0, 12), expand = c(0, 0.1)) +
[…]

### 2015-12-04 Manuscript/manuscript.tex
Addition of caveats and future work

@@
 
-From a management perspective, it is critical to understand this shift in the effect of habitat quality in order to avoid falling into the metapopulation trap created by high quality habitat and its interaction with pathogen spread.
+These results have general management implications that highlight the potential importance of low-quality habitat for the persistence of wildlife populations and the importance of high-quality habitat for the persistence of wildlife disease.
+It is critical to understand these implications in order to avoid falling into the metapopulation trap created by high quality habitat and its interaction with pathogen spread.
 Previous theoretical work suggests that management should focus on maintaining patches where conditions are most favourable for the host (i.e.
@@
 Our model adds to this literature by demonstrating the potential for environmentally persistent pathogens to provide the kind of disturbance under which low quality habitat is beneficial, and thus should be conserved.
+
+The framework used here provides general insights, but generating predictions for specific systems requires a careful consideration of the model's simplifying assumptions.
+In particular, the stochastic patch occupancy model, though well suited for capturing realistic spatial structure, assumes that within-patch dynamics (i.e.
+growth to carrying capacity and pathogen transmission) are rapid relative to among-patch dynamics (i.e.
+colonization and extinction).
+This is often a reasonable assumption, especially for relatively small populations in a highly fragmented landscape (\cite{Hanski2003}), but may not capture the non-linearities and nuances of host-pathogen dynamics in a single population.
+In addition, we assume that patch quality only influences population size and the resulting colonization and extinction processes, but quality could also influence pathogen transmission itself through effects on host density or condition (\cite{Becker2015}).
+Patch quality may also influence the size or infectiousness of the environmental reservoir, either directly, by influencing the decay rate of the pathogen in the environment, or indirectly, through the effects on population size and overall shedding rate.
+The potential importance of these additional mechanisms remains an open question for future work and need to be considered before applying this framework to any given system of interest.
+Fortunately, the SPOM framework is fairly flexible and future work could extend the model presented here to include additional processes, e.g.
+by making transmission parameters also function of patch quality.
+
+Though detailed predictions for specific systems will likely require additional model development, the three scenarios identified here nonetheless provide a coarse framework that we can use to identify the potential risk of metapopulation-level traps in wildlife populations currently affected by an environmentally persistent pathogen.
+For instance, high quality habitat seems unlikely to form metapopulation traps for bat populations affected by white nose syndrome due to the widespread infection both within and between hibernaculum and the scarcity of host refuges (\cite{Langwig2014}, \cite{ORegan2015}).
+Similarly, from these results, we would not expect the spread of chronic wasting disease to turn high quality habitat into traps for mule deer metapopulations, due again to widespread infection (albeit at low prevalence) across population units (\cite{Conner2004}) and the relative importance of local-scale transmission within winter ranges (\cite{Farnsworth2006}).
+However, we might expect to see high quality traps in the plague-prairie dog system, where the infection is widespread and disease-induced mortality is high, but uninfected colonies are still able to persist (\cite{Stapp2004}).
+Indeed, studies have suggested that large colonies are more likely to become infected (\cite{Snall2008}), and thus high quality habitat capable of supporting large populations may help facilitate plague persistence and spread.
+Thus, by considering the characteristics of a host-pathogen system and their effect on the roles played by high and low quality habitat, we can begin to diagnose where empirical systems fall relative to these three scenarios and the corresponding consequences for management.
 
@@
 decreased movement, low quality habitat).
-
-The three scenarios identified here provide a rough framework that we can use to identify the potential risk of metapopulation-level traps in wildlife populations currently affected by an environmentally persistent pathogen.
-For instance, high quality habitat seems unlikely to form metapopulation traps for bat populations affected by white nose syndrome due to the widespread infection both within and between hibernaculum and the scarcity of host refuges (\cite{Langwig2014}, \cite{ORegan2015}).
-Similarly, from these results, we would not expect the spread of chronic wasting disease to turn high quality habitat into traps for mule deer metapopulations, due again to widespread infection (albeit at low prevalence) across population units (\cite{Conner2004}) and the relative importance of local-scale transmission within winter ranges (\cite{Farnsworth2006}).
-However, we might expect to see high quality traps in the plague-prairie dog system, where the infection is widespread and disease-induced mortality is high, but uninfected colonies are still able to persist (\cite{Stapp2004}).
-Indeed, studies have suggested that large colonies are more likely to become infected (\cite{Snall2008}), and thus high quality habitat capable of supporting large populations may help facilitate plague persistence and spread.
-Thus, by considering the characteristics of a host-pathogen system and their effect on the roles played by high and low quality habitat, we can begin to diagnose where empirical systems fall relative to these three scenarios and the corresponding consequences for management.
-As we have seen, failing to do so risks overlooking the importance of low-quality habitat for the persistence of wildlife populations and the importance of high-quality habitat for the persistence of wildlife disease.
 

### 2015-12-04 Manuscript/manuscript.tex
SPOM assumptions and justification in methods

@@
 
-To address the above questions, we developed a stochastic patch occupancy model (SPOM) in which each patch can be in one of three possible states: occupied by susceptible hosts (S), occupied by infectious hosts (I), and unoccupied by the host ($\emptyset$).
+To address the above questions, we developed a model based on the stochastic patch occupancy model (SPOM) framework.
+This approach, based on a well-developed body of theory, models the occupancy state of discrete patches of suitable habitat.
+Though this prevents us from modeling the dynamics of local populations, this approach is able to capture realistic spatial structure while remaining relatively tractable (\cite{Ovaskainen2004}).
+To incorporate disease transmission, we expand on existing SPOM models and allow each patch to be in one of three possible states: occupied by susceptible hosts (S), occupied by infectious hosts (I), and unoccupied by the host ($\emptyset$).
 State transitions are governed by host colonization, extinction, and transmission rates, where a susceptible population can become infectious either through direct contact with infectious immigrants, or through a local environmental pathogen reservoir (Table~\ref{transitions}).

### 2015-12-08 Manuscript/manuscript.tex
Switching lattice and fully-connected results
Presenting lattice results as the main results and pushing fully-connected results to the supp info, with changes to figures, captions, and the text.

@@
 (2) Migrants preferentially select high quality habitat, that is, the rate at which a patch is colonized scales with quality ($\xi_{im}=0.5$, \cite{Hanski2003}, though we relax this below).  
-(3)  Connectivity among patches is determined entirely by quality ($d_{ij}=1$ for all $i \neq j$).  
-In addition to these assumptions, we chose parameters so that without infection, approximately 0.75 of the patches were occupied ($\mu = 0.1$, $D=5$).
+(3)  Patches are arranged in a lattice and only accessible from their four nearest neighbors ($d_{ij}=1$ for all $i$, $j$ neighbors).  
+In addition to these assumptions, we chose parameters so that without infection, approximately 0.75 of the patches were occupied ($\mu = 0.1$, $D=2$).
 We varied infectious survival ($\nu$), probability of direct transmission ($\delta$), and environmental longevity to explore the dynamics of a range of possible pathogens.
 
-The parameterization above assumes that all patches are equally accessible from all other patches (i.e.
-are separated by effective distance 1).
-To examine the effects of a more rigid spatial structure, we also implemented a model with patches arranged in a square lattice such that each patch is only accessible from its four neighbouring patches (in which case, average migration distance was adjusted to $D = 2$ to maintain roughly equivalent occupancy).
+To examine the effects of a less rigid spatial structure, we also implemented a model in which all patches are equally accessible from all other patches (i.e.
+$d_{ij} = 1$ for all $i \neq j$, in which case average migration distance was adjusted to $D = 5$ to maintain roughly equivalent occupancy).
+The lattice and fully-connected models book-end the range of possible landscape structures, with most realistic metapopulations lying somewhere in between.
 
@@
 
-For each simulation -- 100 replicates for each combination of disease parameters ($\nu$, $\delta$, and $\beta$), habitat quality distribution (high or low), and landscape structure (fully connected or lattice), qualities for 100 patches were drawn from a uniform distribution on the given habitat quality range (above).
+For each simulation -- 100 replicates for each combination of disease parameters ($\nu$, $\delta$, and $\beta$), habitat quality distribution (high or low), and landscape structure (lattice or fully connected), qualities for 100 patches were drawn from a uniform distribution on the given habitat quality range (above).
 In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infected.
@@
 In these simulations, we varied longevity over the values used above and explored the consequences of host preference for high quality habitat (and the corresponding potential for ecological traps) by choosing two different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating that migrants select all patches with equal probability).
-To isolate the effects of habitat preference and pathogen longevity, we fixed the direct transmission probability ($\delta = 0.3$), infectious survival ($\nu = 0.2$), and landscape structure (fully connected).
+To isolate the effects of habitat preference and pathogen longevity, we fixed the direct transmission probability ($\delta = 0.5$), infectious survival ($\nu = 0.2$), and landscape structure (lattice).
 We performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by susceptible and infectious hosts, and the final total host population size.
@@
 d-f).
-Metapopulations with a lattice structure exhibited qualitatively similar patterns, but with generally more limited pathogen spread and larger regions of susceptible maintenance (Supplement).
+Metapopulations with a fully-connected structure exhibited qualitatively similar patterns, but with generally easier pathogen spread (Supplement).
 
@@
 In these cases, high quality habitat served as a sort of metapopulation-level trap, that is, otherwise beneficial habitat that is detrimental to the overall host population in the presence of disease.
-As pathogen longevity increased, the range of the parameter space where we observed this metapopulation-level trap became smaller, apparent only for the most virulent pathogens.
-Again, metapopulations with a lattice structure displayed similar patterns (Supplement).
+As pathogen longevity increased, the range of the parameter space where we observed this metapopulation-level trap shifted towards more virulent pathogens.
+Again, metapopulations with a fully-connected structure displayed similar patterns (Supplement).
 
-To better understand the patch-level mechanisms behind these effects, we further examined within patch infection dynamics for parameters in this intermediate range ($\nu = 0.2$ and $\delta = 0.3$).  
+To better understand the patch-level mechanisms behind these effects, we further examined within patch infection dynamics for parameters in this intermediate range ($\nu = 0.2$ and $\delta = 0.5$).  
 Because of their lower extinction rate and higher recolonization rate, high quality patches were more consistently occupied than low quality patches (Fig~\ref{simvis}, Supplement).  
-However, because of this stability, disease risk, measured by the proportion of time a patch spent occupied by infectious hosts, increased with patch quality (Supplement).
+However, because of this stability, disease risk, measured by the mean proportion of time a patch spent occupied by infectious hosts, increased with patch quality (Supplement).
 Indeed, as an epidemic progressed, high quality patches increasingly supported infectious occupants, while low quality patches supported infectious hosts only briefly and infrequently (Fig~\ref{simvis}).
@@
 $\xi_{im} = 0.5$) increasing this disease risk further (Supplement).
-Moreover, assuming preferential selection of high quality habitat had dramatic consequences for the stability of the host metapopulation when the pathogen is long-lived in the environment, increasing the probability of extinction substantially compared to random habitat selection (0.29 vs.
-0.03).
+
+These qualitative trends are consistent across both landscape structures, but the effects of quality and preference are more pronounced for the fully-connected case (Supplement).
+In the fully connected model, high quality patches attract infectious colonists from the entire metapopulation, not just the local neighborhood, and thus are more consistently infected than in the lattice model (Supplement).
 
@@
 
-However, we find that as the environmental longevity of the pathogen increases, the region in which we observe this reversal becomes smaller, that is, high quality habitat is a metapopulation-level trap for a smaller range of potential pathogens (Fig~\ref{highvlow}, blue region shrinks from panels a.
-to c.).
-Environmentally long-lived pathogens are able to spread widely under a larger range of conditions, largely overwhelming the ability of low quality habitat to provide refuges for susceptible hosts. 
+Given that the presence of this ecological trap depends on the relative magnitude of pathogen spread, we find that pathogens with long-lived environmental reservoirs only generate traps when infectious survival is low enough that the pathogen is not able to spread widely.
+Because environmentally long-lived pathogens are able to spread and persist more easily, increasing longevity can overwhelm the ability of low quality habitat to slow spread and provide refuges for susceptible hosts.
 These dynamics reflect the findings of Gog \emph{et al.} (\cite{Gog2002}), who found that under strong background infection from an alternative host, increasing movement always increased total occupancy, despite facilitating pathogen spread.
@@
 \caption{Boxplots showing the range, over 100 replicate simulations, of susceptible (a), infectious (b), and total (c) host population sizes for high quality (red) and low quality (blue) habitat distributions.
-Infectious survival ($\nu$) and direct transmission rate ($\delta$) are fixed at values of 0.2 and 0.3, respectively.
+Infectious survival ($\nu$) and direct transmission rate ($\delta$) are fixed at values of 0.2 and 0.5, respectively.
 Coloured points show medians, while vertical lines indicate the inner-quartile ranges, with horizontal lines indicating the minimum and maximum.
@@
 \centering
-\caption{Results from a single representative simulation with habitat quality uniformly distributed between 0.23 and 1.77, and a pathogen with an infectious survival of $\nu = 0.2$, a direct transmission rate of $\delta = 0.3$, and an envioronmental longevity equal to the expected residence time of a susceptible population on a unit quality patch (10 units of time here).
+\caption{Results from a single representative simulation with habitat quality uniformly distributed between 0.23 and 1.77, and a pathogen with an infectious survival of $\nu = 0.2$, a direct transmission rate of $\delta = 0.5$, and an envioronmental longevity of $\beta = 3.2$.
 The top panel shows total proportion of patches occupied by susceptible (blue) and infectious (red) hosts over time.

### 2015-12-08 Manuscript/supplementary.tex
Switching lattice and fully-connected results
Presenting lattice results as the main results and pushing fully-connected results to the supp info, with changes to figures, captions, and the text.

@@
 
-Supplemental figures 1 and 2 are analogous to figures 1 and 2 from the main text, but with habitat patches arranged in a lattice structure (i.e.
-each habitat patch is accessible only from its four nearest neighbors).
+Supplemental figures 1, 2, and 3 are analogous to figures 1, 2, and 3 from the main text, but with all habitat patches equally connected (i.e.
+each habitat patch is equally accessible from all other habitat patches).
 
-Supplemental figures 3 and 4 show within-patch occupancy for a given $\nu$ and $\delta$.
+Supplemental figures 4 and 5, and 6 and 7 show within-patch occupancy for the lattice model and fully connected models, respectively, both with a fixed $\nu$ and $\delta$.
 
 \begin{figure}
-\includegraphics[scale=0.5]{figure/supplement_1.pdf}
+\includegraphics[angle = 270, scale=0.5]{figure/supplement_1.eps}
 \caption{Median total susceptible population size as a function of infectious survival ($\nu$), and the probability of direct transmission ($\delta$).
@@
 \begin{figure}
-\includegraphics[scale=0.5]{figure/supplement_2.pdf}
+\includegraphics[angle = 270, scale=0.5]{figure/supplement_2.eps}
 \centering
@@
 \begin{figure}
-\includegraphics[scale=0.6]{figure/supplement_3}
+\includegraphics[angle=270, scale=0.5]{figure/supplement_3.eps}
 \centering
-\caption{Probability of susceptible occupancy (measured as the proportion of time spent occupied by susceptible hosts over the last 500 time steps) as a function of patch quality.
-Red points indicate host preference for high quality habitat ($\xi_{im} = 0.5$), while black points indicate no preference ($\xi_{im} = 0$).
-Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.
-Infectious survival and probability of direct infection both fixed at $\nu = 0.2$ and $\delta = 0.3$, with fully connected landscape structure.}
+\caption{Boxplots showing the range, over 100 replicate simulations, of susceptible (a), infectious (b), and total (c) host population sizes for high quality (red) and low quality (blue) habitat distributions.
+Infectious survival ($\nu$) and direct transmission rate ($\delta$) are fixed at values of 0.2 and 0.3, respectively.
+Coloured points show medians, while vertical lines indicate the inner-quartile ranges, with horizontal lines indicating the minimum and maximum.
+At low longevities, this pathogen is unable to invade and high quality habitat supports larger populations of susceptible hosts.
+However, low quality habitat is better able to maintain these susceptible hosts as the pathogen's longevity increases, leading to larger total population sizes.}
+\label{popsizes}
 \end{figure}
@@
 \begin{figure}
-\includegraphics[scale=0.6]{figure/supplement_4}
+\includegraphics[angle=270, scale=0.6]{figure/supplement_4.eps}
 \centering
-\caption{Probability of infectious occupancy (measured as the proportion of time spent occupied by infectious hosts over the last 500 time steps) as a function of patch quality.
+\caption{Mean probability of susceptible occupancy (measured as the proportion of time spent occupied by susceptible hosts over the last 500 time steps) as a function of patch quality.
 Red points indicate host preference for high quality habitat ($\xi_{im} = 0.5$), while black points indicate no preference ($\xi_{im} = 0$).
 Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.
-Infectious survival and probability of direct infection both fixed at $\nu = 0.2$ and $\delta = 0.3$, with fully connected landscape structure.}
+Infectious survival and probability of direct infection both fixed at $\nu = 0.2$ and $\delta = 0.5$, with lattice landscape structure.}
 \end{figure}
 
+\begin{figure}
+\includegraphics[angle=270, scale=0.6]{figure/supplement_5.eps}
+\centering
+\caption{Mean probability of infectious occupancy (measured as the proportion of time spent occupied by infectious hosts over the last 500 time steps) as a function of patch quality.
+Red points indicate host preference for high quality habitat ($\xi_{im} = 0.5$), while black points indicate no preference ($\xi_{im} = 0$).
+Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.
+Infectious survival and probability of direct infection both fixed at $\nu = 0.2$ and $\delta = 0.5$, with lattice landscape structure.}
+\end{figure}
+
+\begin{figure}
+\includegraphics[angle=270, scale=0.6]{figure/supplement_6.eps}
+\centering
+\caption{Mean probability of susceptible occupancy (measured as the proportion of time spent occupied by susceptible hosts over the last 500 time steps) as a function of patch quality.
+Red points indicate host preference for high quality habitat ($\xi_{im} = 0.5$), while black points indicate no preference ($\xi_{im} = 0$).
+Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.
+Infectious survival and probability of direct infection both fixed at $\nu = 0.2$ and $\delta = 0.3$, with fully-connected landscape structure.}
+\end{figure}
+
+\begin{figure}
+\includegraphics[angle=270, scale=0.6]{figure/supplement_7.eps}
+\centering
+\caption{Mean probability of infectious occupancy (measured as the proportion of time spent occupied by infectious hosts over the last 500 time steps) as a function of patch quality.
+Red points indicate host preference for high quality habitat ($\xi_{im} = 0.5$), while black points indicate no preference ($\xi_{im} = 0$).
+Panel columns show low (0.3, a), medium (1.0, b), and high (3.2, c) pathogen longevities.
+Infectious survival and probability of direct infection both fixed at $\nu = 0.2$ and $\delta = 0.3$, with fully-connected landscape structure.}
+\end{figure}
 \end{document}

### 2015-12-08 Analysis/pubplots.R

@@ -6,6 +6,7 @@ library(RColorBrewer)
 library(ggplot2)
 library(ggthemes)
 library(plyr)
+library(grid)
 
 #===============================================================================
 ### Figure 1
@@ -249,33 +250,70 @@ grid.text("Longevity = 3.2", x = unit(0.74, units = "npc"), y = unit(0.96, units
 dev.off()
 
 #===============================================================================
-### Supplemental figure 3
+### Supplemendtal figure 3
 
 rm(list = ls())
 
-postscript("Manuscript/figure/supplement_3.eps", width = 8, height = 4)
+postscript("Manuscript/figure/supplement_3.eps", width = 10, height = 5)
 
-load(paste(getwd(), "/Output/trap(lattice).RData", sep = ""))
+load(paste(getwd(), "/Output/pathgrid.RData", sep = ""))
+sub <- out[abs(out$delta - 0.3) < 0.001 & abs(out$nu - 0.2) < 0.001, ]
+sub$longevity <- round(sub$longevity, 1)
 
-longs <- unique(out$longevity)[c(4, 7, 10)]
-labels <- c("a", "b", "c")
+S <- ddply(sub, c("longevity", "treatment"), summarise, 
+           med = median(S.pop), 
+           lowq = quantile(S.pop, 0.25),
+           highq = quantile(S.pop, 0.75),
+           min = min(S.pop),
+           max = max(S.pop))
 
-dat <- subset(out, xi_em == 0.5 & longevity %in% longs)
+I <- ddply(sub, c("longevity", "treatment"), summarise, 
+           med = median(I.pop), 
+           lowq = quantile(I.pop, 0.25),
+           highq = quantile(I.pop, 0.75),
+           min = min(I.pop),
+           max = max(I.pop))
 
-S <- ddply(dat, .(xi_im, longevity, quality), summarise, occ = median(S))
+tot <- ddply(sub, c("longevity", "treatment"), summarise, 
+             med = median(S.pop + I.pop), 
+             lowq = quantile(S.pop + I.pop, 0.25),
+             highq = quantile(S.pop + I.pop, 0.75),
+             min = min(S.pop + I.pop),
+             max = max(S.pop + I.pop))
 
-xyplot(occ ~ quality | longevity, groups = xi_im, data = S,
-        xlab = list(label = "Quality", cex = 1.2),
-        ylab = list(label = "Probability susceptible", cex = 1.2),
[…]

### 2015-12-08 Analysis/simvis.r

@@ -3,7 +3,7 @@
 
 source("Code/contSPOM(rates).r")
 
-set.seed(8358)
+set.seed(8689)
 
 # Defines fixed inputs for diseaseSPOM
 parms<-data.frame("xi_im" = 0.5, 
@@ -31,7 +31,7 @@ initial[51:100]<-"E"
 
 timesteps<-5000
 
-longevity <- 1
+longevity <- 10 ^ 0.5
 variance <- 0.2
 
 #===============================================================================

### 2015-12-15 Manuscript/manuscript.tex
Edits for clarity throughout

@@
 \begin{abstract} 
-The benefits of high quality habitat for host persistence may be modulated by the presence of an environmentally persistent pathogen.
+Habitat quality plays an important role in the dynamics and stability of wildlife metapopulations.
+However, the benefits of high quality habitat may be modulated by the presence of an environmentally persistent pathogen.
 In some cases, the presence of environmental pathogen reservoirs on high quality habitat may lead to the creation of ecological traps, wherein host individuals preferentially colonize high quality habitat, but are then exposed to increased infection risk and disease-induced mortality.
@@
 
-Since environmental transmission is spatially explicit (i.e.
-environmental reservoirs can only infect local residents), its role in disease dynamics may further depend on the spatial structure of the host population.
-In particular, we expect that host population structure and movement will be influenced by heterogeneity in quality among the habitat patches in a metapopulation.
+Since environmental transmission is necessarily a spatial process, its role in disease dynamics may further depend on the spatial structure of the host population.
+Many of the hosts listed above have populations that are spread across small patches of suitable habitat in a highly fragmented landscape.
+Taking a metapopulation perspective then, we expect that host population structure and movement will be influenced by heterogeneity in the quality of these habitat patches.
 Indeed, the quality of a habitat patch can affect its extinction and colonization rates, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).
@@
 
-We expect that high quality habitat patches, which support greater host density and traffic than lower quality habitat, might be more likely to form pathogen reservoirs and support consistently infectious populations.
+We expect that high quality habitat patches, which are more attractive and support greater host density than lower quality habitat, might be more likely to form pathogen reservoirs and support consistently infectious populations.
 As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Robertson2006}, \cite{Almberg2011}).
@@
 (1) The rate of emigration from a patch increases with the local population size, while the extinction rate of a patch decreases with populations size ($\xi_{em}=0.5$, $\alpha=1$, \cite{Hanski2003}).
-(2) Migrants preferentially select high quality habitat, that is, the rate at which a patch is colonized scales with quality ($\xi_{im}=0.5$, \cite{Hanski2003}, though we relax this below).  
+(2) Migrants preferentially select high quality habitat, that is, the rate at which a patch is colonized scales with quality ($\xi_{im}=0.5$, \cite{Hanski2003}).  
 (3)  Patches are arranged in a lattice and only accessible from their four nearest neighbors ($d_{ij}=1$ for all $i$, $j$ neighbors).  
@@
 To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with patch quality distributions shifted towards either high or low quality habitat.
-These two distributions were chosen to have equal variance and an overlapping core of patches around unit quality, with either the minimum or maximum shifted to simulate regions with more, or less, high quality habitat (yielding distributions with qualities ranging from 0.23 to 1.24 and from 0.76 to 1.77).
+These two distributions were chosen to explore a possible conservation scenario in which managers might need to prioritize habitat for preservation or restoration.
+To capture the consequences of shifting priorities to either low or high quality habitat, we generated patch quality from uniform distributions with equal variance and limits shifted to favor either low or high quality habitat (yielding distributions with qualities ranging from 0.23 to 1.24 and from 0.76 to 1.77).
 We then implemented a full-factorial set of simulations on each quality distribution in which infectious survival ($\nu$) and probability of direct transmission ($\delta$) were each varied over 10 values (ranging from 0.1 to 1, and 0 to 0.9, respectively).
@@
 
-For each simulation -- 100 replicates for each combination of disease parameters ($\nu$, $\delta$, and $\beta$), habitat quality distribution (high or low), and landscape structure (lattice or fully connected), qualities for 100 patches were drawn from a uniform distribution on the given habitat quality range (above).
+For each combination of disease parameters ($\nu$, $\delta$, and $\beta$), habitat quality distribution (high or low), and landscape structure (lattice or fully connected), we ran 100 replicate simulations in which qualities for 100 patches were drawn from a uniform distribution on the given habitat quality range (above).
 In each case, an entirely susceptible population was simulated until it reached an approximate steady state, at which point a randomly chosen occupied patch was infected.
@@
 
-To more carefully explore the role of habitat quality in within-patch dynamics, we performed an additional simulation experiment on metapopulations with patch qualities evenly spaced over the full range explored above (from 0.2 to 1.8).
+To more carefully explore the role of habitat quality in the occupancy of a given patch, we performed an additional simulation experiment on metapopulations with patch qualities evenly spaced over the full range explored above (from 0.2 to 1.8).
 In these simulations, we varied longevity over the values used above and explored the consequences of host preference for high quality habitat (and the corresponding potential for ecological traps) by choosing two different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating that migrants select all patches with equal probability).
@@
 A wide range of disease dynamics were observed over the range of parameters and habitat quality distributions explored.  
-Epidemic outcome was influenced most strongly by infectious survival, with high infectious survival (i.e.
+The ability of the pathogen to spread (and of susceptible hosts to persist) was influenced most strongly by infectious survival, with high infectious survival (i.e.
 low disease-induced mortality) facilitating widespread infection (in which no susceptible host populations persist), and very low infectious survival (i.e.
@@
 The distribution of habitat quality in the metapopulation also played an important role in determining disease dynamics.
-Increasing overall habitat quality facilitated widespread infection for a larger range of parameters (Fig~\ref{endemic}, a-c vs.
+Prioritizing high quality habitat facilitated widespread infection for a larger range of parameters (Fig~\ref{endemic}, a-c vs.
 d-f).
@@
 However, for pathogens falling between these extremes (e.g.
-with intermediate infectious survival), low quality habitat was a net benefit to the host, supporting larger total population sizes than high quality habitat (blue regions of Fig ~\ref{highvlow}).
+with intermediate infectious survival), favoring low quality habitat was a net benefit to the host, with low quality distributions supporting larger total population sizes than high quality distributions (blue regions of Fig ~\ref{highvlow}).
 The spread of these pathogens was strongly influenced by the habitat quality distribution, with low quality habitat limiting disease spread and better maintaining susceptible occupancy relative to high quality habitat (Fig~\ref{popsizes}, Supplemental).
@@
 This work demonstrates that the habitat quality in a metapopulation can have substantial impacts on the dynamics of a variety of environmentally transmitted pathogens and the resulting size and stability of the host population.
-In some cases, the presence of high quality habitat can enhance the spread and impact of environmentally persistent pathogens.  
+In some cases, prioritizing the preservation or restoration of high quality habitat can enhance the spread and impact of environmentally persistent pathogens.  
 In addition, these patches with many resources may become metapopulation-scale ecological traps that are normally beneficial and attractive to hosts, but become a net drain on the metapopulation due to the impacts of disease.
@@
 
-This expands on previous studies of disease spread in a homogeneous metapopulation that focused on movement rate as the driver of this trade-off, with increased movement both improving metapopulation stability and increasing disease spread (\cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}).
-Shifting the patch quality distribution towards high quality habitat in our model is conceptually similar to increasing movement in these models.
-And, as in these models, the net consequences of this trade-off for the host metapopulation depend on the characteristics of the disease.
+The net consequences of this trade-off for the host metapopulation depend on the characteristics of the disease.
 Broadly, we can divide the observed dynamics into three different scenarios: limited, intermediate, and widespread infection (possibly leading to host extinction). 
@@
 Ovaskainen and Hanski (\cite{Ovaskainen2003a}) similarly demonstrate that correlated extinctions reduce the contributions of well-connected habitat to metapopulation persistence and increase the contributions of more isolated habitat.
-Our model adds to this literature by demonstrating the potential for environmentally persistent pathogens to provide the kind of disturbance under which low quality habitat is beneficial, and thus should be conserved.
+Our model adds to this literature by demonstrating the potential for environmentally persistent pathogens to provide the kind of disturbance under which low quality habitat is beneficial, and should thus be conserved.
 
@@
 \centering
-\caption{Differences between the median total effective population size of hosts on low quality patch distributions (quality between 0.23 to 1.24) and high quality patch distributions (quality between 0.76 and 1.77) as a function of infectious survival ($\nu$) and the probability of direct transmission ($\delta$).
+\caption{Differences between the median total population size of hosts on low quality patch distributions (quality between 0.23 to 1.24) and high quality patch distributions (quality between 0.76 and 1.77) as a function of infectious survival ($\nu$) and the probability of direct transmission ($\delta$).
 Reds indicate that high quality habitat distributions support higher median host population sizes than low quality habitat, primarily corresponding to pathogens with either very limited or very extensive spread.
@@
 \centering
-\caption{Results from a single representative simulation with habitat quality uniformly distributed between 0.23 and 1.77, and a pathogen with an infectious survival of $\nu = 0.2$, a direct transmission rate of $\delta = 0.5$, and an envioronmental longevity of $\beta = 3.2$.
+\caption{Results from a single representative simulation with habitat quality uniformly distributed between 0.23 and 1.77, and a pathogen with an infectious survival of $\nu = 0.2$, a direct transmission rate of $\delta = 0.5$, and an environmental longevity of $\beta = 3.2$.
 The top panel shows total proportion of patches occupied by susceptible (blue) and infectious (red) hosts over time.
@@
 Patches are stacked vertically with the lowest quality at the bottom and the highest quality at the top.
-Note the overall more consistent occupancy on high quality habitat and the only very brief spurts of infectious occupancy on low quality habitat.}
+Note the overall more consistent infectious occupancy on high quality habitat and the only very brief spurts of infectious occupancy on low quality habitat.}
 \label{simvis}

### 2015-12-15 Manuscript/supplementary.tex
Edits for clarity throughout

@@
 \centering
-\caption{Differences between the median total effective population size of hosts on high quality patch distributions (quality between 0.76 and 1.77) and low quality patch distributions (quality between 0.23 to 1.24) as a function of infectious survival ($\nu$) and the probability of direct transmission ($\delta$).
+\caption{Differences between the median total population size of hosts on high quality patch distributions (quality between 0.76 and 1.77) and low quality patch distributions (quality between 0.23 to 1.24) as a function of infectious survival ($\nu$) and the probability of direct transmission ($\delta$).
 Reds indicate higher median host population sizes in high quality patch distributions, while blues indicate higher median host population sizes in low quality patch distributions.

### 2016-01-20 Manuscript/manuscript.tex
Lots of clarity edits

@@
 Since environmental transmission is necessarily a spatial process, its role in disease dynamics may further depend on the spatial structure of the host population.
-Many of the hosts listed above have populations that are spread across small patches of suitable habitat in a highly fragmented landscape.
+For many wildlife hosts with environmentally persistent pathogens (e.g.
+prairie dog colonies affected by plague, \cite{George2013}), this spatial structure is well-described as a metapopulation, where host populations occupy small patches of suitable habitat in a highly fragmented landscape.
 Taking a metapopulation perspective then, we expect that host population structure and movement will be influenced by heterogeneity in the quality of these habitat patches.
@@
 We expect that high quality habitat patches, which are more attractive and support greater host density than lower quality habitat, might be more likely to form pathogen reservoirs and support consistently infectious populations.
-As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants, effectively creating an ecological trap (\cite{Robertson2006}, \cite{Almberg2011}).
-In addition, the greater traffic through high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (\cite{Paull2012}).
+As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants.
+The resulting fitness consequences of increased disease risk would then effectively create an ecological trap for migrant hosts preferentially selecting high quality habitat (\cite{Robertson2006}, \cite{Almberg2011}).
+In addition, the greater stability of high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (\cite{Paull2012}).
 Similarly, we expect that low quality patches will be less likely to develop pathogen reservoirs and may serve as refuges where susceptible hosts can escape infection.
-The combination of these effects may then lead to the creation of metapopulation-scale traps, where the inclusion of otherwise high quality habitat in the metapopulation actually reduces host population size relative to low quality habitat.
 
-Many, if not all, of the pathogens listed above affect spatially structured host populations (e.g.
-plague in prairie dog colonies, \cite{George2013}), and thus understanding how pathogen transmission interacts with patterns of habitat quality is critical to managing disease in these systems.
-In this study, we explore how pathogen spread interacts with the distribution of habitat quality to influence host occupancy.
+The combination of these effects may then lead to the creation of a sort of management trap at the metapopulation-scale, where the conservation or restoration of otherwise high quality habitat in the metapopulation actually reduces total host population size relative to actions focused on low quality habitat.
+Thus, understanding how pathogen transmission interacts with patterns of habitat quality is critical to managing these systems.
+In this study, we explore this interaction and its influence on host occupancy and overall population size.
 We specifically focus on the relative importance of high and low quality habitats to host persistence in the presence of pathogens with different levels of environmental longevity.
@@
 To address the above questions, we developed a model based on the stochastic patch occupancy model (SPOM) framework.
-This approach, based on a well-developed body of theory, models the occupancy state of discrete patches of suitable habitat.
-Though this prevents us from modeling the dynamics of local populations, this approach is able to capture realistic spatial structure while remaining relatively tractable (\cite{Ovaskainen2004}).
-To incorporate disease transmission, we expand on existing SPOM models and allow each patch to be in one of three possible states: occupied by susceptible hosts (S), occupied by infectious hosts (I), and unoccupied by the host ($\emptyset$).
-State transitions are governed by host colonization, extinction, and transmission rates, where a susceptible population can become infectious either through direct contact with infectious immigrants, or through a local environmental pathogen reservoir (Table~\ref{transitions}).
+This framework, based on a well-developed body of theory (\cite{Hanski2003}, \cite{Ovaskainen2004}), models the occupancy state of discrete patches of suitable habitat.
+The SPOM approach is flexible and able to capture realistic features of a landscape, including spatial structure and heterogeneity in habitat quality, while remaining relatively tractable (\cite{Ovaskainen2004}).
+To incorporate disease processes, we expand on existing SPOM models (\cite{Hanski2003}, \cite{Ovaskainen2004}) and allow each patch to be in one of three possible states: occupied by susceptible hosts (S), occupied by infectious hosts (I), and unoccupied by the host ($\emptyset$).
+Transitions between states are governed by host colonization and extinction rates, and a susceptible population can become infected either through direct contact with infectious immigrants, or through a local environmental pathogen reservoir (Table~\ref{transitions}).
 
@@
 
-Susceptible populations are infected via direct contact with infectious colonists at rate $\delta C_{Ii}$, where $\delta$ is the direct transmission probability.
-In addition, environmental transmission can take place when a susceptible population occupies a previously infected patch, and occurs at rate:
+Susceptible host populations can become infected through two different routes: via direct contact with infectious immigrants and via an environmental reservoir left behind by a previously infectious population.
+Infectious immigrants arriving on a susceptible patch (at rate $C_{Ii}$) infect the resident population with probability $\delta$, while susceptible populations occupying a contaminated patch become infected at rate:
 \begin{equation}
@@
 \end{equation}
-where $\gamma_0$ is the initial transmission rate of the pathogen reservoir, $\tau_{i}$ is the time since last infectious occupancy ($\tau_{i} = \infty$, and thus $\gamma(\tau_i) = 0$, if the patch has never been occupied by infectious hosts), and $r$ is the pathogen's decay rate in the environment, determined by the pathogen's longevity, $\beta$ (defined as its half-life in the environment relative to the expected residence time of a population on a unit quality patch):
+where $\gamma_0$ is the initial infection rate from the pathogen reservoir, $\tau_{i}$ is the time since last infectious occupancy ($\tau_{i} = \infty$, and thus $\gamma(\tau_i) = 0$, if the patch has never been occupied by infectious hosts), and $r$ is the pathogen's decay rate in the environment, determined by the pathogen's longevity, $\beta$ (defined as its half-life in the environment relative to the expected residence time of a population on a unit quality patch):
 \begin{equation}
@@
 
-To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with patch quality distributions shifted towards either high or low quality habitat.
+To explore the relative influence of low and high quality habitat on host and pathogen dynamics, we simulated the spread of a range of pathogens in metapopulations with patch quality distributions that favor either high or low quality habitat.
 These two distributions were chosen to explore a possible conservation scenario in which managers might need to prioritize habitat for preservation or restoration.
@@
 The state of the metapopulation was then tracked for 5000 time steps.
-We recorded the proportion of patches in each state at the end of each simulation and also computed the mean total population size of susceptible and infectious hosts over the last 500 time steps of the simulation (summing the local population sizes for all occupied patches, assuming a local population size of $A_i$ for patches that are occupied and susceptible and $\nu A_i$ for occupied infectious patches).
+We recorded the proportion of patches in each state at the end of each simulation and also computed the mean total population size of susceptible and infectious hosts over the last 500 time steps of the simulation to get an estimate of the equilibrium state (summing the local population sizes for all occupied patches, assuming a local population size of $A_i$ for occupied susceptible patches and $\nu A_i$ for occupied infectious patches).
 
@@
 In these simulations, we varied longevity over the values used above and explored the consequences of host preference for high quality habitat (and the corresponding potential for ecological traps) by choosing two different values for $\xi_{im}$ (with $\xi_{im} = 0.5$ indicating preference for high quality habitat, and  $\xi_{im} = 0$ indicating that migrants select all patches with equal probability).
-To isolate the effects of habitat preference and pathogen longevity, we fixed the direct transmission probability ($\delta = 0.5$), infectious survival ($\nu = 0.2$), and landscape structure (lattice).
+To isolate the effects of habitat preference and pathogen longevity, we fixed the direct infection probability ($\delta = 0.5$), infectious survival ($\nu = 0.2$), and landscape structure (lattice).
 We performed 100 replicate simulations for each set of model parameters (three values of pathogen longevity and two values of $\xi_{im}$) and recorded the proportion of time each patch spent occupied by susceptible and infectious hosts, and the final total host population size.
@@
 d-f).
-Metapopulations with a fully-connected structure exhibited qualitatively similar patterns, but with generally easier pathogen spread (Supplement).
+Metapopulations with a fully-connected structure exhibited qualitatively similar patterns, but with generally easier pathogen spread (Supplement Fig 1).
 
-For a wide range of parameter values, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions despite generally facilitating pathogen spread (red shaded regions in Fig~\ref{highvlow}).
-In the case of pathogens that were unable to invade due to low infectious survival and weak transmission, high quality habitat distributions supported larger total host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat (bottom left corner of the panels in Fig~\ref{highvlow}).
-Similarly, when high infectious survival and strong transmission (from either environmental or direct transmission) allowed a pathogen to spread widely (i.e.
-at a pandemic level) in both habitat quality distributions, high quality distributions again supported larger total host populations than low quality distributions (top portion of panels in Fig~\ref{highvlow}).
+For two large regions of parameter values, high quality patch distributions had a net positive effect on total host population size relative to low quality distributions (red shaded regions in Fig~\ref{highvlow}).
+In the first region (bottom left corner of the panels in Fig~\ref{highvlow}), corresponding to pathogens that were unable to invade due to low infectious survival and weak transmission, high quality habitat distributions supported larger total host populations than low quality distributions due to the greater stability and connectivity afforded by high quality habitat (e.g.
+Fig~\ref{popsizes} for $\beta = 0.3$).
+In the second region (top portion of panels in Fig~\ref{highvlow}), corresponding to pathogens that were able to spread widely (i.e.
+at a pandemic level) due to high infectious survival and strong transmission (from either environmental or direct transmission), high quality distributions again supported larger total host populations than low quality distributions.
 
 However, for pathogens falling between these extremes (e.g.
-with intermediate infectious survival), favoring low quality habitat was a net benefit to the host, with low quality distributions supporting larger total population sizes than high quality distributions (blue regions of Fig ~\ref{highvlow}).
-The spread of these pathogens was strongly influenced by the habitat quality distribution, with low quality habitat limiting disease spread and better maintaining susceptible occupancy relative to high quality habitat (Fig~\ref{popsizes}, Supplemental).
-In these cases, high quality habitat served as a sort of metapopulation-level trap, that is, otherwise beneficial habitat that is detrimental to the overall host population in the presence of disease.
+with intermediate infectious survival), having low quality habitat was a net benefit to the host, with low quality distributions supporting larger total population sizes than high quality distributions (blue regions of Fig ~\ref{highvlow}).
+The spread of these pathogens was strongly influenced by the habitat quality distribution, with low quality habitat limiting disease spread and better maintaining susceptible occupancy relative to high quality habitat (Fig~\ref{endemic}).
+As an example, for a particular pathogen in this region ($\nu = 0.2$, $\delta = 0.5$, and $\beta = 3.2$, Fig~\ref{popsizes}), we see that low quality habitat distributions supported very small infectious populations and moderate susceptible populations.
+In contrast, high quality habitat distributions better facilitated the maintenance and spread of infection, substantially reducing the population of susceptibles and thereby the total population size (Fig~\ref{popsizes}).
+In this case, high quality habitat served as a sort of metapopulation-level trap, that is, otherwise beneficial habitat that is detrimental to the overall host population in the presence of disease.
 As pathogen longevity increased, the range of the parameter space where we observed this metapopulation-level trap shifted towards more virulent pathogens.
-Again, metapopulations with a fully-connected structure displayed similar patterns (Supplement).
+Again, metapopulations with a fully-connected structure displayed similar patterns (Supplement Figs 2, 3).
 
-To better understand the patch-level mechanisms behind these effects, we further examined within patch infection dynamics for parameters in this intermediate range ($\nu = 0.2$ and $\delta = 0.5$).  
-Because of their lower extinction rate and higher recolonization rate, high quality patches were more consistently occupied than low quality patches (Fig~\ref{simvis}, Supplement).  
-However, because of this stability, disease risk, measured by the mean proportion of time a patch spent occupied by infectious hosts, increased with patch quality (Supplement).
+To better understand the mechanisms behind these effects, we further examined patch-level occupancy for parameters in this intermediate range ($\nu = 0.2$ and $\delta = 0.5$).  
+Because of their lower extinction rate and higher recolonization rate, high quality patches were more consistently occupied than low quality patches (Fig~\ref{simvis}, Supplement Figs 4, 5).  
+However, because of this stability, disease risk, measured by the mean proportion of time a patch spent occupied by infectious hosts, increased with patch quality (Supplement Fig 5).
 Indeed, as an epidemic progressed, high quality patches increasingly supported infectious occupants, while low quality patches supported infectious hosts only briefly and infrequently (Fig~\ref{simvis}).
-Thus, from the perspective of a migrant individual, colonizing high quality habitat posed increased disease risk, with the preferential selection of high quality habitat (i.e.
-$\xi_{im} = 0.5$) increasing this disease risk further (Supplement).
+In addition, migrant preference for high quality habitat (i.e. $\xi_{im} = 0.5$) increased this disease risk further (Supplement Fig 5).
 
-These qualitative trends are consistent across both landscape structures, but the effects of quality and preference are more pronounced for the fully-connected case (Supplement).
-In the fully connected model, high quality patches attract infectious colonists from the entire metapopulation, not just the local neighborhood, and thus are more consistently infected than in the lattice model (Supplement).
+These qualitative trends are consistent across both landscape structures, but the effects of quality and preference are more pronounced for the fully-connected case (Supplement Fig 6, 7).
+In the fully connected model, high quality patches attract infectious colonists from the entire metapopulation, not just the local neighborhood, and thus are more consistently infected than in the lattice model (Supplement Fig 7).
 
@@
 
-This work demonstrates that the habitat quality in a metapopulation can have substantial impacts on the dynamics of a variety of environmentally transmitted pathogens and the resulting size and stability of the host population.
-In some cases, prioritizing the preservation or restoration of high quality habitat can enhance the spread and impact of environmentally persistent pathogens.  
-In addition, these patches with many resources may become metapopulation-scale ecological traps that are normally beneficial and attractive to hosts, but become a net drain on the metapopulation due to the impacts of disease.
+This work demonstrates that the distribution of habitat quality in a metapopulation can have substantial impacts on the dynamics of a variety of environmentally transmitted pathogens and the resulting size and stability of the host population.
+In particular, prioritizing the preservation or restoration of high quality habitat can enhance the spread and impact of environmentally persistent pathogens.  
+In systems with endemic but not widespread disease, these patches with many resources may become metapopulation-scale management traps that are normally beneficial and attractive to hosts, but become a net drain on the metapopulation due to the impacts of disease.
 This suggests that the presence of a disease can lead to a trade-off between otherwise high quality habitat that facilitates metapopulation stability at the expense of increased disease spread, and low quality habitat that hinders disease spread at the expense of metapopulation stability.
@@
 There is an intermediate scenario, however, where low quality habitats are more likely to remain uninfected and as a result boost the total population size while high quality patches that are heavily infected tend to reduce the total population size.
-In this scenario, the ability of pathogens to invade and spread is influenced substantially by the colonization and extinction rates of high and low quality habitat, and the contrasting roles that these habitat types play in the metapopulation.
+In this scenario, the ability of pathogens to invade and spread is influenced substantially by the colonization and extinction rates of both high and low quality habitat, and the contrasting roles that these habitat types play in the metapopulation.
 
@@
 Due to this additional stability, high quality habitat is better able to both maintain a consistent pool of susceptible hosts to which infection can spread, and support infectious host populations under disease-induced mortality.
-As a result, the probability of infection (i.e. proportion of time occupied by infectious hosts) increases with patch quality (Supplement).
+As a result, the probability of infectious occupancy (i.e. proportion of time occupied by infectious hosts) increases with patch quality (Supplement Figs 5, 7).
 From the perspective of a migrant susceptible host then, selection of high quality habitat leads to increased disease risk relative to other habitats, either because the patch is likely already occupied by infectious hosts, or because the local susceptible hosts are likely to soon become infected (either through the presence of an environmental reservoir or direct contact).
@@
 $\xi_{em} > 0$), it forms an ecological trap at the individual scale, that is, preferentially selected habitat that reduces individual fitness relative to other habitats (\cite{Robertson2006}).
-Moreover, the preferential selection of high quality habitat further exacerbates the increase in disease risk with habitat quality (Supplement) by increasing the risk of direct transmission from infectious migrants also selecting high quality habitat.
-This echoes results from the contact network literature that show that highly connected nodes have higher infection risk (\cite{Christley2005}, \cite{Keeling2005}).
+Moreover, host preference for high quality habitat further exacerbates the increase in disease risk on high quality habitat (Supplement Figs 5, 7) by increasing the rate of pathogen import from infectious migrants also selecting high quality habitat.
+This echoes results from the contact network literature that show that highly connected nodes have higher risk of becoming infected (\cite{Christley2005}, \cite{Keeling2005}).
 
-The relatively low extinction rate on high quality patches, and the resulting increased probability of infection, helps to create a relatively stable platform from which the pathogen can spread through the rest of the metapopulation.
-Indeed, for pathogens with spread in this intermediate range, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).  
-This helps to maintain infectious occupancy throughout the metapopulation, which, when coupled with preference for high quality habitat, feeds back on high quality patches to ensure a steady stream of new infectious colonists.
-Through these two interacting processes -- the ecological trap created by the preference of individuals for high quality habitat, and the superspreader role emerging from the increased connectivity and stability -- the presence of high quality patches serves to significantly increase pathogen spread, even for more virulent pathogens (Fig~\ref{endemic}, \ref{popsizes}).
-The consequences of widespread disease-induced mortality can outweigh the positive consequences of increased metapopulation connectivity and stability, leading to the creation of a metapopulation-scale trap and a net decrease in total host population size on high quality habitat distributions (Fig~\ref{highvlow}, \ref{popsizes}).
+The relatively low extinction rate on high quality patches, and the resulting increased probability of infectious occupancy, helps to create a relatively stable platform from which the pathogen can spread through the rest of the metapopulation.
+Indeed, for pathogens with intermediate spread, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).  
+This superspreader role helps to maintain infectious occupancy throughout the metapopulation, which, when coupled with preference for high quality habitat, feeds back on high quality patches to ensure a steady stream of new infectious colonists.
+Through these two interacting processes -- the ecological trap created by the preference of individuals for high quality habitat, and the superspreader role emerging from the increased connectivity and stability generated by high quality habitat -- the presence of high quality patches serves to substantially increase pathogen spread, even for more virulent pathogens (Figs~\ref{endemic}, \ref{popsizes}).
+The consequences of widespread disease-induced mortality can then outweigh the positive consequences of increased metapopulation connectivity and stability, leading to the creation of a metapopulation-scale trap and a net decrease in total host population size on high quality habitat distributions (Fig~\ref{highvlow}, \ref{popsizes}).
 
@@
 
-In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy in the metapopulation (Fig~\ref{endemic}, \ref{popsizes}). 
-Even though individual low quality patches, due to their high extinction and low colonization rates, are only able to support transient occupancy (Fig~\ref{simvis}, Supplement), the presence of low quality habitat facilitates more widespread susceptible persistence for a greater range of pathogens (Fig~\ref{endemic}).
+In contrast to high quality habitat, low quality patches help to limit pathogen spread and increase susceptible occupancy in the metapopulation (Figs~\ref{endemic}, \ref{popsizes}). 
+Even though individual low quality patches, due to their high extinction and low colonization rates, are only able to support transient occupancy (Fig~\ref{simvis}, Supplement Figs 4-7), the presence of low quality habitat facilitates more widespread susceptible persistence for a greater range of pathogens (Fig~\ref{endemic}).
 This phenomenon represents the other side of the role played by high quality patches in that low quality patches are relatively unstable (i.e.
@@
 
-These results have general management implications that highlight the potential importance of low-quality habitat for the persistence of wildlife populations and the importance of high-quality habitat for the persistence of wildlife disease.
-It is critical to understand these implications in order to avoid falling into the metapopulation trap created by high quality habitat and its interaction with pathogen spread.
+These results have general management implications that highlight the potential importance of low-quality habitat for the persistence of wildlife populations and the potential importance of high-quality habitat for the persistence of wildlife disease.
+It is critical to understand these implications in order to avoid falling into the management trap created by high quality habitat and its interaction with pathogen spread.
 Previous theoretical work suggests that management should focus on maintaining patches where conditions are most favourable for the host (i.e.
@@
 Ovaskainen and Hanski (\cite{Ovaskainen2003a}) similarly demonstrate that correlated extinctions reduce the contributions of well-connected habitat to metapopulation persistence and increase the contributions of more isolated habitat.
-Our model adds to this literature by demonstrating the potential for environmentally persistent pathogens to provide the kind of disturbance under which low quality habitat is beneficial, and should thus be conserved.
+Our model adds to this literature by demonstrating the potential for environmentally persistent pathogens to provide the kind of disturbance under which low quality habitat is beneficial and should thus be conserved.
 
@@
 In particular, the stochastic patch occupancy model, though well suited for capturing realistic spatial structure, assumes that within-patch dynamics (i.e.
-growth to carrying capacity and pathogen transmission) are rapid relative to among-patch dynamics (i.e.
+growth to carrying capacity and local pathogen spread) are rapid relative to among-patch dynamics (i.e.
 colonization and extinction).
 This is often a reasonable assumption, especially for relatively small populations in a highly fragmented landscape (\cite{Hanski2003}), but may not capture the non-linearities and nuances of host-pathogen dynamics in a single population.
-In addition, we assume that patch quality only influences population size and the resulting colonization and extinction processes, but quality could also influence pathogen transmission itself through effects on host density or condition (\cite{Becker2015}).
+In addition, we assume that patch quality only influences population size and the resulting colonization and extinction processes, but quality could also influence pathogen dynamics within a patch through effects on host density or condition (\cite{Becker2015}).
 Patch quality may also influence the size or infectiousness of the environmental reservoir, either directly, by influencing the decay rate of the pathogen in the environment, or indirectly, through the effects on population size and overall shedding rate.
-The potential importance of these additional mechanisms remains an open question for future work and need to be considered before applying this framework to any given system of interest.
+The potential importance of these additional mechanisms remains an open question for future work and needs to be considered before applying this framework to any given system of interest.
 Fortunately, the SPOM framework is fairly flexible and future work could extend the model presented here to include additional processes, e.g.
@@
 
-Though detailed predictions for specific systems will likely require additional model development, the three scenarios identified here nonetheless provide a coarse framework that we can use to identify the potential risk of metapopulation-level traps in wildlife populations currently affected by an environmentally persistent pathogen.
-For instance, high quality habitat seems unlikely to form metapopulation traps for bat populations affected by white nose syndrome due to the widespread infection both within and between hibernaculum and the scarcity of host refuges (\cite{Langwig2014}, \cite{ORegan2015}).
+Though detailed predictions for specific systems will likely require additional model development, the three scenarios identified here nonetheless provide a coarse framework that we can use to identify the potential risk of management traps in wildlife populations currently affected by an environmentally persistent pathogen.
+For instance, high quality habitat seems unlikely to form management traps for bat populations affected by white nose syndrome due to the widespread infection both within and between hibernaculum and the scarcity of host refuges (\cite{Langwig2014}, \cite{ORegan2015}).
 Similarly, from these results, we would not expect the spread of chronic wasting disease to turn high quality habitat into traps for mule deer metapopulations, due again to widespread infection (albeit at low prevalence) across population units (\cite{Conner2004}) and the relative importance of local-scale transmission within winter ranges (\cite{Farnsworth2006}).
@@
 \hline
-$S \rightarrow I$ & Transmission (from contact or reservoir) & $\delta C_{Ii} + \gamma(\tau_i)$\\
+$S \rightarrow I$ & Infection (from contact or reservoir) & $\delta C_{Ii} + \gamma(\tau_i)$\\
 $S \rightarrow \emptyset $ & Extinction (of susceptible) & $E_{Si}$\\
@@
 \hline
-$\delta$ & Probability of direct transmission & 0 - 0.9 \\
-$\gamma_0$ & Initial rate of transmission from reservoir patch & 0.5\\
+$\delta$ & Probability of direct infection & 0 - 0.9 \\
+$\gamma_0$ & Initial rate of infection from reservoir patch & 0.5\\
 $\beta$ & Pathogen longevity (half-life in environment, relative to $1/\mu$) & 0.3 - 3

### 2016-01-21 Manuscript/manuscript.tex
Additions to future work and preference discussion

@@
 $\xi_{em} > 0$), it forms an ecological trap at the individual scale, that is, preferentially selected habitat that reduces individual fitness relative to other habitats (\cite{Robertson2006}).
-Moreover, host preference for high quality habitat further exacerbates the increase in disease risk on high quality habitat (Supplement Figs 5, 7) by increasing the rate of pathogen import from infectious migrants also selecting high quality habitat.
+Moreover, host preference for high quality habitat further exacerbates this increase in disease risk (Supplement Figs 5, 7) by increasing the rate of pathogen import from infectious migrants also selecting high quality habitat.
+The magnitude of this effect depends on the landscape structure and the scale at which individuals sense and select habitat.
+In a fully connected metapopulation, where high quality patches attract migrants from the entire landscape, the effect of preference on the relationship between quality and disease risk is the most pronounced (Supplement Fig 7).
+However, the fact that the effect persists in a lattice metapopulation (Supplement Fig 5), where migrants can only select among the four neighboring patches, suggests that host preference for high quality habitat likely serves to further increase the disease risk on high quality patches even in more realistic landscapes.
 This echoes results from the contact network literature that show that highly connected nodes have higher risk of becoming infected (\cite{Christley2005}, \cite{Keeling2005}).
@@
 Patch quality may also influence the size or infectiousness of the environmental reservoir, either directly, by influencing the decay rate of the pathogen in the environment, or indirectly, through the effects on population size and overall shedding rate.
+More generally, the dynamics and establishment of environmental reservoirs could be influenced by a number of other factors, such as the length of infectious occupancy or the occupancy history of a patch.
 The potential importance of these additional mechanisms remains an open question for future work and needs to be considered before applying this framework to any given system of interest.

### 2016-01-25 Manuscript/manuscript.tex
Formatting citations
Deleting parentheses around citations

@@
 
-\begin{quote}
-This draft manuscript is distributed solely for purposes of scientific peer review.
-Its content is deliberative and pre decisional, so it must not be disclosed or released by reviewers.
-Because the manuscript has not yet been approved for publication by the U.S.
-Geological Survey (USGS), it does not represent any official USGS finding or policy.
-\end{quote}
-
-\newpage
-
 \begin{abstract} 
@@
 Many prominent and problematic wildlife diseases are caused by pathogens that can persist, and remain infectious, for long periods of time in the environment.
-Examples include chronic wasting disease (\cite{Miller2006}), anthrax (\emph{Bacillus anthracis}, \cite{Dragon1995}), plague (\emph{Yersinia pestis}, \cite{Eisen2008}), and white nose syndrome (\emph{Pseudogymnoascus destructans}, \cite{Reynolds2015}), among others.
+Examples include chronic wasting disease \cite{Miller2006}, anthrax (\emph{Bacillus anthracis}, \cite{Dragon1995}), plague (\emph{Yersinia pestis}, \cite{Eisen2008}), and white nose syndrome (\emph{Pseudogymnoascus destructans}, \cite{Reynolds2015}), among others.
 This environmental persistence creates environmental pathogen reservoirs from which susceptible hosts can become infected without direct contact with an infectious individual.
-This additional transmission pathway can have important consequences for disease dynamics, with models showing that increased environmental longevity generally facilitates increased pathogen persistence and spread relative to direct transmission alone (\cite{Almberg2011}, \cite{Sharp2011}, \cite{Breban2009}).
+This additional transmission pathway can have important consequences for disease dynamics, with models showing that increased environmental longevity generally facilitates increased pathogen persistence and spread relative to direct transmission alone \cite{Almberg2011, Sharp2011, Breban2009}.
 
@@
 Taking a metapopulation perspective then, we expect that host population structure and movement will be influenced by heterogeneity in the quality of these habitat patches.
-Indeed, the quality of a habitat patch can affect its extinction and colonization rates, as well as its contribution to the colonization of other empty patches (\cite{Moilanen1998}).
+Indeed, the quality of a habitat patch can affect its extinction and colonization rates, as well as its contribution to the colonization of other empty patches \cite{Moilanen1998}.
 These processes will influence how a pathogen spreads, where environmental pathogen reservoirs get established, and the resulting effects on host occupancy and population size.
@@
 As these reservoirs are undetectable to the host, we predict that high quality patches will continue to attract -- and infect -- susceptible immigrants.
-The resulting fitness consequences of increased disease risk would then effectively create an ecological trap for migrant hosts preferentially selecting high quality habitat (\cite{Robertson2006}, \cite{Almberg2011}).
-In addition, the greater stability of high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders (\cite{Paull2012}).
+The resulting fitness consequences of increased disease risk would then effectively create an ecological trap for migrant hosts preferentially selecting high quality habitat \cite{Robertson2006, Almberg2011}.
+In addition, the greater stability of high quality patches may further facilitate pathogen spread by positioning high quality patches as metapopulation-scale superspreaders \cite{Paull2012}.
 Similarly, we expect that low quality patches will be less likely to develop pathogen reservoirs and may serve as refuges where susceptible hosts can escape infection.
@@
 To address the above questions, we developed a model based on the stochastic patch occupancy model (SPOM) framework.
-This framework, based on a well-developed body of theory (\cite{Hanski2003}, \cite{Ovaskainen2004}), models the occupancy state of discrete patches of suitable habitat.
-The SPOM approach is flexible and able to capture realistic features of a landscape, including spatial structure and heterogeneity in habitat quality, while remaining relatively tractable (\cite{Ovaskainen2004}).
-To incorporate disease processes, we expand on existing SPOM models (\cite{Hanski2003}, \cite{Ovaskainen2004}) and allow each patch to be in one of three possible states: occupied by susceptible hosts (S), occupied by infectious hosts (I), and unoccupied by the host ($\emptyset$).
+This framework, based on a well-developed body of theory \cite{Hanski2003, Ovaskainen2004}, models the occupancy state of discrete patches of suitable habitat.
+The SPOM approach is flexible and able to capture realistic features of a landscape, including spatial structure and heterogeneity in habitat quality, while remaining relatively tractable \cite{Ovaskainen2004}.
+To incorporate disease processes, we expand on existing SPOM models \cite{Hanski2003, Ovaskainen2004} and allow each patch to be in one of three possible states: occupied by susceptible hosts (S), occupied by infectious hosts (I), and unoccupied by the host ($\emptyset$).
 Transitions between states are governed by host colonization and extinction rates, and a susceptible population can become infected either through direct contact with infectious immigrants, or through a local environmental pathogen reservoir (Table~\ref{transitions}).
@@
 
-Continuous time stochastic simulations of the above model were implemented in the R language (\cite{R2014}) using the Gillespie algorithm (\cite{Gillespie1977}).  Code is available at \\
+Continuous time stochastic simulations of the above model were implemented in the R language \cite{R2014} using the Gillespie algorithm \cite{Gillespie1977}.  Code is available at \\
 \url{https://github.com/clint-leach/Metapop-Disease}.
@@
 Thus, if high quality habitat is preferred by migrants (i.e.
-$\xi_{em} > 0$), it forms an ecological trap at the individual scale, that is, preferentially selected habitat that reduces individual fitness relative to other habitats (\cite{Robertson2006}).
+$\xi_{em} > 0$), it forms an ecological trap at the individual scale, that is, preferentially selected habitat that reduces individual fitness relative to other habitats \cite{Robertson2006}.
 Moreover, host preference for high quality habitat further exacerbates this increase in disease risk (Supplement Figs 5, 7) by increasing the rate of pathogen import from infectious migrants also selecting high quality habitat.
@@
 However, the fact that the effect persists in a lattice metapopulation (Supplement Fig 5), where migrants can only select among the four neighboring patches, suggests that host preference for high quality habitat likely serves to further increase the disease risk on high quality patches even in more realistic landscapes.
-This echoes results from the contact network literature that show that highly connected nodes have higher risk of becoming infected (\cite{Christley2005}, \cite{Keeling2005}).
+This echoes results from the contact network literature that show that highly connected nodes have higher risk of becoming infected \cite{Christley2005, Keeling2005}.
 
 The relatively low extinction rate on high quality patches, and the resulting increased probability of infectious occupancy, helps to create a relatively stable platform from which the pathogen can spread through the rest of the metapopulation.
-Indeed, for pathogens with intermediate spread, high quality patches are effectively metapopulation scale superspreaders (\cite{Lloyd-Smith2005}, \cite{Paull2012}).  
+Indeed, for pathogens with intermediate spread, high quality patches are effectively metapopulation scale superspreaders \cite{Lloyd-Smith2005, Paull2012}.  
 This superspreader role helps to maintain infectious occupancy throughout the metapopulation, which, when coupled with preference for high quality habitat, feeds back on high quality patches to ensure a steady stream of new infectious colonists.
 Through these two interacting processes -- the ecological trap created by the preference of individuals for high quality habitat, and the superspreader role emerging from the increased connectivity and stability generated by high quality habitat -- the presence of high quality patches serves to substantially increase pathogen spread, even for more virulent pathogens (Figs~\ref{endemic}, \ref{popsizes}).
-The consequences of widespread disease-induced mortality can then outweigh the positive consequences of increased metapopulation connectivity and stability, leading to the creation of a metapopulation-scale trap and a net decrease in total host population size on high quality habitat distributions (Fig~\ref{highvlow}, \ref{popsizes}).
+The consequences of widespread disease-induced mortality can then outweigh the positive consequences of increased metapopulation connectivity and stability, leading to the creation of a metapopulation-scale trap and a net decrease in total host population size on high quality habitat distributions (Figs~\ref{highvlow}, \ref{popsizes}).
 
@@
 As a result, low quality patches effectively represent a dead-end for the pathogen, reducing the number of patches from which it can spread, and allowing susceptible hosts to more easily persist and avoid infection throughout the metapopulation.
-The opportunity for pathogen reservoir decay provided by the ephemeral occupancy of low quality habitat echoes the 'migratory escape' hypothesis (\cite{Loehle1995}), which suggests that annual migration allows hosts to vacate contaminated sites and return once they are clean.
+The opportunity for pathogen reservoir decay provided by the ephemeral occupancy of low quality habitat echoes the 'migratory escape' hypothesis \cite{Loehle1995}, which suggests that annual migration allows hosts to vacate contaminated sites and return once they are clean.
 Taken with this literature, our results highlight the importance of transient/temporary habitat in mitigating host exposure to environmentally transmitted pathogens.
@@
 high quality habitat), but Strasser et al.
-(\cite{Strasser2010}) show that stochastic disturbance (e.g.
+\cite{Strasser2010} show that stochastic disturbance (e.g.
 disease-induced mortality) can lead to cases where focusing on low quality habitat is more effective in increasing population growth rate.
-Ovaskainen and Hanski (\cite{Ovaskainen2003a}) similarly demonstrate that correlated extinctions reduce the contributions of well-connected habitat to metapopulation persistence and increase the contributions of more isolated habitat.
+Ovaskainen and Hanski \cite{Ovaskainen2003a} similarly demonstrate that correlated extinctions reduce the contributions of well-connected habitat to metapopulation persistence and increase the contributions of more isolated habitat.
 Our model adds to this literature by demonstrating the potential for environmentally persistent pathogens to provide the kind of disturbance under which low quality habitat is beneficial and should thus be conserved.
@@
 colonization and extinction).
-This is often a reasonable assumption, especially for relatively small populations in a highly fragmented landscape (\cite{Hanski2003}), but may not capture the non-linearities and nuances of host-pathogen dynamics in a single population.
-In addition, we assume that patch quality only influences population size and the resulting colonization and extinction processes, but quality could also influence pathogen dynamics within a patch through effects on host density or condition (\cite{Becker2015}).
+This is often a reasonable assumption, especially for relatively small populations in a highly fragmented landscape \cite{Hanski2003}, but may not capture the non-linearities and nuances of host-pathogen dynamics in a single population.
+In addition, we assume that patch quality only influences population size and the resulting colonization and extinction processes, but quality could also influence pathogen dynamics within a patch through effects on host density or condition \cite{Becker2015}.
 Patch quality may also influence the size or infectiousness of the environmental reservoir, either directly, by influencing the decay rate of the pathogen in the environment, or indirectly, through the effects on population size and overall shedding rate.
@@
 Though detailed predictions for specific systems will likely require additional model development, the three scenarios identified here nonetheless provide a coarse framework that we can use to identify the potential risk of management traps in wildlife populations currently affected by an environmentally persistent pathogen.
-For instance, high quality habitat seems unlikely to form management traps for bat populations affected by white nose syndrome due to the widespread infection both within and between hibernaculum and the scarcity of host refuges (\cite{Langwig2014}, \cite{ORegan2015}).
-Similarly, from these results, we would not expect the spread of chronic wasting disease to turn high quality habitat into traps for mule deer metapopulations, due again to widespread infection (albeit at low prevalence) across population units (\cite{Conner2004}) and the relative importance of local-scale transmission within winter ranges (\cite{Farnsworth2006}).
-However, we might expect to see high quality traps in the plague-prairie dog system, where the infection is widespread and disease-induced mortality is high, but uninfected colonies are still able to persist (\cite{Stapp2004}).
-Indeed, studies have suggested that large colonies are more likely to become infected (\cite{Snall2008}), and thus high quality habitat capable of supporting large populations may help facilitate plague persistence and spread.
+For instance, high quality habitat seems unlikely to form management traps for bat populations affected by white nose syndrome due to the widespread infection both within and between hibernaculum and the scarcity of host refuges \cite{Langwig2014, ORegan2015}.
+Similarly, from these results, we would not expect the spread of chronic wasting disease to turn high quality habitat into traps for mule deer metapopulations, due again to widespread infection (albeit at low prevalence) across population units \cite{Conner2004} and the relative importance of local-scale transmission within winter ranges \cite{Farnsworth2006}.
+However, we might expect to see high quality traps in the plague-prairie dog system, where the infection is widespread and disease-induced mortality is high, but uninfected colonies are still able to persist \cite{Stapp2004}.
+Indeed, studies have suggested that large colonies are more likely to become infected \cite{Snall2008}, and thus high quality habitat capable of supporting large populations may help facilitate plague persistence and spread.
 Thus, by considering the characteristics of a host-pathogen system and their effect on the roles played by high and low quality habitat, we can begin to diagnose where empirical systems fall relative to these three scenarios and the corresponding consequences for management.
@@
 This partitioning of pathogen dynamics into three regions based on whether high quality habitat is a net benefit or detriment to the host metapopulation echoes the work of \cite{Hess1996}, \cite{Gog2002}, \cite{Park2012}.
-In particular, for a given pathogen, Hess (\cite{Hess1996}) similarly identified three regions of disease dynamics which determined whether increased movement had a positive or negative effect on host occupancy.
+In particular, for a given pathogen, Hess \cite{Hess1996} similarly identified three regions of disease dynamics which determined whether increased movement had a positive or negative effect on host occupancy.
 In cases where the host movement rate was either low enough that the pathogen was unable to invade or high enough to cause a widespread infection, Hess found that further increasing movement increased host occupancy, while at intermediate levels that resulted in moderate disease spread, increasing movement decreased host occupancy.
@@
 Because environmentally long-lived pathogens are able to spread and persist more easily, increasing longevity can overwhelm the ability of low quality habitat to slow spread and provide refuges for susceptible hosts.
-These dynamics reflect the findings of Gog \emph{et al.} (\cite{Gog2002}), who found that under strong background infection from an alternative host, increasing movement always increased total occupancy, despite facilitating pathogen spread.
-Park (\cite{Park2012}) expanded on this by adding environmental transmission to the mix and found that in cases where background and environmental transmission were strong relative to direct transmission, increasing movement was again a net benefit to the host metapopulation.
+These dynamics reflect the findings of Gog \emph{et al.} \cite{Gog2002}, who found that under strong background infection from an alternative host, increasing movement always increased total occupancy, despite facilitating pathogen spread.
+Park \cite{Park2012} expanded on this by adding environmental transmission to the mix and found that in cases where background and environmental transmission were strong relative to direct transmission, increasing movement was again a net benefit to the host metapopulation.
 These studies, coupled with the results presented here, suggest that when there is a persistent source of infection throughout the metapopulation (e.g.

### 2016-02-17 Manuscript/manuscript.tex
Edits to discussion

@@
 
+In addition to helping to guide the prioritization of conservation efforts, our model may also help to inform the deployment of disease control measures, including the treatment of the environmental reservoir itself.
+In the case of African herbivores and their tick-borne pathogens, for example, the environmental reservoir created by ticks and livestock hosts could be controlled through the treatment of cattle with acaricides \cite{Keesing2013}.
+Our work suggests that such efforts would likely be most effective if focused on cattle that share high quality habitat with the wildlife hosts of interest.
+
 The framework used here provides general insights, but generating predictions for specific systems requires a careful consideration of the model's simplifying assumptions.
-In particular, the stochastic patch occupancy model, though well suited for capturing realistic spatial structure, assumes that within-patch dynamics (i.e.
-growth to carrying capacity and local pathogen spread) are rapid relative to among-patch dynamics (i.e.
-colonization and extinction).
-This is often a reasonable assumption, especially for relatively small populations in a highly fragmented landscape \cite{Hanski2003}, but may not capture the non-linearities and nuances of host-pathogen dynamics in a single population.
+In particular, the stochastic patch occupancy model, though well suited for capturing realistic spatial structure, does not account for within-patch dynamics.
+As a result, patches are modeled as either unoccupied or occupied by a host population with a fixed size and discrete disease status -- either susceptible or infected.
+This occupancy framework assumes that within-patch dynamics (i.e. growth to carrying capacity and local pathogen spread) are rapid relative to among-patch dynamics (i.e. colonization and extinction).
+Moreover, it assumes that the pathogen spreads widely within a patch and affects host dynamics primarily by reducing population size.
+These are often reasonable assumptions, especially for relatively small populations in a highly fragmented landscape \cite{Hanski2003}, but may not capture the non-linearities and nuances of host-pathogen dynamics in a single population.
+
 In addition, we assume that patch quality only influences population size and the resulting colonization and extinction processes, but quality could also influence pathogen dynamics within a patch through effects on host density or condition \cite{Becker2015}.

### 2016-02-17 Manuscript/manuscript.tex
Adding citation to Zenodo archived code

@@
 
-Continuous time stochastic simulations of the above model were implemented in the R language \cite{R2014} using the Gillespie algorithm \cite{Gillespie1977}.  Code is available at \\
-\url{https://github.com/clint-leach/Metapop-Disease}.
+Continuous time stochastic simulations of the above model were implemented in the R language \cite{R2014} using the Gillespie algorithm \cite{Gillespie1977}.
+Code is available from the project's Github repository \cite{Leach2016}.
 
@@
 \subsection*{Data accessibility}
-Simulation source code and generated data are available at \\  \url{https://github.com/clint-leach/Metapop-Disease}.
+Simulation source code and generated data are available on Github and archived at \url{http://dx.doi.org/10.5281/zenodo.46194}.
 
